# Baseline Characterisitcs for Participants included in Microbiome Analysis

```{r}
#| echo: false
#| eval: true

# This initial code block loads custom utility functions and environment variables
# These are stored in separate files to keep the main analysis clean and organized
suppressWarnings(suppressMessages({
  if (file.exists("utils/r-utils.r")) {
    source("utils/r-utils.r")
  }
  if (file.exists("_environment.required")) {
    source("_environment.required")
  }

  library(promicrobialTools)

  prefix <- ""

  filepath <- get_results_dir(directory_naming = "parent")
}))

baseline_metadata <- readRDS("data/metadata/baseline-metadata.rds")
```

```{r}
#| eval: true
#| label:  metadata comparison data prep
m1 <- baseline_metadata %>%
  rownames_to_column(var = "RowNames") %>%
  select(
    `Age (months)` = agemonths_base,
    Sex = sexf,
    `Intervention arm` = arm,
    `Birth mode vaginal (vs. Caesarean)` = del_type,
    `Birth weight (kg)` = bwt,
    `Current weight (kg)` = cwt,
    `Mid-upper arm circumference (cm)` = muac_avg,
    `Head circumference (cm)` = hc_avg,
    `Head circumference-for-age Z-score` = zhc,
    `Length (cm)` = ln_avg,
    `Length-for-age Z-score (LAZ)` = zlen,
    `Stunting (LAZ < -2)` = stunt_base,
    `Weight-for-age Z-score (WAZ) ` = zwei,
    `Underweight (WAZ < -2)` = underwt_base,
    `Weight-for-length Z-score (WLZ) ` = zwfl,
    `Wasting (WLZ < -2)` = wast_base,
    `Ferritin (ng/ml)` = fer_base,
    `Iron deficiency (<12 ng/ml) ` = fer_def12_base,
    `Ferritin (ng/ml)* adjusted for CRP` = fer_crpadjbr, # using BRINDA
    `Iron deficiency (<12 ng/ml)* adjusted for CRP` = fer_crpbr_def12, # using BRINDA
    `Zinc (µg/dL)` = zinc,
    `Zinc deficiency (<70 µg/dL) ` = zinc_def70, # <70 µg/dL
    `Hemoglobin (g/dl)` = hb_base,
    `Anemia (hemoglobin < 11 g/dl) ` = anemia_base,
    `C-reactive protein (CRP) ≥ 5 mg/L ` = crp_elev5,
    `GI symptoms` = gi_yn,
    `Diarrhea today or within past 4 weeks` = freq3diarr,
    Fever = fever_yn,
    `Fever today or within past 4 weeks` = freqfever,
    `Respiratory symptoms` = resp_yn,
    `Cough today or within past 4 weeks` = freqcough,
    `Observed features` = observed,
    `Chao 1 Index` = chao1,
    `Shannon Entropy` = diversity_shannon,
    Evenness = evenness_pielou,
    `Faith's Phylogenetic Diversity` = pd
  ) %>%
  mutate_at(vars(everything()), str_trim) %>%
  type.convert(., as.is = TRUE) %>%
  mutate(across(
    c(`GI symptoms`, Fever, `Respiratory symptoms`),
    ~ ifelse(.x == 0, "No", "Yes")
  )) %>%
  mutate(across(
    c(
      `Diarrhea today or within past 4 weeks`,
      `Fever today or within past 4 weeks`,
      `Cough today or within past 4 weeks`
    ),
    ~ case_when(
      .x == "1" ~ "Today",
      .x == "2" ~ "Last 7 days",
      .x == "3" ~ "Last 4 weeks",
      .x == "4" ~ "None",
      .x == "8" ~ "Don't know"
    )
  )) %>%
  mutate(
    Sex = ifelse(Sex == 0, "Male", "Female"),
    `Birth weight (kg)` = (`Birth weight (kg)`) / 1000,
    `Current weight (kg)` = (`Current weight (kg)`),
    `Low birth weight (<2.5 kg)` = (`Birth weight (kg)`) <= 2.5,
    `Birth mode vaginal (vs. Caesarean)` = case_when(
      `Birth mode vaginal (vs. Caesarean)` == "1" ~ "Vaginal",
      `Birth mode vaginal (vs. Caesarean)` == "2" ~ "Caesarian"
    ),
    `Diarrhea today or within past 4 weeks` = ifelse(
      `GI symptoms` == "No",
      "None",
      `Diarrhea today or within past 4 weeks`
    ),
    `Fever today or within past 4 weeks` = ifelse(
      Fever == "No",
      "None",
      `Fever today or within past 4 weeks`
    ),
    `Cough today or within past 4 weeks` = ifelse(
      `Respiratory symptoms` == "No",
      "None",
      `Cough today or within past 4 weeks`
    )
  )
```

```{r}
#| eval: true
#| label: tbl-basemb
#| tbl-cap: "Baseline characteristics of pearl millet trial participants who provided fecal samples for microbiome analysis, stratified by intervention arm. Continuous variables are presented as median (interquartile range; minimum-maximum). Categorical variables are presented as n/N (%). P-values from Wilcoxon rank sum test for continuous variables and Pearson's chi-squared test or Fisher's exact test for categorical variables. Q-values represent false discovery rate-adjusted p-values for multiple testing correction."
m1 %>%
  tbl_summary(
    by = `Intervention arm`,
    missing = "no",
    statistic = list(
      all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
      all_categorical() ~ "{n} / {N}  \n({p}%)"
    )
  ) %>% # stats and format for categorical columns
  add_n() %>%
  add_overall() %>%
  add_p() %>%
  add_q() %>%
  bold_p(q = T) %>%
  as_kable_extra(booktabs = TRUE) %>%
  kable_styling(latex_options = "scale_down") %>%
  landscape()
```

## Results

### Intervention Group Comparison

Of the 168 participants with available microbiome sequencing data, 82 (49%) were allocated to the control pearl millet (CPM) group and 86 (51%) to the iron and zinc fortified pearl millet (FeZnPM) intervention group. Baseline characteristics were well-balanced between intervention arms, with no statistically significant differences observed after correction for multiple testing (all q-values > 0.90) (@tbl-basemb).

Demographic characteristics were similar between groups, with median ages of 15.03 months in the CPM group compared to 14.39 months in the FeZnPM group (p = 0.44). Sex distribution was balanced, with females comprising 48% and 52% of the CPM and FeZnPM groups, respectively (p = 0.54). Birth mode and birth weight were comparable between arms (p = 0.54 and p = 0.57, respectively).

Anthropometric measurements and nutritional status indicators showed no significant differences. Stunting prevalence was 24% in the CPM group versus 33% in the FeZnPM group (p = 0.24), whilst underweight (24% vs. 23%, p = 0.86) and wasting (9.8% vs. 7.0%, p = 0.51) were similar between groups.

Micronutrient status was comparable between intervention arms. Iron deficiency affected 63% of the CPM group compared to 57% of the FeZnPM group (p = 0.44), whilst zinc deficiency was present in 15% and 22% of participants, respectively (p = 0.31). Anaemia prevalence was similar between groups (66% vs. 64%, p = 0.80).

Clinical symptoms showed similar patterns across intervention arms. Gastrointestinal symptoms were reported by 13% of the CPM group versus 9.5% of the FeZnPM group (p = 0.49), fever by 37% versus 39% (p = 0.82), and respiratory symptoms by 46% versus 47% (p = 0.91), respectively.

Microbiome diversity metrics were well-balanced between groups, with median observed features of 1,941 in the CPM group compared to 1,840 in the FeZnPM group (p = 0.45). Shannon entropy (3.34 vs. 3.34, p = 0.98) and other alpha diversity measures showed no significant differences, confirming appropriate randomisation for microbiome analyses.