## Baseline to Endpoint Functional Analysis

```{r}
#| echo: false
#| eval: true
#| label: setup
#| cache: false

# This initial code block loads custom utility functions and environment variables
# These are stored in separate files to keep the main analysis clean and organized

suppressWarnings(suppressMessages({
  source(normalizePath("utils/r-utils.r"))
  source(normalizePath("_environment.required"))
  load(normalizePath("data/seqdata.RData"))

  metadata <- read_delim(normalizePath(METADATA))
  bac_base_end_meta <- readRDS(normalizePath(
    "data/metadata/base-end-metadata.rds"
  ))

  kegg <- readRDS(file.path(WOLTKA_OUT, "kegg/kegg.rds"))

  library(edgeR)
  library(ComplexHeatmap)
  library(pairwiseAdonis)
  library(Glimma)
  library(GUniFrac)
  library(ANCOMBC)
  library(limma)
  library(promicrobialTools)

  #to save for consistancy:
  pathway_class_col <- c(
    "Carbohydrate metabolism" = "#A6CEE3",
    "Lipid metabolism" = "#5399C6",
    "Metabolism of cofactors and vitamins" = "#3F8EAA",
    "Amino acid metabolism" = "#99CD91",
    "Nucleotide metabolism" = "#79C360",
    "Biosynthesis of other secondary metabolites" = "#3E9F32",
    "Xenobiotics biodegradation and metabolism" = "#B89B74",
    "Metabolism of other amino acids" = "#F47676",
    "Glycan biosynthesis and metabolism" = "#E52829",
    "Metabolism of terpenoids and polyketides" = "#F06C45",
    "Energy metabolism" = "#FDB762",
    "Metabolic pathways" = "#FE901E",
    "Microbial metabolism in diverse environments" = "#ED8F47",
    "Carbon metabolism" = "#CCAFCA",
    "2-Oxocarboxylic acid metabolism" = "#9471B4",
    "Degradation of aromatic compounds" = "#825D99",
    "Biosynthesis of amino acids" = "#DDD399",
    "Biosynthesis of cofactors" = "#E0BE6D",
    "Biosynthesis of nucleotide sugars" = "#B15928"
  )

  base_end_paired_samples <- read.csv(
    "data/metadata/base-end-paired-samples.csv"
  )

  # kegg pathway annotation files
  pathways <- read.csv(
    file.path(REF, "/WoL2-Woltka/function", "pathways.csv")
  )

  prefix <- "EvsB-KO-"
  filepath <- get_results_dir(directory_naming = "both")
}))
```

Results will be saved to ``r filepath``.

### Pathways with 70% or more completeness

#### Percentage of pathways in each timepoint

```{r}
#| eval: true
#| label: baseline-endpoint pathway coverage filtering

kegg_be <-
  subset(
    subset(
      kegg,
      sample %in% base_end_paired_samples$sample
    ),
    timepoint %in% c("Baseline", "Endpoint")
  ) %>%
  mutate(
    timepoint = case_when(
      str_detect(timepoint, "Baseline") ~ "Baseline",
      str_detect(timepoint, "Endpoint") ~ "Endpoint"
    )
  ) %>%
  dplyr::filter(percent_cov >= 70) %>%
  mutate(n = 1)

```

::: report

`r report::report(kegg_be)`

:::

```{r}
#| eval: true
#| label: tbl-unique-ko
#| tbl-cap: Total number of unique pathways (ko terms) at baseline and endpoint in each intervention arm.

kegg_be %>%
  group_by(timepoint, arm) %>%
  summarise(n_distinct(ko)) %>%
  kable()
```

```{r}
#| eval: true

tmp <- kegg_be %>%
  dplyr::select(!c(trt, pathway_group)) %>%
  arrange(desc(percent_cov)) %>%
  group_by(timepoint, arm) %>%
  mutate(total_samples = n_distinct(sample)) %>%
  ungroup() %>%
  reframe(
    samples_with_pathway = sum(n),
    percent_samples = samples_with_pathway / total_samples * 100,
    .by = c(timepoint, arm, pathway_name, pathway_class, total_samples)
  )

tmp2 <- tmp %>%
  group_by(pathway_name, arm, timepoint) %>%
  summarise(path_presence = mean(percent_samples)) %>%
  dplyr::filter(path_presence < 95)
```


::: {#fig-box}

```{r}
#| eval: true
#| label: fig-ms-4
#| fig-width: 22
#| fig-height: 18

tmp %>%
  dplyr::filter(pathway_name %in% tmp2$pathway_name) %>%
  ggplot(aes(
    y = reorder(pathway_name, dplyr::desc(pathway_class)),
    x = percent_samples,
    fill = pathway_class
  )) +
  geom_col(position = "dodge") +
  xlab("Percentage of Samples") +
  ylab(NULL) +
  scale_fill_manual(values = pathway_class_col) +
  ggh4x::facet_nested_wrap(~ timepoint + arm)
```

**Percentage of KEGG pathways present in samples among intervention arms at baseline and endpoint:** There were `r length(unique(kegg_be$ko))` unique functional pathways with $\geq$ 70% completeness identified across all samples. Those present in $\geq$ 95% of samples are plotted in graph, separated by timepoint and intervention arm. Pathway terms are color coded according to their parent pathway class.

:::

#### Differential KEGG pathway abundance analysis

##### Cross-sectional comparisons

```{r}
#| eval: true

base <- kegg_be %>% filter(timepoint == "Baseline")
end <- kegg_be %>% filter(timepoint == "Endpoint")
```

###### Baseline arm differences

####### Data preparation

```{r}
#| eval: true
#| label: testing for de abundant pathways
base_counts <- base %>%
  dplyr::select(ko, sample, count) %>%
  pivot_wider(names_from = sample, values_from = count, values_fill = 0) %>%
  as.data.frame()

base_counts <- DGEList(as.data.frame(base_counts))

base_counts$samples$arm <- as.factor(kegg$arm[match(
  rownames(base_counts$samples),
  kegg$sample
)])

base_counts$samples$group <- NULL

base_counts$genes <- base_counts$genes %>%
  mutate(
    pathway_name = kegg$pathway_name[match(ko, kegg$ko)],
    pathway_class = kegg$pathway_class[match(ko, kegg$ko)]
  )

base_counts.keep <- rowSums(base_counts$counts == 0) == ncol(base_counts$counts)
base_counts <- base_counts[base_counts.keep == FALSE, ]

base_counts <- normLibSizes(base_counts)
```

####### Contrast design

:::h0

$H_0$: KEGG pathways are not differentially abundant between arms at baseline

:::

::: h1

$H_1$: There is a difference between KEGG pathway abundance between arms at baseline
    $H_1a$: KEGG pathway abundance *increases* between arms at baseline
    $H_1b$: KEGG pathway abundance *decreases* between arms at baseline

:::

```{r}
#| eval: true
#| label: KEGG pathway differences between baseline and endpoint
#Linear Model
#Contrasts
design <- stats::model.matrix(~arm, data = base_counts$samples)

contrasts <- makeContrasts(
  CPMvsFeZnPM = armFeZnPM,
  levels = design
)
```

```{r}
#| eval: true
#| label: fig-base-glmQLfit
#| fig-cap: KEGG pathway differences in FeZnPM at baseline compared to CPM based on a negative binomial generalized log-linear model with quasi-dispersions and empirical Bayes moderation.

fit <- glmQLFit(base_counts, design)

qlf <- glmQLFTest(fit, contrast = contrasts)
glm_results <- topTags(qlf, n = Inf) %>%
  as.data.frame() %>%
  arrange(ko)

glm_results %>%
  filter(abs(logFC) >= log2(2)) %>%
  ggplot(aes(y = reorder(pathway_name, logFC), x = logFC)) +
  geom_bar(
    position = position_dodge(),
    stat = "identity",
    aes(fill = pathway_class)
  ) +
  xlab(bquote(bold("Fold change (" ~ Log[2] ~ "FC )"))) +
  ylab(NULL) + #labels the y axis
  scale_fill_manual(values = pathway_class_col)
```

####### Empirical Bayes Statistics for Differential Expression 

```{r}
#| eval: true
#Transform raw count data to log2-base_counts per million (logCPM), estimate the mean-variance relationship and use this to compute appropriate observation-level weights. The data are then ready for linear modelling.

#If the DGEList object has been scale-normalized in edgeR (or before), then leave normalize.method="none" in voom, i.e., the logCPM values should not usually be re-normalized in the voom call.
#KEGG abundance values have not been normalised
# use updated edgeR voom procedure
v <- voomLmFit(base_counts, design, normalize.method = "none")
#v <- voom(base_counts, design, normalize.method="none") #Transforms count data to log2-base_counts per million
#vfit <- lmFit(v, design)
vfit <- contrasts.fit(v, contrasts = contrasts)
efit <- eBayes(vfit) #estimation of QL dispersions
```

```{r}
#| eval: true
#| label: tbl-fdr5
#| tbl-cap: OGUs that are significantly differentially abundant from zero for each contrast from a fit object (eBayes) containing p-values and test statistics. An FDR of 0.05 was used.

knitr::kable(summary(
  FDR5 <- decideTests(efit, adjust.method = "BH", p.value = 0.05, fc = 0)
))

write.fit(
  fit = efit,
  results = FDR5,
  file = file.path(
    filepath,
    paste0(prefix, "limma-write-fit-baseline-results-FDR5-eBayes.csv")
  ),
  row.names = F
)
```

####### Ordination Analysis and PERMANOVA of differences between arms at baseline

```{r}
#| eval: true
#| label: tbl-mds
#| tbl-cap: Results of PERMANOVA test investigating abundance differences between arms at baseline in paired samples.
#unsupervised learning: MDS plot
group_colours <- base_counts$samples$arm
levels(group_colours) <- c("CPM" = "#FC8D62", "FeZnPM" = "#8DA0CB")
group_colours <- as.character(group_colours)
df <- as.data.frame(t(base_counts$counts))
df <- cbind(df, group = base_counts$samples$arm)

#PERMANOVA testing differenital abundances between timepoints
mds_stat <- pairwiseAdonis::pairwise.adonis2(
  df[, 1:(ncol(df) - 1)] ~ group,
  data = df,
  method = 'eu'
) #returns non-adjusted p-values

#output table
knitr::kable(mds_stat[2])

```

```{r}
#| eval: true
#| label: tbl-mds-sum
#| tbl-cap: Results of PERMANOVA test investigating abundance differences between arms at baseline in paired samples.

pv <- lapply(2:length(mds_stat), function(x) {
  w <- mds_stat[[x]][1, 4]
  y <- mds_stat[[x]][1, 5]
  z <- names(mds_stat)[x]
  data.frame(contrast = z, p = y, F.stat = w)
})
pv <- bind_rows(pv)
knitr::kable(
  pv %>%
    mutate(p.adj = round(p.adjust(p, method = "BH"), 4)) %>%
    mutate(
      sig = case_when(
        p.adj <= 0.0001 ~ "****",
        p.adj <= 0.001 ~ "***",
        p.adj <= 0.01 ~ "**",
        p.adj <= 0.05 ~ "*",
        p.adj > 0.05 ~ "ns"
      )
    )
)
```

The p-value ($Pr(>F)$) for `r base_counts$samples$arm` is `r pv$p`, which is greater than the conventional significance level of 0.05. This suggests that we cannot reject the null hypothesis, implying that there is no significant overall difference in microbial pathway abundance between baseline and endpoint.

The `plotMDS()` function accepts "any data object that can be coerced to a matrix of log-expression values, for example an ExpressionSet or an EList" [@Gentleman2025]. If the `edgeR::voomLmFit()` function was used the `EList` object can be obtained using `fit$EList`

```{r}
#| eval: true
#| label: fig-mds-baseArms
#| fig-cap: MDS plot comparing arms at baseline in paired samples.

xmax = max(
  plotMDS(
    v$EList,
    col = group_colours,
    dim.plot = c(1, 2),
    pch = 16,
    gene.selection = "common"
  )$x
) -
  0.05
ymax = max(
  plotMDS(
    v$EList,
    col = group_colours,
    dim.plot = c(1, 2),
    pch = 16,
    gene.selection = "common"
  )$y
) -
  0.05

par(mar = c(7.1, 5.1, 5.1, 5.1), xpd = TRUE)
plotMDS(
  v$EList,
  col = group_colours,
  dim.plot = c(1, 2),
  pch = 16,
  gene.selection = "common"
)
legend(
  "bottom",
  pch = 16,
  col = unique(group_colours),
  legend = levels(v$targets$arm),
  inset = c(0, -0.165),
  bty = "n",
  ncol = 2
)
text(
  x = xmax - 0.05,
  y = ymax,
  labels = bquote(R^2 == .(round(as.data.frame(mds_stat[2])[1, 3], 3)))
)
text(
  x = xmax - 0.05,
  y = ymax - 0.12,
  labels = bquote(p == .(round(as.data.frame(mds_stat[2])[1, 5], 3)))
)
```

There are two obvious clusters that are not due to arm in the MDS graph.

###### Endpoint arm differences

####### Data preparation

```{r}
#| eval: true
#| label: testing for de abundant pathways
end_counts <- end %>%
  dplyr::select(ko, sample, count) %>%
  pivot_wider(names_from = sample, values_from = count, values_fill = 0) %>%
  as.data.frame()

end_counts <- DGEList(as.data.frame(end_counts))

end_counts$samples$arm <- as.factor(kegg$arm[match(
  rownames(end_counts$samples),
  kegg$sample
)])

end_counts$samples$group <- NULL

end_counts$genes <- end_counts$genes %>%
  mutate(
    pathway_name = kegg$pathway_name[match(ko, kegg$ko)],
    pathway_class = kegg$pathway_class[match(ko, kegg$ko)]
  )

end_counts.keep <- rowSums(end_counts$counts == 0) == ncol(end_counts$counts)
end_counts <- end_counts[end_counts.keep == FALSE, ]

end_counts <- normLibSizes(end_counts)
```

####### Contrast design

:::h0

$H_0$: KEGG pathways are not differentially abundant between arms at endline

:::

::: h1

$H_1$: There is a difference between KEGG pathway abundance between arms at endline
    $H_1a$: KEGG pathway abundance *increases* between arms at endline
    $H_1b$: KEGG pathway abundance *decreases* between arms at endline

:::

```{r}
#| eval: true
#| label: KEGG pathway differences between arms at endpoint
#Linear Model
#Contrasts
design <- stats::model.matrix(~arm, data = end_counts$samples)

contrasts <- makeContrasts(
  CPMvsFeZnPM = armFeZnPM,
  levels = design
)
```

```{r}
#| eval: true
#| label: fig-end-glmQLfit
#| fig-cap: KEGG pathway differences in FeZnPM at endpoint compared to CPM based on a negative binomial generalized log-linear model with quasi-dispersions and empirical Bayes moderation.
fit <- glmQLFit(end_counts, design)

qlf <- glmQLFTest(fit, contrast = contrasts)
glm_results <- topTags(qlf, n = Inf) %>%
  as.data.frame() %>%
  arrange(ko)

glm_results %>%
  filter(abs(logFC) >= log2(2)) %>%
  ggplot(aes(y = reorder(pathway_name, logFC), x = logFC)) +
  geom_bar(
    position = position_dodge(),
    stat = "identity",
    aes(fill = pathway_class)
  ) +
  xlab(bquote(bold("Fold change (" ~ Log[2] ~ "FC )"))) +
  ylab(NULL) + #labels the y axis
  scale_fill_manual(values = pathway_class_col)
```

####### Empirical Bayes Statistics for Differential Expression 

```{r}
#| eval: true
#Transform raw count data to log2-end_counts per million (logCPM), estimate the mean-variance relationship and use this to compute appropriate observation-level weights. The data are then ready for linear modelling.

#If the DGEList object has been scale-normalized in edgeR (or before), then leave normalize.method="none" in voom, i.e., the logCPM values should not usually be re-normalized in the voom call.
#KEGG abundance values have not been normalised
# use updated edgeR voom procedure
v <- voomLmFit(end_counts, design, normalize.method = "none")
#v <- voom(end_counts, design, normalize.method="none") #Transforms count data to log2-end_counts per million
#vfit <- lmFit(v, design)
vfit <- contrasts.fit(v, contrasts = contrasts)
efit <- eBayes(vfit) #estimation of QL dispersions
```

```{r}
#| eval: true
#| label: tbl-fdr5
#| tbl-cap: OGUs that are significantly differentially abundant from zero for each contrast from a fit object (eBayes) containing p-values and test statistics. An FDR of 0.05 was used.

knitr::kable(summary(
  FDR5 <- decideTests(efit, adjust.method = "BH", p.value = 0.05, fc = 0)
))

write.fit(
  fit = efit,
  results = FDR5,
  file = file.path(
    filepath,
    paste0(prefix, "limma-write-fit-endpoint-results-FDR5-eBayes.csv")
  ),
  row.names = F
)
```

####### Ordination Analysis and PERMANOVA of differences between arms at endline

```{r}
#| eval: true
#| label: tbl-mds
#| tbl-cap: Results of PERMANOVA test investigating abundance differences between arms at endline in paired samples.
#unsupervised learning: MDS plot
group_colours <- end_counts$samples$arm
levels(group_colours) <- c("CPM" = "#FC8D62", "FeZnPM" = "#8DA0CB")
group_colours <- as.character(group_colours)
df <- as.data.frame(t(end_counts$counts))
df <- cbind(df, group = end_counts$samples$arm)

#PERMANOVA testing differenital abundances between timepoints
mds_stat <- pairwiseAdonis::pairwise.adonis2(
  df[, 1:(ncol(df) - 1)] ~ group,
  data = df,
  method = 'eu'
) #returns non-adjusted p-values

#output table
knitr::kable(mds_stat[2])
```

```{r}
#| eval: true
#| label: tbl-mds-sum
#| tbl-cap: Results of PERMANOVA test investigating abundance differences between arms at endline in paired samples.

pv <- lapply(2:length(mds_stat), function(x) {
  w <- mds_stat[[x]][1, 4]
  y <- mds_stat[[x]][1, 5]
  z <- names(mds_stat)[x]
  data.frame(contrast = z, p = y, F.stat = w)
})
pv <- bind_rows(pv)
knitr::kable(
  pv %>%
    mutate(p.adj = round(p.adjust(p, method = "BH"), 4)) %>%
    mutate(
      sig = case_when(
        p.adj <= 0.0001 ~ "****",
        p.adj <= 0.001 ~ "***",
        p.adj <= 0.01 ~ "**",
        p.adj <= 0.05 ~ "*",
        p.adj > 0.05 ~ "ns"
      )
    )
)
```

The p-value ($Pr(>F)$) for `r end_counts$samples$arm` is `r pv$p`, which is greater than the conventional significance level of 0.05. This suggests that we cannot reject the null hypothesis, implying that there is no significant overall difference in microbial pathway abundance between endline and endpoint.

The `plotMDS()` function accepts "any data object that can be coerced to a matrix of log-expression values, for example an ExpressionSet or an EList" [@Gentleman2025]. If the `edgeR::voomLmFit()` function was used the `EList` object can be obtained using `fit$EList`

```{r}
#| eval: true
#| label: fig-mds-endArm
#| fig-cap: MDS plot comparing arms at endline in paired samples.

xmax = max(
  plotMDS(
    v$EList,
    col = group_colours,
    dim.plot = c(1, 2),
    pch = 16,
    gene.selection = "common"
  )$x
) -
  0.05
ymax = max(
  plotMDS(
    v$EList,
    col = group_colours,
    dim.plot = c(1, 2),
    pch = 16,
    gene.selection = "common"
  )$y
) -
  0.05

par(mar = c(7.1, 5.1, 5.1, 5.1), xpd = TRUE)
plotMDS(
  v$EList,
  col = group_colours,
  dim.plot = c(1, 2),
  pch = 16,
  gene.selection = "common"
)
legend(
  "bottom",
  pch = 16,
  col = unique(group_colours),
  legend = levels(v$targets$arm),
  inset = c(0, -0.165),
  bty = "n",
  ncol = 2
)
text(
  x = xmax - 0.05,
  y = ymax,
  labels = bquote(R^2 == .(round(as.data.frame(mds_stat[2])[1, 3], 3)))
)
text(
  x = xmax - 0.05,
  y = ymax - 0.12,
  labels = bquote(p == .(round(as.data.frame(mds_stat[2])[1, 5], 3)))
)
```

There are two obvious clusters that are not due to arm in the MDS graph.

##### Change in abundances overtime

###### Data preparation

```{r}
#| eval: true
#| label: testing for de abundant pathways
counts <- kegg_be %>%
  dplyr::select(ko, sample, count) %>%
  pivot_wider(names_from = sample, values_from = count, values_fill = 0) %>%
  as.data.frame()

counts <- DGEList(counts)

counts$samples$timepoint <- as.factor(str_remove_all(
  kegg$timepoint[match(rownames(counts$samples), kegg$sample)],
  "1|3"
))

counts$samples$group <- as.factor(kegg$arm[match(
  rownames(counts$samples),
  kegg$sample
)])

counts$genes <- counts$genes %>%
  mutate(
    pathway_name = kegg$pathway_name[match(ko, kegg$ko)],
    pathway_class = kegg$pathway_class[match(ko, kegg$ko)]
  )

counts.keep <- rowSums(counts$counts == 0) == ncol(counts$counts)
counts <- counts[counts.keep == FALSE, ]

counts <- normLibSizes(counts)
counts <- estimateDisp(counts)
```

###### Contrast design

:::h0

$H_0$: KEGG pathways are not differentially abundant between baseline and endpoint

$H_0: \mu_{\text{Endpoint}} - \mu_{\text{Baseline}} = 0$

:::

::: h1

$H_1$: There is a difference between KEGG pathway abundance between baseline and endpoint
    $H_1a$: KEGG pathway abundance *increases* between baseline and endpoint
    $H_1b$: KEGG pathway abundance *decreases* between baseline and endpoint

Two-Sided Alternative:
$$
H_A: \mu_{\text{Endpoint}} - \mu_{\text{Baseline}} \neq 0
$$

Interpretation: KEGG pathways are differentially abundant between timepoints (could be higher or lower at endpoint)

One-Sided Alternatives:

Alternative 1: Increase at Endpoint
$$
H_A: \mu_{\text{Endpoint}} - \mu_{\text{Baseline}} > 0
$$

Interpretation: KEGG pathways are more abundant at endpoint than baseline

Alternative 2: Decrease at Endpoint
$$
H_A: \mu_{\text{Endpoint}} - \mu_{\text{Baseline}} < 0
$$

Interpretation: KEGG pathways are less abundant at endpoint than baseline
:::

The intercept represents the expected value at baseline for the reference group (CPM), while
`timepointEndpoint` represents the change from baseline to endpoint (i.e. the difference between timepoints averaged across groups)


```{r}
#| eval: true
#| label: KEGG pathway differences between baseline and endpoint
#Linear Model
#Contrasts
(design <- stats::model.matrix(~timepoint, data = counts$samples))
colnames(design) <- c("Intercept", "Endpoint")

contrasts <- makeContrasts(
  EvsB = Endpoint,
  levels = design
)
```

```{r}
#| eval: true
#| label: fig-EvsB-glmQLfit
#| fig-cap: KEGG pathway differences at endpoint compared to baseline based on a negative binomial generalized log-linear model with quasi-dispersions and empirical Bayes moderation.
fit <- glmQLFit(counts, design)

qlf <- glmQLFTest(fit, contrast = contrasts)
glm_results <- topTags(qlf, n = Inf) %>%
  as.data.frame() %>%
  arrange(ko)

glm_results %>%
  filter(abs(logFC) >= log2(2)) %>%
  ggplot(aes(y = reorder(pathway_name, logFC), x = logFC)) +
  geom_bar(
    position = position_dodge(),
    stat = "identity",
    aes(fill = pathway_class)
  ) +
  xlab(bquote(bold("Fold change (" ~ Log[2] ~ "FC )"))) +
  ylab(NULL) + #labels the y axis
  scale_fill_manual(values = pathway_class_col)
```

###### Empirical Bayes Statistics for Differential Expression 

```{r}
#| eval: true
#Transform raw count data to log2-counts per million (logCPM), estimate the mean-variance relationship and use this to compute appropriate observation-level weights. The data are then ready for linear modelling.

#If the DGEList object has been scale-normalized in edgeR (or before), then leave normalize.method="none" in voom, i.e., the logCPM values should not usually be re-normalized in the voom call.
#KEGG abundance values have not been normalised
# use updated edgeR voom procedure
v <- voomLmFit(counts, design, normalize.method = "none") #Transforms count data to log2-counts per million
vfit <- contrasts.fit(v, contrasts = contrasts)
efit <- eBayes(vfit) #estimation of QL dispersions
fc1.2 <- treat(vfit, fc = 1.2)
fc2 <- treat(vfit, fc = 2)
```

```{r}
#| eval: true

top_efit <- as.data.frame(topTable(efit, n = Inf)) %>%
  select(
    ko,
    EvsB = logFC,
    PValue = P.Value,
    FDR = adj.P.Val
  )

top_efit %>%
  arrange(ko) %>%
  mutate(EvsB_FC = 2^abs(EvsB)) %>%
  tibble()
```

```{r}
#| eval: true
#| label: tbl-fdr5
#| tbl-cap: OGUs that are significantly differentially abundant from zero for each contrast from a fit object (eBayes) containing p-values and test statistics. An FDR of 0.05 was used.

knitr::kable(summary(
  FDR5 <- decideTests(efit, adjust.method = "BH", p.value = 0.05, fc = 0)
))

```

```{r}
#| eval: true
#| label: tbl-fdr5-fc1.2
#| tbl-cap: OGUs with a fold change of 1.2 or more that are significantly differentially abundant for each contrast. An FDR of 0.05 was used.

knitr::kable(summary(
  FDR5.FC1.2 <- decideTests(fc1.2, adjust.method = "BH", p.value = 0.05)
))

```

```{r}
#| eval: true
#| label: tbl-fdr-fc2
#| tbl-cap: OGUs with a fold change of 2 or more that are significantly differentially abundant for each contrast. An FDR of 0.05 was used.

knitr::kable(summary(
  FDR5.FC2 <- decideTests(fc2, adjust.method = "BH", p.value = 0.05)
))

write.fit(
  fit = fc2,
  results = FDR5.FC2,
  file = file.path(
    filepath,
    paste0(prefix, "limma-write-fit-results-FDR5-FC2.csv")
  ),
  row.names = F
)
```

###### Raw abundance changes

Check results make sense by comparing the average abundance of features that have changed

```{r}
#| eval: true
#| label: tbl-total-da
#| tbl-cap: Based on raw counts alone, fold OGU differences at endpoint compared to baseline are shown.
(cnt <- kegg_be %>%
  group_by(ko, timepoint) %>%
  summarise(average_count = mean(count)) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = average_count,
    values_fill = 0
  ) %>%
  mutate(EvsB = `Endpoint` - `Baseline`) %>%
  mutate(
    fold_difference = fc(Endpoint, Baseline),
    difference = case_when(
      EvsB < 0 ~ "Decrease",
      EvsB > 0 ~ "Increase",
      EvsB == 0 ~ "None",
      TRUE ~ NA
    )
  ))
```

```{r}
#| eval: true
#| label: tbl-cnt-sum
#| tbl-cap: Based on raw counts alone, the total number of OGUs increases and decreases at endpoint compared to baseline are shown.
cnt %>%
  group_by(difference) %>%
  summarise(total = n()) %>%
  kable()

```

Numbers are very similar to those found in DA.

###### Volcano Plot

```{r}
#| eval: true
#| label: fig-vol-fdr5-fc2
#| fig-cap: Volcano plot of the limma-voom differential abundance multiple-testing results.
par(
  # Margins: bottom, left, top, right
  mar = c(5, 5, 4, 2),
  # White background
  bg = "white",
  # Horizontal axis labels
  las = 1,
  # L-shaped box
  bty = "l",
  # Bold axis labels
  font.lab = 2,
  # Smaller text size
  cex.axis = 0.8,
  # Default point size
  cex = 1.2
)
volcanoplot(
  fc2,
  coef = 1,
  names = fc2$genes$ko,
  highlight = nrow(fc2),
  style = "p-value"
)

grid(
  # Line type
  lty = "dotted",
  # Grid line color
  col = "gray90"
)

box(
  # Box type
  bty = "l",
  # Line width
  lwd = 2
)

abline(
  h = 1.3, # Horizontal line at -log10(0.05)
  col = "indianred", # Line color
  lty = 2, # Line type (dashed)
  lwd = 1 # Line width
)
```

###### Annotation of Significantly Differentially Abundant Pathways

```{r}
#| eval: true
#| label: identify diff abund pathways FRD5 FC>2

dekegg <- read.delim(
  file = file.path(
    filepath,
    paste0(prefix, "limma-write-fit-results-FDR5-FC2.csv")
  ),
  header = T
)

colnames(dekegg) <- str_remove(colnames(dekegg), "Genes.")

dekegg <- dekegg %>%
  dplyr::select(
    AveExpr,
    Coef = `Coef`,
    `t-statistic` = `t`,
    `P-value` = `P.value`,
    Result = `Results`,
    ko
  ) %>%
  mutate(
    fold_change = case_when(
      Coef >= 0 ~ 2^Coef,
      Coef < 0 ~ -(2^abs(Coef))
    ),
    # For validation
    raw_fc = cnt$fold_difference[match(ko, cnt$ko)]
  ) %>%
  dplyr::filter(Result != 0) %>%
  mutate(
    Result = case_when(
      Result == -1 ~ "Decreased abundance",
      Result == 1 ~ "Increased abundance"
    )
  )

dekegg <- dekegg %>%
  mutate(
    pathway_group = pathways$group[match(.$ko, pathways$id)],
    pathway_class = pathways$class[match(.$ko, pathways$id)],
    pathway_name = pathways$name[match(.$ko, pathways$id)]
  ) %>%
  mutate(pathway_class = coalesce(pathway_class, pathway_name)) %>%
  mutate(pathway_group = coalesce(pathway_group, pathway_class))
```

```{r}
#| eval: true
#| label: save-fdr5
#| results: asis

save_and_link(
  data = dekegg,
  filename = paste0(prefix, "limma-treat-significant-results-FDR5-FC2.csv"),
  description = "Differentially abundant pathways (FRD5 FC>2) identified by limma treat",
  directory_naming = "both"
)
```

::: {#fig-box}

Changes in differential abundance of KEGG pathways over time: `r length(unique(dekegg$ko))` KEGG pathways with at least 70% completeness were found to be differentially abundant at endpoint compared to baseline. All pathways changed a minimum of two-fold over time, and changes were not different between intervention arm. Pathway terms are color coded according to their parent class.

```{r}
#| eval: true
#| label: fig-ms-10
#| fig-height: 18
#| fig-width: 14

dekegg %>%
  ggplot(aes(y = reorder(pathway_name, Coef), x = Coef)) +
  geom_bar(
    position = position_dodge(),
    stat = "identity",
    aes(fill = pathway_class)
  ) +
  xlab(bquote(bold("Fold change (" ~ Log[2] ~ "FC )"))) +
  ylab(NULL) + #labels the y axis
  scale_fill_manual(values = pathway_class_col)

```

:::

###### Ordination Analysis and PERMANOVA of differences between timepoints

```{r}
#| eval: true
#| label: tbl-mds
#| tbl-cap: Results of PERMANOVA test investigating abundance differences between timepoints.
#unsupervised learning: MDS plot
group_colours <- counts$samples$timepoint
levels(group_colours) <- c("Baseline" = "#66C2A5", "Endpoint" = "#E78AC3")
group_colours <- as.character(group_colours)
df <- as.data.frame(t(counts$counts))
df <- cbind(df, group = counts$samples$timepoint)

#PERMANOVA testing differenital abundances between timepoints
mds_stat <- pairwiseAdonis::pairwise.adonis2(
  df[, 1:(ncol(df) - 1)] ~ group,
  data = df,
  method = 'eu'
) #returns non-adjusted p-values

#output table
knitr::kable(mds_stat[2])
```

```{r}
#| eval: true
#| label: tbl-mds-sum
#| tbl-cap: Results of PERMANOVA test investigating abundance differences between timepoints.

pv <- lapply(2:length(mds_stat), function(x) {
  w <- mds_stat[[x]][1, 4]
  y <- mds_stat[[x]][1, 5]
  z <- names(mds_stat)[x]
  data.frame(contrast = z, p = y, F.stat = w)
})
pv <- bind_rows(pv)
knitr::kable(
  pv %>%
    mutate(p.adj = round(p.adjust(p, method = "BH"), 4)) %>%
    mutate(
      sig = case_when(
        p.adj <= 0.0001 ~ "****",
        p.adj <= 0.001 ~ "***",
        p.adj <= 0.01 ~ "**",
        p.adj <= 0.05 ~ "*",
        p.adj > 0.05 ~ "ns"
      )
    )
)
```

The p-value ($Pr(>F)$) for `r counts$samples$timepoint` is `r pv$p`, which is greater than the conventional significance level of 0.05. This suggests that we cannot reject the null hypothesis, implying that there is no significant overall difference in microbial pathway abundance between baseline and endpoint.

The `plotMDS()` function accepts "any data object that can be coerced to a matrix of log-expression values, for example an ExpressionSet or an EList" [@Gentleman2025]. If the `edgeR::voomLmFit()` function was used the `EList` object can be obtained using `fit$EList`

```{r}
#| eval: true
#| label: MDS plot comparing timepoints

xmax = max(
  plotMDS(
    v$EList,
    col = group_colours,
    dim.plot = c(1, 2),
    pch = 16,
    gene.selection = "common"
  )$x
) -
  0.05
ymax = max(
  plotMDS(
    v$EList,
    col = group_colours,
    dim.plot = c(1, 2),
    pch = 16,
    gene.selection = "common"
  )$y
) -
  0.05

par(mar = c(7.1, 5.1, 5.1, 5.1), xpd = TRUE)
plotMDS(
  v$EList,
  col = group_colours,
  dim.plot = c(1, 2),
  pch = 16,
  gene.selection = "common"
)
legend(
  "bottom",
  pch = 16,
  col = unique(group_colours),
  legend = levels(v$targets$timepoint),
  inset = c(0, -0.165),
  bty = "n",
  ncol = 2
)
text(
  x = xmax - 0.05,
  y = ymax,
  labels = bquote(R^2 == .(round(as.data.frame(mds_stat[2])[1, 3], 3)))
)
text(
  x = xmax - 0.05,
  y = ymax - 0.12,
  labels = bquote(p == .(round(as.data.frame(mds_stat[2])[1, 5], 3)))
)
```

There are two obvious clusters that are not due to timepoint in the MDS graph. Another PERMANOVA test to 

```{r}
#| eval: false
#| label: tbl-mds-time
#| tbl-cap: Results of PERMANOVA test investigating abundance differences between timepoints.
#unsupervised learning: MDS plot
group_colours <- v$targets$arm
levels(group_colours) <- c("Baseline" = "#66C2A5", "Endpoint" = "#E78AC3")
group_colours <- as.character(group_colours)
df <- as.data.frame(t(v$EList$E))
df$group <- v$targets$arm[match(rownames(df), rownames(v$targets))]

#PERMANOVA testing differenital abundances between timepoints
mds_stat <- pairwiseAdonis::pairwise.adonis2(
  df[, 1:(ncol(df) - 1)] ~ group,
  data = df,
  method = 'eu'
) #returns non-adjusted p-values

#output table
knitr::kable(mds_stat[2])
```

```{r}
#| eval: false
#| label: cluster analysis

metaRDA <- bac_base_end_meta %>%
  select(
    timepoint,
    arm,
    chao1,
    diversity_shannon,
    evenness_pielou,
    pd,
    avg_iron,
    avg_zinc,
    avg_bajra,
    agemonths_base,
  )

metaRDA <- bac_base_end_meta %>%
  select(
    `Age (months)` = agemonths_base,
    Female = sexf,
    `Intervention arm` = trt,
    `Birth mode vaginal (vs. Caesarean)` = del_type,
    `Birth weight (kg)` = bwt,
    `Current weight (kg)` = cwt,
    `Mid-upper arm circumference (cm)` = muac_avg,
    `Head circumference (cm)` = hc_avg,
    `Head circumference-for-age Z-score` = zhc,
    `Length (cm)` = ln_avg,
    `Length-for-age Z-score (LAZ)` = zlen,
    `Stunting (LAZ < -2)` = stunt_base,
    `Weight-for-age Z-score (WAZ) ` = zwei,
    `Underweight (WAZ < -2)` = underwt_base,
    `Weight-for-length Z-score (WLZ) ` = zwfl,
    `Wasting (WLZ < -2)` = wast_base,
    `Ferritin (ng/ml)` = fer_base,
    `Iron deficiency (<12 ng/ml) ` = fer_def12_base,
    `Ferritin (ng/ml)* adjusted for CRP` = fer_crpadjbr, # using BRINDA
    `Iron deficiency (<12 ng/ml)* adjusted for CRP` = fer_crpbr_def12, # using BRINDA
    `Zinc (µg/dL)` = zinc,
    `Zinc deficiency (<70 µg/dL) ` = zinc_def70, # <70 µg/dL
    `Hemoglobin (g/dl)` = hb_base,
    `Anemia (hemoglobin < 11 g/dl) ` = anemia_base,
    `C-reactive protein (CRP) ≥ 5 mg/L ` = crp_elev5,
    `Diarrhea today or within past 4 weeks ` = freq3diarr, # 1,2&3
    `Fever today or within past 4 weeks ` = freqfever, # 1,2&3
    `Cough today or within past 4 weeks` = freqcough, # 1,2&3
    `Observed features` = observed,
    `Chao 1 Index` = chao1,
    `Shannon Entropy` = diversity_shannon,
    Evenness = evenness_pielou,
    `Faith's Phylogenetic Diversity` = pd
  )

rda(X = da_mat, Y = metaRDA)

da <- data.frame(v$EList$genes, v$EList$E)
rownames(da) <- NULL

da_mat <- da %>% column_to_rownames("ko") %>% t() #is now a data matrix because of t() function
rownames(da_mat) <- str_remove(rownames(da_mat), "SO_8465_")
df <- data.frame(Timepoint = v$targets$timepoint, Arm = v$targets$arm)

row_annotation <- HeatmapAnnotation(
  df = df,
  show_legend = T,
  col = list(
    timepoint = c("Baseline" = "#66C2A5", "Endpoint" = "#E78AC3"),
    arm = c("CPM" = "#FC8D62", "FeZnCPM" = "#8DA0CB")
  ),
  which = "row",
  annotation_legend_param = list(
    title_gp = gpar(fontsize = 15, fontface = "bold"),
    labels_gp = gpar(fontsize = 15),
    grid_height = unit(8, "mm"),
    grid_width = unit(4, "mm")
  )
)

Heatmap(
  da_mat,
  width = unit(0.5, "npc"),
  cluster_rows = T,
  show_row_dend = T,
  col = c('#0571b0', '#92c5de', '#f4a582', '#ca0020'),
  row_names_side = "left",
  row_names_gp = gpar(cex = 1, fontface = "bold"),
  show_column_names = F,
  show_heatmap_legend = T,
  right_annotation = row_annotation,
  heatmap_legend_param = list(
    title = "log2 fold change",
    title_gp = gpar(fontsize = 15, fontface = "bold"),
    labels_gp = gpar(fontsize = 15),
    grid_height = unit(8, "mm"),
    grid_width = unit(4, "mm")
  ),
)

```

#### KEGG Differential Abundace with ANCOM-BC

##### Change in abundances overtime (ignoring arm)

```{r}
#| eval: true
#| label: data-prep
#| fig-cap: NA
#| tbl-cap: NA

otu <- kegg_be %>%
  select(ko, sample, count) %>%
  pivot_wider(names_from = sample, values_from = count, values_fill = 0) %>%
  column_to_rownames(var = "ko") %>%
  otu_table(taxa_are_rows = TRUE)

meta <- bac_base_end_meta %>%
  column_to_rownames(var = "sample_names") %>%
  sample_data()

kegg_phy <- phyloseq(otu, meta)
```

```{r}
#| eval: true
#| label: ancombc
#| fig-cap: NA
#| tbl-cap: NA

time_results <- ancombc2(
  kegg_phy,
  taxa_are_rows = TRUE,
  assay_name = "abundance",
  rank = tax_level,
  tax_level = NULL,
  fix_formula = "timepoint",
  p_adj_method = "BH",
  pseudo = 0,
  pseudo_sens = TRUE,
  prv_cut = 0.1,
  lib_cut = 0,
  s0_perc = 0.05,
  group = "timepoint",
  struc_zero = TRUE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 1,
  verbose = TRUE,
  global = FALSE,
  pairwise = FALSE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(tol = 0.01, max_iter = 20, verbose = FALSE),
  em_control = list(tol = 1e-05, max_iter = 100),
  lme_control = lme4::lmerControl(),
  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
  trend_control = list(contrast = NULL, node = NULL, solver = "ECOS", B = 100)
)

time_prim <- time_results$res
```

```{r}
#| eval: true
#| label: save
#| results: asis

save_and_link(
  data = time_prim,
  filename = paste0(prefix, "ancombc.csv"),
  description = "Differential abundances between Endpoint and Baseline",
  directory_naming = "both"
)
```

```{r}
#| eval: true
#| label: tbl-EvsB
#| tbl-cap: Differentially abundant KEGG terms at Endpoint Compared to Baseline. Significant terms were determined to be so if they had a FDR adjusted p-value $\leq$ 0.05 and a fold change of 2 or more
tmp <- time_prim %>%
  select(ko = taxon, ends_with("_timepointEndpoint")) %>%
  filter(passed_ss_timepointEndpoint == TRUE)

colnames(tmp) <- str_remove(colnames(tmp), "_timepointEndpoint")

EvsB <- tmp %>%
  filter(abs(lfc) >= log(2) & q <= 0.05) %>%
  mutate(
    pathway_group = pathways$group[match(.$ko, pathways$id)],
    pathway_class = pathways$class[match(.$ko, pathways$id)],
    pathway_name = pathways$name[match(.$ko, pathways$id)]
  ) %>%
  mutate(pathway_class = coalesce(pathway_class, pathway_name)) %>%
  mutate(pathway_group = coalesce(pathway_group, pathway_class))

kable(EvsB)
```

```{r}
#| eval: true
#| label: fig-EvsB
#| fig-cap: Differentially abundant KEGG terms at Endpoint Compared to Baseline. Significant terms were determined to be so if they had a FDR adjusted p-value $\leq$ 0.05 and a fold change of 2 or more
plotDAvol(
  tmp,
  title = "Differentially abundant KEGG terms at Endpoint Compared to Baseline",
  p.fdr = q
)

```

```{r}
#| eval: true
#| label: fig-ancom-anno
#| fig-cap: Differentially abundant KEGG pathways at Endpoint Compared to Baseline
EvsB %>%
  ggplot(aes(y = reorder(pathway_name, lfc), x = lfc)) +
  geom_bar(
    position = position_dodge(),
    stat = "identity",
    aes(fill = pathway_class)
  ) +
  xlab(bquote(bold("Fold change (" ~ Log[2] ~ "FC )"))) +
  ylab(NULL) + #labels the y axis
  scale_fill_manual(values = pathway_class_col)

```

## Results Summary

```{r}
#| eval: true
#| echo: false
#| results: asis

list_all_results()
```

```{r}
#| eval: true
#| echo: false
#| results: asis
generate_results_section(directory_naming = "both")
```

## Appendix

### ANCOM-BC Fold Change 1.2

```{r}
#| eval: true
#| label: ancom-fc-1.2
#| fig-cap: NA
#| tbl-cap: NA
tmp %>%
  filter(abs(lfc) >= log(1.2) & q <= 0.05 & passed_ss == TRUE) %>%
  mutate(
    pathway_group = pathways$group[match(.$ko, pathways$id)],
    pathway_class = pathways$class[match(.$ko, pathways$id)],
    pathway_name = pathways$name[match(.$ko, pathways$id)]
  ) %>%
  mutate(pathway_class = coalesce(pathway_class, pathway_name)) %>%
  mutate(pathway_group = coalesce(pathway_group, pathway_class)) %>%
  ggplot(aes(y = reorder(pathway_name, lfc), x = lfc)) +
  geom_bar(
    position = position_dodge(),
    stat = "identity",
    aes(fill = pathway_class)
  ) +
  xlab(bquote(bold("Fold change (" ~ Log[2] ~ "FC )"))) +
  ylab(NULL) + #labels the y axis
  scale_fill_manual(values = pathway_class_col)
```

```{r}
#| eval: true
#| echo: false
#| label: info

print(sessionInfo(), locale = FALSE)
```
