# Investigating changes in OGU abundance over time using Zero-inflated Regression Models with Random Effects

```{r setup}
#| echo: false
#| eval: true
#| label: setup

# This initial code block loads custom utility functions and environment variables
# These are stored in separate files to keep the main analysis clean and organized
suppressMessages({
  source("utils/r-utils.r")
  source("scripts/OGeeU.r")
  source("scripts/zeroGUS.r")

  source("_environment.required")
  load("data/seqdata.RData")

  library(glmmTMB)
  library(DHARMa)
  library(broom)
  library(patchwork)
  library(emmeans)

  library(promicrobialTools)

  dir.create(
    file.path(RESULTS, "EvsB-OGU-ZIBR"),
    showWarnings = FALSE,
    recursive = TRUE
  )
})
```

## Prepare Outcome and Predictor Variables

```{r data-prep}
#| eval: true
#| label: data prep
#| results: asis

data <- prepare_ogu_data(
  absAbund,
  filter_prevalence = TRUE,
  prevalence_cutoff = 0.4,
  filter_abundance = FALSE
)

data$taxa_data <- data$taxa_data %>%
  mutate(Phylum = ifelse(str_detect(Phylum, "Bacillota"), "Bacillota", Phylum))

outcome <- data$ogu_data
rownames(outcome) <- str_replace(
  str_replace(str_remove(rownames(outcome), "_G"), "_B_B", "_B"),
  "_B_E",
  "_E"
)

predictors <- prepare_predictors(data$meta_data)
predictors <- predictors %>%
  mutate(arm = case_when(arm == "CPM" ~ 0, arm == "FeZnPM" ~ 1)) %>%
  rownames_to_column(var = "sample") %>%
  mutate(
    baseline_sample = str_replace(sample, "_G_E|_B_E|_E|_G_B|_B_B", "_B"),
    sample = str_remove(sample, "_G"),
  ) %>%
  mutate(sample = str_replace(sample, "_B_B", "_B")) %>%
  mutate(sample = str_replace(sample, "_B_E", "_E"))
```

## Determining Direction of Change in Abundance over Time

```{r}
#| eval: true
#| label: tbl-abundance-changes
#| tbl-cap: Count Changes between Baseline and Endpoint

joined_data <- predictors %>%
  left_join(rownames_to_column(outcome, "sample"), by = "sample") %>%
  pivot_longer(
    cols = !c(
      "sample",
      "subject",
      "arm",
      "timepoint",
      "time",
      "age_B",
      "baseline_sample",
      "zlen"
    ),
    names_to = "outcome_id"
  )

joined_data_changes <- joined_data %>%
  group_by(outcome_id, timepoint) %>%
  summarise(average_count = mean(value)) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = average_count,
    values_fill = 0
  ) %>%
  mutate(EvsB = Endpoint - Baseline) %>%
  mutate(
    difference = case_when(
      EvsB < 0 ~ "Decrease",
      EvsB > 0 ~ "Increase",
      EvsB == 0 ~ "None",
      TRUE ~ NA
    )
  ) %>%
  group_by(difference) %>%
  summarise(total = n()) %>%
  kable()
```

The purpose of these models will be to determine if these changes are significant.

## Variable Characteristics

### Predictor Variables

```{r}
#| eval: true
#| label: predictor-describe
#| results: asis

Hmisc::describe(predictors)

```

### OGU Outcome Variables

```{r describe}
#| eval: true
#| label: outcome-describe

Hmisc::describe(outcome, condense = TRUE)

```

```{r zero-hist}
#| eval: true
#| label: fig-outcome-zeroProp
#| fig-cap: Distribution of zero proportions across variables

analyze_zeros(outcome, "density")
```

```{r disp}
#| eval: true
#| label: fig-outcome-dispersion
#| fig-cap: Distribution of dispearsion, defined as the variance-to-mean ratio. A log10 scale is used.

outdisp <- varDisp(outcome, threshold = 2, verbose = FALSE)

plot(outdisp, col = "#8CAAEE")

```

Generally, if the dispersion ratio is > 2 (0.3 on the log_10 scale), fitting a Poisson model will not accurately model the data (Poisson assumes that variance $\approx$ mean).

|      Model       |          Continuous component           | ZI component |                   Test                    |
|:----------------:|:---------------------------------------:|:------------:|:-----------------------------------------:|
|   noZi_models    |                  y ~ 1                  |     ~ 0      |  Is the zero-inflated component needed?   |
|   null_models    |                  y ~ 1                  |     ~ 1      | Tests if predictors affect zero-inflation |
| noZi_rand_models |         y ~ 1 + (1 \| subject)          |     ~ 0      |      Is the random intercept needed?      |
| null_rand_models |         y ~ 1 + (1 \| subject)          |     ~ 1      |                                           |
|        m1        | y ~ arm * time + age_B + (1 \| subject) | ~ arm * time |                                           |
|                  |                                         |              |                                           |

: List of model formulae tested

## Testing different distributions

The following distribution families were used to fit a zero-inflated model:

- Poisson
- Simple negative binomial (linear parameterization) [@Hardin2007]
- Complex negative binomial (quadratic parameterization) [@Hardin2007]
- Gaussian
- Log normal

```{r fit}
#| eval: false
nb2_results <- genZI(
  outcome,
  predictors,
  formula = "time + arm + age_B + arm:time + (1 | subject)",
  zi = "~ arm * time",
  family = "nbinom2"
)

nb1_results <- genZI(
  outcome,
  predictors,
  formula = "time + arm + age_B + arm:time + (1 | subject)",
  zi = "~ arm * time",
  family = "nbinom1"
)

poi_results <- genZI(
  outcome,
  predictors,
  formula = "time + arm + age_B + arm:time + (1 | subject)",
  zi = "~ arm * time",
  family = "poisson"
)

gau_results <- genZI(
  outcome,
  predictors,
  formula = "time + arm + age_B + arm:time + (1 | subject)",
  zi = "~ arm * time",
  family = "gaussian"
)

lognorm_results <- genZI(
  outcome,
  predictors,
  formula = "time + arm + age_B + arm:time + (1 | subject)",
  zi = "~ arm * time",
  family = "lognormal"
)
```

```{r save}
#| eval: false
#| echo: false
saveRDS(
  nb2_results,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "genZI-OGU-nb2_results.rds")
)

saveRDS(
  nb1_results,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "genZI-OGU-nb1_results.rds")
)

saveRDS(
  poi_results,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "genZI-OGU-poi_results.rds")
)

saveRDS(
  gau_results,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "genZI-OGU-gau_results.rds")
)

saveRDS(
  lognorm_results,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "genZI-OGU-lognorm_results.rds")
)
```

```{r load}
#| eval: true
#| echo: false
nb2_results <- readRDS(file.path(
  RESULTS,
  "EvsB-OGU-ZIBR",
  "genZI-OGU-nb2_results.rds"
))

nb1_results <- readRDS(file.path(
  RESULTS,
  "EvsB-OGU-ZIBR",
  "genZI-OGU-nb1_results.rds"
))

poi_results <- readRDS(file.path(
  RESULTS,
  "EvsB-OGU-ZIBR",
  "genZI-OGU-poi_results.rds"
))

gau_results <- readRDS(file.path(
  RESULTS,
  "EvsB-OGU-ZIBR",
  "genZI-OGU-gau_results.rds"
))

lognorm_results <- readRDS(file.path(
  RESULTS,
  "EvsB-OGU-ZIBR",
  "genZI-OGU-lognorm_results.rds"
))
```

### Model diagnostic failures

```{r}
#| eval: true
#| label: list

list <- list(
  gau = gau_results,
  lognorm = lognorm_results,
  poi = poi_results,
  nb1 = nb1_results,
  nb2 = nb2_results
)
```

```{r}
#| eval: true
#| label: diagnose-mods
#| include: false

poor_fits <- lapply(1:length(list), function(x) {
  d <- diagnoseZI(list[[x]]$all_models)
  d$dist <- names(list)[x]

  return(d)
})

poor_fits <- data.table::rbindlist(poor_fits)
```

### Comparing Model AIC and BIC values

```{r}
#| eval: true
#| label: summaries

summaries <- list(
  gau = gau_results$model_summary,
  lognorm = lognorm_results$model_summary,
  poi = poi_results$model_summary,
  nb1 = nb1_results$model_summary,
  nb2 = nb2_results$model_summary
)

summaries <- data.table::rbindlist(summaries, idcol = "dist")

summaries <- left_join(summaries, poor_fits, by = c("outcome_id", "dist"))
```

```{r}
#| eval: true
#| label: any-errors

if (any(summaries$error_message != "")) {
  print("Some models returned an error message")
}

```

```{r}
#| eval: true
#| label: diag-compare
summaries <- summaries %>%
  filter(pass == TRUE) %>%
  select(!c(error_message, message)) %>%
  mutate(index = 1:nrow(.)) %>% #make it easier to distinguish between outcome ids
  pivot_longer(
    cols = !c(index, outcome_id, dist, pass),
    names_to = "criterion",
    values_to = "value"
  )
```

```{r}
#| eval: true
#| label: best-AIC

aic <- summaries %>%
  filter(criterion == "AIC") %>%
  group_by(criterion, outcome_id) %>%
  arrange(value) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, minAIC = value, distAIC = dist)
```

```{r}
#| eval: true
#| label: best-BIC

bic <- summaries %>%
  filter(criterion == "BIC") %>%
  group_by(criterion, outcome_id) %>%
  arrange(value) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, minBIC = value, distBIC = dist)

```

```{r}
#| eval: true
#| label: tbl-best-fits
#| tbl-cap: Best fit models based on AIC and BIC

(bestDist <- left_join(aic, bic) %>%
  mutate(dist = ifelse(distAIC == distBIC, distAIC, "Undetermined")))
```

### Checking for deviations in model assumptions

::: {.callout-note}

The larger the dispersion value, $\theta$ in `nbinom2` models, the less variance there is in the model [@Brooks2017].
:::

`dharmaDisp` represents the dispersion ratio, which should ideally be close to 1

```{r}
#| eval: true
#| label: tbl-best-disp
#| tbl-cap: lowest disp value for models that passed diagnostics

dispRatio <- summaries %>%
  filter(criterion == "dharmaDisp") %>%
  group_by(criterion, outcome_id) %>%
  mutate(skew = abs(value - 1)) %>%
  arrange(skew) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, bestDisp = value, distDisp = dist)
```

```{r}
#| eval: true
#| label: tbl-pvals
#| tbl-cap: NA

pvals <- left_join(
  bestDist,
  summaries,
  by = c("index", "outcome_id", "dist")
) %>%
  filter(str_detect(criterion, "Pvalue"), value <= 0.05) %>%
  arrange(index)
```

`r n_distinct(pvals$outcome_id)` models have a significant test result making the model fit problematic

### Final Model Distributions

```{r}
#| eval: true
#| label: best-dist-filtered
#| fig-cap: NA
#| tbl-cap: NA

bestDist_filtered <- bestDist %>%
  filter(!outcome_id %in% pvals$outcome_id)
```

## Fitting models to best distribution

#### Zero-inflated Negative Binomial 1 Models

```{r}
#| eval: true
#| label: nbinom1 models

nb1_models <- keep_at(
  nb1_results$all_models,
  filter(bestDist_filtered, dist == "nb1")$outcome_id
)

nb1_unfit <- diagnoseZI(nb1_models)
```

Are there any models that fail diagnostics: `r any(nb1_unfit$pass == FALSE)`

```{r}
#| eval: true
#| label: nb1-summary

nb1_summary <- data.frame(
  outcome_id = names(nb1_models),
  continuous_distribution = "nbinom1",
  continuous_link = "log",
  covaraince_structure = "unstructured",
  zero_distribution = "binary",
  zero_link = "logit",
  optimizer = "nlminb"
)
```

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for nbinom1
nb1_dharma_results <- process_dharma_results(nb1_models)

kable(nb1_dharma_results$summary)

```

#### Zero-inflated Negative Binomial Polynomial Models

```{r}
#| eval: true
#| label: nbinom2 models

nb2_models <- keep_at(
  nb2_results$all_models,
  filter(bestDist_filtered, dist == "nb2")$outcome_id
)

nb2_unfit <- diagnoseZI(nb2_models)
```

Are there any models that fail diagnostics: `r any(nb2_unfit$pass == FALSE)`

```{r}
#| eval: true
#| label: nb2-summary

nb2_summary <- data.frame(
  outcome_id = names(nb2_models),
  continuous_distribution = "nbinom2",
  continuous_link = "log",
  covaraince_structure = "unstructured",
  zero_distribution = "binary",
  zero_link = "logit",
  optimizer = "nlminb"
)
```

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for nbinom1
nb2_dharma_results <- process_dharma_results(nb2_models)

kable(nb2_dharma_results$summary)

```

#### Zero-inflated Poisson Models

```{r}
#| eval: true
#| label: poi models

poi_models <- keep_at(
  poi_results$all_models,
  filter(bestDist_filtered, dist == "poi")$outcome_id
)

poi_unfit <- diagnoseZI(poi_models)
```

Are there any models that fail diagnostics: `r any(poi_unfit$pass == FALSE)`

```{r}
#| eval: true
#| label: poi-summary

poi_summary <- data.frame(
  outcome_id = names(poi_models),
  continuous_distribution = "poisson",
  continuous_link = "log",
  covaraince_structure = "unstructured",
  zero_distribution = "binary",
  zero_link = "logit",
  optimizer = "nlminb"
)
```

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for Poisson models
poi_dharma_results <- process_dharma_results(poi_models)

kable(poi_dharma_results$summary)

```

#### Zero-inflated Gaussian Models

```{r}
#| eval: true
#| label: gau unfit models

gau_models <- keep_at(
  gau_results$all_models,
  filter(bestDist_filtered, dist == "gau")$outcome_id
)

gau_unfit <- diagnoseZI(gau_models)
```

Are there any models that fail diagnostics: `r any(gau_unfit$pass == FALSE)`

```{r}
#| eval: true
#| label: poi-summary

gau_summary <- data.frame(
  outcome_id = names(poi_models),
  continuous_distribution = "gaussian",
  continuous_link = "log",
  covaraince_structure = "unstructured",
  zero_distribution = "binary",
  zero_link = "logit",
  optimizer = "nlminb"
)
```

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for Poisson models
gau_dharma_results <- process_dharma_results(gau_models)

kable(gau_dharma_results$summary)

```

#### Zero-inflated Lognormal Models

```{r}
#| eval: true
#| label: lognorm unfit models

lognorm_models <- keep_at(
  lognorm_results$all_models,
  filter(bestDist_filtered, dist == "lognorm")$outcome_id
)

```

There were no models that could be fit with a log normal distribution

## Significant OGUs

```{r}
#| eval: true
#| label: model-ogus
nb1_ogu_models <- nb1_results$adj_results %>%
  filter(outcome_id %in% names(nb1_models))

nb2_ogu_models <- nb2_results$adj_results %>%
  filter(outcome_id %in% names(nb2_models))

poi_ogu_models <- poi_results$adj_results %>%
  filter(outcome_id %in% names(poi_models))

gau_ogu_models <- gau_results$adj_results %>%
  filter(outcome_id %in% names(gau_models))
```

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: Summary table of OGUs successfully fit by a model, and their corresponding coefficient p-values

poi_summary <- left_join(poi_summary, poi_ogu_models)
nb1_summary <- left_join(nb1_summary, nb1_ogu_models)

nb2_summary <- left_join(nb2_summary, nb2_ogu_models)
gau_summary <- left_join(gau_summary, gau_ogu_models)

kable(
  summary <- rbind(
    nb1_summary,
    nb2_summary,
    poi_summary,
    gau_summary
  )
)
```

```{r}
#| eval: true
#| label: save-tbl-summary
#| echo: false

write.csv(
  summary,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "zero-inflated-ogu-model-summaries.csv"),
  row.names = FALSE
)
```

Of the `r ncol(outcome)` OGUs that passed filtering (`r ncol(outcome)/ncol(data$ogu_data)`), `r nrow(summary)` were sucessfully fit to a zero-inflated model (`r nrow(summary)/ncol(outcome)*100`%)

```{r}
#| eval: true
#| label: zi-sig

ziSig <- identify_significant_ogus(
  summary,
  data$taxa_data
)

colnames(ziSig)[colnames(ziSig) == "rowname"] <- "outcome_id"
```

## Individual abundance changes

```{r}
#| eval: true
#| label: tmp
#| collapse: true
#| tbl-cap: Individual abundance changes for each OGU

tmp <- joined_data %>%
  dplyr::filter(outcome_id %in% ziSig$outcome_id) %>%
  group_by(sample, outcome_id) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = value,
    values_fill = 0
  ) %>%
  mutate(EvsB = Endpoint - Baseline) %>%
  mutate(
    difference = case_when(
      EvsB < 0 ~ "Decrease",
      EvsB > 0 ~ "Increase",
      EvsB == 0 ~ "None",
      TRUE ~ NA
    )
  )

tmp %>%
  group_by(outcome_id, difference) %>%
  summarise(total = n()) %>%
  kable()
```

### Average abundance changes

```{r}
#| eval: true
#| label: tbl-average-sig-changes

aveChange <- joined_data %>%
  dplyr::filter(outcome_id %in% ziSig$outcome_id) %>%
  group_by(outcome_id, timepoint) %>%
  summarise(average_count = mean(value)) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = average_count,
    values_fill = NA
  ) %>%
  mutate(EvsB = Endpoint - Baseline) %>%
  mutate(
    difference = case_when(
      EvsB < 0 ~ "Decrease",
      EvsB > 0 ~ "Increase",
      EvsB == 0 ~ "None",
      TRUE ~ NA
    )
  )

aveChange %>%
  group_by(difference) %>%
  summarise(total = n()) %>%
  kable()
```

### Demonstrating the importance of random effects

OGU G012221805 has an average abundance change of zero, so why is the EvsB term significant? Random effects!

```{r}
#| eval: true
#| label: fig-indv-changes
#| fig-cap: Changes in count abundance for OGU G012221805, showing the clear impact of individual random effects and the importance of including such in a model. Note the black dotted line would be the standard regression line indicating no change over time.

joined_data %>%
  filter(outcome_id == "G012221805") %>%
  group_by(subject) %>%
  ggplot(aes(x = time, y = value, )) +
  geom_point() +
  geom_line(
    aes(
      group = subject,
      col = as.factor(arm)
    ),
    linewidth = 1
  ) +
  geom_smooth(
    method = "lm",
    se = F,
    color = "black",
    formula = y ~ x,
    linetype = 'dotted',
    linewidth = 3
  ) +
  xlab("Time Since Feed Start (days)") +
  ylab("Count") +
  scale_color_manual(values = armColours) +
  labs(color = "Intervention Arm")

```

## Save Significant Model Results

```{r}
#| eval: true
ziSig <- left_join(ziSig, aveChange, by = "outcome_id")

write.csv(
  ziSig,
  file.path(
    RESULTS,
    "EvsB-OGU-ZIBR",
    "significant-ogus-zero-inflated-model.csv"
  ),
  row.names = FALSE
)
```

## Zero-inflated component

```{r}
#| eval: true
#| label: zi-comp
#| fig-cap: NA
#| tbl-cap: NA

zi <- lapply(
  list(nb1_models, nb2_models, poi_models, gau_models),
  function(l) {
    lapply(l, function(x) {
      summary(x)$coefficients$zi[, 4]
    })
  }
)

#FIX make naming automatic using indexing
zip <- bind_rows(
  lapply(zi, function(model_type) {
    bind_rows(model_type, .id = "outcome_id")
  }),
  .id = "model_type"
) %>%
  # Add better names for model types
  mutate(
    model_type = case_when(
      model_type == "1" ~ "nb1",
      model_type == "2" ~ "nb2",
      model_type == "3" ~ "poi",
      model_type == "4" ~ "gau",
      TRUE ~ model_type
    )
  )

if (any(zip <= 0.05)) {
  zip_sig <- identify_significant_ogus(zip, data$taxa_data)
}
```

## Using EMMEANS to determine significant changes in taxa abundances

```{r}
#| eval: true
#| label: fit-mods

fits <- list(nb1_models, nb2_models, poi_models, gau_models)

fits <- list_c(fits)
```

### Conditional Model EMMs

#### OGUs Estimated marginal means

The below estimates the mean for each `arm` at each `timepoint`, controlling for age at baseline

```{r}
#| eval: true
#| label: tbl-emms
#| tbl-cap: Estimated marginal means for OGU abundances within each intervention arm

emm <- lapply(fits, function(x) {
  # Get EMMs for arm at each time point
  emmeans(
    x,
    ~ arm | time,
    at = list(time = c(0, 1)),
    adjust = "sidak",
    weights = "proportional",
    type = "scale"
  )
})

```

##### OGU EMM differences between intervention arm at baseline and endpoint

```{r}
#| eval: true
#| label: tbl-emm-pairs
#| tbl-cap: Comparison of abundance marginal mean estimates (EMMs) betweeen intervention arm at baseline and endpoint

emm_bind(emm_pairs <- lapply(emm, pairs, reverse = TRUE))

emm_pairs_df <- emm_bind(emm_pairs, kable = F)

sigArms <- emm_pairs_df %>%
  filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(
    change = ifelse(estimate < 0, "Decrease", "Increase"),
    time = ifelse(time == 1, "Endpoint", "Basepoint")
  )
```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint

plotPhylaEMMs(sigArms) +
  facet_wrap(~time, scales = "free") +
  theme(axis.text.y = element_text(size = 10))
```

##### Effect Size

The default value used to standardise the effect size is the model's residual standard deviation (sigma; within-group variation). However, this may underestimate effect sizes in mixed models where combined variance may be more accurate. Here, the random effect variance is used to standardise effect sizes.

```{r}
#| eval: true
#| label: sigArm-eff
#| fig-cap: NA
#| tbl-cap: Effect sizes calculated as a standardized mean difference similar to Cohen's d.

emmEff <- lapply(1:length(emm_pairs), function(i) {
  eff_size(
    emm_pairs[[i]],
    sigma = attr(VarCorr(fits[[i]])$cond$subject, "stddev"),
    edf = df.residual(fits[[i]]),
    method = "identity"
  )
})

```

#### OGU EMM changes over time within each intervention arm

```{r}
#| eval: true
#| label: tbl-emm-arm-trend
#| tbl-cap: Rate of change per unit increase of abundance marginal mean estimates over time. The time.trend column represents the slope of the relationship with time.

emm_bind(
  trend <- lapply(fits, function(x) {
    emtrends(x, ~ arm | time, var = "time")
  })
)
```

```{r}
#| eval: true
#| label: tbl-emm-arm-trend-compare
#| tbl-cap: Comparison of the rates of change from baseline to endpoint in abundance EMMs between intervention arms.

emm_bind(slopes <- lapply(trend, pairs, reverse = TRUE, type = "response"))

slopes_df <- emm_bind(slopes, kable = F)

sigArmsTime <- slopes_df %>%
  filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(time = ifelse(time == 1, "Endpoint", "Basepoint"))
```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint

plotPhylaEMMs(sigArmsTime) +
  theme(axis.text.y = element_text(size = 10))
```

##### Effect Size

### Zero-inflated Componet EMMs

```{r}
#| eval: true
#| label: tbl-emms
#| tbl-cap: Estimated marginal means for OGU abundances within each intervention arm

emm_zi <- lapply(fits, function(x) {
  # Get EMMs for arm at each time point
  emmeans(
    x,
    ~ arm | time,
    at = list(time = c(0, 1)),
    adjust = "sidak",
    weights = "proportional",
    type = "scale",
    component = "zi"
  )
})

# Convert to probability scale
zi_probs <- lapply(emm_zi, regrid, transform = "logit")
```

```{r}
#| eval: true
#| label: odds-ratio
#| tbl-cap: Odds-ratio for zero-inflation

ZIodds <- lapply(fits, function(model) {
  exp(fixef(model)$zi)
})
```


##### OGU EMM differences between intervention arm at baseline and endpoint

```{r}
#| eval: true
#| label: tbl-emm-pairs
#| tbl-cap: Comparison of abundance marginal mean estimates (EMMs) betweeen intervention arm at baseline and endpoint

emm_bind(emm_zi_pairs <- lapply(emm_zi, pairs, reverse = TRUE))

emm_zi_pairs_df <- emm_bind(emm_zi_pairs, kable = F)

ZIsigArms <- emm_zi_pairs_df %>%
  filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(
    change = ifelse(estimate < 0, "Decrease", "Increase"),
    time = ifelse(time == 1, "Endpoint", "Basepoint")
  )
```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint

plotPhylaEMMs(ZIsigArms) +
  facet_wrap(~time, scales = "free") +
  theme(axis.text.y = element_text(size = 10))
```

## Effect of Stunting on OGU Abundances

```{r}
#| eval: false
#| label: stunt-model

y <- outcome %>% select(names(poi_models))

poi_stunt <- genZI(
  y,
  predictors,
  formula = "time + arm + age_B + arm:time",
  zi = "~ arm * time",
  family = "poisson"
)

compare_nested_models(list(
  poi_models = keep_at(poi_results, names(poi_models)),
  poi_stunt = poi_stunt
))

```


## Note on applying Structured Covariance Matricies to Zero-inflated data

Structured covariance matricies are typically avoided when the data is sparse, zero-inflated or have few repeated measures available. See notes on "Approaches to Modelling Covariance".