## Investigating changes in OGU abundance over time using Zero-inflated Beta Regression Models with Random Effects

```{r}
#| echo: false
#| eval: true
#| label: setup

# This initial code block loads custom utility functions and environment variables
# These are stored in separate files to keep the main analysis clean and organized
source("utils/r-utils.r")
source("scripts/OGeeU.r")
source("scripts/zeroGUS.r")

source("_environment.required")
load("data/seqdata.RData")

suppressPackageStartupMessages({
  library(ZIBR)
  library(glmmTMB)
  })

dir.create(file.path(RESULTS, "EvsB-OGU-ZIBR"), showWarnings = FALSE, recursive = TRUE)

```

```{r}
#| eval: true
#| label: data prep

data <- prepare_ogu_data(absAbund) # converted to relative abundance data after filtering
outcome <- data$ogu_data

# checks prevalence
# Removes rare species
# counts how many non-zero values are in each column
# Returns TRUE if more than 40% of values are non-zero
filter_index1 <- apply(outcome, 2, function(X) {
  sum(X > 0) > 0.4 * length(X)
})

# checks abundance
# Removes low abundance species
# finds the 90th percentile value and returns TRUE if greater than 1

# filter_index2 <- apply(outcome, 2, function(X) {
#   quantile(X, 0.9) > 1
# })

# outcome <- outcome[, filter_index1 & filter_index2]
outcome <- outcome[, filter_index1]

outcome <- as.matrix(outcome)

predictors <- prepare_predictors(data$meta_data)
predictors <- predictors %>%
  mutate(arm = case_when(arm == "CPM" ~ 0, arm == "FeZnPM" ~ 1)) %>%
  rownames_to_column(var = "sample") %>%
  mutate(
    baseline_sample = str_replace(sample, "_G_E|_B_E|_E|_G_B|_B_B", "_B"),
    sample = str_remove(sample, "_G"),
  ) %>%
  mutate(sample = str_replace(sample, "_B_B", "_B")) %>%
  mutate(sample = str_replace(sample, "_B_E", "_E")) 

rownames(outcome) <- str_replace(
  str_replace(str_remove(rownames(outcome), "_G"), "_B_B", "_B"),
  "_B_E",
  "_E"
)

names <- colnames(outcome)

tmp <- glmmTMB(outcome[,1] ~ arm*time+age_B + (1|subject), zi = ~arm*time, data=predictors, family = poisson)

compareZI(outcome[,1] ~ arm*time+age_B + (1|subject), zi = ~arm*time, data=predictors)

if (diagnose(tmp) == FALSE) {
  diagnose(tmp)

}
summary(tmp)

ff <- fixef(tmp)$zi
round(plogis(c(CPM=unname(ff[1]),ff[-1]+ff[1])),3)

plot(simulateResiduals(tmp))

```

```{r}
models <- list()
p_taxa_list_zibr <- list()

for (j in names) {
  if (any(is.na(predictors) | is.null(predictors))) {
    stop("Predictor variables cannot contain NA values")
  }

  df <- data.frame(
    Baseline = outcome[predictors$baseline_sample, j],
    predictors[, c("time", "arm")]
  )

  rownames(df) <- predictors$sample

  if (any(is.na(df) | is.null(df))) {
    stop("Predictor variables cannot contain NA values")
    view(df)
  }

  # df <- data.frame(y = y, predictors) %>%
  #   arrange(subject, time) %>% select(y, arm, time)
  subject_ind <- predictors$subject
  time_ind <- predictors$time
  if (length(subject_ind) != length(time_ind) && nrow(predictors)) {
    stop("Input variables are not the same length")
  }

  y <- outcome[predictors$sample, j]

  if (!all(data$y > 0 & data$y < 1)) {
    stop("Response must be between 0 and 1")
  }

  if (any(is.na(y) | is.null(y))) {
    stop("Outcome variable contains NA values")
  } else {
    est <- tryCatch(
      {
        ZIBR::zibr(
          logistic_cov = df,
          beta_cov = df,
          Y = y,
          subject_ind = subject_ind,
          time_ind = time_ind,
          quad_n = 30,
          verbose = TRUE
        )
      },
      error = function(e) {
        # Print dimensions of problematic matrices
        print(paste("Error:", e))
        return(NULL)
      }
    )
    # models[j] <- est
    # p_taxa_list_zibr[j, ] <- est$joint.p
  }
}
```

```{r}
#| eval: true
#| label: auto

results <- run_zibr_pipeline (
  phyloseq_obj = bac_base_end,
  output_dir = file.path(RESULTS, "EvsB-OGU-GEE")
)

# 3. Get summary of results
cat("Total OGUs analyzed:", nrow(results$model_results$convergence_summary), "\n")
cat("Significant OGUs:", nrow(results$significant_ogus), "\n")

# 4. Create summary plots
plots <- create_summary_plots(results)
print(plots$p_value_dist)
print(plots$volcano_plot)

# 5. Examine a specific significant OGU
# Replace "OGU_ID" with an actual OGU ID from your significant results
ogu_details <- examine_ogu(results, "OGU_ID")

# 6. Access detailed results for further analysis
significant_ogus <- results$significant_ogus
model_summaries <- results$model_results$convergence_summary

# 7. Export results to Excel for easier viewing
if (require(openxlsx)) {
  write.xlsx(
    list(
      Convergence_Summary = results$model_results$convergence_summary,
      Significant_OGUs = results$significant_ogus
    ),
    file = file.path(output_dir, "ZIBR_Results.xlsx")
  )
}