# Investigating changes in OGU abundance over time using Zero-inflated Regression Models with Random Effects

```{r setup}
#| echo: false
#| eval: true
#| label: setup

# This initial code block loads custom utility functions and environment variables
# These are stored in separate files to keep the main analysis clean and organized
suppressMessages({
  source("utils/r-utils.r")
  source("scripts/OGeeU.r")
  source("scripts/zeroGUS.r")

  source("_environment.required")
  load("data/seqdata.RData")

  library(glmmTMB)
  library(DHARMa)
  library(broom)
  library(patchwork)
  library(emmeans)
  library(ggtext)

  library(promicrobialTools)

  dir.create(
    file.path(RESULTS, "EvsB-OGU-ZIBR"),
    showWarnings = FALSE,
    recursive = TRUE
  )
})
```

## Prepare Outcome and Predictor Variables

```{r data-prep}
#| eval: true
#| label: data prep

data <- prepare_ogu_data(
  absAbund,
  filter_prevalence = TRUE,
  prevalence_cutoff = 0.4,
  filter_abundance = FALSE
)

data$taxa_data <- data$taxa_data %>%
  mutate(Phylum = ifelse(str_detect(Phylum, "Bacillota"), "Bacillota", Phylum))

outcome <- data$ogu_data
rownames(outcome) <- str_replace(
  str_replace(str_remove(rownames(outcome), "_G"), "_B_B", "_B"),
  "_B_E",
  "_E"
)

predictors <- prepare_predictors(data$meta_data)
predictors <- predictors %>%
  mutate(arm = case_when(arm == "CPM" ~ 0, arm == "FeZnPM" ~ 1)) %>%
  rownames_to_column(var = "sample") %>%
  mutate(
    baseline_sample = str_replace(sample, "_G_E|_B_E|_E|_G_B|_B_B", "_B"),
    sample = str_remove(sample, "_G"),
  ) %>%
  mutate(sample = str_replace(sample, "_B_B", "_B")) %>%
  mutate(sample = str_replace(sample, "_B_E", "_E"))
```

## Determining Direction of Change in Abundance over Time

```{r}
#| eval: true
#| label: tbl-abundance-changes
#| tbl-cap: Count Changes between Baseline and Endpoint

joined_data <- predictors %>%
  left_join(rownames_to_column(outcome, "sample"), by = "sample") %>%
  pivot_longer(
    cols = !c(
      "sample",
      "subject",
      "arm",
      "timepoint",
      "time",
      "age_B",
      "baseline_sample",
      "zlen"
    ),
    names_to = "outcome_id"
  )

joined_data_changes <- joined_data %>%
  group_by(outcome_id, timepoint) %>%
  summarise(average_count = mean(value)) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = average_count,
    values_fill = 0
  ) %>%
  mutate(EvsB = Endpoint - Baseline) %>%
  mutate(
    difference = case_when(
      EvsB < 0 ~ "Decrease",
      EvsB > 0 ~ "Increase",
      EvsB == 0 ~ "None",
      TRUE ~ NA
    )
  ) %>%
  group_by(difference) %>%
  summarise(total = n()) %>%
  kable()
```

The purpose of these models will be to determine if these changes are significant.

## Variable Characteristics

### Predictor Variables

```{r}
#| eval: true
#| label: predictor-describe
#| results: asis

Hmisc::describe(predictors)

```

### OGU Outcome Variables

```{r zero-hist}
#| eval: true
#| label: fig-outcome-zeroProp
#| fig-cap: Distribution of zero proportions across variables

analyze_zeros(outcome, "density")
```

```{r disp}
#| eval: true
#| label: fig-outcome-dispersion
#| fig-cap: Distribution of dispearsion, defined as the variance-to-mean ratio. A log10 scale is used.

outdisp <- varDisp(outcome, threshold = 2, verbose = FALSE)

plot(outdisp, col = "#8CAAEE")

```

Generally, if the dispersion ratio is > 2 (0.3 on the log_10 scale), fitting a Poisson model will not accurately model the data (Poisson assumes that variance $\approx$ mean).

|      Model       |          Continuous component           | ZI component |                   Test                    |
|:----------------:|:---------------------------------------:|:------------:|:-----------------------------------------:|
|   noZi_models    |                  y ~ 1                  |     ~ 0      |  Is the zero-inflated component needed?   |
|   null_models    |                  y ~ 1                  |     ~ 1      | Tests if predictors affect zero-inflation |
| noZi_rand_models |         y ~ 1 + (1 \| subject)          |     ~ 0      |      Is the random intercept needed?      |
| null_rand_models |         y ~ 1 + (1 \| subject)          |     ~ 1      |                                           |
|        m1        | y ~ arm * time + age_B + (1 \| subject) | ~ arm * time |                                           |
|                  |                                         |              |                                           |

: List of model formulae tested

## Testing different distributions

The following distribution families were used to fit a zero-inflated model:

- Poisson
- Simple negative binomial (linear parameterization) [@Hardin2007]
- Complex negative binomial (quadratic parameterization) [@Hardin2007]
- Gaussian
- Log normal

```{r fit}
#| eval: false
nb2_results <- genZI(
  outcome,
  predictors,
  formula = "time + arm + age_B + arm:time + (1 | subject)",
  zi = "~ arm * time",
  family = "nbinom2"
)

nb1_results <- genZI(
  outcome,
  predictors,
  formula = "time + arm + age_B + arm:time + (1 | subject)",
  zi = "~ arm * time",
  family = "nbinom1"
)

poi_results <- genZI(
  outcome,
  predictors,
  formula = "time + arm + age_B + arm:time + (1 | subject)",
  zi = "~ arm * time",
  family = "poisson"
)

gau_results <- genZI(
  outcome,
  predictors,
  formula = "time + arm + age_B + arm:time + (1 | subject)",
  zi = "~ arm * time",
  family = "gaussian"
)

lognorm_results <- genZI(
  outcome,
  predictors,
  formula = "time + arm + age_B + arm:time + (1 | subject)",
  zi = "~ arm * time",
  family = "lognormal"
)
```

```{r save}
#| eval: false
#| echo: false
saveRDS(
  nb2_results,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "genZI-OGU-nb2_results.rds")
)

saveRDS(
  nb1_results,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "genZI-OGU-nb1_results.rds")
)

saveRDS(
  poi_results,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "genZI-OGU-poi_results.rds")
)

saveRDS(
  gau_results,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "genZI-OGU-gau_results.rds")
)

saveRDS(
  lognorm_results,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "genZI-OGU-lognorm_results.rds")
)
```

```{r load}
#| eval: true
#| echo: false
nb2_results <- readRDS(file.path(
  RESULTS,
  "EvsB-OGU-ZIBR",
  "genZI-OGU-nb2_results.rds"
))

nb1_results <- readRDS(file.path(
  RESULTS,
  "EvsB-OGU-ZIBR",
  "genZI-OGU-nb1_results.rds"
))

poi_results <- readRDS(file.path(
  RESULTS,
  "EvsB-OGU-ZIBR",
  "genZI-OGU-poi_results.rds"
))

gau_results <- readRDS(file.path(
  RESULTS,
  "EvsB-OGU-ZIBR",
  "genZI-OGU-gau_results.rds"
))

lognorm_results <- readRDS(file.path(
  RESULTS,
  "EvsB-OGU-ZIBR",
  "genZI-OGU-lognorm_results.rds"
))
```

### Model diagnostic failures

```{r}
#| eval: true
#| label: list

list <- list(
  gau = gau_results,
  lognorm = lognorm_results,
  poi = poi_results,
  nb1 = nb1_results,
  nb2 = nb2_results
)
```

```{r}
#| eval: true
#| label: diagnose-mods
#| include: false

poor_fits <- lapply(1:length(list), function(x) {
  d <- diagnoseZI(list[[x]]$all_models)
  d$dist <- names(list)[x]

  return(d)
})

poor_fits <- data.table::rbindlist(poor_fits)
```

### Comparing Model AIC and BIC values

```{r}
#| eval: true
#| label: summaries

summaries <- list(
  gau = gau_results$model_summary,
  lognorm = lognorm_results$model_summary,
  poi = poi_results$model_summary,
  nb1 = nb1_results$model_summary,
  nb2 = nb2_results$model_summary
)

summaries <- data.table::rbindlist(summaries, idcol = "dist")

summaries <- left_join(summaries, poor_fits, by = c("outcome_id", "dist"))
```

```{r}
#| eval: true
#| label: any-errors

if (any(summaries$error_message != "")) {
  print("Some models returned an error message")
}

```

```{r}
#| eval: true
#| label: diag-compare
summaries <- summaries %>%
  filter(pass == TRUE) %>%
  select(!c(error_message, message)) %>%
  mutate(index = 1:nrow(.)) %>% #make it easier to distinguish between outcome ids
  pivot_longer(
    cols = !c(index, outcome_id, dist, pass),
    names_to = "criterion",
    values_to = "value"
  )
```

```{r}
#| eval: true
#| label: best-AIC

aic <- summaries %>%
  filter(criterion == "AIC") %>%
  group_by(criterion, outcome_id) %>%
  arrange(value) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, minAIC = value, distAIC = dist)
```

```{r}
#| eval: true
#| label: best-BIC

bic <- summaries %>%
  filter(criterion == "BIC") %>%
  group_by(criterion, outcome_id) %>%
  arrange(value) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, minBIC = value, distBIC = dist)

```

```{r}
#| eval: true
#| label: tbl-best-fits
#| tbl-cap: Best fit models based on AIC and BIC

(bestDist <- left_join(aic, bic) %>%
  mutate(dist = ifelse(distAIC == distBIC, distAIC, "Undetermined")))
```

### Checking for deviations in model assumptions

::: {.callout-note}

The larger the dispersion value, $\theta$ in `nbinom2` models, the less variance there is in the model [@Brooks2017].
:::

`dharmaDisp` represents the dispersion ratio, which should ideally be close to 1

```{r}
#| eval: true
#| label: tbl-best-disp
#| tbl-cap: lowest disp value for models that passed diagnostics

dispRatio <- summaries %>%
  filter(criterion == "dharmaDisp") %>%
  group_by(criterion, outcome_id) %>%
  mutate(skew = abs(value - 1)) %>%
  arrange(skew) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, bestDisp = value, distDisp = dist)
```

```{r}
#| eval: true
#| label: tbl-pvals
#| tbl-cap: NA

pvals <- left_join(
  bestDist,
  summaries,
  by = c("index", "outcome_id", "dist")
) %>%
  filter(str_detect(criterion, "Pvalue"), value <= 0.05) %>%
  arrange(index)
```

`r n_distinct(pvals$outcome_id)` models have a significant test result making the model fit problematic

### Final Model Distributions

```{r}
#| eval: true
#| label: best-dist-filtered
#| fig-cap: NA
#| tbl-cap: NA

bestDist_filtered <- bestDist %>%
  filter(!outcome_id %in% pvals$outcome_id)
```

## Fitting models to best distribution

#### Zero-inflated Negative Binomial 1 Models

```{r}
#| eval: true
#| label: nbinom1 models

nb1_models <- keep_at(
  nb1_results$all_models,
  filter(bestDist_filtered, dist == "nb1")$outcome_id
)

nb1_unfit <- diagnoseZI(nb1_models)
```

Are there any models that fail diagnostics: `r any(nb1_unfit$pass == FALSE)`

```{r}
#| eval: true
#| label: nb1-summary

nb1_summary <- data.frame(
  outcome_id = names(nb1_models),
  continuous_distribution = "nbinom1",
  continuous_link = "log",
  covaraince_structure = "unstructured",
  zero_distribution = "binary",
  zero_link = "logit",
  optimizer = "nlminb"
)
```

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for nbinom1
nb1_dharma_results <- process_dharma_results(nb1_models)

kable(nb1_dharma_results$summary)

```

#### Zero-inflated Negative Binomial Polynomial Models

```{r}
#| eval: true
#| label: nbinom2 models

nb2_models <- keep_at(
  nb2_results$all_models,
  filter(bestDist_filtered, dist == "nb2")$outcome_id
)

nb2_unfit <- diagnoseZI(nb2_models)
```

Are there any models that fail diagnostics: `r any(nb2_unfit$pass == FALSE)`

```{r}
#| eval: true
#| label: nb2-summary

nb2_summary <- data.frame(
  outcome_id = names(nb2_models),
  continuous_distribution = "nbinom2",
  continuous_link = "log",
  covaraince_structure = "unstructured",
  zero_distribution = "binary",
  zero_link = "logit",
  optimizer = "nlminb"
)
```

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for nbinom1
nb2_dharma_results <- process_dharma_results(nb2_models)

kable(nb2_dharma_results$summary)

```

#### Zero-inflated Poisson Models

```{r}
#| eval: true
#| label: poi models

poi_models <- keep_at(
  poi_results$all_models,
  filter(bestDist_filtered, dist == "poi")$outcome_id
)

poi_unfit <- diagnoseZI(poi_models)
```

Are there any models that fail diagnostics: `r any(poi_unfit$pass == FALSE)`

```{r}
#| eval: true
#| label: poi-summary

poi_summary <- data.frame(
  outcome_id = names(poi_models),
  continuous_distribution = "poisson",
  continuous_link = "log",
  covaraince_structure = "unstructured",
  zero_distribution = "binary",
  zero_link = "logit",
  optimizer = "nlminb"
)
```

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for Poisson models
poi_dharma_results <- process_dharma_results(poi_models)

kable(poi_dharma_results$summary)

```

#### Zero-inflated Gaussian Models

```{r}
#| eval: true
#| label: gau unfit models

gau_models <- keep_at(
  gau_results$all_models,
  filter(bestDist_filtered, dist == "gau")$outcome_id
)

gau_unfit <- diagnoseZI(gau_models)
```

Are there any models that fail diagnostics: `r any(gau_unfit$pass == FALSE)`

```{r}
#| eval: true
#| label: poi-summary

gau_summary <- data.frame(
  outcome_id = names(poi_models),
  continuous_distribution = "gaussian",
  continuous_link = "log",
  covaraince_structure = "unstructured",
  zero_distribution = "binary",
  zero_link = "logit",
  optimizer = "nlminb"
)
```

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for Poisson models
gau_dharma_results <- process_dharma_results(gau_models)

kable(gau_dharma_results$summary)

```

#### Zero-inflated Lognormal Models

```{r}
#| eval: true
#| label: lognorm unfit models

lognorm_models <- keep_at(
  lognorm_results$all_models,
  filter(bestDist_filtered, dist == "lognorm")$outcome_id
)

```

There were no models that could be fit with a log normal distribution

```{r}
#| eval: false
#| label: save
#| echo: false

save(
  nb1_models,
  nb2_models,
  poi_models,
  gau_models,
  lognorm_models,
  file = file.path(RESULTS, "EvsB-OGU-ZIBR", "finalised-fits.RData")
)
```

```{r}
#| eval: true
#| label: load
#| echo: false

load(file.path(RESULTS, "EvsB-OGU-ZIBR", "finalised-fits.RData"))

```

## Significant OGUs

```{r}
#| eval: true
#| label: model-ogus
nb1_ogu_models <- nb1_results$adj_results %>%
  filter(outcome_id %in% names(nb1_models))

nb2_ogu_models <- nb2_results$adj_results %>%
  filter(outcome_id %in% names(nb2_models))

poi_ogu_models <- poi_results$adj_results %>%
  filter(outcome_id %in% names(poi_models))

gau_ogu_models <- gau_results$adj_results %>%
  filter(outcome_id %in% names(gau_models))
```

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: Summary table of OGUs successfully fit by a model, and their corresponding coefficient p-values

poi_summary <- left_join(poi_summary, poi_ogu_models)
nb1_summary <- left_join(nb1_summary, nb1_ogu_models)

nb2_summary <- left_join(nb2_summary, nb2_ogu_models)
gau_summary <- left_join(gau_summary, gau_ogu_models)

kable(
  summary <- rbind(
    nb1_summary,
    nb2_summary,
    poi_summary,
    gau_summary
  )
)
```

```{r}
#| eval: true
#| label: save-tbl-summary
#| echo: false

write.csv(
  summary,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "zero-inflated-ogu-model-summaries.csv"),
  row.names = FALSE
)
```

Of the `r ncol(outcome)` OGUs that passed filtering (`r ncol(outcome)/ncol(data$ogu_data)`), `r nrow(summary)` were sucessfully fit to a zero-inflated model (`r nrow(summary)/ncol(outcome)*100`%)

```{r}
#| eval: true
#| label: zi-sig

ziSig <- identify_significant_ogus(
  summary,
  data$taxa_data
)

colnames(ziSig)[colnames(ziSig) == "rowname"] <- "outcome_id"
```

## Individual abundance changes

```{r}
#| eval: true
#| label: tmp
#| collapse: true
#| tbl-cap: Individual abundance changes for each OGU

tmp <- joined_data %>%
  dplyr::filter(outcome_id %in% ziSig$outcome_id) %>%
  group_by(sample, outcome_id) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = value,
    values_fill = 0
  ) %>%
  mutate(EvsB = Endpoint - Baseline) %>%
  mutate(
    difference = case_when(
      EvsB < 0 ~ "Decrease",
      EvsB > 0 ~ "Increase",
      EvsB == 0 ~ "None",
      TRUE ~ NA
    )
  )

tmp %>%
  group_by(outcome_id, difference) %>%
  summarise(total = n()) %>%
  kable()
```

### Average abundance changes

```{r}
#| eval: true
#| label: tbl-average-sig-changes

aveChange <- joined_data %>%
  dplyr::filter(outcome_id %in% ziSig$outcome_id) %>%
  group_by(outcome_id, timepoint) %>%
  summarise(average_count = mean(value)) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = average_count,
    values_fill = NA
  ) %>%
  mutate(EvsB = Endpoint - Baseline) %>%
  mutate(
    difference = case_when(
      EvsB < 0 ~ "Decrease",
      EvsB > 0 ~ "Increase",
      EvsB == 0 ~ "None",
      TRUE ~ NA
    )
  )

aveChange %>%
  group_by(difference) %>%
  summarise(total = n()) %>%
  kable()
```

### Demonstrating the importance of random effects

OGU G012221805 has an average abundance change of zero, so why is the EvsB term significant? Random effects!

```{r}
#| eval: true
#| label: fig-indv-changes
#| fig-cap: Changes in count abundance for OGU G012221805, showing the clear impact of individual random effects and the importance of including such in a model. Note the black dotted line would be the standard regression line indicating no change over time.

joined_data %>%
  filter(outcome_id == "G012221805") %>%
  group_by(subject) %>%
  ggplot(aes(x = time, y = value, )) +
  geom_point() +
  geom_line(
    aes(
      group = subject,
      col = as.factor(arm)
    ),
    linewidth = 1
  ) +
  geom_smooth(
    method = "lm",
    se = F,
    color = "black",
    formula = y ~ x,
    linetype = 'dotted',
    linewidth = 3
  ) +
  xlab("Time Since Feed Start (days)") +
  ylab("Count") +
  scale_color_manual(values = armColours) +
  labs(color = "Intervention Arm")

```

## Save Significant Model Results

```{r}
#| eval: true
ziSig <- left_join(ziSig, aveChange, by = "outcome_id")

write.csv(
  ziSig,
  file.path(
    RESULTS,
    "EvsB-OGU-ZIBR",
    "significant-ogus-zero-inflated-model.csv"
  ),
  row.names = FALSE
)
```

## Zero-inflated component

```{r}
#| eval: true
#| label: zi-comp
#| fig-cap: NA
#| tbl-cap: NA

zi <- lapply(
  list(nb1_models, nb2_models, poi_models, gau_models),
  function(l) {
    lapply(l, function(x) {
      summary(x)$coefficients$zi[, 4]
    })
  }
)

#FIX make naming automatic using indexing
zip <- bind_rows(
  lapply(zi, function(model_type) {
    bind_rows(model_type, .id = "outcome_id")
  }),
  .id = "model_type"
) %>%
  # Add better names for model types
  mutate(
    model_type = case_when(
      model_type == "1" ~ "nb1",
      model_type == "2" ~ "nb2",
      model_type == "3" ~ "poi",
      model_type == "4" ~ "gau",
      TRUE ~ model_type
    )
  )

if (any(zip <= 0.05)) {
  zip_sig <- identify_significant_ogus(zip, data$taxa_data)
}
```

## Using EMMEANS to determine significant changes in taxa abundances

```{r}
#| eval: true
#| label: fit-mods

fits <- list(nb1_models, nb2_models, poi_models, gau_models)

fits <- list_c(fits)
```

### Conditional Model EMMs

#### OGUs Estimated marginal means

The below estimates the mean for each `arm` at each `timepoint`, controlling for age at baseline

```{r}
#| eval: true
#| label: Estimated marginal means for OGU abundances within each intervention arm

emm <- lapply(fits, function(x) {
  # Get EMMs for arm at each time point
  emmeans(
    x,
    ~ arm | time,
    at = list(time = c(0, 1)),
    adjust = "sidak",
    weights = "proportional",
    type = "scale"
  )
})

```

##### OGU EMM differences between intervention arm at baseline and endpoint

```{r}
#| eval: true
#| label: tbl-emm-pairs
#| tbl-cap: Comparison of abundance marginal mean estimates (EMMs) betweeen intervention arm at baseline and endpoint. Only those with p $\leq$ 0.05 are show for brevity.

emm_pairs <- lapply(emm, pairs, reverse = TRUE)

emm_pairs_df <- emm_bind(emm_pairs, kable = F)

sigArms <- emm_pairs_df %>%
  filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(
    change = ifelse(estimate < 0, "Decrease", "Increase"),
    timepoint = ifelse(time == 1, "Endpoint", "Basepoint")
  )
```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint

plotPhylaEMMs(sigArms) +
  facet_wrap(~timepoint, scales = "free")
```

##### Effect Size

The default value used to standardise the effect size is the model's residual standard deviation (sigma; within-group variation). However, this may underestimate effect sizes in mixed models where combined variance may be more accurate. Here, the random effect variance is used to standardise effect sizes.

```{r}
#| eval: true
#| label: sigArm-eff
#| fig-cap: NA
#| tbl-cap: Effect sizes calculated as a standardized mean difference similar to Cohen's d.

emmEff <- lapply(1:length(emm_pairs), function(i) {
  eff_size(
    emm_pairs[[i]],
    sigma = attr(VarCorr(fits[[i]])$cond$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    method = "identity"
  )
})

names(emmEff) <- names(emm_pairs)

emmEff_df <- emm_bind(emmEff, kable = F)

emmEff_df <- emmEff_df %>%
  filter(Model %in% sigArms$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    ),
    lower.CL = ifelse(is.na(lower.CL), asymp.LCL, lower.CL),
    upper.CL = ifelse(is.na(upper.CL), asymp.UCL, upper.CL)
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, time, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

sigArmsEff <- left_join(sigArms, emmEff_df, by = c("Model", "time"))

```

```{r}
#| eval: true
#| label: fig-phyla-emms-effsize
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint. Significant differences with an effect size > 0.5 (medium effect size) are shown in the plot.

plotPhylaEMMs(sigArmsEff) +
  facet_wrap(~timepoint, scales = "free")
```

#### OGU EMM changes over time within each intervention arm

```{r}
#| eval: true
#| label: tbl-emm-arm-trend
#| tbl-cap: Rate of change per unit increase of abundance marginal mean estimates over time. The time.trend column represents the slope of the relationship with time.

emm_bind(
  trend <- lapply(fits, function(x) {
    emtrends(x, ~ arm | time, var = "time")
  })
)
```

```{r}
#| eval: true
#| label: tbl-emm-arm-trend-compare
#| tbl-cap: Comparison of the rates of change from baseline to endpoint in abundance EMMs between intervention arms.

emm_bind(slopes <- lapply(trend, pairs, reverse = TRUE))

slopes_df <- emm_bind(slopes, kable = F)

sigArmsTime <- slopes_df %>%
  filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(timepoint = ifelse(time == 1, "Endpoint", "Basepoint"))
```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint

plotPhylaEMMs(sigArmsTime, text_size = 8)
```

##### Effect Size

The default value used to standardise the effect size is the model's residual standard deviation (sigma; within-group variation). However, this may underestimate effect sizes in mixed models where combined variance may be more accurate. Here, the random effect variance is used to standardise effect sizes.

```{r}
#| eval: true
#| label: sigArm-eff
#| fig-cap: NA
#| tbl-cap: Effect sizes calculated as a standardized mean difference similar to Cohen's d.

emmEff <- lapply(1:length(slopes), function(i) {
  eff_size(
    slopes[[i]],
    sigma = attr(VarCorr(fits[[i]])$cond$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    method = "identity"
  )
})

names(emmEff) <- names(slopes)

emmEff_df <- emm_bind(emmEff, kable = F)

emmEff_df <- emmEff_df %>%
  filter(Model %in% sigArmsTime$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    ),
    lower.CL = ifelse(is.na(lower.CL), asymp.LCL, lower.CL),
    upper.CL = ifelse(is.na(upper.CL), asymp.UCL, upper.CL)
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, time, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

sigArmsTimeEff <- left_join(sigArmsTime, emmEff_df, by = c("Model", "time"))

```

```{r}
#| eval: true
#| label: fig-phyla-emms-effsize
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint. Significant differences with an effect size > 0.5 (medium effect size) are shown in the plot.

plotPhylaEMMs(filter(sigArmsTimeEff, effect.size >= 0.5))
```

### Zero-inflated Componet EMMs

```{r}
#| eval: true
#| label: tbl-emms
#| tbl-cap: Estimated marginal means for OGU abundances within each intervention arm

emm_zi <- lapply(fits, function(x) {
  # Get EMMs for arm at each time point
  emmeans(
    x,
    ~ arm | time,
    at = list(time = c(0, 1)),
    adjust = "sidak",
    weights = "proportional",
    type = "scale",
    component = "zi"
  )
})

# Convert to probability scale
zi_probs <- lapply(emm_zi, regrid, transform = "logit")
```

##### OGU EMM differences between intervention arm at baseline and endpoint

```{r}
#| eval: true
#| label: tbl-emm-pairs
#| tbl-cap: Comparison of abundance marginal mean estimates (EMMs) betweeen intervention arm at baseline and endpoint

emm_zi_pairs <- lapply(emm_zi, pairs, reverse = TRUE)

emm_zi_pairs_df <- emm_bind(emm_zi_pairs, kable = F)

ZIsigArms <- emm_zi_pairs_df %>%
  filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(
    change = ifelse(estimate < 0, "Decrease", "Increase"),
    timepoint = ifelse(time == 1, "Endpoint", "Basepoint")
  )
```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint

plotPhylaEMMs(ZIsigArms) +
  facet_wrap(~timepoint, scales = "free")
```

##### Effect Size

> [!quote]
> When a logistic regression is calculated, the regression coefficient (b1) is the estimated increase in the log odds of the outcome per unit increase in the value of the exposure. In other words, the exponential function of the regression coefficient (eb1) is the odds ratio associated with a one-unit increase in the exposure. [@Szumilas2010]

i.e. if $\beta_x = log(odds)$, then the odds ratio is $OR = \exp(\beta_x) = e^{\beta_x}$

```{r}
#| eval: true
#| label: odds-ratio
#| tbl-cap: Effect size calculated as odds-ratio for zero-inflation component
ZIodds <- lapply(fits, function(model) {
  exp(fixef(model)$zi)
})

ZIodds <- do.call(rbind, ZIodds) %>%
  as.data.frame() %>%
  rownames_to_column("Model")
```

```{r}
#| eval: true
#| label: cohens-d
#| fig-cap: NA
#| tbl-cap: NA
ZIemmEff <- lapply(1:length(emm_zi_pairs), function(i) {
  eff_size(
    emm_zi_pairs[[i]],
    sigma = attr(VarCorr(fits[[i]])$cond$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    method = "identity"
  )
})

names(ZIemmEff) <- names(emm_zi_pairs)

ZIemmEff <- emm_bind(ZIemmEff, kable = F)

ZIemmEff <- ZIemmEff %>%
  filter(Model %in% sigArms$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    ),
    lower.CL = ifelse(is.na(lower.CL), asymp.LCL, lower.CL),
    upper.CL = ifelse(is.na(upper.CL), asymp.UCL, upper.CL)
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, time, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

ZIsigArmsEff <- left_join(ZIsigArms, ZIemmEff, by = c("Model", "time"))
kable(ZIsigArmsEff <- left_join(ZIsigArmsEff, ZIodds, by = c("Model")))
```

```{r}
#| eval: true
#| label: fig-phyla-emms-effsize
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint. Significant differences with an effect size > 0.5 (medium effect size) are shown in the plot.

plotPhylaEMMs(ZIsigArmsEff) +
  facet_wrap(~timepoint, scales = "free")
```

#### OGU EMM changes over time within each intervention arm

```{r}
#| eval: true
#| label: tbl-emm-arm-trend
#| tbl-cap: Rate of change per unit increase of abundance marginal mean estimates over time. The time.trend column represents the slope of the relationship with time.

trend_zi <- lapply(fits, function(x) {
  emtrends(x, ~ arm | time, var = "time")
})
```

```{r}
#| eval: true
#| label: tbl-emm-arm-trend-compare
#| tbl-cap: Comparison of the rates of change from baseline to endpoint in abundance EMMs between intervention arms.

slopes_zi <- lapply(trend_zi, pairs, reverse = TRUE)

slopes_zi_df <- emm_bind(slopes_zi, kable = F)

ZIsigArmsTime <- slopes_zi_df %>%
  filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(timepoint = ifelse(time == 1, "Endpoint", "Basepoint"))
```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint

plotPhylaEMMs(ZIsigArmsTime)
```

##### Effect Size

The default value used to standardise the effect size is the model's residual standard deviation (sigma; within-group variation). However, this may underestimate effect sizes in mixed models where combined variance may be more accurate. Here, the random effect variance is used to standardise effect sizes.

```{r}
#| eval: true
#| label: sigArm-eff
#| fig-cap: NA
#| tbl-cap: Effect sizes calculated as a standardized mean difference similar to Cohen's d.

emmEff_zi <- lapply(1:length(slopes_zi), function(i) {
  eff_size(
    slopes[[i]],
    sigma = attr(VarCorr(fits[[i]])$zi$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    method = "identity"
  )
})

names(emmEff_zi) <- names(slopes_zi)

emmEff_zi_df <- emm_bind(emmEff_zi, kable = F)

emmEff_zi_df <- emmEff_df_zi %>%
  filter(Model %in% ZIsigArmsTime$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    ),
    lower.CL = ifelse(is.na(lower.CL), asymp.LCL, lower.CL),
    upper.CL = ifelse(is.na(upper.CL), asymp.UCL, upper.CL)
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, time, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

ZIsigArmsTimeEff <- left_join(
  ZIsigArmsTime,
  emmEff_zi_df,
  by = c("Model", "time")
)

```

```{r}
#| eval: true
#| label: fig-phyla-emms-effsize
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint.

plotPhylaEMMs(filter(ZIsigArmsTimeEff))
```

### OGUs Common between GLMMs and differential abundance changes over time

```{r}
#| eval: true
#| label: changes in relative abundace plot
#common <- read.csv("07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/bac_base_end_timepoint_up_down_reg_OGUs_FDR5_FC2_LMM_Mumbai_ID14768_Feb_07_2024_02:35_NC.csv", header = T)

#updated results file 22-08-24
common <- read.csv(
  "~/Work/05-Projects/01-Precision_nutrition/07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/bac_base_end_timepoint_up_down_reg_OGUs_FDR5_FC2_LMM_Mumbai_ID14768_Aug_22_2024_04:23_NC.csv",
  header = T
)

dim(common)

# some OGUs represent the same species, therefore I will combine them and take the sum of their FC
common_combined <- common %>%
  ungroup() %>%
  group_by(Phylum, Species) %>%
  dplyr::summarise(logFC = sum(logFC), FC = fc(logFC)) %>%
  arrange(desc(abs(logFC)), .by_group = T) %>%
  as.data.frame()

phylum_anno <- assign_unique_colors(
  as.vector(common_combined$Phylum),
  phyla_colours
)

paste0(
  "Some OGUs represent the same species. When combined there are ",
  nrow(common_combined),
  " total species"
)
```

```{r}
#| eval: true
#| label: Combined OGU fold change table
#| tab-cap: For significant fold changes of 2 or more, the number of OGUs common between differential abundance results and LMM results is 106, representing 91 unique species

common_combined %>%
  arrange(desc(logFC)) %>%
  kable()
```

```{r}
#| eval: true
#| label: Combined OGU fold change table
#| tab-cap: For significant fold changes of 2 or more, the percentage of OGUs in each phyla that are common between differential abundance results and LMM results.

common_combined %>%
  group_by(Phylum) %>%
  summarise(total = n(), percent = n() / nrow(common_combined) * 100) %>%
  kable()
```

```{r}
#| eval: true
#| label: plots
#| fig-cap: Log2 fold change of OGUs shared across LMM and DA results

#png(paste("~/Work/05-Projects/01-Precision_nutrition/07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/bac_base_end_timepoint_up_down_reg_OGUs_FDR5_FC2_LMM_barplot", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.png", sep = "_"), 1200, 1500)

common_combined %>%
  arrange(logFC) %>%
  ggplot(aes(y = reorder(Species, logFC), x = logFC, fill = Phylum)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  xlab(NULL) +
  ylab(NULL) + #labels the y axis
  scale_color_manual(values = phylum_anno, aesthetics = 'fill') +
  geom_vline(xintercept = 0) +
  theme(
    axis.text.x = element_text(
      angle = 0,
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    axis.text.y = element_text(size = 12, face = 'bold', color = 'black')
  )

#dev.off()
```

```{r}
#| eval: true
#| label: top50 OGU barplot
#| fig-cap: Log2 fold change of the top 50 OGUs shared across LMM and DA results

#png(paste("~/Work/05-Projects/01-Precision_nutrition/07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/bac_base_end_timepoint_up_down_reg_OGUs_FDR5_FC2_LMM_barplot_top50", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.png", sep = "_"), 1200, 800)

common_combined %>%
  arrange(desc(logFC)) %>%
  slice(1:50) %>%
  ggplot(aes(y = reorder(Species, logFC), x = logFC, fill = Phylum)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  xlab(NULL) +
  ylab(NULL) + #labels the y axis
  scale_color_manual(values = phylum_anno, aesthetics = 'fill') +
  geom_vline(xintercept = 0) +
  theme(
    axis.text.x = element_text(
      angle = 0,
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    axis.text.y = element_text(size = 12, face = 'bold', color = 'black')
  )

#dev.off()
```

The below chunk is no longer needed with the new reuslts only having 106 OGUs rather
than 209.

```{r}
#| eval: false
#| label: breakdown of barplot for better visualisation

p1 <- common_combined %>%
  arrange(logFC) %>%
  slice(1:45) %>%
  ggplot(aes(y = reorder(Species, logFC), x = logFC, fill = Phylum)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  xlab(NULL) +
  ylab(NULL) + #labels the y axis
  scale_color_manual(values = phylum_anno, aesthetics = 'fill') +
  geom_vline(xintercept = 0) +
  theme(
    axis.text.x = element_text(
      angle = 0,
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    axis.text.y = element_text(size = 8, face = 'bold', color = 'black')
  )

p2 <- common_combined %>%
  arrange(logFC) %>%
  slice(45:90) %>%
  ggplot(aes(y = reorder(Species, logFC), x = logFC, fill = Phylum)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  xlab(NULL) +
  ylab(NULL) + #labels the y axis
  scale_color_manual(values = phylum_anno, aesthetics = 'fill') +
  geom_vline(xintercept = 0) +
  theme(
    axis.text.x = element_text(
      angle = 0,
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    axis.text.y = element_text(size = 8, face = 'bold', color = 'black')
  )

p3 <- common_combined %>%
  arrange(logFC) %>%
  slice(90:135) %>%
  ggplot(aes(y = reorder(Species, logFC), x = logFC, fill = Phylum)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  xlab(NULL) +
  ylab(NULL) + #labels the y axis
  scale_color_manual(values = phylum_anno, aesthetics = 'fill') +
  geom_vline(xintercept = 0) +
  theme(
    axis.text.x = element_text(
      angle = 0,
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    axis.text.y = element_text(size = 8, face = 'bold', color = 'black')
  )

p4 <- common_combined %>%
  arrange(logFC) %>%
  slice(135:185) %>%
  ggplot(aes(y = reorder(Species, logFC), x = logFC, fill = Phylum)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  xlab(NULL) +
  ylab(NULL) + #labels the y axis
  scale_color_manual(values = phylum_anno, aesthetics = 'fill') +
  geom_vline(xintercept = 0) +
  theme(
    axis.text.x = element_text(
      angle = 0,
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    axis.text.y = element_text(size = 8, face = 'bold', color = 'black')
  )

#png(paste("07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/bac_base_end_timepoint_up_down_reg_OGUs_FDR5_FC2_LMM_barplot", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.png", sep = "_"), 1500, 1300)

ggarrange(p1, p2, p3, p4, legend.grob = get_legend(p))

#dev.off()
```

## Appendix

### Effect of Stunting on OGU Abundances

```{r}
#| eval: false
#| label: stunt-model

y <- outcome %>% select(names(poi_models))

poi_stunt <- genZI(
  y,
  predictors,
  formula = "time + arm + age_B + arm:time",
  zi = "~ arm * time",
  family = "poisson"
)

compare_nested_models(list(
  poi_models = keep_at(poi_results, names(poi_models)),
  poi_stunt = poi_stunt
))

```


### Note on applying Structured Covariance Matricies to Zero-inflated data

Structured covariance matricies are typically avoided when the data is sparse, zero-inflated or have few repeated measures available. See notes on "Approaches to Modelling Covariance".