## Baseline Sample Gene Ontology Enrichment Analysis

```{r}
#| echo: false
#| eval: true
#| label: setup

# This initial code block loads custom utility functions and environment variables
# These are stored in separate files to keep the main analysis clean and organized
source(normalizePath("utils/r-utils.r"))
source(normalizePath("_environment.required"))
load(normalizePath("data/seqdata.RData"))

metadata <- read_delim(normalizePath(METADATA))
baseline_metadata <- readRDS(normalizePath("data/metadata/baseline-metadata.rds"))

suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(pairwiseAdonis))
suppressPackageStartupMessages(library(patchwork))

dir.create(file.path(RESULTS, "baseline-go"), showWarnings = FALSE, recursive = TRUE)

#goTermsHierarchy <- file.path(REF, "WoL2-Woltka/function/go/parent/all.map")
go2kegg <- read.delim(file.path(REF, "WoL2-Woltka/function/go/idmaps/kegg.map"),header=T)
```

### Load data

```{r}
#| eval: true
#| label: load data
go <- read.delim(file.path(WOLTKA_OUT, "go/all.generic.tsv"), header = T) %>%
      mutate_at(vars(everything()), str_trim, side = "both") %>%
      pivot_longer(cols = !c(Name, X.FeatureID), names_to = "sample", values_to = "count") %>%
      rename(Terms = Name, go_id = X.FeatureID) %>%
      mutate(sample = str_remove(sample, "_filtered")) %>%
      mutate_at(vars(everything()), type.convert, as.is = TRUE)
```

### Add additional metadata to data

```{r}
#| eval: true
#| label: assign ontology to GO IDs
go_onto <- clusterProfiler::go2ont(go$go_id)

go <- go %>%
    mutate(count = ifelse(is.na(count), 0, count)) %>%
    mutate(
        timepoint = metadata$timepoint[match(.$sample, metadata$sample_names)],
        trt = metadata$trt[match(.$sample, metadata$sample_names)],
        kegg_rxn_id = go2kegg$kegg_rxn_id[match(.$go_id, go2kegg$go_id)],
        arm = case_when(trt == 0 ~ "CPM", trt == 1 ~ "FeZnPM"),
        ontology = go_onto$Ontology[match(go_id, go_onto$go_id)]
    ) %>%
    mutate(ontology = case_when(
        ontology == "MF" ~ "Molecular function",
        ontology == "BP" ~ "Biological process",
        ontology == "CC" ~ "Cellular component"
    )) 
  
go <- go  %>% 
    dplyr::filter(timepoint == "Baseline")

write.csv(go, file.path(RESULTS, "baseline-go/baseline-annotated-go-terms.csv"), row.names = F)
```

There are `r length(unique(go$Terms))` unique GO terms in the baseline dataset.

### Dataset filtering

Filtering GO database to contain only bacterial GO terms using [GO slim prokaryote](https://geneontology.org/docs/download-ontology/). *downloaded 08-Oct-24*

```{bash}
#| eval: false
#| label: download GO slim

wget -P ref/ https://current.geneontology.org/ontology/subsets/goslim_prokaryote.tsv
```


```{r}
#| eval: true
#| label: annotation with only bacterial GO terms 

annoBac <- read.delim(file.path(REF, "goslim_prokaryote.tsv"), col.names = c("go_id", "terms")) %>% mutate(go_id = str_replace(str_extract(go_id, "GO_\\d*"), "_", ":"))

# amigo <- go %>%
#   filter(!(go_id %in% annoBac$go_id))

slim <- go %>%
  filter(go_id %in% annoBac$go_id) 
```

### Calculate the percentage of baseline samples with each GO term

**defines `go_base` variable and calculates percentage of samples with particular GO terms**

```{r}
#| eval: true
#| label: calculate percentage of samples with particular GO terms

go_base <- slim %>%
  filter(ontology != "Cellular Component") %>%
    arrange(desc(count)) %>%
    group_by(go_id) %>% 
    mutate(n = 1) %>%
    mutate(percent_samples = sum(n)/length(unique(.$sample)) * 100,)

onto_col <- assign_unique_colors(go_base$ontology)

#write.csv(go_base, paste0(WORK,paste("05b-woltka_GO_analysis/go/all_go_combined_BASELINE_annotated", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.csv", sep = "_")), row.names = F)

tmp2 <- go_base %>% 
    group_by(Terms, arm) %>%
    summarise(term_presence = mean(percent_samples)) %>%
  dplyr::filter(term_presence<95) 
```

There are `r length(unique(go_base$Terms))` unique GO terms (collapsed into their REVIGO heirarchies i.e. "generic" GO terms) in the baseline dataset.

On [REVIGO](http://revigo.irb.hr/), the list of baseline go terms and theri absolute count values were pasted into the prompt box. REVIGO was run with the following parameters:

- medium (0.7)
- Higher absolute value is better
- remove obsolete GO terms
- use default whole UniProt database
- use SimRel defualt semantic similarity measure

This returns the exact same list as the generic terms (for the full set also even when I change the settings to large)

```{r}
#| eval: true
#| label: fig-ms-sf2
#| fig-cap: The percentage of GO terms present in samples at baseline. Terms are separated by ontology, with biological processes in red and molecular functions in blue.
#| fig-width: 18

p1 <- go_base %>%
  dplyr::filter(ontology == "Biological process") %>%
    filter(Terms %in% tmp2$Terms)%>%
    ggplot(aes(y=reorder(Terms, dplyr::desc(ontology)), x = percent_samples, fill = ontology)) +
        geom_col(position = "dodge") +
        xlab("Percentage of Samples") +
        ylab(NULL) +
        scale_fill_manual(values = onto_col) +
        theme(axis.text.y = element_text(size = 14, face='bold', color='black'))
p2 <- go_base %>%
  dplyr::filter(ontology == "Molecular function") %>%
    filter(Terms %in% tmp2$Terms)%>%
    ggplot(aes(y=reorder(Terms, dplyr::desc(ontology)), x = percent_samples, fill = ontology)) +
        geom_col(position = "dodge") +
        xlab("Percentage of Samples") +
        ylab(NULL) +
        scale_fill_manual(values = onto_col) +
        theme(axis.text.y = element_text(size = 14, face='bold', color='black'))

p1 + p2 + plot_layout(guides='collect') & theme(legend.position='bottom')
```


```{r}
#| eval: true
#| label: fig-base-go-arm
#| fig-cap: The percentage of GO terms present in samples at baseline, separated by intervention arm. Terms are separated by ontology, with biological processes in red and molecular functions in blue.
#| fig-width: 18
p1 <- go_base %>%
  dplyr::filter(ontology == "Biological process") %>%
    #filter(Terms %in% tmp2$Terms)%>%
    ggplot(aes(y=reorder(Terms, dplyr::desc(ontology)), x = percent_samples, fill = ontology)) +
        geom_col(position = "dodge") +
        xlab("Percentage of Samples") +
        ylab(NULL) +
        scale_fill_manual(values = onto_col) +
        theme(axis.text.y = element_text(size = 14, face='bold', color='black')) +
        facet_wrap(~arm)

p2 <- go_base %>%
  dplyr::filter(ontology == "Molecular function") %>%
    #filter(Terms %in% tmp2$Terms)%>%
    ggplot(aes(y=reorder(Terms, dplyr::desc(ontology)), x = percent_samples, fill = ontology)) +
        geom_col(position = "dodge") +
        xlab("Percentage of Samples") +
        ylab(NULL) +
        scale_fill_manual(values = onto_col) +
        theme(axis.text.y = element_text(size = 14, face='bold', color='black')) +
        facet_wrap(~arm)

p1 + p2 + plot_layout(guides='collect') & theme(legend.position='bottom')

```

The percentage of GO terms present in samples at baseline. Terms are separated by intervention arm and ontology, with biological processes in red and molecular functions in blue. No significant differences in GO term abundance between intervention arms at baseline were highlighted in #differential-abundance analysis. See (@baselineGOde)

### Differential GO term analysis between intervention groups at baseline {#baselineGOde}

#### Assinging GO term counts to DGEList object

```{r}
#| eval: true
#| label: assinging GO term counts to DGEList object
#| results: asis
counts <- go_base %>% 
      dplyr::select(go_id, sample, count) %>%
        pivot_wider(names_from = sample, values_from = count, values_fill = 0) %>%
        as.data.frame()

counts <- edgeR::DGEList(as.data.frame(counts))

dim(counts)
    
counts$samples
counts$samples$arm <- as.factor(go_base$arm[match(rownames(counts$samples), go_base$sample)])
counts$samples$timepoint <- go_base$timepoint[match(rownames(counts$samples), go_base$sample)]

counts$go_id <- go_base %>% select(go_id, Terms, ontology) %>% as.data.frame()

counts.keep <- rowSums(counts$counts==0)==ncol(counts$counts)
counts <- counts[counts.keep == FALSE,]
dim(counts)
```

#### Contrast testing: Difference between Intervention arm at Baseline

$H_0$: GO functional terms are not differentially abundant between intervention arm at baseline
$H_1$: There is a difference between GO functional terms pathway abundance between intervention arm at baseline

```{r}
#| eval: true
#| label: GO functional terms differences between treatment arms at baseline
#Linear Model
#Contrasts
design <- stats::model.matrix(~0+counts$samples$arm)
colnames(design) <- levels(counts$samples$arm)

contrasts <- makeContrasts(CPMvsFeZnPM = CPM-FeZnPM, levels = design)

#Transform raw count data to log2-counts per million (logCPM), estimate the mean-variance relationship and use this to compute appropriate observation-level weights. The data are then ready for linear modelling.

#If the DGEList object has been scale-normalized in #edgeR (or before), then leave normalize.method="none" in voom, i.e., the logCPM values should not usually be re-normalized in the voom call. 
#KEGG abundance values have not been normalised 
v <- voom(counts, design, normalize.method="none") #Transforms count data to log2-counts per million
vfit <- lmFit(v, design)
vfit <- contrasts.fit(vfit, contrasts = contrasts)
efit <- eBayes(vfit) #estimation of QL dispersions

#FDR of 0.05
#Identify which genes are significantly differentially expressed for each contrast from a fit object (eBayes) containing p-values and test statistics.
knitr::kable(summary(FDR5 <- decideTests(efit, adjust.method = "BH", p.value = 0.05, lfc = 0)), caption = "GO terms with significant differences in abundance between intervention arms at baseline")

#write.fit(fit = efit, results = FDR5, file = , paste("./07-Research_Q_analysis/06-Functional_analysis/02-Baseline_Endpoint_analysis/01-Bacteria/02-GO/baseline_up_down_reg_GO_term_table_FDR5_eBayes", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.csv", sep = "_"), row.names=F)
```

There were no GO terms with significant differences in abundance between intervention arms at baseline. 

#### Unsupervised ordination: MDS plot of GO term abundace differences between intervention arm at baseline

```{r}
#| eval: true
#| label: baseline MDS plot by intervention arm and GO term PERMANOVA results
#| fig-cap: Multidimensional scaling plot comparing the abundance of GO terms between intervention arms at baseline
#| tab-cap: Summary table of PERMANOVA results for differences between intervention arm GO terms at baseline

groups <- counts$samples$arm
group_colours <- groups
levels(group_colours) <- c("#FC8D62", "#8DA0CB")
group_colours <- as.character(group_colours)
df <- as.data.frame(t(counts$counts))
df <- cbind(df, groups)

#PERMANOVA testing differenital abundances between intervention arms
mds_stat <- pairwiseAdonis::pairwise.adonis2(df[,1:(ncol(df)-1)] ~ groups, data = df, method='eu') #returns non-adjusted p-values
xmax = max(plotMDS(v, col = group_colours, dim.plot = c(1,2), pch=16, gene.selection = "common")$x)-0.05
ymax = max(plotMDS(v, col = group_colours, dim.plot = c(1,2), pch=16, gene.selection = "common")$y)-0.05

par(mar=c(7.1, 5.1, 5.1, 5.1), xpd=TRUE)
plotMDS(v, labels = groups, col = group_colours, dim.plot = c(1,2), pch=16, gene.selection = "common")
legend("left", pch=16, col = unique(group_colours), legend = levels(groups), inset=c(0,-0.165), bty = "n", ncol=2)
text(
    x = xmax-0.05, 
    y = ymax, 
    labels = bquote(R^2 == .(round(as.data.frame(mds_stat[2])[1,3],3)))
)
text(
    x = xmax-0.05, 
    y = ymax-0.12, 
    labels = bquote(p == .(round(as.data.frame(mds_stat[2])[1,5],3)))
)

pv <- lapply(2:length(mds_stat), function(x){
  w <- mds_stat[[x]][1,4]
  y <- mds_stat[[x]][1,5]
  z <- names(mds_stat)[x]
  data.frame(contrast = z, p = y, F.stat = w)
})

pv <- bind_rows(pv)

kable(pv %>% mutate(p.adj = round(p.adjust(p, method = "BH"), 4)) %>% mutate(sig = case_when(
  p.adj <= 0.0001 ~ "****",
  p.adj <= 0.001 ~ "***",
  p.adj <= 0.01 ~ "**",
  p.adj <= 0.05 ~ "*",
  p.adj > 0.05 ~ "ns"
)), format = "markdown")

#output table
knitr::kable(mds_stat[2], caption = "raw output table")
```

|         |  Df|          SumOfSqs|    R2|      F| Pr(>F)|
|:--------|---:|-----------------:|-----:|------:|------:|
|groups   |   1|    22735574867274| 0.002| 0.3248|   0.62|
|Residual | 166| 11620959426396938| 0.998|       |       |
|Total    | 167| 11643695001264222| 1.000|       |       |

Based on the PERMANOVA output, we can conclude that the variable arm has no significant impact on the variability of the microbial pathway abundances ($R^2 =$ `r mds_stat[[2]][1,3]` $\Rightarrow$ 0.2% varaibility explained). The residuals $R^2$ value is `r mds_stat[[2]][2,3]`, indicating that the variation we see is random/due to other factors.

The p-value ($Pr(>F)$) for `counts$samples$arm` is `r pv$p`, which is greater than the conventional significance level of 0.05. This suggests that we cannot reject the null hypothesis, implying that there is no significant difference in microbial pathway abundance between intervention arms at baseline.


### Analysis of the whole GO term list

*EDIT: 15-08-24*
I had the thought that there may have been little point to performing #differential-abundance analysis with the filtered generic GO term list. For visualising I think it was ok but the PERMANOVA and DA analysis can handle large numbers and may actually help identify any differences between intervention arms. Here I repeat the analysis with the full dataset.

**assinging GO term counts to DGEList object**

```{r}
#| eval: true
#| label: assinging full GO term dataset counts to DGEList object
#| results: asis
counts <- go %>% 
      dplyr::select(go_id, sample, count) %>%
        mutate_at(vars(everything()), type.convert, as.is =T) %>%
        pivot_wider(names_from = sample, values_from = count, values_fill = 0) %>%
        as.data.frame()

counts <- DGEList(as.data.frame(counts))

dim(counts)
    
counts$samples
counts$samples$arm <- as.factor(go$arm[match(rownames(counts$samples), go$sample)])
counts$samples$timepoint <- go$timepoint[match(rownames(counts$samples), go$sample)]

counts$go_id <- go %>% select(go_id, Terms, ontology) %>% as.data.frame()

counts.keep <- rowSums(counts$counts==0)==ncol(counts$counts)
counts <- counts[counts.keep == FALSE,]
dim(counts)
```

$H_0$: GO functional terms are not differentially abundant between intervention arm at baseline
$H_1$: There is a difference between GO functional terms pathway abundance between intervention arm at baseline

```{r}
#| eval: true
#| label: Full GO dataset functional terms differences between treatment arms at baseline
#Linear Model
#Contrasts
design <- stats::model.matrix(~0+counts$samples$arm)
colnames(design) <- levels(counts$samples$arm)

contrasts <- makeContrasts(CPMvsFeZnPM = CPM-FeZnPM, levels = design)

#Transform raw count data to log2-counts per million (logCPM), estimate the mean-variance relationship and use this to compute appropriate observation-level weights. The data are then ready for linear modelling.

#If the DGEList object has been scale-normalized in #edgeR (or before), then leave normalize.method="none" in voom, i.e., the logCPM values should not usually be re-normalized in the voom call. 
#abundance values have not been normalised 
v <- voom(counts, design, normalize.method="none") #Transforms count data to log2-counts per million
vfit <- lmFit(v, design)
vfit <- contrasts.fit(vfit, contrasts = contrasts)
efit <- eBayes(vfit) #estimation of QL dispersions

#FDR of 0.05
FDR5 <- decideTests(efit, adjust.method = "BH", p.value = 0.05, lfc = 0) #Identify which genes are significantly differentially expressed for each contrast from a fit object (eBayes) containing p-values and test statistics.
knitr::kable(summary(FDR5))

#write.fit(fit = efit, results = FDR5, file = , paste("./07-Research_Q_analysis/06-Functional_analysis/02-Baseline_Endpoint_analysis/01-Bacteria/02-GO/baseline_up_down_reg_GO_term_table_FDR5_eBayes", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.csv", sep = "_"), row.names=F)
```

There were no GO terms with significant differences in abundance between intervention arms at baseline. 

```{r}
#| eval: true
#| label: unsupervised learning- MDS plot
#| tab-cap: GO term unsupervised multidimensional scaling PERMANOVA results
group_colours <- counts$samples$arm
levels(group_colours) <- c("#FC8D62", "#8DA0CB")
group_colours <- as.character(group_colours)
df <- as.data.frame(t(counts$counts))
df <- cbind(df, group = counts$samples$arm)

#PERMANOVA testing differenital abundances between intervention arms
mds_stat <- pairwiseAdonis::pairwise.adonis2(df[,1:(ncol(df)-1)] ~ group, data = df, method='eu') #returns non-adjusted p-values

#output table
knitr::kable(mds_stat[2])
```

Based on the PERMANOVA output, we can conclude that the variable arm has no significant impact on the variability of the microbial pathway abundances ($R^2 =$ `r mds_stat[[2]][1,3]` $\Rightarrow$ 0.2% varaibility explained). The residuals $R^2$ value is `r mds_stat[[2]][2,3]`, indicating that the variation we see is random/due to other factors.

```{r}
#| eval: true
#| tab-cap: Summary table of PERMANOVA results for differences between intervention arm GO terms at baseline
pv <- lapply(2:length(mds_stat), function(x){
  r <- mds_stat[[x]][1,3]
  w <- mds_stat[[x]][1,4]
  y <- mds_stat[[x]][1,5]
  z <- names(mds_stat)[x]
  data.frame(contrast = z, r2 = r, F.stat = w, p = y)
})
pv <- bind_rows(pv)
knitr::kable(
    pv %>% 
    mutate(p.adj = round(p.adjust(p, method = "BH"), 4)) %>% 
    mutate(sig = case_when(
        p.adj <= 0.0001 ~ "****",
        p.adj <= 0.001 ~ "***",
        p.adj <= 0.01 ~ "**",
        p.adj <= 0.05 ~ "*",
        p.adj > 0.05 ~ "ns"
       )
    )
)
```

The p-value ($Pr(>F)$) for `counts$samples$arm` is `r pv$p`, which is greater than the conventional significance level of 0.05. So, even with the full GO term dataset, we cannot reject the null hypothesis, implying that there is no significant difference in microbial pathway abundance between intervention arms at baseline.

**MDS plot comparing arms at baseline**

```{r}
#| eval: true
#| label: MDS plot comparing arms at baseline
#| fig-cap: Multidimensional scaling plot comparing the abundance of GO terms between intervention arms at baseline
#| fig-width: 14

xmax = max(plotMDS(v, col = group_colours, dim.plot = c(1,2), pch=16, gene.selection = "common")$x)-0.05
ymax = max(plotMDS(v, col = group_colours, dim.plot = c(1,2), pch=16, gene.selection = "common")$y)-0.05

par(mar=c(7.1, 5.1, 5.1, 5.1), xpd=TRUE)
plotMDS(v, col = group_colours, dim.plot = c(1,2), pch=16, gene.selection = "common")
legend("bottom", pch=16, col = unique(group_colours), legend = levels(counts$samples$arm), inset=c(0,-0.165), bty = "n", ncol=2)
text(
    x = xmax-0.05, 
    y = ymax, 
    labels = bquote(R^2 == .(round(as.data.frame(mds_stat[2])[1,3],3)))
)
text(
    x = xmax-0.05, 
    y = ymax-0.12, 
    labels = bquote(p == .(round(as.data.frame(mds_stat[2])[1,5],3)))
)
```