## Baseline Diversity Analysis

```{r setup}
#| echo: false
#| eval: true
#| label: setup
#| cache: false

# This initial code block loads custom utility functions and environment variables
suppressWarnings(suppressMessages({
  source("utils/r-utils.r")
  source("_environment.required")

  library(promicrobialTools)

  prefix <- "baseDiv-"

  filepath <- get_results_dir(directory_naming = "both")
}))
```

Results will be saved to ``r filepath``.

### Load and Prepare Rarefied Reads

The phyloseq object saved as "rarefied_bacteria.rds" in [Section 3](../03-rarefaction/rarefaction.qmd) is the starting point for this analysis, and must be loaded into the R environment:

```{r}
#| eval: true
#| label: load rarefied read data
#| cache: false

(seqdata_bac <- readRDS(paste0(RARE_RESULTS, "/rarefied_bacteria.rds")))
sample_data(seqdata_bac)$timepoint <- str_remove(
  sample_data(seqdata_bac)$timepoint,
  "1|2|3"
)
```

::: {.callout-note}
Metadata was modified as per [Appendix 2]("analysis/100-Appendix/Appendix-2-Metadata.qmd)
:::

Next, data is filtered to contain only baseline samples.

```{r}
#| eval: true
#| label: extract only baseline data

# subset data
# Filters data to include only samples from the study's starting point ("baseline")
baseline <- phyloseq::subset_samples(seqdata_bac, timepoint == "Baseline")

# Creates a new column with sample names by reassigning rownames (containing sample names)
sample_data(baseline) <- cbind(
  sample_names = rownames(sample_data(baseline)),
  sample_data(baseline)
)

# Ensures the treatment variable (arm) is properly formatted as a categorical factor
sample_data(baseline)$trt <- as.factor(as.character(sample_data(baseline)$trt))
```

### Bacterial Relative Abundances

Relative abundances are calculated to show the proportion of each bacterial group in a sample.

```{r}
#| eval: true
#| label: calculate relative abundances

#create new object to avoid changing baseline phyloseq object
baseTmp <- baseline

#OGU table to relative abundances
otu_table(baseTmp) <- otu_table(
  microbiome::abundances(baseTmp, "compositional"),
  taxa_are_rows = T
)
```

#### Top Phyla at Baseline

Create an ampvis object from phyloseq object

```{r}
#| eval: true
#| label: create ampvis object

library(ampvis2)

d <- amp_load(
  otutable = otu_table(baseTmp),
  metadata = microbiome::meta(baseTmp),
  taxonomy = tax_table(baseTmp)
)
```


```{r}
#| eval: true
#| label: top bacterial phyla at baseline
#| fig-cap: Relative abundance of bacterial phyla at baseline.

#default settings plot relative abundances (normalise = TRUE)
p <- amp_boxplot(
  d,
  normalise = FALSE,
  tax_aggregate = "Phylum",
  tax_show = 20,
  point_size = 0,
  plot_flip = T
)

p +
  geom_point(colour = "white") +
  geom_boxplot(
    aes(fill = Display),
    colour = "black",
    position = position_dodge(5),
    outlier.shape = NA
  ) +
  geom_jitter(color = "black", size = 1.5, stat = "identity", width = 0.35) +
  scale_color_manual(values = phyla_colours, aesthetics = "fill") +
  theme(legend.position = "none")

```

#### Top Genera at Baseline

```{r}
#| eval: true
#| label: top genera at baseline
#| fig-cap: Relative abundance of the top 20 bacterial genera at baseline.
p <- amp_boxplot(
  d,
  normalise = FALSE,
  tax_aggregate = "Genus",
  tax_add = "Phylum",
  point_size = 0,
  tax_show = 20,
  plot_flip = T
)

p +
  geom_point(colour = "white") +
  geom_boxplot(
    aes(fill = Display),
    colour = "black",
    position = position_dodge(5),
    outlier.shape = NA
  ) +
  geom_jitter(color = "black", size = 1.5, stat = "identity", width = 0.35) +
  scale_color_manual(values = pairedColours(n = 20), aesthetics = "fill") +
  theme(legend.position = "none")
```

### Calulation of Baseline Bacteria Alpha Diversity Statistics

Multiple diversity metrics were calculated for each sample to capture different aspects of community structure:

- Observed features: Simple count of bacterial species
- Chao1: Estimates true species richness, accounting for unobserved species
- Shannon: Measures both richness and evenness of bacterial communities
- Pielou's Evenness: Quantifies how equally abundant different species are
- Faith's PD: Considers evolutionary relationships between species

::: {.callout-note}
#### A note on Shannon Diverity

Shannon's Entropy Index ($H$) (aka Shannon Diversity Index; SDI) measures the uncertainty of being able to predict the next sampled OTU in a single indvidual. It calculates the poroprtion of species relative to the total number of species (multiplied by the natural log). While theoretically this can range from 0 to $\inf$, in biological community analysis, this does not seem to be larger that 5.0. 

$H$ is a measure of information content and should only be used on random samples drawn from a large community in which the total number of species is known. $H$ gives more importance to rare species and greater weight on species richness that species evenness. The larger the value of $H$ the more diverse the community is (the larger the uncertainty).

In practice, the true number of total seceis in a microbiome sample is unknown due to absolute quantificaion constraints, varying sequencing depth. This is one limitation of using SDI. 

:::


```{r}
#| eval: true
#| label: alpha diversity calculation

#bacterial baseline samples as defiend above
#calculate abundance based alpha diversity metrics
# absolute abundances need to be used as the microbiome package calculates relative abundances in-function
alpha.baseline <- microbiome::alpha(baseline, index = "all")

pd <- picante::pd(
  samp = as.matrix(otu_table(baseline)),
  tree = phy_tree(baseline),
  include.root = FALSE
)
colnames(pd) <- c("pd", "sr")

#merge alpha metrics with metadata
baseline_metadata <- microbiome::meta(baseline)

baseline_metadata <- microbiome::meta(baseline) %>%
  left_join(
    .,
    rownames_to_column(alpha.baseline),
    by = c("sample_names" = "rowname")
  ) %>%
  left_join(., rownames_to_column(pd), by = c("sample_names" = "rowname")) %>%
  mutate(
    arm = case_when(
      trt == 0 ~ "CPM",
      trt == 1 ~ "FeZnPM"
    )
  )

rownames(baseline_metadata) <- baseline_metadata$sample_names

#update metadata table
sample_data(baseline) <- baseline_metadata
```

```{r}
#| eval: true
#| label: differences between intervention arms

#compare between intervention arms
unpaired_intervention_wilcox <- baseline_metadata %>%
  dplyr::select(
    arm,
    observed,
    chao1,
    diversity_shannon,
    evenness_pielou,
    pd
  ) %>%
  pivot_longer(cols = !arm, names_to = "diversity_metric") %>%
  group_by(diversity_metric) %>%
  wilcox_test(value ~ arm, paired = FALSE) %>%
  add_xy_position(x = "arm", step.increase = 0)
```

```{r}
#| eval: true
#| label: fig-baseline-arm-alpha
#| fig-cap: Alpha diversity statistics among baseline samples separated by intervention arm (n = 168).

plot_names <- c(
  'pd' = "Faith's Phylogentic Diversity",
  'observed' = "Observed Features",
  'diversity_shannon' = "Shannon Diversity Index",
  'evenness_pielou' = "Pielou Evenness",
  "chao1" = "Chao diversity Index"
)


baseline_metadata %>%
  dplyr::select(
    arm,
    observed,
    chao1,
    diversity_shannon,
    evenness_pielou,
    pd
  ) %>%
  pivot_longer(cols = !arm, names_to = "diversity_metric") %>%
  ggplot(aes(x = as.factor(arm), y = value)) +
  geom_boxplot(aes(fill = as.factor(arm))) +
  stat_boxplot(geom = "errorbar") +
  geom_jitter(size = 1.5, stat = "identity", width = 0.35) +
  stat_pvalue_manual(
    unpaired_intervention_wilcox,
    label = "p = {p}",
    step.increase = 0.1
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  xlab("Intervention Arm") +
  ylab(NULL) +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    panel.spacing = unit(2, "lines"),
    strip.background = element_blank(),
    strip.placement = "outside"
  ) +
  facet_wrap(
    diversity_metric ~ .,
    scales = "free",
    ncol = 2,
    labeller = as_labeller(plot_names)
  )
```

```{r}
#| eval: true
#| label: baseline alpha diversity plots
#| fig-cap: Alpha diversity statistics among baseline samples (n = 168).

baseline_metadata %>%
  dplyr::select(
    timepoint,
    observed,
    chao1,
    diversity_shannon,
    evenness_pielou,
    pd
  ) %>%
  pivot_longer(cols = !timepoint, names_to = "diversity_metric") %>%
  ggplot(aes(x = timepoint, y = value, fill = diversity_metric)) +
  geom_boxplot(position = position_dodge(5), outlier.shape = NA) +
  stat_boxplot(geom = "errorbar") +
  geom_jitter(aes(color = ), size = 1.5, stat = "identity", width = 0.35) +
  xlab(NULL) +
  ylab(NULL) +
  scale_x_discrete(label = "Baseline") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "none"
  ) +
  facet_wrap(diversity_metric ~ ., scales = "free", ncol = 2)

```

#### Phyla Relative Abundances across all Individuals

```{r}
#| eval: true
#| label: phyla relative abundance per indv
#| fig-cap: Relative abundance of bacterial phyla at baseline across all individuals (n = 168).

#baseTmp as defined in the realtive abundance boxplots above
#collapse into phyla to make computation easier
base_genus <- microbiome::aggregate_taxa(baseTmp, level = "Genus")
base_phyla <- microbiome::aggregate_taxa(baseTmp, level = "Phylum")

#top phyla at baseline for each individual
p <- plot_bar(base_phyla, fill = "Phylum")
p$data$Sample <- str_remove_all(
  p$data$Sample,
  "SO_8465_|_B_wol_alignment|_B|_G"
)

#order by high to low Pseudomonadota/Bacillota
levels <- p$data %>%
  dplyr::filter(OTU == "Pseudomonadota") %>%
  arrange(desc(Abundance))
p$data$Sample <- factor(p$data$Sample, levels = levels$Sample)

p +
  geom_bar(
    aes(fill = forcats::fct_reorder(Phylum, Abundance)),
    stat = "identity",
    position = "stack"
  ) +
  scale_fill_manual(values = phyla_colours) +
  theme(
    axis.text.x = element_text(size = 8),
    plot.margin = unit(c(1, 0.1, 0, 0), "cm"),
    legend.direction = "horizontal"
  ) + #unit(c(top, right, bottom, left), units)
  scale_y_continuous(expand = c(0, 0)) + #removes white space between plot and x axis
  ylab("Relative Abundance")

```

```{r}
#| eval: true
#| label: alpha diversity plots
#| fig-cap: Relative abundance of bacterial phyla at baseline across all individuals ordered by age (days) (n = 168).

#order samples by age
age_labels <- baseline_metadata %>%
  dplyr::select(displayid, agedays_rounded) %>%
  arrange(agedays_rounded)
p$data$Sample <- factor(p$data$Sample, levels = age_labels$displayid)

p +
  geom_bar(
    aes(fill = forcats::fct_reorder(Phylum, Abundance)),
    stat = "identity",
    position = "stack"
  ) +
  scale_fill_manual(values = phyla_colours) +
  geom_text(
    aes(y = max(Abundance + 0.05), label = agedays_rounded, group = Sample),
    angle = 90,
    size = 3,
    check_overlap = T
  ) +
  annotate(
    geom = "text",
    x = 84,
    y = 1.06,
    label = "Age (days)",
    fontface = "bold"
  ) +
  theme(
    axis.text.x = element_text(size = 8),
    plot.margin = unit(c(1, 0.1, 0, 0), "cm"),
    legend.direction = "horizontal"
  ) + #unit(c(top, right, bottom, left), units)
  scale_y_continuous(expand = c(0, 0)) + #removes white space between plot and x axis
  ylim(0, 1.06) +
  ylab("Relative Abundance")
```

#### Alpha Diveristy Summary Table

```{r}
#| eval: true
#| label: tbl-ms-2
#| tbl-cap: Microbiome composition and diversity statistics at baseline (n = 168)

top10_genera_list <- names(sort(
  apply(as.data.frame(t(otu_table(base_genus))), 2, median),
  decreasing = T
)[1:10])
top_phyla_list <- colnames(as.data.frame(t(otu_table(base_phyla))))

top10_genera <- as.data.frame(t(otu_table(base_genus))) %>%
  dplyr::select(all_of(top10_genera_list))

baseline_metadata <- baseline_metadata %>%
  mutate(
    timepoint = case_when(
      str_detect(timepoint, "Baseline") ~ "Baseline"
    )
  ) %>%
  left_join(
    .,
    rownames_to_column(as.data.frame(t(otu_table(base_phyla)))),
    by = c("sample_names" = "rowname")
  ) %>%
  left_join(
    .,
    rownames_to_column(top10_genera),
    by = c("sample_names" = "rowname")
  )

saveRDS(baseline_metadata, "data/metadata/baseline-metadata.rds")

(p <- baseline_metadata %>%
  dplyr::select(
    arm,
    `Observed features` = observed,
    `Chao 1 Index` = chao1,
    `Shannon Diversity Index` = diversity_shannon,
    Evenness = evenness_pielou,
    `Faiths Phylogenetic Diversity` = pd,
    all_of(top_phyla_list),
    all_of(top10_genera_list)
  ) %>%
  dplyr::select(!c(Unknown)) %>%
  mutate_at(vars(everything()), type.convert, as.is = TRUE) %>%
  tbl_summary(
    by = arm,
    missing = "no",
    statistic = list(
      all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
      all_categorical() ~ "{n} / {N}  \n({p}%)"
    )
  ) %>% # stats and format for categorical columns
  add_n() %>%
  add_overall() %>%
  add_p(pvalue_fun = function(x) pvalue_format(x)) %>%
  add_q(pvalue_fun = function(x) pvalue_format(x)) %>%
  bold_p(q = T) %>%
  as_gt() %>%
  fmt_markdown(columns = where(~ is.character(.x))) %>%
  fmt_number(
    columns = where(~ is.numeric(.x)),
    drop_trailing_zeros = TRUE,
    use_seps = FALSE
  ) %>%
  fmt_number(
    columns = where(~ is.numeric(.x) < 0.0001),
    pattern = "<0.0001"
  ) %>%
  fmt_number(columns = where(~ is.numeric(.x) < 0.001), pattern = "<0.001") %>%
  fmt_number(columns = where(~ is.numeric(.x) < 0.01), pattern = "<0.01") %>%
  fmt_number(columns = where(~ is.numeric(.x) < 0.05), pattern = "<0.05"))

htmltools::save_html(
  gt::as_raw_html(p),
  file.path(filepath, paste0(prefix, "alpha-div.tbl"))
)

```

### Baseline Bacteria Beta diversity

phyloseq provides a wrapper function around ggplot and NeatMapand uses ordination methods to organize the rows and columns instead of (hierarchical) cluster analysis. In many cases the ordination-based ordering does a much better job than h-clustering

```{r}
#| eval: true
#| label: MS Figure 9
#| fig-cap: "β-diversity ordination plots comparing intervention arms at baseline."

#phyloseq wraps a number of vegan package functions and can be used to call vegdist()
bray <- phyloseq::distance(baseline, method = "bray") #calculated on abundance
jac <- phyloseq::distance(baseline, method = "jaccard")
uunifrac <- phyloseq::distance(baseline, method = "uunifrac")
wunifrac <- phyloseq::distance(baseline, method = "wunifrac")

bray.ord <- phyloseq::ordinate(baseline, method = "NMDS", distance = bray)
jac.ord <- phyloseq::ordinate(baseline, method = "NMDS", distance = jac)
uunifrac.ord <- phyloseq::ordinate(
  baseline,
  method = "NMDS",
  distance = uunifrac
)
wunifrac.ord <- phyloseq::ordinate(
  baseline,
  method = "NMDS",
  distance = wunifrac
)

bray.plot <- plot_ordination(baseline, bray.ord, color = "arm") +
  labs(title = "Bray-Curtis Distance")
jac.plot <- plot_ordination(baseline, jac.ord, color = "arm") +
  labs(title = "Jaccard Disimilarity")
uunifrac.plot <- plot_ordination(baseline, uunifrac.ord, color = "arm") +
  labs(title = "Unweighted UniFrac")
wunifrac.plot <- plot_ordination(baseline, wunifrac.ord, color = "arm") +
  labs(title = "Weighted UniFrac")

#set layout
ggarrange(
  bray.plot,
  jac.plot,
  uunifrac.plot,
  wunifrac.plot,
  ncol = 2,
  nrow = 2,
  common.legend = T,
  legend = "bottom"
)

```

```{r}
#| eval: true
#| label: MS Table 2 cont.
#| tbl-cap: Beta diversity summary table
#| results: asis

dissimilarity_by_time_treatment <- list()

timepoints <- unique(sample_data(baseline)$timepoint)

distance_methods <- c("bray", "jaccard", "uunifrac", "wunifrac")

# Calculate dissimilarity for each timepoint
for (t in timepoints) {
  index <- timepoints == t
  for (method in distance_methods) {
    # Print out current timepoint and distance method for debugging
    print(paste("Timepoint:", t, "Method:", method))

    # Subset the data by timepoint
    subset_physeq_time <- phyloseq::subset_samples(baseline, index)

    # Get unique treatment groups within this timepoint
    treatments_in_timepoint <- unique(sample_data(subset_physeq_time)$arm)

    # Calculate Bray-Curtis dissimilarity for each treatment group within this timepoint
    for (treat in treatments_in_timepoint) {
      # Subset the data by treatment group within the timepoint
      subset_physeq_treatment <- phyloseq::subset_samples(
        subset_physeq_time,
        arm == treat
      )

      # Calculate dissimilarity
      dist_subset <- distance(subset_physeq_treatment, method = method)

      # Store the dissimilarity matrix in the list
      dissimilarity_by_time_treatment[[paste(
        t,
        treat,
        method,
        sep = "_"
      )]] <- as.numeric(dist_subset)
    }
  }
}

baseline_betadivCPM <- data.frame(
  dissimilarity_by_time_treatment$`Baseline_CPM_bray`,
  dissimilarity_by_time_treatment$`Baseline_CPM_jaccard`,
  dissimilarity_by_time_treatment$`Baseline_CPM_uunifrac`,
  dissimilarity_by_time_treatment$`Baseline_CPM_wunifrac`
)

baseline_betadivCPM$id <- seq_len(nrow(baseline_betadivCPM))

baseline_betadivCPM <- baseline_betadivCPM %>%
  pivot_longer(cols = -id, names_to = "variable", values_to = "value") %>%
  mutate(
    variable = str_remove_all(variable, "dissimilarity_by_time_treatment|\\.")
  ) %>%
  separate(variable, into = c("timepoint", "arm", "metric"), sep = "_") %>%
  pivot_wider(names_from = metric) %>%
  dplyr::select(!id)

baseline_betadivFeZnPM <- data.frame(
  dissimilarity_by_time_treatment$`Baseline_FeZnPM_bray`,
  dissimilarity_by_time_treatment$`Baseline_FeZnPM_jaccard`,
  dissimilarity_by_time_treatment$`Baseline_FeZnPM_uunifrac`,
  dissimilarity_by_time_treatment$`Baseline_FeZnPM_wunifrac`
)

baseline_betadivFeZnPM$id <- seq_len(nrow(baseline_betadivFeZnPM))
baseline_betadivFeZnPM <- baseline_betadivFeZnPM %>%
  pivot_longer(cols = -id, names_to = "variable", values_to = "value") %>%
  mutate(
    variable = str_remove_all(variable, "dissimilarity_by_time_treatment|\\.")
  ) %>%
  separate(variable, into = c("timepoint", "arm", "metric"), sep = "_") %>%
  pivot_wider(names_from = metric) %>%
  dplyr::select(!id)

baseline_betadiv <- bind_rows(baseline_betadivCPM, baseline_betadivFeZnPM)

saveRDS(
  baseline_betadiv,
  file.path(filepath, paste0(prefix, "baseline-beta-div.rds"))
)

(p <- baseline_betadiv %>%
  tbl_strata(
    strata = timepoint,
    .tbl_fun = ~ .x %>%
      tbl_summary(
        by = arm,
        missing = "no",
        statistic = list(
          all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
          all_categorical() ~ "{n} / {N}  \n({p}%)"
        )
      ) %>% # stats and format for categorical columns
      add_n() %>%
      add_overall() %>%
      add_p(test = everything() ~ "wilcox.test", pvalue_fun = function(x) {
        pvalue_format(x)
      }) %>%
      add_q(pvalue_fun = function(x) pvalue_format(x)) %>%
      bold_p(q = T) %>%
      add_stat(fns = all_continuous() ~ vda) %>%
      modify_fmt_fun(list(vda_magnitude ~ as.character)) %>%
      modify_header(list(
        vda_estimate ~ "**Effect Size Estimate**",
        vda_magnitude ~ "**Magnitude**"
      )),
    .header = "**{strata}**, N = {n}"
  ) %>%
  as_gt() %>%
  fmt_markdown(columns = where(~ is.character(.x))) %>%
  fmt_number(
    columns = where(~ is.numeric(.x)),
    drop_trailing_zeros = TRUE,
    use_seps = FALSE
  ) %>%
  fmt_number(
    columns = where(~ is.numeric(.x) < 0.0001),
    pattern = "<0.0001"
  ) %>%
  fmt_number(columns = where(~ is.numeric(.x) < 0.001), pattern = "<0.001") %>%
  fmt_number(columns = where(~ is.numeric(.x) < 0.01), pattern = "<0.01") %>%
  fmt_number(columns = where(~ is.numeric(.x) < 0.05), pattern = "<0.05"))


htmltools::save_html(
  gt::as_raw_html(p),
  file.path(filepath, paste0(prefix, "beta-div.tbl"))
)

```
