## Investigating changes in OGU abundance over time using Generalised Estimating Equations

```{r}
#| echo: false
#| eval: true
#| label: setup

# This initial code block loads custom utility functions and environment variables
# These are stored in separate files to keep the main analysis clean and organized
source("utils/r-utils.r")
source("scripts/gee.r")

source("_environment.required")
load("data/seqdata.RData")

suppressPackageStartupMessages(library(glmtoolbox))
suppressPackageStartupMessages(library(geepack))
suppressPackageStartupMessages(library(rms))

dir.create(file.path(RESULTS, "EvsB-OGU-GEE"), showWarnings = FALSE, recursive = TRUE)

```

This analysis focuses on univariate analysis of OGU abundances using generalised
estimating equations.

### Data preparation

In general unclassified taxa and taxa with low abundance are removed from the dataset
before modelling and can be built using read counts or realtive abundances [@Xia2023].
In this analysis, filtering was performed to remove OGUs with 0 counts across all
samples.

```{r}
#| eval: true
#| label: data prep

tmp <- as.data.frame(t(otu_table(bac_base_end))) #taxa(OGUs) are rows so transform to columns

bac_base_end_ogu <- tmp[,colSums(tmp)!=0] #remove OGUs with 0 abundance across all sampels
bac_base_end_taxa <- as.data.frame(tax_table(bac_base_end))
bac_base_end_meta <- as.data.frame(as.matrix(sample_data(bac_base_end)))
```

Are the rownames of the OGU and metadata dataframes in the same order:
`r all(rownames(bac_base_end_ogu) == rownames(bac_base_end_meta))`

```{r}
#| eval: true
#| label: tbl-variables
#| tbl-cap: Summary of LMM outcome and predictor variables.

outcome <- as.matrix(bac_base_end_ogu)

#remove any NAs and replace with 0
outcome <- ifelse(is.na(outcome), 0, outcome)
#create matrix of OGU IDs
ogu_id <- colnames(outcome)

#identify covariates for the model
predictors <- bac_base_end_meta %>% 
                mutate(
                  time = scale_values(as.numeric(timesincefeedstart))
                ) %>%
                group_by(displayid) %>%
                mutate(agedays_rounded = as.numeric(agedays_rounded)) %>%
                arrange(agedays_rounded, .by_group = TRUE) %>%
                mutate(age_B = case_when(
                  timepoint == "Baseline" ~ agedays_rounded,
                  FALSE ~ NA)) %>%
                fill(age_B) %>%
                ungroup() %>%
                mutate(
                  age_B = scale_values(age_B),
                  arm = as.factor(arm),
                  subject = str_remove(displayid,"-G|-B")
                ) %>%
                column_to_rownames(var = "sample_names") %>%
                select(
                  subject,
                  arm,
                  time, #numeric (days)
                  age_B #numeric (days)
                ) 

# Generate summary table
describe(predictors)
```

```{r}
#| eval: true
#| label: tbl-multi-cor
#| tbl-cap: Correlation coefficients from multicolinearity analysis of predictor varaibles

tmp <- predictors %>% select(time, age_B) %>% mutate(across(everything(), as.numeric))

kable(tmp %>% cor(method = "spearman", use = "pairwise.complete.obs"))

```

### Identifying significant changes in OGU relative abundances overtime

The below loop will run individual random intercept model for each OGU/taxon
($y = relAbund$)

```{r}
#| eval: true
#| label: changes in individual OGUs

#create an empty list to hold the model results in
model <- list()
index = intercept = rest1 = rest2 = rest3 = rest4 = matrix(NA, ncol(outcome),1)

if(is.matrix(outcome)){
  for(j in 1:ncol(outcome)){
    tryCatch({
      y <- outcome[,j]
      df <- data.frame(y=y, predictors) %>% arrange(subject, time) # order for gee

      model <- glmgee(y ~ time * arm + age_B, id = subject, data = df, verbose = FALSE, family = poisson(link ="log")) 
      
      index[j,] <- j
      #extract p-values
      intercept[j,] <- summary(model)$coefficients[1,4] #intercept
      rest1[j,] <- summary(model)$coefficients[2,4] #time
      rest2[j,] <- summary(model)$coefficients[3,4] #arm
      rest3[j,] <- summary(model)$coefficients[4,4] #age_B
      rest4[j,] <- summary(model)$coefficients[5,4] #time*arm

    }, error=function(e){cat ("ERROR at j =", j, ":", conditionMessage(e), "\n")})
  }
} else message("Outcome variable should be in a matrix")

#adjust P-values: Benjimani Hochberg
all_gee <- as.data.frame(cbind(intercept, rest1, rest2, rest3, rest4))
index <- index
intercept_gee_adj <- round(p.adjust(intercept, method = "BH"), 4)
time_gee_adj <- round(p.adjust(rest1, method = "BH"), 4)
arm_gee_adj <- round(p.adjust(rest2, method = "BH"), 4)
age_gee_adj <- round(p.adjust(rest3, method = "BH"), 4)
interaction_gee_adj <- round(p.adjust(rest4, method = "BH"), 4)

all_gee_adj <- as.data.frame(cbind(index, ogu_id, intercept_gee_adj, arm_gee_adj,time_gee_adj,interaction_gee_adj, age_gee_adj))

#export results table
write.csv(all_gee_adj, file.path(RESULTS, "EvsB-OGU-GEE/OGU-GEE-adj-p-values.csv"), row.names = FALSE)

apply(all_gee, 2, function(x) any(x <= 0.05, na.rm = TRUE)) #determine which column if any has significant p-values
apply(all_gee_adj, 2, function(x) any(x <= 0.05, na.rm = TRUE)) #determine which column if any has significant p-values

bac_taxa_signif <- left_join(rownames_to_column(bac_base_end_taxa), all_gee_adj, by=c("rowname" = "ogu_id")) %>% 
  dplyr::filter(if_any(c("time_gee_adj", "arm_gee_adj", "interaction_gee_adj", "age_gee_adj"),~. <= 0.05)) 

#export significant OGUs
write.csv(bac_taxa_signif, file.path(RESULTS, "EvsB-OGU-GEE/significant-OGU-GEE-adj-p-values.csv"), row.names = FALSE)
```

```{r}
#| eval: true
#| label: summary statistics for significant OGUs
ogu_gee_results <- lapply(bac_taxa_signif$rowname, function(x) {
    y <- outcome[,as.character(x)]

    df <- data.frame(y=y, predictors) %>% arrange(subject, time) # order for gee

    model <- glmgee(y ~ time * arm + age_B, id = subject, data = df, family = poisson(link ="log")) 

    stats <- summary(model)
    list <- list(model = model, stats = stats)
    
    return(list)
})

which(colnames(outcome) %in% c("G000379625"))

df <- data.frame(y=outcome[,1012], predictors) %>% arrange(subject, time) # order for gee

m1 <- glmgee(y ~ time * arm + age_B, id = subject, data = df, corstr="Exchangeable", verbose = FALSE) 
summary(model)
m2 <- update(model, corstr="Independence")
m3 <- update(model, corstr="Unstructured")
m4 <- update(model, corstr="Stationary-M-dependent(1)")
m5 <- update(model, corstr="Non-Stationary-M-dependent(1)")

lapply(list(m1,m2,m3,m4,m5), summary)

a <- glmtoolbox::CIC(m1, m2, m3, m4, m5,verbose=FALSE)
b <- glmtoolbox::QIC(m1, m2, m3, m4, m5,verbose=FALSE)
c <- glmtoolbox::GHYC(m1, m2, m3, m4,m5,  verbose=FALSE)
d <- glmtoolbox::RJC(m1, m2, m3, m4, m5,verbose=FALSE)
e <- glmtoolbox::AGPC(m1, m2, m3, m4,m5,  verbose=FALSE)
f <- glmtoolbox::SGPC(m1, m2, m3, m4,m5,  verbose=FALSE)

cbind(a,QIC=b[,"QIC"],GHYC=c[,"GHYC"],RJC=d[,"RJC"],AGPC=e[,"AGPC"],SGPC=f[,"SGPC"])

anova(m3, test="wald")
anova(m3a, test="score")

cbind(model=diag(vcov(m3, type="model")), robust=diag(vcov(m3, type="robust")), bias.corrected=diag(vcov(m3, type="bias-corrected")), jackknife=diag(vcov(m3, type="jackknife")))

residuals(m3, type="deviance", plot.it=TRUE, pch=16)
residuals(m3, type="mahalanobis", plot.it=TRUE, identify=3)

#saveRDS(ogu_gee_results, paste("07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/Base_End_Bacteria_gee_top_significant_OGUs_model_summaries", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.rds", sep = "_"))

ogu_gee_results <- readRDS("07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/Base_End_Bacteria_gee_top_significant_OGUs_model_summaries_Mumbai_ID14768_Jan_19_2024_04:52_NC.rds")
```

### OGUs Common between LMMs and differential abundance changes over time

Update: New results as of 26-08-24 having been rectified due to the mistake in
contrast assignment (`Baseline-Endpoint` instead of `Endpoint-Baseline`)

```{r}
#| eval: true
#| label: changes in relative abundace plot
#common <- read.csv("07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/bac_base_end_timepoint_up_down_reg_OGUs_FDR5_FC2_gee_Mumbai_ID14768_Feb_07_2024_02:35_NC.csv", header = T)

#updated results file 22-08-24 
common <- read.csv("~/Work/05-Projects/01-Precision_nutrition/07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/bac_base_end_timepoint_up_down_reg_OGUs_FDR5_FC2_gee_Mumbai_ID14768_Aug_22_2024_04:23_NC.csv", header = T)

dim(common)

# some OGUs represent the same species, therefore I will combine them and take the sum of their FC
common_combined <- common %>%
  ungroup() %>%
  group_by(Phylum, Species) %>%
  dplyr::summarise(logFC = sum(logFC), FC = fc(logFC)) %>%
  arrange(desc(abs(logFC)), .by_group = T) %>%
  as.data.frame()

phylum_anno <- assign_unique_colors(as.vector(common_combined$Phylum),phyla_colours)

paste0("Some OGUs represent the same species. When combined there are ", nrow(common_combined), " total species")
```

```{r}
#| eval: true
#| label: Combined OGU fold change table
#| tab-cap: For significant fold changes of 2 or more, the number of OGUs common between differential abundance results and LMM results is 106, representing 91 unique species

common_combined %>% 
    arrange(desc(logFC)) %>%
    kable()
```

```{r}
#| eval: true
#| label: Combined OGU fold change table
#| tab-cap: For significant fold changes of 2 or more, the percentage of OGUs in each phyla that are common between differential abundance results and LMM results.

common_combined %>% 
  group_by(Phylum) %>%
  summarise(total = n(), percent = n()/nrow(common_combined)*100) %>%
  kable()
```

```{r}
#| eval: true
#| label: plots
#| fig-cap: Log2 fold change of OGUs shared across LMM and DA results

#png(paste("~/Work/05-Projects/01-Precision_nutrition/07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/bac_base_end_timepoint_up_down_reg_OGUs_FDR5_FC2_gee_barplot", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.png", sep = "_"), 1200, 1500)

common_combined %>% 
    arrange(logFC) %>%
    ggplot(aes(y=reorder(Species,logFC), x=logFC, fill = Phylum))+  
      geom_bar(position=position_dodge(), stat="identity")+
      xlab(NULL) + 
      ylab(NULL) + #labels the y axis
      scale_color_manual(values= phylum_anno, aesthetics= 'fill')+
      geom_vline(xintercept = 0) +
      theme(axis.text.x = element_text(angle = 0, size = fontsize, face='bold', color='black'), 
        axis.text.y = element_text(size = 12, face='bold', color='black'))

#dev.off()
```

```{r}
#| eval: true
#| label: top50 OGU barplot
#| fig-cap: Log2 fold change of the top 50 OGUs shared across LMM and DA results

#png(paste("~/Work/05-Projects/01-Precision_nutrition/07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/bac_base_end_timepoint_up_down_reg_OGUs_FDR5_FC2_gee_barplot_top50", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.png", sep = "_"), 1200, 800)

common_combined %>% 
    arrange(desc(logFC)) %>%
    slice(1:50) %>%
    ggplot(aes(y=reorder(Species,logFC), x=logFC, fill = Phylum))+  
      geom_bar(position=position_dodge(), stat="identity")+
      xlab(NULL) + 
      ylab(NULL) + #labels the y axis
      scale_color_manual(values= phylum_anno, aesthetics= 'fill')+
      geom_vline(xintercept = 0) +
      theme(axis.text.x = element_text(angle = 0, size = fontsize, face='bold', color='black'), 
        axis.text.y = element_text(size = 12, face='bold', color='black'))

#dev.off()
```

The below chunk is no longer needed with the new reuslts only having 106 OGUs rather
than 209.

```{r}
#| eval: false
#| label: breakdown of barplot for better visualisation

p1 <- common_combined %>% 
    arrange(logFC) %>%
    slice(1:45) %>%
    ggplot(aes(y=reorder(Species,logFC), x=logFC, fill = Phylum))+  
      geom_bar(position=position_dodge(), stat="identity")+
      xlab(NULL) + 
      ylab(NULL) + #labels the y axis
      scale_color_manual(values= phylum_anno, aesthetics= 'fill')+
      geom_vline(xintercept = 0) +
      theme(axis.text.x = element_text(angle = 0, size = fontsize, face='bold', color='black'), 
        axis.text.y = element_text(size = 8, face='bold', color='black'))

p2 <-  common_combined %>% 
    arrange(logFC) %>%
    slice(45:90) %>%
    ggplot(aes(y=reorder(Species,logFC), x=logFC, fill = Phylum))+  
      geom_bar(position=position_dodge(), stat="identity")+
      xlab(NULL) + 
      ylab(NULL) + #labels the y axis
      scale_color_manual(values= phylum_anno, aesthetics= 'fill')+
      geom_vline(xintercept = 0) +
      theme(axis.text.x = element_text(angle = 0, size = fontsize, face='bold', color='black'), 
        axis.text.y = element_text(size = 8, face='bold', color='black'))

p3 <-  common_combined %>% 
    arrange(logFC) %>%
    slice(90:135) %>%
    ggplot(aes(y=reorder(Species,logFC), x=logFC, fill = Phylum))+  
      geom_bar(position=position_dodge(), stat="identity")+
      xlab(NULL) + 
      ylab(NULL) + #labels the y axis
      scale_color_manual(values= phylum_anno, aesthetics= 'fill')+
      geom_vline(xintercept = 0) +
      theme(axis.text.x = element_text(angle = 0, size = fontsize, face='bold', color='black'), 
        axis.text.y = element_text(size = 8, face='bold', color='black'))

p4 <-  common_combined %>% 
    arrange(logFC) %>%
    slice(135:185) %>%
    ggplot(aes(y=reorder(Species,logFC), x=logFC, fill = Phylum))+  
      geom_bar(position=position_dodge(), stat="identity")+
      xlab(NULL) + 
      ylab(NULL) + #labels the y axis
      scale_color_manual(values= phylum_anno, aesthetics= 'fill')+
      geom_vline(xintercept = 0) +
      theme(axis.text.x = element_text(angle = 0, size = fontsize, face='bold', color='black'), 
        axis.text.y = element_text(size = 8, face='bold', color='black'))

#png(paste("07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/bac_base_end_timepoint_up_down_reg_OGUs_FDR5_FC2_gee_barplot", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.png", sep = "_"), 1500, 1300)

ggarrange(p1,p2,p3,p4, legend.grob = get_legend(p))

#dev.off()
```