## Calculating Zero-inflated Model Diagnostics outside of the genZI function

```{r}
#| eval: false
#| label: diagnostics
diagnostics <- lapply(list, function(x) {
  lapply(x$all_models, function(i) {
    diagnostics <- zinb_diagnostics(i, predictors)
  })
})

model_names <- names(list)

all_summaries <- lapply(model_names, function(name) {
  sum <- create_model_summary(diagnostics[[name]], name)
  sum$converged <- list[[name]]$model_summary$converged
  sum$zero_prop <- list[[name]]$model_summary$zero_prop
  sum$error_message <- NULL

  return(sum)
})

# Combine all summaries
combined_summary <- do.call(rbind, all_summaries)

# Wide format for criteria comparison
criteria_comparison <- combined_summary %>%
  pivot_longer(
    cols = !c(outcome_id, model),
    names_to = "criterion",
    values_to = "value"
  ) %>%
  pivot_wider(names_from = model, values_from = value)

# Diagnostic test comparison
diagnostic_comparison <- combined_summary %>%
  select(outcome_id, model, contains("dharma"), contains("ratio")) %>%
  pivot_longer(
    cols = -c(outcome_id, model),
    names_to = "test",
    values_to = "value"
  ) %>%
  pivot_wider(names_from = model, values_from = value)

# Create visualization plots
plots <- list()

# AIC/BIC comparison
plots$criteria <- ggplot(combined_summary, aes(x = model, y = AIC)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Model Comparison by AIC", y = "AIC Value")

# Diagnostic test p-values
plots$diagnostics <- combined_summary %>%
  select(model, contains("Pvalue")) %>%
  pivot_longer(
    cols = contains("Pvalue"),
    names_to = "test",
    values_to = "p_value"
  ) %>%
  ggplot(aes(x = model, y = p_value)) +
  geom_boxplot() +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  facet_wrap(~test) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Diagnostic Test P-values by Model")


```


```{r}
#| eval: false
names <- names(list)

modCompare <- lapply(1:length(list), function(i) {
  name <- names[[i]]
  mod <- list[[i]]

  mod$model_summary %>%
    dplyr::select(outcome_id, AIC, BIC, dispersion) %>%
    mutate(model = name) %>%
    pivot_longer(cols = c(AIC, BIC, dispersion)) %>%
    pivot_wider(names_from = model, values_from = value) %>%
    mutate_at(vars(everything()), str_trim) %>%
    mutate_at(vars(everything()), type.convert, as.is = T)
})

modCompare <- reduce(modCompare, left_join, by = c("outcome_id", "name"))

kable(modCompare)

lapply(noZi_rand_models$all_models, getME, name = "theta")
lapply(noZi_rand_models$all_models, sigma)

results <- compare_zi_models(list)
```


## Building Null Models

```{r fit-null}
#| eval: false
#| label: null-models

noZi_models <- genZI(
  outcome,
  predictors,
  formula = "1",
  zi = "~ 0"
)

noZi_rand_models <- genZI(
  outcome,
  predictors,
  formula = "1 + (1 | subject)",
  zi = "~ 0"
)

null_models <- genZI(outcome, predictors, formula = "1", zi = "~ 1")

null_rand_models <- genZI(
  outcome,
  predictors,
  formula = "1 + (1 | subject)",
  zi = "~ 1"
)
```

```{r save}
#| eval: false
#| echo: false
saveRDS(
  noZi_models,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "genZI-OGU-noZi-nb2_results.rds")
)
saveRDS(
  noZi_rand_models,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "genZI-OGU-noZi-rand-nb2_results.rds")
)
saveRDS(
  null_models,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "genZI-OGU-null-nb2_results.rds")
)
saveRDS(
  null_rand_models,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "genZI-OGU-null-rand-nb2_results.rds")
)
```

```{r load}
#| eval: true
#| echo: false

noZi_models <- readRDS(file.path(
  RESULTS,
  "EvsB-OGU-ZIBR",
  "genZI-OGU-noZi-nb2_results.rds"
))

noZi_rand_models <- readRDS(file.path(
  RESULTS,
  "EvsB-OGU-ZIBR",
  "genZI-OGU-noZi-rand-nb2_results.rds"
))

null_models <- readRDS(file.path(
  RESULTS,
  "EvsB-OGU-ZIBR",
  "genZI-OGU-null-nb2_results.rds"
))

null_rand_models <- readRDS(file.path(
  RESULTS,
  "EvsB-OGU-ZIBR",
  "genZI-OGU-null-rand-nb2_results.rds"
))
```

```{r nested-anova}
#| eval: true
#| label: null-anovas
#| tbl-cap: NA

nested_null <- compare_nested_models(list(
  noZi_models = noZi_models,
  noZi_rand_models = noZi_rand_models,
  null_models = null_models,
  null_rand_models = null_rand_models
))

nested_null$comparisons %>%
  filter(model1 == "noZi_models", significant == TRUE) %>%
  n_distinct("outcome_id")
```

### Model diagnostic failures

```{r}
#| eval: true
#| label: list

list <- list(
  noZi_models = noZi_models,
noZi_rand_models = noZi_rand_models,
null_models = null_models,
null_rand_models = null_rand_models,
  gau = gau_results,
  lognorm = lognorm_results,
  poi = poi_results,
  nb1 = nb1_results,
  nb2 = nb2_results
)
```

```{r}
#| eval: true
#| label: diagnose-mods

poor_fits <- lapply(1:length(list), function(x) {
  d <- diagnoseZI(list[[x]]$all_models)
  d$dist <- names(list)[x]

  return(d)
})

poor_fits <- data.table::rbindlist(poor_fits)
```

## Previous model fitting attempts

### Fitting the best model

#### Zero-inflated Negative Binomial 1 Models

```{r}
#| eval: true
#| label: nbinom1 unfit models

fitted_models <- keep_at(
  nb1_results$all_models,
  filter(bestDist_filtered, dist == "nb1")$outcome_id
)

nb1_unfit <- diagnoseZI(fitted_models)

any(nb1_unfit$pass == FALSE)

```

```{r}
#| eval: true
#| label: prune

nb1_models <- discard_at(fitted_models, nb1_unfit)

nb1_summary <- data.frame(
  outcome_id = names(nb1_models),
  continuous_distribution = "nbinom1",
  continuous_link = "log",
  covaraince_structure = "unstructured",
  zero_distribution = "binary",
  zero_link = "logit",
  optimizer = "nlminb"
)
```

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for nbinom1
nb1_dharma_results <- process_dharma_results(nb1_models)

kable(nb1_dharma_results$summary)

```

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Significant Dharma residual tests
nb1_dharma_signif <- filter(nb1_dharma_results$complete, p.value <= 0.05)

kable(
  nb1_results$model_summary %>%
    filter(outcome_id %in% names(nb1_models)) %>%
    filter(
      ratioObsSim_Pvalue <= 0.05 |
        dharmaDisp_Pvalue <= 0.05 |
        dharmaKS_Pvalue <= 0.05 |
        dharmaOutliers_Pvalue <= 0.05
    )
)

```

A total of `r length(nb1_unfit)` models did not converge appropriately and were refit with a negative binomial (nbinom2) distribution.

#### Zero-inflated Negative Binomial Polynomial Models

```{r}
#| eval: true
#| label: refits
nb2_outcomes <- outcome %>%
  select(all_of(nb1_unfit))

ncol(nb2_outcomes) == length(nb1_unfit)

```

```{r}
#| eval: true
#| label: tbl-model-diagnostics
fitted_models <- keep_at(nb2_results$all_models, nb1_unfit)

nb2_unfit <- diagnoseZI(fitted_models)

nb2_unfit <- nb2_unfit %>%
  filter(pass == FALSE)

nb2_unfit <- nb2_unfit[, 1]

length(nb2_unfit)
```

```{r}
#| eval: true
#| label: prune

nb2_models <- discard_at(fitted_models, nb2_unfit)

nb2_summary <- data.frame(
  outcome_id = names(nb2_models),
  continuous_distribution = "nbinom2",
  continuous_link = "log",
  covaraince_structure = "unstructured",
  zero_distribution = "binary",
  zero_link = "logit",
  optimizer = "nlminb"
)
```

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for nbinom2
nb2_dharma_results <- process_dharma_results(nb2_models)

kable(nb2_dharma_results$summary)

```

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Significant Dharma residual tests fo nbinom2
nb2_dharma_signif <- filter(nb2_dharma_results$complete, p.value <= 0.05)

kable(
  nb2_results$model_summary %>%
    filter(outcome_id %in% names(nb2_models)) %>%
    filter(
      ratioObsSim_Pvalue <= 0.05 |
        dharmaDisp_Pvalue <= 0.05 |
        dharmaKS_Pvalue <= 0.05 |
        dharmaOutliers_Pvalue <= 0.05
    )
)

```

A total of `r length(nb2_unfit)` models did not converge appropriately and were refit with a Poisson distribution.

#### Zero-inflated Poisson Models

```{r}
#| eval: true
#| label: refits
poi_outcomes <- outcome %>%
  select(all_of(nb2_unfit))

ncol(poi_outcomes) == length(nb2_unfit)

```

```{r}
#| eval: true
#| label: poi unfit models

fitted_models <- keep_at(poi_results$all_models, nb2_unfit)

poi_unfit <- diagnoseZI(fitted_models)

poi_unfit <- poi_unfit %>%
  filter(pass == FALSE)

poi_unfit <- poi_unfit[, 1]

```

```{r}
#| eval: true
#| label: prune

length(poi_models <- discard_at(fitted_models, poi_unfit))

poi_summary <- data.frame(
  outcome_id = names(poi_models),
  continuous_distribution = "lognormal",
  continuous_link = "log",
  covaraince_structure = "unstructured",
  zero_distribution = "binary",
  zero_link = "logit",
  optimizer = "nlminb"
)
```

```{r}
poi_dharma_results <- process_dharma_results(poi_models)

kable(poi_dharma_results$summary)

```

A total of `r length(poi_unfit)` models did not converge appropriately and were refit with a negative binomial polynomial (nbinom2) distribution.

#### Zero-inflated Gaussian Models

```{r}
#| eval: true
#| label: refits
gau_outcomes <- outcome %>%
  select(all_of(nb2_unfit))

ncol(gau_outcomes) == length(nb2_unfit)
```

```{r}
#| eval: true
#| label: poisson unfit models

fitted_models <- keep_at(gau_results$all_models, nb2_unfit)

gau_unfit <- diagnoseZI(fitted_models)

gau_unfit <- gau_unfit %>%
  filter(pass == FALSE)

gau_unfit <- gau_unfit[, 1]

length(gau_unfit)
```

```{r}
#| eval: true
#| label: prune

length(gau_models <- discard_at(fitted_models, gau_unfit))

gau_summary <- data.frame(
  outcome_id = names(gau_models),
  continuous_distribution = "gaussian",
  continuous_link = "identity",
  covaraince_structure = "unstructured",
  zero_distribution = "binary",
  zero_link = "logit",
  optimizer = "nlminb"
)

gau_dharma_results <- process_dharma_results(gau_models)

kable(gau_dharma_results$summary)

```

A total of `r length(gau_unfit)` models did not converge appropriately and were refit with a lognormal distribution.

#### Zero-inflated Lognormal Models

```{r}
#| eval: true
#| label: refits
lognorm_outcomes <- outcome %>%
  select(all_of(gau_unfit))

ncol(lognorm_outcomes) == length(gau_unfit)
```

```{r}
#| eval: true
#| label: lognormal unfit models

fitted_models <- keep_at(lognorm_results$all_models, gau_unfit)

lognorm_unfit <- diagnoseZI(fitted_models)

lognorm_unfit <- lognorm_unfit %>%
  filter(pass == FALSE)

lognorm_unfit <- lognorm_unfit[, 1]

length(lognorm_unfit)
```

```{r}
#| eval: true
#| label: prune

length(lognorm_models <- discard_at(fitted_models, lognorm_unfit))

lognorm_summary <- data.frame(
  outcome_id = names(lognorm_models),
  continuous_distribution = "lognormal",
  continuous_link = "log",
  covaraince_structure = "unstructured",
  zero_distribution = "binary",
  zero_link = "logit",
  optimizer = "nlminb"
)

lognorm_dharma_results <- process_dharma_results(lognorm_models)

kable(lognorm_dharma_results$summary)

```

#### Comparing Model AIC and BIC values

```{r}
#| eval: true
#| label: summaries

summaries <- list(
  gau = gau_results$model_summary,
  lognorm = lognorm_results$model_summary,
  poi = poi_results$model_summary,
  nb1 = nb1_results$model_summary,
  nb2 = nb2_results$model_summary
)

summaries <- data.table::rbindlist(summaries, idcol = "dist")

summaries <- left_join(summaries, poor_fits, by = c("outcome_id", "dist"))
```

```{r}
#| eval: true
#| label: any-errors

if (any(summaries$error_message != "")) {
  print("Some models returned an error message")
}

```

```{r}
#| eval: true
#| label: diag-compare
summaries <- summaries %>%
  filter(pass == TRUE) %>%
  select(!c(error_message, message)) %>%
  mutate(index = 1:nrow(.)) %>% #make it easier to distinguish between outcome ids
  pivot_longer(
    cols = !c(index, outcome_id, dist, pass),
    names_to = "criterion",
    values_to = "value"
  )


```

```{r}
#| eval: true
#| label: tbl-best-AIC
#| tbl-cap: lowest AIC value for models that passed diagnostics

aic <- summaries %>%
  filter(criterion == "AIC") %>%
  group_by(criterion, outcome_id) %>%
  arrange(value) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, minAIC = value, distAIC = dist)
```

```{r}
#| eval: true
#| label: tbl-best-BIC
#| tbl-cap: lowest BIC value for models that passed diagnostics

bic <- summaries %>%
  filter(criterion == "BIC") %>%
  group_by(criterion, outcome_id) %>%
  arrange(value) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, minBIC = value, distBIC = dist)

```

```{r}
#| eval: true
#| label: tbl-best-fits
#| tbl-cap: Best fit models based on AIC and BIC

(bestDist <- left_join(aic, bic) %>%
  mutate(dist = ifelse(distAIC == distBIC, distAIC, "Undetermined")))
```

```{r}
#| eval: true
#| label: tbl-undetermined
#| tbl-cap: Undetermined fit

bestDist %>% filter(dist == "Undetermined")

```

Where there was an undetermined best distribution to use based on AIC and BIC, other metrics were examined e.g significant deviations from assumptions of uniformity, low dispersion and few outliers. 

> [!note]
> The larger the dispersion value, $\theta$ in `nbinom2` models, the less variance there is in the model [@Brooks2017].

`dharmaDisp` represents the dispersion ratio, which should ideally be close to 1

```{r}
#| eval: true
#| label: tbl-best-disp
#| tbl-cap: lowest disp value for models that passed diagnostics

dispRatio <- summaries %>%
  filter(criterion == "dharmaDisp") %>%
  group_by(criterion, outcome_id) %>%
  mutate(skew = abs(value - 1)) %>%
  arrange(skew) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, bestDisp = value, distDisp = dist)
```

```{r}
#| eval: true
#| label: tbl-pvals
#| tbl-cap: NA

pvals <- left_join(
  bestDist,
  summaries,
  by = c("index", "outcome_id", "dist")
) %>%
  filter(str_detect(criterion, "Pvalue"), value <= 0.05) %>%
  arrange(index)

```

`r n_distinct(pvals$outcome_id)` models have a significant test result making the model fit problematic

```{r}
#| eval: true
#| label: best-dist-filtered
#| fig-cap: NA
#| tbl-cap: NA

bestDist_filtered <- bestDist %>%
  filter(!outcome_id %in% pvals$outcome_id)
```

```{r}
#| eval: true
#| label: tbl-summary

poi_summary <- left_join(poi_summary, poi_ogu_models)
nb1_summary <- left_join(nb1_summary, nb1_ogu_models)

nb2_summary <- left_join(nb2_summary, nb2_ogu_models)
gau_summary <- left_join(gau_summary, gau_ogu_models)
lognorm_summary <- left_join(lognorm_summary, lognorm_ogu_models)

summary <- rbind(
  nb1_summary,
  nb2_summary,
  poi_summary,
  gau_summary,
  lognorm_summary
)

write.csv(
  summary,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "zero-inflated-ogu-model-summaries.csv"),
  row.names = FALSE
)
```

#### Log liklihood tests for nested models

> [!cite]
> The best way to judge whether the ZIP is acceptable is to compare it with a ZINB as these models are nested [@Zuur2009].

```{r}
#| eval: true
#| label: model-anova
test <- list(
  nb1_results = nb1_results,
  nb2_results = nb2_results,
  poi_results = poi_results
)

nested_anova <- compare_nested_models(test)
```

```{r}
#| eval: true
#| label: poi-anova

# Print summary
poi_anova <- nested_anova$comparisons %>%
  filter(
    outcome_id %in% names(poi_models),
    model1 == "poi_results" |
      model2 == "poi_results"
  )
```

Most of the data is highly dispersed so a NB model would be more appropriate than Poisson

## Significant OGUs

### Negative Binomial Distribution

```{r}
#| eval: true
#| label: tbl-nb-sig-ogus
nb1_ogu_models <- nb1_results$adj_results %>%
  filter(outcome_id %in% names(nb1_models))

nb1_sig_ogus <- identify_significant_ogus(
  nb2_ogu_models,
  data$taxa_data
)

dim(nb1_sig_ogus)

```

### Negative Binomial Polynomial Distribution

```{r}
#| eval: true
#| label: tbl-nb-sig-ogus
nb2_ogu_models <- nb2_results$adj_results %>%
  filter(outcome_id %in% names(nb2_models))

nb2_sig_ogus <- identify_significant_ogus(
  nb2_ogu_models,
  data$taxa_data
)

dim(nb2_sig_ogus)

```

### Poisson Distribution

```{r}
#| eval: true
#| label: tbl-poi-sig-ogus
poi_ogu_models <- poi_results$adj_results %>%
  filter(outcome_id %in% names(poi_models))

poi_sig_ogus <- identify_significant_ogus(
  poi_ogu_models,
  data$taxa_data
)

dim(poi_sig_ogus)

```

### Gaussian Distribution

```{r}
#| eval: true
#| label: tbl-gau-sig-ogus
gau_ogu_models <- gau_results$adj_results %>%
  filter(outcome_id %in% names(gau_models))

gau_sig_ogus <- identify_significant_ogus(
  gau_ogu_models,
  data$taxa_data
)

dim(gau_sig_ogus)

```

### Lognormal Distribution

```{r}
#| eval: true
#| label: tbl-lognorm-sig-ogus
lognorm_ogu_models <- lognorm_results$adj_results %>%
  filter(outcome_id %in% names(lognorm_models))

lognorm_sig_ogus <- identify_significant_ogus(
  lognorm_ogu_models,
  data$taxa_data
)

dim(lognorm_sig_ogus)

```

## Summary

```{r}
#| eval: true
#| label: tbl-summary

poi_summary <- left_join(poi_summary, poi_ogu_models)
nb1_summary <- left_join(nb1_summary, nb1_ogu_models)

nb2_summary <- left_join(nb2_summary, nb2_ogu_models)
gau_summary <- left_join(gau_summary, gau_ogu_models)
lognorm_summary <- left_join(lognorm_summary, lognorm_ogu_models)

summary <- rbind(
  nb1_summary,
  nb2_summary,
  poi_summary,
  gau_summary,
  lognorm_summary
)

write.csv(
  summary,
  file.path(RESULTS, "EvsB-OGU-ZIBR", "zero-inflated-ogu-model-summaries.csv"),
  row.names = FALSE
)
```

Of the `r ncol(outcome)` OGUs that passed filtering, `r nrow(summary)` were sucessfully fit to a xero-inflated model (`r nrow(summary)/ncol(outcome)*100`%)

```{r}
#| eval: true
#| label: zi-sig

ziSig <- identify_significant_ogus(
  summary,
  data$taxa_data
)

colnames(ziSig)[colnames(ziSig) == "rowname"] <- "outcome_id"

tmp <- joined_data %>%
  dplyr::filter(outcome_id %in% ziSig$outcome_id) %>%
  group_by(sample, outcome_id) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = value,
    values_fill = 0
  ) %>%
  mutate(EvsB = Endpoint - Baseline) %>%
  mutate(
    difference = case_when(
      EvsB < 0 ~ "Decrease",
      EvsB > 0 ~ "Increase",
      EvsB == 0 ~ "None",
      TRUE ~ NA
    )
  )

tmp %>%
  group_by(outcome_id, difference) %>%
  summarise(total = n())
```

```{r}
#| eval: true
#| label: tbl-average-sig-changes

aveChange <- joined_data %>%
  dplyr::filter(outcome_id %in% ziSig$outcome_id) %>%
  group_by(outcome_id, timepoint) %>%
  summarise(average_count = mean(value)) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = average_count,
    values_fill = 0
  ) %>%
  mutate(EvsB = Endpoint - Baseline) %>%
  mutate(
    difference = case_when(
      EvsB < 0 ~ "Decrease",
      EvsB > 0 ~ "Increase",
      EvsB == 0 ~ "None",
      TRUE ~ NA
    )
  )

aveChange %>%
  group_by(difference) %>%
  summarise(total = n())

ziSig <- left_join(ziSig, aveChange, by = "outcome_id")

write.csv(
  ziSig,
  file.path(
    RESULTS,
    "EvsB-OGU-ZIBR",
    "significant-ogus-zero-inflated-model.csv"
  ),
  row.names = FALSE
)

ziSig <- read.csv(file.path(
  RESULTS,
  "EvsB-OGU-ZIBR",
  "significant-ogus-zero-inflated-model.csv"
))

```

```{r}
#| eval: true
#| label: fig-alluvial

# Modify abundance categories
create_alluvial(joined_data, "G004561445", min_value = 1) # Remove zeros

# Add arm information
create_alluvial <- function(data, outcome_id, min_value = 0) {
  alluvial_data <- data %>%
    filter(outcome_id == !!outcome_id) %>%
    filter(value > min_value) %>%
    mutate(
      abundance = cut(
        value,
        breaks = c(-Inf, 1, 10, 100, Inf),
        labels = c("None (0)", "Low (1-10)", "Medium (11-100)", "High (>100)")
      ),
      timepoint = factor(timepoint, levels = c("Baseline", "Endpoint"))
    )

  ggplot(
    alluvial_data,
    aes(
      x = timepoint,
      stratum = abundance,
      alluvium = subject,
      fill = abundance,
      label = abundance
    )
  ) +
    geom_flow(aes(fill = arm), alpha = 0.5) + # Color by arm
    geom_stratum(alpha = 0.8) +
    geom_text(
      stat = "stratum",
      aes(label = after_stat(n)),
      size = 3
    ) +
    facet_wrap(~arm) + # Separate by arm
    theme_minimal() +
    labs(
      title = paste("Abundance Changes for", outcome_id),
      subtitle = "Separated by Treatment Arm",
      x = "Timepoint",
      y = "Number of Subjects",
      fill = "Abundance Level"
    )
}

```


```{r}
#| eval: true
#| label: fig-alluvial

joined_data %>%
  filter(outcome_id == "G012221805") %>%
  select(subject, timepoint, outcome_id, value) %>%
  mutate(
    abundance = cut(
      value,
      breaks = c(-Inf, 1, 10, 100, Inf),
      labels = c("None (0)", "Low (1-10)", "Medium (11-100)", "High (>100)")
    ),
    timepoint = factor(timepoint, levels = c("Baseline", "Endpoint"))
  )

alluvial_data <- joined_data %>%
  filter(outcome_id == "G012221805") %>%
  select(subject, timepoint, outcome_id, value) %>%
  mutate (abundance = case_when(
      value < 0 ~ "Decrease",
      value > 0 ~ "Low",
      value == 0 ~ "None",
      TRUE ~ NA
    )

    )) %>%
  pivot_wider(names_from = timepoint, values_from = value) %>%
  mutate(EvsB = Endpoint - Baseline) %>%
  mutate(
    difference = case_when(
      EvsB < 0 ~ "Decrease",
      EvsB > 0 ~ "Increase",
      EvsB == 0 ~ "None",
      TRUE ~ NA
    )
  ) %>%
  pivot_longer(cols = c(Endpoint, Baseline), names_to = "timepoint") %>%
  mutate(
    difference = ifelse(
      timepoint == "Baseline",
      "Baseline",
      difference
    )
  )

alluvial_data <- to_alluvia_form(
  alluvial_data,
  key = "timepoint",
  value = "difference",
  id = "subject"
)

alluvial_data %>%
  ggplot(
    aes(
      axis1 = Baseline,
      axis2 = Endpoint
    )
  ) +
  geom_alluvium(
    aes(fill = as.factor(Endpoint)),
    width = 2 / 5,
    discern = TRUE
  ) +
  geom_stratum(width = 2 / 5, discern = TRUE) +
  geom_text(
    stat = "stratum",
    discern = TRUE,
    aes(label = after_stat(stratum))
  ) +
  theme(legend.position = "none")


geom_flow(aes(fill = difference), alpha = 0.5) + # Color by arm
  geom_stratum(alpha = 0.8) +
  geom_text(
    stat = "stratum",
    aes(label = after_stat(n)),
    size = 3
  ) +
  theme_minimal() +
  labs(
    title = paste("Abundance Changes for", outcome_id),
    subtitle = "Separated by Treatment Arm",
    x = "Timepoint",
    y = "Number of Subjects",
    fill = "Abundance Level"
  )

```


```{r}
#| eval: true
#| label: ggplot

joined_data %>%
  dplyr::filter(outcome_id %in% ziSig$outcome_id) %>%
  filter(
    outcome_id == "G000712635" |
      outcome_id == "G000772105" |
      outcome_id == "G000771955" |
      outcome_id ==
        "G000196255
  "
  ) %>%
  group_by(outcome_id) %>%
  ggplot(aes(
    x = time,
    y = as.numeric(value),
    colour = arm
  )) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = F,
    color = "black",
    formula = y ~ x,
    linetype = 'dotted',
    linewidth = 3
  ) + #overall regression arm
  xlab("Time Since Feed Start (days)") +
  ylab("Abundance") +
  theme(legend.position = "none") +
  labs(color = "Intervention Arm") +
  facet_grid(~outcome_id, scales = "free")

joined_data %>%
  dplyr::filter(outcome_id %in% ziSig$outcome_id) %>%
  filter(
    outcome_id == "G900155635" |
      outcome_id == "G002236805" |
      outcome_id == "G000814825" |
      outcome_id == "G000786655"
  ) %>%
  group_by(outcome_id) %>%
  ggplot(aes(
    x = time,
    y = as.numeric(value),
    colour = arm
  )) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = F,
    color = "black",
    formula = y ~ x,
    linetype = 'dotted',
    linewidth = 3
  ) + #overall regression arm
  xlab("Time Since Feed Start (days)") +
  ylab("Abundance") +
  theme(legend.position = "none") +
  labs(color = "Intervention Arm") +
  facet_grid(~outcome_id, scales = "free")

```

#### Log liklihood tests for nested models

> [!cite]
> The best way to judge whether the ZIP is acceptable is to compare it with a ZINB as these models are nested [@Zuur2009].

```{r}
#| eval: true
#| label: model-anova

lapply(names(nb1_models), function(x) {
  anova(
    nb1_models[[x]],
    # null_models$all_models[[x]],
    # null_rand_models$all_models[[x]]
    # noZi_models$all_models[[x]],
    noZi_rand_models$all_models[[x]]
  )
})
```

```{r}
#| eval: true
#| label: poi-anova

# Print summary
poi_anova <- nested_anova$comparisons %>%
  filter(
    outcome_id %in% names(poi_models),
    model1 == "poi_results" |
      model2 == "poi_results"
  )
```

Most of the data is highly dispersed so a NB model would be more appropriate than Poisson


### extra

```{r}
#| eval: true
#| label: fig-emms-arm
#| fig-cap: Changes in the estimated marginal means of alpha diversity statistics within each intervention arm

# Create an empty list to store plots
plot_list <- list()

# Loop through each item in the list
for (i in names(emmeans)) {
  # Create a plot for each item (replace with your actual plot creation code)
  plot <- emmip(emmeans[[i]], arm ~ time, CIs = T) +
    ggtitle(paste(str_extract(i, "^[^_]*"))) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(size = 10, face = "bold")
    )

  # Add the plot to the list
  plot_list[[i]] <- plot
}

#Arrange and display the plots using grid.arrange
ggpubr::annotate_figure(ggpubr::ggarrange(
  plotlist = plot_list,
  ncol = 2,
  nrow = 2,
  common.legend = T
))
```

1. Did diversity change over the 9-month period?
   
  - Compared average diversity at baseline vs endpoint
  - Combined both treatment groups

::: question

Does 9 months daily feeding of PM-based foods result in differences in the gut microbiome in terms of mean $\alpha$ diversity metrics i.e. does average at baseline = average at endpoint?

:::

::: h0

$$
\begin{array}{cc}
\frac{Arm0B + Arm1B}{2} = \frac{Arm0E + Arm1E}{2} \\
\Rightarrow \frac{Arm0B}{2} + \frac{Arm1B}{2} - \frac{Arm0E}{2} - \frac{Arm1E}{2} = 0 \\
\Rightarrow \frac{Arm0E}{2} + \frac{Arm1E}{2} - \frac{Arm0B}{2} - \frac{Arm1B}{2} = 0
\end{array}
$$

:::

The coeficients for the contrast are as follows (**depends on the order of the emmeans output**): 

- time 0 = baseline
- time 1 = endpoint
- Coefficients = -0.5, -0.5, 0.5, 0.5

```{r}
#| eval: true
#| label: tbl-emm-arm
#| results: asis
#| tbl-cap: Changes in alpha diveristy estimated marginal means over time by intervention arm.

#store results for graph
emmeans_BvsE <- lapply(emmeans, function(x) {
  contrast(x, method = list(c(-0.5, -0.5, 0.5, 0.5)), adjust = "bonferroni")
})

# Convert each emmeans result to a data frame and add a label column
labeled_tables <- Map(
  function(x, name) {
    df <- as.data.frame(x)
    df$Comparison <- name
    return(df)
  },
  emmeans_BvsE,
  names(emmeans_BvsE)
)

# Combine all tables into one
combined_df <- do.call(rbind, labeled_tables)

# Reorder columns to put Comparison first
combined_df <- combined_df[, c(
  "Comparison",
  setdiff(names(combined_df), "Comparison")
)]

# Print as one table
knitr::kable(combined_df, digits = 4)
```

```{r}
#| eval: true
#| label: tbl-emm-arm-eff
#| tbl-cap: The effect size of temporal changes in alpha diveristy estimated marginal means by intervention arm.

lapply(
  lapply(emmeans, function(x) {
    contrast(x, method = "eff", adjust = "bonferroni")
  }),
  knitr::kable
)
```

2. Did treatments have different effects?
  
  - Examined if changes in diversity differed between CPM and FeZnPM
  - Considered interaction between time and treatment

::: question

Is there a difference between the mean $\alpha$ diversity of individuals fed either PM or FeZnPM-based foods across 9 months of feeding? i.e. Is the change in mean $\alpha$ diversity different between intervention arms?

:::

::: h0

$$
\begin{array}{cc}
(Arm0B - Arm0E) = (Arm1B - Arm1E) \\
\Rightarrow Arm0B - Arm0E - Arm1B + Arm1E = 0
\end{array}
$$

:::

The coeficients for the contrast are as follows (**depends on the order of the emmeans output**): 

- time 0 = baseline
- time 1 = endpoint
- Coefficients = -1, 1, 1, -1

```{r}
#| eval:  true
#| label: tbl-emm-arm-contrasts
#| tbl-cap: Contrast coeficients
#| results: asis

emmeans_CPMvsFeZnPM <- lapply(emmeans, function(x) {
  contrast(x, method = list(c(-1, 1, 1, -1)), adjust = "bonferroni")
})

lapply(
  lapply(emmeans, function(x) {
    contrast(x, method = list(c(-1, 1, 1, -1)), adjust = "bonferroni")
  }),
  knitr::kable
)
```

**Estimate**: Represents the estimated difference between the two levels being compared. It suggests that the first level has a lower value (-50.59 units) than the average of the other three levels combined.

The impact of age at baseline was examined by:

- Creating estimates at different age percentiles (0, 25, 50, 75, 100%)
- Checking if treatment effects varied with starting age
- Visualizing age-related patterns in diversity changes

::: question

Does FeZnPM change/improve the gut microbiome in some children more than others based on age at baseline?

:::

```{r}
#| eval: false
#| label: effect of baseline age in model estimation
age_emmeans <- lapply(alpha_lmms_transformed, function(x) {
  emmeans(
    x,
    ~ arm + time + age_B,
    at = list(age_B = c(0, 0.25, 0.5, 0.75, 1), time = c(0, 1)),
    type = "response",
    adjust = "sidak"
  )
})
```

```{r}
#| eval:  true
#| label: fig-emms-age
#| fig-cap: Changes in the estimated marginal means of alpha diversity statistics within each intervention arm adjusted for age at baseline.

plot_list <- list()

# Loop through each item in the list
for (i in names(age_emmeans)) {
  # Create a plot for each item (replace with your actual plot creation code)
  plot <- emmip(age_emmeans[[i]], arm ~ age_B, CIs = T) +
    ggtitle(paste(str_extract(i, "^[^_]*"))) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(size = 10, face = "bold")
    )

  # Add the plot to the list
  plot_list[[i]] <- plot
}

#Arrange and display the plots
#basically the same as the ggplots below except not to the correct scale as the orginal data. Covariates not backtransformed
ggpubr::annotate_figure(
  ggpubr::ggarrange(
    plotlist = plot_list,
    ncol = 2,
    nrow = 2,
    common.legend = T
  ),
  top = "Change in diversity statistics"
)
```

```{r}
#| eval:  true
#| label: tbl-age-contrasts
#| tbl-cap: Estimated marginal means adjusting for age at baseline.
#| results: asis

lapply(
  lapply(age_emmeans, function(x) {
    pairs(x, adjust = "bonferroni")
  }),
  knitr::kable
)
```

The resulting tables are in units on a 0-1 scale for the number of days for the time and age covariates. The response column is the already back-transformed diversity statistic (done by emmeans). The following can be used to backtransform them, where `x` is the desired diversity metric column:

```{r}
#| eval: false
backtransform_scaled_values(
  scaled_value = x,
  original_data = as.numeric(bac_base_end_meta$timesincefeedstart)
)
backtransform_scaled_values(scaled_value = x, original_data = age_B_unscaled)
```

An ANOVA/ANCOVA can be perfomed to determine if the grouping variable, in this case the intervention arm, is an important category for determining the outcome variable.

::: h0

$H_0$: the mean outcome is the same between groups i.e.

$$
\mu_x1 = \mu_x2 = \mu_x3 = \mu_x4
$$

:::

::: h1

$H_1$: the mean outcome is different between groups i.e.

$$
\mu_x1 \neq \mu_x2 \neq \mu_x3 \neq \mu_x4
$$

:::

Here only the intercept and time variable are significant. Therefore, we must reject $H_1$ and accept $H_0$. The mean diversity statistic response variable is equal between intervention arm 0 and 1

```{r}
#| eval: false
#| label: tbl-anova
#| tbl-cap: ANOVA
#| results: asis

lapply(lapply(alpha_lmms_transformed, anova), knitr::kable)
```

###### Graphing alpha diversity statistic LMMs

```{r}
#| eval: false
#| label: fig-emms-alpha
#| fig-cap: Alpha diversity EMMs plots
#| fig-width: 18
#| fig-height: 12

shannon_p <- bac_base_end_meta %>%
  ggplot(aes(
    x = backtransform_scaled_values(
      scaled_value = time,
      original_data = as.numeric(bac_base_end_meta$timesincefeedstart)
    ),
    y = as.numeric(diversity_shannon),
    colour = arm
  )) +
  geom_point() +
  geom_line(
    aes(
      y = sqrt(predict(alpha_lmms_transformed$shannon_lmm_transformed)),
      group = displayid,
      colour = arm
    ),
    alpha = 0.5
  ) + # sqrt used to backtransform
  geom_smooth(
    method = "lm",
    se = F,
    color = "black",
    formula = y ~ x,
    linetype = 'dotted',
    linewidth = 3
  ) + #overall regression arm
  annotate(
    "text",
    x = -Inf,
    y = Inf,
    hjust = -1,
    vjust = 2,
    label = bquote(atop(
      "Baseline vs Endpoint:" ~
        p ==
          .(ifelse(
            round(
              summary(emmeans_BvsE$shannon_lmm_transformed[1])[, "p.value"],
              3
            ) <
              0.0001,
            "<0.0001",
            round(
              summary(emmeans_BvsE$shannon_lmm_transformed[1])[, "p.value"],
              3
            )
          )),
      "CPM vs FeZnPM:" ~
        p ==
          .(ifelse(
            round(
              summary(emmeans_CPMvsFeZnPM$shannon_lmm_transformed[1])[,
                "p.value"
              ],
              3
            ) <
              0.0001,
            "<0.0001",
            round(
              summary(emmeans_CPMvsFeZnPM$shannon_lmm_transformed[1])[,
                "p.value"
              ],
              3
            )
          ))
    ))
  ) +
  xlab("Time Since Feed Start (days)") +
  ylab("Shannon Diversity Index") +
  theme(legend.position = "none") +
  scale_color_manual(values = armColours) +
  labs(color = "Intervention Arm")

evenness_p <- bac_base_end_meta %>%
  ggplot(aes(
    x = backtransform_scaled_values(
      scaled_value = time,
      original_data = as.numeric(bac_base_end_meta$timesincefeedstart)
    ),
    y = as.numeric(evenness_pielou),
    colour = arm
  )) +
  geom_point() +
  geom_line(
    aes(
      y = sqrt(predict(alpha_lmms_transformed$evenness_lmm_transformed)),
      group = displayid,
      colour = arm
    ),
    alpha = 0.5
  ) +
  geom_smooth(
    method = "lm",
    se = F,
    color = "black",
    formula = y ~ x,
    linetype = 'dotted',
    linewidth = 3
  ) +
  annotate(
    "text",
    x = -Inf,
    y = Inf,
    hjust = -1,
    vjust = 2,
    label = bquote(atop(
      "Baseline vs Endpoint:" ~
        p ==
          .(ifelse(
            round(
              summary(emmeans_BvsE$evenness_lmm_transformed[1])[, "p.value"],
              3
            ) <
              0.0001,
            "<0.0001",
            round(
              summary(emmeans_BvsE$evenness_lmm_transformed[1])[, "p.value"],
              3
            )
          )),
      "CPM vs FeZnPM:" ~
        p ==
          .(ifelse(
            round(
              summary(emmeans_CPMvsFeZnPM$evenness_lmm_transformed[1])[,
                "p.value"
              ],
              3
            ) <
              0.0001,
            "<0.0001",
            round(
              summary(emmeans_CPMvsFeZnPM$evenness_lmm_transformed[1])[,
                "p.value"
              ],
              3
            )
          ))
    ))
  ) +
  xlab("Time Since Feed Start (days)") +
  ylab("Pielou Evenness Index") +
  theme(legend.position = "none") +
  scale_color_manual(values = armColours) +
  labs(color = "Intervention Arm")

chao_p <- bac_base_end_meta %>%
  ggplot(aes(
    x = backtransform_scaled_values(
      scaled_value = time,
      original_data = as.numeric(bac_base_end_meta$timesincefeedstart)
    ),
    y = as.numeric(chao1),
    colour = arm
  )) +
  geom_point() +
  geom_line(
    aes(
      y = exp(predict(alpha_lmms_transformed$chao_lmm_transformed)),
      group = displayid,
      colour = arm
    ),
    alpha = 0.5
  ) +
  geom_smooth(
    method = "lm",
    se = F,
    color = "black",
    formula = y ~ x,
    linetype = 'dotted',
    linewidth = 3
  ) +
  annotate(
    "text",
    x = -Inf,
    y = Inf,
    hjust = -1,
    vjust = 2,
    label = bquote(atop(
      "Baseline vs Endpoint:" ~
        p ==
          .(ifelse(
            round(
              summary(emmeans_BvsE$chao_lmm_transformed[1])[, "p.value"],
              3
            ) <
              0.0001,
            "<0.0001",
            round(summary(emmeans_BvsE$chao_lmm_transformed[1])[, "p.value"], 3)
          )),
      "CPM vs FeZnPM:" ~
        p ==
          .(ifelse(
            round(
              summary(emmeans_CPMvsFeZnPM$chao_lmm_transformed[1])[, "p.value"],
              3
            ) <
              0.0001,
            "<0.0001",
            round(
              summary(emmeans_CPMvsFeZnPM$chao_lmm_transformed[1])[, "p.value"],
              3
            )
          ))
    ))
  ) +
  xlab("Time Since Feed Start (days)") +
  ylab("Chao Abundance Index") +
  theme(legend.position = "none") +
  scale_color_manual(values = armColours) +
  labs(color = "Intervention Arm")

pd_p <- bac_base_end_meta %>%
  ggplot(aes(
    x = backtransform_scaled_values(
      scaled_value = time,
      original_data = as.numeric(bac_base_end_meta$timesincefeedstart)
    ),
    y = as.numeric(pd),
    colour = arm
  )) +
  geom_point() +
  geom_line(
    aes(
      y = exp(predict(alpha_lmms_transformed$pd_lmm_transformed)),
      group = displayid,
      colour = arm
    ),
    alpha = 0.5
  ) +
  geom_smooth(
    method = "lm",
    se = F,
    color = "black",
    formula = y ~ x,
    linetype = 'dotted',
    linewidth = 3
  ) +
  annotate(
    "text",
    x = -Inf,
    y = Inf,
    hjust = -1,
    vjust = 2,
    label = bquote(atop(
      "Baseline vs Endpoint:" ~
        p ==
          .(ifelse(
            round(summary(emmeans_BvsE$pd_lmm_transformed[1])[, "p.value"], 3) <
              0.0001,
            "<0.0001",
            round(summary(emmeans_BvsE$pd_lmm_transformed[1])[, "p.value"], 3)
          )),
      "CPM vs FeZnPM:" ~
        p ==
          .(ifelse(
            round(
              summary(emmeans_CPMvsFeZnPM$pd_lmm_transformed[1])[, "p.value"],
              3
            ) <
              0.0001,
            "<0.0001",
            round(
              summary(emmeans_CPMvsFeZnPM$pd_lmm_transformed[1])[, "p.value"],
              3
            )
          ))
    ))
  ) +
  xlab("Time Since Feed Start (days)") +
  ylab("Faith's Phylogenetic diversity Index") +
  theme(legend.position = "none") +
  scale_color_manual(values = armColours) +
  labs(color = "Intervention Arm")

ggpubr::ggarrange(
  chao_p,
  shannon_p,
  evenness_p,
  pd_p,
  ncol = 2,
  nrow = 2,
  common.legend = T,
  legend = "bottom",
  labels = "AUTO"
)
```

## Appendix

```{r}
#| eval: false
#| label: unfit2

unfit2 <- tmp$adj_results %>%
  filter(if_any(everything(), is.na))

unfit2 <- unfit2[, 1]

outcome <- outcome %>%
  select(all_of(unfit2))

ncol(outcome) == length(unfit2)

tmp2 <- genZI(outcome, predictors, family = gaussian)

```

```{r}
#| eval: false
#| label: unfit3

unfit3 <- tmp$adj_results %>%
  filter(if_any(everything(), is.na))

unfit3 <- unfit3[, 1]

outcome <- outcome %>%
  select(all_of(unfit3))

ncol(outcome) == length(unfit3)

tmp3 <- genZI(outcome, predictors, family = lognormal)

lapply(tmp3$all_models, diagnose)
```

A total of `r length(unfit2)` models did not converge appropriately and were refit with a Log normal distribution. A single OGU (G001054195) resulted in large coefficients in the zero-inflated model component. A better fit could not be found.

```{r}
#| eval: false
#| label: auto

results <- run_zibr_pipeline(
  phyloseq_obj = bac_base_end,
  output_dir = file.path(RESULTS, "EvsB-OGU-GEE")
)

# Create summary plots
plots <- create_summary_plots(results)
print(plots$p_value_dist)
print(plots$volcano_plot)

# 5. Examine a specific significant OGU
# Replace "OGU_ID" with an actual OGU ID from your significant results
ogu_details <- examine_ogu(results, "OGU_ID")

# 6. Access detailed results for further analysis
significant_ogus <- results$significant_ogus
model_summaries <- results$model_results$convergence_summary

# 7. Export results to Excel for easier viewing
if (require(openxlsx)) {
  write.xlsx(
    list(
      Convergence_Summary = results$model_results$convergence_summary,
      Significant_OGUs = results$significant_ogus
    ),
    file = file.path(output_dir, "ZIBR_Results.xlsx")
  )
}

```


## ROC Curves

```{r roc}
#| eval: true
#| label: roc-curves
#| fig-cap: NA

roc <- lapply(
  list(poi_models, nb1_models, nb2_models, lognorm_models),
  function(x) {
    tmp <- list()
    tmp$all_models <- x
    create_zi_roc(tmp)
  }
)
```
