## Comparison of Results

```{r}
#| echo: false
#| eval: true
#| label: setup

# This initial code block loads custom utility functions and environment variables
# These are stored in separate files to keep the main analysis clean and organized
source("utils/r-utils.r")
source("_environment.required")
```

This code aims to compare the results of the original analysis on data that was filtered with a single human reference genome, to the new analysis pipeline that decontaminated raw reads using a human pangenome approach [@Knight2024].

### QC and WoL2-OGU Classification

```{r}
#| eval: true
#| label: compare alignment stats
library(dataCompareR)
old <- read.csv("analysis/100-Appendix/previous-analysis-data/metagenome_results_master_summary.csv", header = T)

new <- read.csv("results/metagenome_results_summary.csv", header = TRUE)

compDetails <- rCompare(old, new, keys = "sample", trimChars = TRUE, roundDigits = 2)
(compDetailsSum <- summary(compDetails, mismatchCount = 5))

saveReport(compDetails, reportName = "alignment-ogu-stat-comparison-report.rmd", reportLocation = "results/analysis-comparison", HTMLreport = TRUE, printAll = TRUE, showInViewer = FALSE)

```

```{r}
#| eval: true
#| label: statistical testing
#| tbl-cap: Comparison of QC and Alignment statistics
#| results: asis
library(gtsummary)

old <- old %>% mutate_at(vars(!c(sample)), as.numeric) %>% mutate(group = "old")
new <- new %>% mutate_at(vars(!c(sample)), as.numeric) %>% mutate(group = "new")

df <- dplyr::bind_rows(old,new)

(p <- df %>% select(!sample) %>%
	  tbl_summary(
			by = group,
			missing = "no", 
			statistic = list(
				all_continuous() ~ "{median}  \n({IQR}; {min}-{max})",
				all_categorical() ~ "{n} / {N}  \n({p}%)")) %>%
		add_stat(fns = all_continuous() ~ vda) %>%
		modify_header(list(estimate ~ "**Effect Size Estimate**", magnitude ~ "**Magnitude**")) %>%
		add_p(pvalue_fun = function(x) pvalue_format(x)) %>%
	  add_q(pvalue_fun = function(x) pvalue_format(x)) %>%
		bold_p(q = TRUE) %>%
		as_gt() %>% 
			fmt_markdown(columns=where(~is.character(.x))) %>%  
			fmt_number(columns = where(~is.numeric(.x)<0.05), pattern = "<0.05") %>%
			fmt_number(columns = where(~is.numeric(.x)<0.0001), pattern = "<0.0001") %>% 
			fmt_number(columns = where(~is.numeric(.x)<0.001), pattern = "<0.001")%>% 
			fmt_number(columns = where(~is.numeric(.x)<0.01), pattern = "<0.01")) 


htmltools::save_html(gt::as_raw_html(p), "analysis/100-Appendix/previous-analysis-data/qc-woltka-comparison.tbl")

```

### Baseline Diveristy Statistics

#### Alpha Diversity Statistics

```{r}
#| eval: true
#| label: tbl-alpha-div-base
#| tbl-cap: Comparison of baseline alpha diversity statistics
#| results: asis

old <- readRDS("analysis/100-Appendix/previous-analysis-data/baseline-metadata.rds") %>% 
				select(sample_names, trt, observed,  chao1, diversity_shannon, evenness_pielou, Actinobacteriota,Bacteroidota,
     Campylobacterota,Firmicutes,Other,Proteobacteria,Unclassified,     Verrucomicrobiota,Unknown,Escherichia, Enterococcus, Prevotella,Anaerococcus,     Klebsiella,Staphylococcus,Corynebacterium,Salmonella,Peptoniphilus,     Enterobacter) %>% 
					mutate(group = "old") %>% 
					rename(Bacillota = Firmicutes,
					Pseudomonadota = Proteobacteria)

new <- readRDS("data/metadata/baseline-metadata.rds")%>% 
				select(sample_names, trt, observed,  chao1, diversity_shannon, evenness_pielou, Actinobacteriota,Bacteroidota,
     Campylobacterota,Bacillota,Other,Pseudomonadota,Unclassified,     Verrucomicrobiota,Unknown,Escherichia, Enterococcus, Prevotella,Anaerococcus,     Klebsiella,Staphylococcus,Corynebacterium,Salmonella,Peptoniphilus,     Enterobacter) %>%
					mutate(group = "new")

compDetails <- rCompare(old, new, keys = "sample_names", trimChars = TRUE, roundDigits = 2)
(compDetailsSum <- summary(compDetails, mismatchCount = 5))

saveReport(compDetails, reportName = "alpha-diversity-comparison-report.rmd", reportLocation = "results/analysis-comparison", HTMLreport = TRUE, printAll = TRUE, showInViewer = FALSE)

df <- dplyr::bind_rows(old,new)

(p <- df %>% select (!c(Unknown, sample_names)) %>%
	  tbl_summary(
			by = group,
			digits = everything() ~ 2,
			missing = "no", 
			statistic = list(
				all_continuous() ~ "{median}  \n({IQR}; {min}-{max})",
				all_categorical() ~ "{n} / {N}  \n({p}%)")) %>%
		add_stat(fns = all_continuous() ~ vda) %>%
		modify_header(list(estimate ~ "**Effect Size Estimate**", magnitude ~ "**Magnitude**")) %>%
		add_p(pvalue_fun = function(x) pvalue_format(x)) %>%
	  add_q(pvalue_fun = function(x) pvalue_format(x)) %>%
		bold_p(q = TRUE) %>%
		as_gt() %>% 
			fmt_markdown(columns=where(~is.character(.x))) %>%  
			fmt_number(columns = where(~is.numeric(.x)<0.0001), pattern = "<0.0001") %>% 
			fmt_number(columns = where(~is.numeric(.x)<0.001), pattern = "<0.001")%>% 
			fmt_number(columns = where(~is.numeric(.x)<0.01), pattern = "<0.01")%>% 
			fmt_number(columns = where(~is.numeric(.x)<0.05), pattern = "<0.05"))

htmltools::save_html(gt::as_raw_html(p), "analysis/100-Appendix/previous-analysis-data/alpha-div-comparison.tbl")
```

```{r}
#| eval: true
#| tbl-cap: Comparison of alpha diversity statistics between intervention arms
#| results: asis

(p <- df %>% select (!c(Unknown, sample_names)) %>%
  tbl_strata(strata = group, .tbl_fun = ~ .x %>% 
	  tbl_summary(
			by = trt,
			missing = "no", 
			statistic = list(
				all_continuous() ~ "{median}  \n({IQR}; {min}-{max})",
				all_categorical() ~ "{n} / {N}  \n({p}%)")) %>%
		add_stat(fns = all_continuous() ~ vda) %>%
		modify_header(list(estimate ~ "**Effect Size Estimate**", magnitude ~ "**Magnitude**")) %>%
		add_p(pvalue_fun = function(x) pvalue_format(x)) %>%
	  add_q(pvalue_fun = function(x) pvalue_format(x)) %>%
		bold_p(q = TRUE),
.header = "**{strata}**, N = {n}")%>%
		as_gt() %>% 
			fmt_markdown(columns=where(~is.character(.x))) %>%  
			fmt_number(columns = where(~is.numeric(.x)<0.0001), pattern = "<0.0001") %>% 
			fmt_number(columns = where(~is.numeric(.x)<0.001), pattern = "<0.001")%>% 
			fmt_number(columns = where(~is.numeric(.x)<0.01), pattern = "<0.01")%>% 
			fmt_number(columns = where(~is.numeric(.x)<0.05), pattern = "<0.05"))


htmltools::save_html(gt::as_raw_html(p), "analysis/100-Appendix/previous-analysis-data/arm-alpha-div-comparison2.tbl")

```

#### Beta Diversity Statistics

```{r}
#| eval: false
#| label: load beta diversity data
#| tbl-cap: Comparison of beta diversity statistics
#| results: asis

old <- readRDS("analysis/100-Appendix/previous-analysis-data/baseline-beta-div.rds") %>%
					mutate(group = "old")

new <- readRDS("results/baseline-diversity-analysis/baseline-beta-div.rds") %>%
					mutate(group = "new")

df <- dplyr::bind_rows(old,new)

(p <- df %>% select (!timepoint) %>%
tbl_strata(strata = trt, .tbl_fun = ~ .x %>%
	tbl_summary(
	  by = group,
			missing = "no", 
			statistic = list(
				all_continuous() ~ "{median}  \n({IQR}; {min}-{max})",
				all_categorical() ~ "{n} / {N}  \n({p}%)")) %>%
		add_stat(fns = all_continuous() ~ vda) %>%
		modify_header(list(estimate ~ "**Effect Size Estimate**", magnitude ~ "**Magnitude**")) %>%
		add_p(pvalue_fun = function(x) pvalue_format(x)) %>%
	  add_q(pvalue_fun = function(x) pvalue_format(x)) %>%
		bold_p(q = TRUE),
.header = "**{strata}**, N = {n}") %>%
		as_gt() %>% 
			fmt_markdown(columns=where(~is.character(.x))) %>%  
			fmt_number(columns = where(~is.numeric(.x)<0.0001), pattern = "<0.0001") %>% 
			fmt_number(columns = where(~is.numeric(.x)<0.001), pattern = "<0.001")%>% 
			fmt_number(columns = where(~is.numeric(.x)<0.01), pattern = "<0.01")%>% 
			fmt_number(columns = where(~is.numeric(.x)<0.05), pattern = "<0.05"))

htmltools::save_html(gt::as_raw_html(p), "analysis/100-Appendix/previous-analysis-data/beta-div-comparison.tbl")

```

