### B12 Status: Baseline Bacterial Relative Sums

```{r}
#| echo: false
#| eval: true
#| label: setup
#| include: false

# This initial code block loads custom utility functions and environment variables

# These are stored in separate files to keep the main analysis clean and organized
source(normalizePath("utils/r-utils.r"))
source(normalizePath("_environment.required"))
load(normalizePath("data/seqdata.RData"))

metadata <- read_delim(normalizePath(METADATA))
bac_base_end_meta <- readRDS(normalizePath("data/metadata/base-end-metadata.rds"))
baseline_metadata <- readRDS(normalizePath("data/metadata/baseline-metadata.rds"))

suppressPackageStartupMessages(library(emmeans))
suppressPackageStartupMessages(library(nlme))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ampvis2))

dir.create(file.path(RESULTS, "B12-status-subgroup-analysis"), showWarnings = FALSE, recursive = TRUE)
```

#### Data preparation

Serum vitamin B12 concentration (`b12_pmol`) is in pg/mL and is converted below to pmol/L using the 0.738 conversion factor.

| Characteristic                       | Median (IQR) or n (%) |
| :----------------------------------- | :-------------------- |
| Serum vitamin B12 (pmol/L)           | 5.45 (5.08, 5.93)     |
|   Tertile 1 (n=42)                     | 4.99 (4.79, 5.08)     |
|   Tertile 2 (n=43)                     | 5.45 (5.34, 5.62)     |
|   Tertile 3 (n=42)                     | 6.13 (5.93, 6.34)     |
| Binary: Deficiency (<148 pmol/L)     | 21 (16.5%)            |
| 3-level: Insufficiency (≥148 & <221) | 34 (26.8%)            |
| Binary: Insufficiency (<221 pmol/L)  | 55 (43.3%)            |
| Sufficient (≥221 pmol/L)             | 72 (56.7%)            |


```{r}
#| eval: true
#| label: subset baseline
b12_base_meta <- baseline_metadata %>%
  mutate(b12_pmol = vitb12*0.738) %>%
  mutate(
    b12_tertiles = tertiles(b12_pmol),
    b12_2Lev = case_when(
      b12_pmol < 148 ~ "deficiency",
      b12_pmol >= 148 ~ "sufficiency",
      TRUE ~ "Unknown"
    ),
    b12_3Lev = case_when(
      b12_pmol < 148 ~ "deficiency",
      b12_pmol >= 148 & b12_pmol < 221 ~ "insufficiency",
      b12_pmol >= 221 ~ "sufficiency",
      TRUE ~ "Unknown"
    ),
    sexf = case_when(
      sexf == 1 ~ "Female",
      sexf == 0 ~ "Male"
    )
  ) %>%
  dplyr::filter(!is.na(b12_pmol)) 

rownames(b12_base_meta) <- b12_base_meta$sample_names

saveRDS(b12_base_meta,file.path(RESULTS, "B12-status-subgroup-analysis/b12-baseline-subset.rds"))
```


Relative abundances are calculated to show the proportion of each bacterial group in a sample.

Create an ampvis object from phyloseq object

```{r}
#| eval: true
#| label: create ampvis object

base_b12 <- phyloseq::subset_samples(seqdata_bac, !is.na(vitb12))
base_b12 <- phyloseq::subset_samples(base_b12, timepoint=="1Baseline")

sample_data(base_b12) <- b12_base_meta

#slight modification to collapse Bacillota subgroups into one. As of Apr 2025 I cannot find any valid publication of this taxonomy. It appears to be based purely on the analysis of MAGs at GTDB
tax_table(base_b12) <- tax_df(base_b12) %>% 
                            mutate(across(where(is.character), str_trim, "both")) %>%
                            mutate(Phylum = str_remove(Phylum, "_[a-zA-Z]\\b")) %>%
                            mutate(Genus = str_remove(Genus, "_[a-zA-Z]\\b")) %>%
                            mutate(Species = str_remove(Species, "_[a-zA-Z]\\b")) %>%
                            mutate(Phylum = case_when(
                              Phylum == "" ~ "Unknown",
                              str_detect(Phylum, "\\d") ~ "Unclassified",
                              TRUE ~ Phylum)) %>%
                            mutate(Genus = case_when(
                              str_detect(Genus, "\\d") ~ "Unclassified",
                              Genus == "" ~ "Unknown",
                              TRUE ~ Genus)) %>%
                            as.matrix()
saveRDS(base_b12, file.path(RESULTS, "B12-status-subgroup-analysis/b12-baseline-seqdata.rds"))

d <- amp_load(otutable = t(otu_table(base_b12)), b12_base_meta, taxonomy = tax_table(base_b12))
```

#### Basic Characteristics of Subgroup

```{r}
#| eval: true
#| label: tbl-base-char
#| tbl-cap: Baseline characterisitcs of subjects with vitamin B12 measurements available.

b12_base_meta %>% select(Sex = sexf, `Intervention Arm`= arm, `B12 Status` = b12_status, `B12 Concentration (pg/mL)` = vitb12) %>% tbl_summary() %>% as_kable()

```

```{r}
#| eval: true
#| label: tbl-base-char
#| tbl-cap: Baseline characterisitcs by sex for subjects with vitamin B12 measurements available.

b12_base_meta %>% select(Sex = sexf, `Intervention Arm`= arm, `B12 Status` = b12_status, `B12 Concentration (pg/mL)` = vitb12) %>% tbl_summary(by=Sex) %>% as_kable()

```

#### Baseline Phyla Abundance

```{r}
#| eval: true
#| label: fig-base-b12-phyla-heatmap
#| fig-cap: Relative abundance of bacterial phyla at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL). Average group relative abundances are shown on the graph.

(heatmap <- amp_heatmap(d,
            group_by = "b12_status",
            tax_aggregate = "Phylum",
            tax_show = 20,
            plot_values = T,
            normalise = T, 
            plot_values_size = 8,
            round = 3,
            tax_empty="OTU",
            order_x_by = "cluster",
            order_y_by = "cluster") +
              theme(axis.text.x = element_text(angle = 45, size = 14, vjust = 1), axis.text.y = element_text(size = 14),  legend.position = "right", legend.title = element_text(color = "black", size = 10, face = "bold")))
```

```{r}
#| eval: true
#| label: tbl-b12-phyla-heatmap
#| tbl-cap: Relative abundance of bacterial phyla at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
heatmap$data %>% 
  group_by(Display, .Group) %>% 
  summarise(mean = round(mean(Sum), 4)) %>% 
  arrange(desc(mean)) %>%
  mutate(mean =  case_when(
    mean >= 0.01 ~ round(mean, 2), 
    TRUE ~ mean
  )) %>%
  mutate(mean = paste0(mean, "%")) %>%
  pivot_wider(names_from = .Group, values_from = mean) %>%
  kable()
```

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: Summary statistics of bacterial relative abundances at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
df <- heatmap$data %>% 
  dplyr::select(phyla = Display, Sample , relAbund = Sum) %>% 
  mutate(timepoint = b12_base_meta$timepoint[match(Sample,b12_base_meta$sample_names)]) %>%
  mutate(b12_status = b12_base_meta$b12_status[match(Sample,b12_base_meta$sample_names)])
  
df %>%
  dplyr::select(phyla, b12_status, relAbund) %>%
  group_by(phyla, b12_status) %>%
  summarise(mean = mean(relAbund), median = median(relAbund), max = max(relAbund), min = min(relAbund))
```

```{r}
#| eval: true
#| label: fig-base-phyla-boxplot
#| fig-cap: Relative abundances of the top five most abundant bacterial phyla at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).

#default settings plot relative abundances (normalise = TRUE)
p <- amp_boxplot(d,
            normalise = TRUE,
            group_by = "b12_status",
            tax_aggregate = "Phylum",
            tax_show = 5,
            point_size = 1,
            plot_flip = T)

p + 
  scale_color_brewer(palette = "Set2")        
```

##### Statistical Comparison between B12 Status group Phyla Sum

::: h0

There is no differnece in the counts of bacterial phyla between vitamin B12 status groups at baseline 

:::

::: h1

Bacterial phyla counts diverge between vitamin B12 status groups at baseline

:::

```{r}
#| eval: true
#| label: tbl-b12-kw
#| tbl-cap: Kruskall-Wallis test results with computed $\eta^2_H$ effect size estimate for bacterial phyla relative abundances at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
tmp <- amp_heatmap(d,
            group_by = "b12_status",
            tax_aggregate = "Phylum",
            tax_show = 20,
            plot_values = T,
            normalise = F, 
            tax_empty="OTU")

df <- tmp$data %>% dplyr::rename(phyla = Display, b12_status = `.Group`)

kw <- df %>% group_by(phyla) %>%
  kruskal_test(Sum ~ b12_status)

kw_eff<-df %>% group_by(phyla) %>% kruskal_effsize(Sum ~ b12_status)

left_join(kw,kw_eff, by="phyla") %>% kable()
```

```{r}
#| eval: true
#| label: tbl-b12-dunn-vda
#| tbl-cap: Kruskall-Wallis post hoc Dunn's test results with computed pairwise VDA effect size estimate for bacterial phyla relative abundances at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
 
dunn <- df %>% group_by(phyla) %>%
  dunn_test(Sum ~ b12_status)

b12_vda <- data.frame()

for(p in unique(df$phyla)) {
  tmp_vda <- df %>% dplyr::filter(phyla==p) %>%
  calculate_pairwise_vda(., "Sum", "b12_status") %>% 
  mutate(phyla=p) %>% 
  relocate(phyla, .before=everything())

  b12_vda <- rbind(b12_vda,tmp_vda)
}

left_join(dunn, b12_vda, by =c("phyla", "group1", "group2")) %>% kable()
```

###### Fusobacteria and B12

```{r}
#| eval: true
#| label: fig-base-fuso
#| fig-cap: Fusobacterium at baseline

df <- df %>%
  mutate(vitb12 = baseline_metadata$vitb12[match(Sample, baseline_metadata$sample_names)],
  sex = baseline_metadata$sexf[match(Sample, baseline_metadata$sample_names)]) 

library(corrplot)
library(Hmisc)

# Create correlation matrix
cor_matrix <- rcorr(as.matrix(df[c("Sum", "b12_status")]), type = "spearman")

# Visualize
corrplot(cor_matrix$r, type = "upper", 
         p.mat = cor_matrix$P, 
         sig.level = 0.05)

fuso <- df %>%
  filter(phyla == "Fusobacteriota")

fuso %>%
  ggplot(aes(x=b12_status, y = Sum)) +
    geom_boxplot() +
    geom_jitter() +
    coord_cartesian(ylim=c(0,3000))

```

```{r}
#| eval: true
#| label: fig-corr-plot

# Reshape data to wide format (each phyla becomes a column)
phyla_wide <- df %>%
  select(Sample, phyla, Sum, b12_status) %>%
  pivot_wider(names_from = phyla, 
              values_from = Sum) %>%
  select(-Sample)  # Remove Sample column if you don't need it

# Calculate correlation matrix
cor_matrix <- rcorr(as.matrix(phyla_wide), type = "spearman")

# Create corrplot
corrplot(cor_matrix$r,
         p.mat = cor_matrix$P,
         method = "color",    # can be "circle", "square", "number"
         type = "upper",      # show only upper triangle
         order = "hclust",    # cluster similar correlations together
         addCoef.col = "black", # add correlation coefficients
         tl.col = "black",    # text label color
         tl.srt = 45,         # rotate text labels
         sig.level = 0.05,    # significance level
         insig = "blank")     # hide insignificant correlations

# Alternative with significance stars
corrplot(cor_matrix$r,
         p.mat = cor_matrix$P,
         method = "color",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45,
         sig.level = c(0.001, 0.01, 0.05),
         pch.cex = 0.9,
         insig = "label_sig")  # show significance stars


phyla_summary <- df %>%
  group_by(phyla, b12_status) %>%
  summarise(total_abundance = sum(Sum)) %>%
  pivot_wider(names_from = b12_status, 
              values_from = total_abundance)

# Calculate correlations
cor_matrix <- cor(phyla_summary[,-1], method = "spearman")  # remove phyla names column
p_values <- cor.mtest(phyla_summary[,-1], method = "spearman")$p

# Create corrplot
corrplot(cor_matrix,
         method = "color",
         type = "full",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45,
         sig.level = 0.05,
         insig = "blank")

```


Poisson distribution and log link function used as this is count data.

```{r}
#| eval: true
#| label: tbl-base-glm-fuso

anova(glm(Sum ~ vitb12*sex, data=fuso, family = "poisson"))
```


#### Baseline Genera Abundance

```{r}
#| eval: true
#| label: fig-base-b12-gen-heatmap
#| fig-cap: Relative abundance of top 20 bacterial genera at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL). Average group relative abundances are shown on the graph.

(heatmap <- amp_heatmap(d,
            group_by = "b12_status",
            tax_aggregate = "Genus",
            tax_show = 20,
            plot_values = T,
            normalise = T, 
            plot_values_size = 8,
            round = 3,
            tax_empty="OTU") +
              theme(axis.text.x = element_text(angle = 45, size = 14, vjust = 1), axis.text.y = element_text(size = 14, face="bold.italic"),  legend.position = "right", legend.title = element_text(color = "black", size = 10, face = "bold")))
```

```{r}
#| eval: true
#| label: tbl-b12-spp-heatmap
#| tbl-cap: Relative abundance of top 20 bacterial genera at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
heatmap$data %>% 
  group_by(Display, .Group) %>% 
  summarise(mean = round(mean(Sum), 4)) %>% 
  arrange(desc(mean)) %>%
  mutate(mean =  case_when(
    mean >= 0.01 ~ round(mean, 2), 
    TRUE ~ mean
  )) %>%
  mutate(mean = paste0(mean, "%")) %>%
  pivot_wider(names_from = .Group, values_from = mean) %>%
  kable()
```

###### Statistical Compariosn between B12 Status group  Genera Sum

::: h0

there is no differnece in count of genera between vitamin B12 status groups at baseline 

:::

::: h1

genera count increases/decreases from vitamin B12 status groups at baseline

:::

```{r}
#| eval: true
#| label: tbl-sum-gen
#| tbl-cap: Summary of counts

tmp <- amp_heatmap(d,
            group_by = "b12_status",
            tax_aggregate = "Genus",
            tax_show = 200,
            normalise = F, 
            tax_empty="OTU")

df <- tmp$data %>% dplyr::rename(genera = Display, b12_status = `.Group`) 

df %>%
  group_by(genera, b12_status) %>%
  summarise(mean = mean(Sum), median = median(Sum), max = max(Sum), min = min(Sum)) %>%
  arrange(desc(genera))
```

```{r}
#| eval: true
#| label: tbl-b12-kw
#| tbl-cap: Kruskall-Wallis test results with computed $\eta^2_H$ effect size estimate for bacterial phyla relative abundances at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).

kw <- df %>% group_by(genera) %>%
  kruskal_test(Sum ~ b12_status)

kw_eff<-df %>% group_by(genera) %>% kruskal_effsize(Sum ~ b12_status)

left_join(kw,kw_eff, by="genera") %>% kable()
```

```{r}
#| eval: true
#| label: tbl-b12-dunn-vda
#| tbl-cap: Kruskall-Wallis post hoc Dunn's test results with computed pairwise VDA effect size estimate for bacterial genera relative abundances at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).

kw_sig <- kw %>% filter(p<=0.05)

dunn <- df %>% filter(genera %in% kw_sig$genera) %>%
  group_by(genera) %>%
  dunn_test(Sum ~ b12_status)

b12_vda <- data.frame()

for(p in unique(df$genera)) {
  tmp_vda <- df %>% dplyr::filter(genera==p) %>%
  calculate_pairwise_vda(., "Sum", "b12_status") %>% 
  mutate(genera=p) %>% 
  relocate(genera, .before=everything())

  b12_vda <- rbind(b12_vda,tmp_vda)
}

left_join(dunn, b12_vda, by =c("genera", "group1", "group2")) %>% kable()
```

#### Baseline Species Abundance

```{r}
#| eval: true
#| label: fig-base-b12-spp-heatmap
#| fig-cap: Relative abundance of top 20 bacterial species at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL). Average group relative abundances are shown on the graph.

(heatmap <- amp_heatmap(d,
            group_by = "b12_status",
            tax_aggregate = "Species",
            tax_show = 20,
            plot_values = T,
            normalise = T, 
            plot_values_size = 8,
            round = 3,
            tax_empty="OTU") +
              theme(axis.text.x = element_text(angle = 45, size = 14, vjust = 1), axis.text.y = element_text(size = 14, face="bold.italic"),  legend.position = "right", legend.title = element_text(color = "black", size = 10, face = "bold")))
```

```{r}
#| eval: true
#| label: tbl-b12-spp-heatmap
#| tbl-cap: Relative abundance of top 20 bacterial species at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
heatmap$data %>% 
  group_by(Display, .Group) %>% 
  summarise(mean = round(mean(Sum), 4)) %>% 
  arrange(desc(mean)) %>%
  mutate(mean =  case_when(
    mean >= 0.01 ~ round(mean, 2), 
    TRUE ~ mean
  )) %>%
  mutate(mean = paste0(mean, "%")) %>%
  pivot_wider(names_from = .Group, values_from = mean) %>%
  kable()
```

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: Summary statistics of bacterial relative abundances at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
df <- heatmap$data %>% 
  dplyr::select(species = Display, Sample , relAbund = Sum) %>% 
  mutate(timepoint = b12_base_meta$timepoint[match(Sample,b12_base_meta$sample_names)]) %>%
  mutate(b12_status = b12_base_meta$b12_status[match(Sample,b12_base_meta$sample_names)])

df %>%
  dplyr::select(species, b12_status, relAbund) %>%
  group_by(species, b12_status) %>%
  summarise(mean = mean(relAbund), median = median(relAbund), max = max(relAbund), min = min(relAbund))
```

###### Statistical Compariosn between B12 Status group  Species Sum

::: h0

there is no differnece in realtive abundance of species between vitamin B12 status groups at baseline 

:::

::: h1

species realtive abundance increases/decreases from vitamin B12 status groups at baseline

:::

```{r}
#| eval: true
#| label: tbl-sum-gen
#| tbl-cap: Summary of counts

tmp <- amp_heatmap(d,
            group_by = "b12_status",
            tax_aggregate = "Species",
            tax_show = 200,
            normalise = F, 
            tax_empty="OTU")

df <- tmp$data %>% dplyr::rename(species = Display, b12_status = `.Group`) 

df %>%
  group_by(species, b12_status) %>%
  summarise(mean = mean(Sum), median = median(Sum), max = max(Sum), min = min(Sum)) %>%
  arrange(desc(species)) %>% kable()
```

```{r}
#| eval: true
#| label: tbl-b12-kw
#| tbl-cap: Kruskall-Wallis test results with computed $\eta^2_H$ effect size estimate for bacterial phyla relative abundances at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).

kw <- df %>% group_by(species) %>%
  kruskal_test(Sum ~ b12_status)

kw_eff<-df %>% group_by(species) %>% kruskal_effsize(Sum ~ b12_status)

left_join(kw,kw_eff, by="species") %>% kable()
```

```{r}
#| eval: true
#| label: tbl-b12-dunn-vda
#| tbl-cap: Kruskall-Wallis post hoc Dunn's test results with computed pairwise VDA effect size estimate for bacterial species relative abundances at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).

kw_sig <- kw %>% filter(p<=0.05)

dunn <- df %>% filter(species %in% kw_sig$species) %>%
  group_by(species) %>%
  dunn_test(Sum ~ b12_status)

b12_vda <- data.frame()

for(p in unique(df$species)) {
  tmp_vda <- df %>% dplyr::filter(species==p) %>%
  calculate_pairwise_vda(., "Sum", "b12_status") %>% 
  mutate(species=p) %>% 
  relocate(species, .before=everything())

  b12_vda <- rbind(b12_vda,tmp_vda)
}

kable(dunn_spp_results <- left_join(dunn, b12_vda, by =c("species", "group1", "group2")))
```

```{r}
#| eval: true
#| label: fig-base-heatmap
#| fig-cap: Heatmap of phyla and species differences between 

ggplot(df, aes(x = b12_status, y = Sum)) +
  # Add boxplot
  geom_boxplot(aes(fill = b12_status), alpha = 0.5) +
  # Add points with size reflecting abundance
  geom_point(aes(size = Abundance), 
             position = position_jitterdodge(),
             alpha = 0.6) +
  # Facet by species
  facet_wrap(~species, scales = "free_y") +
  # Add significance levels from Dunn's test
  geom_text(data = dunn_spp_results, 
            aes(x = group1, y = max(df$Sum) * 1.1,
                label = p.adj.signif),
            size = 3) +
  # Customize appearance
  scale_fill_manual(values = c(
    "deficiency" = "#e78284",
    "insufficiency" = "#f4b8e4",
    "sufficiency" = "#8CAAEE"
  )) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  ) +
  labs(
    x = "B12 Status",
    y = "Sum",
    size = "Abundance",
    fill = "B12 Status",
    title = "Species Abundance by B12 Status",
    subtitle = "Point size indicates abundance; ns: not significant"
  )

#Bubble plot with significance connections
abundance_summary <- count_data %>%
  group_by(species, b12_status) %>%
  summarise(
    mean_abundance = mean(Abundance),
    mean_sum = mean(Sum)
  )
  
ggplot(abundance_summary, 
       aes(x = b12_status, y = species)) +
  # Add bubbles
  geom_point(aes(size = mean_sum, fill = mean_abundance),
             shape = 21, alpha = 0.7) +
  # Add significance lines
  geom_segment(data = dunns_results,
               aes(x = group1, xend = group2,
                   y = species, yend = species,
                   linetype = p.adj.signif),
               position = position_dodge(width = 0.9)) +
  scale_size_continuous(range = c(2, 20)) +
  scale_fill_gradient(
    low = "#8CAAEE",
    high = "#f4b8e4"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### OGU Network Analysis
```{r}
#| eval: true
#| label: fig-base-OGU-net
#| fig-cap: "OGU network graph at Phylum level: A) fruchtermanreingold; B) MDS" 

mds_net <- amp_otu_network(
  d,
  min_abundance = 0,
  color_by = "b12_status",
  tax_aggregate = "Phylum",
  tax_show = 10,
  normalise = TRUE,
  size = "degree",
  label=TRUE,
  mode="mds"
)

fr_net <- amp_otu_network(
  d,
  min_abundance = 0,
  color_by = "b12_status",
  tax_aggregate = "Phylum",
  tax_show = 10,
  normalise = TRUE,
  size = "degree",
  label=TRUE,
  mode="fruchtermanreingold"
)

ggarrange(fr_net, mds_net, ncol=2, common.legend = TRUE, labels = "AUTO")
```

```{r}
#| eval: true
#| label: fig-base-OGU-net
#| fig-cap: "OGU network graph at Species level: A) fruchtermanreingold; B) MDS"

mds_net <- amp_otu_network(
  d,
  min_abundance = 0,
  color_by = "b12_status",
  tax_aggregate = "Species",
  tax_show = 10,
  normalise = TRUE,
  size = "degree",
  label=TRUE,
  mode="mds"
)

fr_net <- amp_otu_network(
  d,
  min_abundance = 0,
  color_by = "b12_status",
  tax_aggregate = "Species",
  tax_show = 10,
  normalise = TRUE,
  size = "degree",
  label=TRUE,
  mode="fruchtermanreingold"
)

ggarrange(fr_net, mds_net, ncol=2, common.legend = TRUE, labels = "AUTO")
```

```{r}
#| eval: false
#| include: false
#| label: fig-base-OGU-statnet
#| fig-cap: OGU network graph

library(statnet)
# Convert distances to similarities
sim_matrix <- 1 - as.matrix(bray)

# Create adjacency matrix based on threshold
adj_matrix <- sim_matrix
adj_matrix[sim_matrix < 0.7] <- 0
diag(adj_matrix) <- 0
    
# Convert to network
network <- network(adj_matrix, matrix.type="adjacency")

gplot(network, vertex.col = 2, displaylabels = TRUE)

get.vertex.attribute(network)
```

