## B12 Status: Baseline Beta Diversity

This analysis examines how bacterial communities differ between samples, using multiple distance metrics to capture different aspects of community similarity.

Four different distance metrics were calculated:

1. *Bray-Curtis*: Considers abundance data, sensitive to dominant species
2. *Jaccard*: Based on presence/absence, treats all species equally
3. *Unweighted UniFrac*: Incorporates phylogenetic relationships, presence/absence
4. *Weighted UniFrac*: Combines phylogenetic relationships with abundance data

```{r}
#| echo: false
#| eval: true
#| label: setup
#| include: false

# This initial code block loads custom utility functions and environment variables

# These are stored in separate files to keep the main analysis clean and organized
source(normalizePath("utils/r-utils.r"))
source(normalizePath("_environment.required"))
load(normalizePath("data/seqdata.RData"))

base_b12 <- readRDS(file.path(RESULTS, "B12-status-subgroup-analysis/b12-baseline-seqdata.rds"))

suppressPackageStartupMessages(library(vegan))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(pairwiseAdonis))

dir.create(file.path(RESULTS, "B12-status-subgroup-analysis"), showWarnings = FALSE, recursive = TRUE)
```

### Beta diversity calculation

```{r}
#| eval: true
#| label: beta diversity distance matrix calculation

#phyloseq wraps a number of vegan package functions and can be used to call vegdist()
bray <- phyloseq::distance(base_b12, method = "bray") #calculated on abundance 
jac <- phyloseq::distance(base_b12, method = "jaccard")
uunifrac <- phyloseq::distance(base_b12, method = "uunifrac")
wunifrac <- phyloseq::distance(base_b12, method = "wunifrac")
```

### Statistial testing for differences between B12 Status Groups

Three complementary approaches were used to test for community differences:

::: {.grid}

::: {.g-col-4}

**ANOSIM (Analysis of Similarity)**

Tests whether between-group differences are larger than within-group differences

- R statistic ranges from -1 to 1
- R ≈ 0: no difference between groups
- R > 0: groups are more different than expected by chance
- Stratification by timepoint accounts for temporal effects

:::

::: {.g-col-4}

**PERMDISP (Permutational Analysis of Multivariate Dispersions)**

Tests for homogeneity of dispersions between groups

- Important assumption check for PERMANOVA
- Analogous to Levene's test for univariate data
- Significant results could indicate different within-group variability

:::

::: {.g-col-4}

**PERMANOVA (Permutational Multivariate Analysis of Variance)**

Tests for differences in community centroids between groups

- Accounts for both treatment and time effects
- More robust than ANOSIM
- Effect sizes calculated using partial eta-squared and omega-squared
- Requires homogeneous dispersions (tested by PERMDISP)

:::

:::

#### ANOSIM: Analysis of Similarity

ANOSIM is a permutation test which calculates R statistic (-1,1) $\approx$ correlation coefficient $\approx$ effect size

::: h0

there is no difference between intervention b12_statuss in terms of beta diversity metrics ($R = 0$)

:::

::: h0

biofortified pearl millet differentlaly changes beta diversity metrics ($R \neq 0$)

:::

```{r}
#| eval:  true
#| label: ANOSIM
#function requires that rows = sample and column = OGU
summary(vegan::anosim(bray, grouping = microbiome::meta(base_b12)$b12_status)) #there is no difference in Bray Curtis distance between treatment groups

summary(vegan::anosim(jac, grouping = microbiome::meta(base_b12)$b12_status)) #there is no difference in Jaccard distance between treatment groups

summary(vegan::anosim(uunifrac, grouping = microbiome::meta(base_b12)$b12_status)) #there is no difference in unweighted unifrac distance between treatment groups

summary(vegan::anosim(wunifrac, grouping = microbiome::meta(base_b12)$b12_status)) #there is no difference in weighted unifrac distance between treatment groups
```

#### PERMDISP

PERMANOVA assumes homogeneity of multivariate dispersions across groups therefore a significant PERMANOVA result could be due to dispersion, location or both many permutational techniqures can be sensitive to dispersion (differences in centroids and/or differences in how data points in clusters scatter around their centroid). In univariate analysis Levene's test can test for this. PERMDISP is an extension of this that can be used in multivariate analysis technique involves calculating the distance from each data point to its group centroid and then testing whether those distances differ among the groups however, certain metrics are non-metric or semi-metric e.g. Bray-Curtis and centroids cannot be directly calculated -> PCoA + cantroid calc on the coordinates ([cite](https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/permdisp/))

```{r}
#| eval:  true
#| label: PERMDISP
bray_disp <- betadisper(d = bray, group = microbiome::meta(base_b12)$b12_status, type = "centroid")
permutest(bray_disp, pairwise = F, group = timepoint)
```

```{r}
#| eval: true
#| label: tbl-anova-sum
#| tbl-cap: ANOVA results summary for Bray-Curtis Distance
report::report_table(anova(bray_disp))
```

```{r}
#| eval: true
jac_disp <- betadisper(d = jac, group = microbiome::meta(base_b12)$b12_status, type = "centroid")
permutest(jac_disp, pairwise = F, group = timepoint)
```

```{r}
#| eval: true
#| label: tbl-anova-sum
#| tbl-cap: ANOVA results summary for Jaccard Distance
report::report_table(anova(jac_disp))
```

```{r}
#| eval: true
uunifrac_disp <- betadisper(d = uunifrac, group = microbiome::meta(base_b12)$b12_status, type = "centroid")
permutest(uunifrac_disp, pairwise = F, group = timepoint)
```

```{r}
#| eval: true
#| label: tbl-anova-sum
#| tbl-cap: ANOVA results summary for Unweighted UniFrac Distance
report::report_table(anova(uunifrac_disp))
```

```{r}
#| eval: true
wunifrac_disp <- betadisper(d = wunifrac, group = microbiome::meta(base_b12)$b12_status, type = "centroid")
permutest(wunifrac_disp, pairwise = F, group = timepoint)
```

```{r}
#| eval: true
#| label: tbl-anova-sum
#| tbl-cap: ANOVA results summary for Weighted UniFrac Distance
report::report_table(anova(wunifrac_disp))
```

#### PERMANOVA

::: h0

there is no change in beta diversity metrics arosss time (baseline to endpoint)
:::

::: h1

beta diversity metrics change (increase) across time (baseline to endpoint)
:::

assumption: variance of dispersions are homogenous across groups

```{r}
#| eval: true
#| label: omnibus PERMANOVA results output
(bray_adon <- adonis2(bray ~ b12_status*sexf, by = "terms", data = microbiome::meta(base_b12)))

(jac_adon <- adonis2(jac ~ b12_status*sexf, by = "terms", data = microbiome::meta(base_b12)))

(wunifrac_adon <- adonis2(wunifrac ~ b12_status*sexf, by = "terms", data = microbiome::meta(base_b12)))

(uunifrac_adon <- adonis2(uunifrac ~ b12_status*sexf, by = "terms", data = microbiome::meta(base_b12)))

```

```{r}
#| eval: true
#| label: pairwise PERMANOVA results output
(bray_adon_pairwise <- pairwise.adonis2(bray ~ b12_status*sexf, by = "terms", data = microbiome::meta(base_b12)))

(jac_adon_pairwise <- pairwise.adonis2(jac ~ b12_status*sexf, by = "terms", data = microbiome::meta(base_b12)))

(wunifrac_adon_pairwise <- pairwise.adonis2(wunifrac ~ b12_status*sexf, by = "terms", data = microbiome::meta(base_b12)))

(uunifrac_adon_pairwise <- pairwise.adonis2(uunifrac ~ b12_status*sexf, by = "terms", data = microbiome::meta(base_b12)))

```

#### Effect Size Calculations

##### Eta-Squared

PERMANOVA effect size is typically measured using the partial eta-squared ($\eta^2_p$) statistic. The partial eta-squared is a measure of the proportion of total variance in the multivariate data explained by the grouping variable (i.e., the factor of interest). In other words, $\eta^2_p$ calculations assess the magnitude of community differences.

::: {.callout-warn}

This measure tends to overestimate effect sizes in small samples

:::

$$
(\eta^2_p = \frac{SS_{\text{between}}}{SS_{\text{total}}})
$$

- Partial eta-squared ranges from 0 to 1.
- Values < 0.6 indicate small effect sizes
- A value of 0 indicates that the grouping variable explains none of the variance in the dependent variable.
- A value of 1 indicates that the grouping variable explains all of the variance in the dependent variable.
- Can be interpreted similarly to $R^2$ in linear regression
- In the context of PERMANOVA, $\eta^2_p$ indicates the proportion of total variance in the multivariate data (e.g., beta diversity distances) that is explained by the grouping variable.

```{r}
#| eval: true
#| label: tbl-eta-squared
#| tbl-cap: Eta-squared effect size calculations for permanova

etasq <- function(adonis){
  SS_between <- adonis$SumOfSqs[1]
  SS_total <- adonis$SumOfSqs[length(adonis$SumOfSqs)]
  etasq <- SS_between / SS_total
  return(etasq)
}

Overall <- c(etasq(bray_adon),etasq(jac_adon),etasq(wunifrac_adon),etasq(uunifrac_adon))

rbind(Overall,do.call(cbind,
  list(
    `Bray-Curtis` = lapply(bray_adon_pairwise[2:length(bray_adon_pairwise)], etasq),
    Jaccard = lapply(jac_adon_pairwise[2:length(jac_adon_pairwise)], etasq),
    `Weighted UniFrac` = lapply(wunifrac_adon_pairwise[2:length(wunifrac_adon_pairwise)], etasq),
    `Unweighted UniFrac` = lapply(uunifrac_adon_pairwise[2:length(uunifrac_adon_pairwise)], etasq)))
) %>% kable()
#all <0.06 => small effect size
```

##### Omega-squared

Omega-squared ($\omega^2$) provides a less biased measure of effect size for ANOVA-type analyses by accounting for the mean-squared error of the observed samples [@Olejnik2009]

$$
\omega^2 = \frac{SS_A - (a-1)\frac{SS_W}{(N-a)}}{SS_T +\frac{SS_W}{(N-a)}}
$$

Where:

- $SS_T$: total sum of squares
- $SS_W$: within-group sum of squares (sum of the squares of distances within groups divided by the number of subjects per group)
- $SS_A$: between-group sum of squares ($SS_A =SS_T − SS_W$)
- $(a − 1)$: degress of freedom of the grouping factor
- $(N - a)$: residual degrees of freedom

```{r}
#| eval: true
#| label: tbl-Omega-squared
#| tbl-cap: Omega-squared effect size calculations for permanova

#source: https://stackoverflow.com/a/74452219/16461035
adonis_OmegaSq <- function(adonisOutput, partial = TRUE){
    if(!(is(adonisOutput, "adonis") || is(adonisOutput, "anova.cca")))
        stop("Input should be an adonis object")
    if (is(adonisOutput, "anova.cca")) {
        aov_tab <- adonisOutput
        aov_tab$MeanSqs <- aov_tab$SumOfSqs / aov_tab$Df
        aov_tab$MeanSqs[length(aov_tab$Df)] <- NA
    } else {
        aov_tab <- adonisOutput$aov.tab
    }
    heading <- attr(aov_tab, "heading")
    MS_res <- aov_tab[pmatch("Residual", rownames(aov_tab)), "MeanSqs"]
    SS_tot <- aov_tab[rownames(aov_tab) == "Total", "SumsOfSqs"]
    N <- aov_tab[rownames(aov_tab) == "Total", "Df"] + 1
    if(partial){
        omega <- apply(aov_tab, 1, function(x) (x["Df"]*(x["MeanSqs"]-MS_res))/(x["Df"]*x["MeanSqs"]+(N-x["Df"])*MS_res))
        aov_tab$parOmegaSq <- c(omega[1:(length(omega)-2)], NA, NA)
    } else {
        omega <- apply(aov_tab, 1, function(x) (x["SumsOfSqs"]-x["Df"]*MS_res)/(SS_tot+MS_res))
        aov_tab$OmegaSq <- c(omega[1:(length(omega)-2)], NA, NA)
    }
    if (is(adonisOutput, "adonis"))
        cn_order <- c("Df", "SumsOfSqs", "MeanSqs", "F.Model", "R2",
                      if (partial) "parOmegaSq" else "OmegaSq", "Pr(>F)")
    else
        cn_order <- c("Df", "SumOfSqs", "F", if (partial) "parOmegaSq" else "OmegaSq",
                      "Pr(>F)")
    aov_tab <- aov_tab[, cn_order]
    attr(aov_tab, "names") <- cn_order
    attr(aov_tab, "heading") <- heading
    if (is(adonisOutput, "adonis"))
        adonisOutput$aov.tab <- aov_tab
    else
        adonisOutput <- aov_tab
    return(adonisOutput)
}

Overall <- c(adonis_OmegaSq(bray_adon)$parOmegaSq[1],adonis_OmegaSq(jac_adon)$parOmegaSq[1],adonis_OmegaSq(wunifrac_adon)$parOmegaSq[1],adonis_OmegaSq(uunifrac_adon)$parOmegaSq[1])

rbind(Overall,do.call(cbind,
  list(
    `Bray-Curtis` = lapply(bray_adon_pairwise[2:length(bray_adon_pairwise)], function(x){
      adonis_OmegaSq(x)$parOmegaSq[1]}),
    Jaccard = lapply(jac_adon_pairwise[2:length(jac_adon_pairwise)], function(x){
      adonis_OmegaSq(x)$parOmegaSq[1]}),
    `Weighted UniFrac` = lapply(wunifrac_adon_pairwise[2:length(wunifrac_adon_pairwise)], function(x){
      adonis_OmegaSq(x)$parOmegaSq[1]}),
    `Unweighted UniFrac` = lapply(uunifrac_adon_pairwise[2:length(uunifrac_adon_pairwise)], function(x){
      adonis_OmegaSq(x)$parOmegaSq[1]})))
) %>% kable()
```

### Ordination Methods

Two different ordination approaches were used to visualize community differences:

#### Metric Multidimensional Scaling (MDS)

MDS creates a visualization where:

- Similar samples are placed closer together
- Distances between points approximate their actual dissimilarities
- `trymax = 100` ensures a stable solution by trying multiple starting configurations

```{r}
#| eval:  true
#| label: beta div MDS ordination

#perform metric multidimensional scaling
bray.ord.mds <- phyloseq::ordinate(base_b12, method = "MDS", distance = bray, trymax = 100)
jac.ord.mds <- phyloseq::ordinate(base_b12, method = "MDS", distance = jac, trymax = 100)
uunifrac.ord.mds <- phyloseq::ordinate(base_b12, method = "MDS", distance = uunifrac, trymax = 100)
wunifrac.ord.mds <- phyloseq::ordinate(base_b12, method = "MDS", distance = wunifrac, trymax = 100)
```

#### Non-metric Multidimensional Scaling (NMDS)

NMDS is an iterative rank based procedure that sets stress values rather than eigenvalues to determine the "success" of the ordination. It is typically to aim for stress values $\leq$ 0.1, which indicates a good representation of the relationship between values.

NMDS is often preferred for ecological data because:

- It works with rank-order distances rather than absolute distances
- More robust to non-linear relationships
- Stress value indicates how well the ordination represents the data

```{r}
#| eval:  true
#| label: beta div NMDS ordination
bray.ord.nmds <- phyloseq::ordinate(base_b12, method = "NMDS", distance = bray, trymax = 100)
jac.ord.nmds <- phyloseq::ordinate(base_b12, method = "NMDS", distance = jac, trymax = 100)
uunifrac.ord.nmds <- phyloseq::ordinate(base_b12, method = "NMDS", distance = uunifrac, trymax = 100)
wunifrac.ord.nmds <- phyloseq::ordinate(base_b12, method = "NMDS", distance = wunifrac, trymax = 100)
```


#### Ordination Plots with PERMANOVA $R^2$ and P-values

##### MDS

::: {#fig-box}

```{r}
#| eval:  true
#| label: fig-base-beta-mds
bray.plot.mds <- plot_ordination(base_b12, bray.ord.mds, color = "vitb12", shape = "sexf") + 
  labs(title="Bray-Curtis Distance", color = "Timepoint", shape = "Intervention b12_status") + geom_point(size=4) +     
  scale_color_brewer(palette = "RdYlBu") +
  theme(legend.title= element_text(size = fontsize, face='bold', color='black'), plot.title = element_text(size = 12)) +
  scale_shape_manual(values = c(16,17), labels = trt_labels) + 
  annotate("text", x = Inf, y = Inf, hjust = 2, vjust = 2, label = bquote(atop(R^2 == .(round(bray_adon[1,3],3)), p == .(round(bray_adon[1,5],3)))))

jac.plot.mds <- plot_ordination(base_b12, jac.ord.mds, color = "b12_status", shape = "sexf") + 
    labs(title="Jaccard Disimilarity", color = "Timepoint", shape = "Intervention b12_status") + 
    geom_point(size=4) + 
    scale_color_brewer(palette = "Set2") +
    theme(legend.title= element_text(size = fontsize, face='bold', color='black'), plot.title = element_text(size = 12)) +
    scale_shape_manual(values = c(16,17), labels = trt_labels)+ 
   annotate("text", x = Inf, y = Inf, hjust = 2, vjust = 2, label = bquote(atop(R^2 == .(round(jac_adon[1,3],3)), p == .(round(jac_adon[1,5],3)))))

uunifrac.plot.mds <- plot_ordination(base_b12, uunifrac.ord.mds, color = "b12_status", shape = "sexf") + 
    labs(title="Unweighted UniFrac", color = "Timepoint", shape = "Intervention b12_status") + 
    geom_point(size=4) + 
    scale_color_brewer(palette = "Set2") +
    theme(legend.title= element_text(size = fontsize, face='bold', color='black'), plot.title = element_text(size = 12)) +
    scale_shape_manual(values = c(16,17), labels = trt_labels)+ 
   annotate("text", x = Inf, y = Inf, hjust = 2, vjust = 2, label = bquote(atop(R^2 == .(round(uunifrac_adon[1,3],3)), p == .(round(uunifrac_adon[1,5],3)))))

wunifrac.plot.mds <- plot_ordination(base_b12, wunifrac.ord.mds, color = "b12_status", shape = "sexf") + 
    labs(title="Weighted UniFrac", color = "Timepoint", shape = "Intervention b12_status") + 
    geom_point(size=4) + 
    scale_color_brewer(palette = "Set2") +
    theme(legend.title= element_text(size = fontsize, face='bold', color='black'), plot.title = element_text(size = 12)) +
    scale_shape_manual(values = c(16,17), labels = trt_labels)+ 
   annotate("text", x = -Inf, y = Inf, hjust = -1, vjust = 2, label = bquote(atop(R^2 == .(round(wunifrac_adon[1,3],3)), p == .(round(wunifrac_adon[1,5],3)))))

ggarrange(bray.plot.mds, jac.plot.mds, uunifrac.plot.mds, wunifrac.plot.mds,ncol = 2, nrow=2, common.legend = T, legend = "bottom", labels = "AUTO")

```

**β-diversity ordination plots comparing intervention b12_statuss at baseline and endpoint:** There were no differences in  community structure, nor dispersion of sequences by treatment group for any β-diversity metric measured (PERMDISP; p > 0.05).  However, overall, there was a small shift in community structure at endpoint compared to baseline (determined by PERMANOVA) for (A) Bray Curtis distances (F statistic: `r bray_adon$F[1]`; p = `r bray_adon$'Pr(>F)'[1]`), (B) Jaccard similarity (F statistic: `r jac_adon$F[1]`; p = `r jac_adon$'Pr(>F)'[1]`), (C) Unweighted UniFrac (F  statistic: `r uunifrac_adon$F[1]`; p = `r uunifrac_adon$'Pr(>F)'[1]`), and (D) Weighted UniFrac (F statistic: `r wunifrac_adon$F[1]`; p = `r wunifrac_adon$'Pr(>F)'[1]`).

:::

#### NMDS

::: {fig-box}

```{r}
#| eval: true
#| label: fig-base-beta-nmds

bray.plot.nmds <- plot_ordination(base_b12, bray.ord.nmds, color = "b12_status", shape = "sexf") + 
              labs(title="Bray-Curtis Distance", color = "Timepoint", shape = "Intervention b12_status") + 
              geom_point(size=4) + 
              #ggalt::geom_encircle(aes(group = b12_status, fill = b12_status), size = 1, alpha = 0.1) +
              scale_color_brewer(palette = "Set2") +
              theme(legend.title= element_text(size = fontsize, face='bold', color='black'), plot.title = element_text(size = 12)) +
              scale_shape_manual(values = c(16,17), labels = trt_labels)+ 
   annotate("text", x = Inf, y = Inf, hjust = 2, vjust = 2, label = bquote(atop(R^2 == .(round(bray_adon[1,3],3)), p == .(round(bray_adon[1,5],3)))))
 
jac.plot.nmds <- plot_ordination(base_b12, jac.ord.nmds, color = "b12_status", shape = "sexf") + 
              labs(title="Jaccard Disimilarity", color = "Timepoint", shape = "Intervention b12_status") + 
              geom_point(size=4) + 
              scale_color_brewer(palette = "Set2") +
              theme(legend.title= element_text(size = fontsize, face='bold', color='black'), plot.title = element_text(size = 12)) +
              scale_shape_manual(values = c(16,17), labels = trt_labels)+ 
   annotate("text", x = Inf, y = Inf, hjust = 2, vjust = 2, label = bquote(atop(R^2 == .(round(jac_adon[1,3],3)), p == .(round(jac_adon[1,5],3)))))

uunifrac.plot.nmds <- plot_ordination(base_b12, uunifrac.ord.nmds, color = "b12_status", shape = "sexf") + 
              labs(title="Unweighted UniFrac", color = "Timepoint", shape = "Intervention b12_status") + 
              geom_point(size=4) + 
              scale_color_brewer(palette = "Set2") +
              theme(legend.title= element_text(size = fontsize, face='bold', color='black'), plot.title = element_text(size = 12)) +
              scale_shape_manual(values = c(16,17), labels = trt_labels)+ 
   annotate("text", x = Inf, y = Inf, hjust = 2, vjust = 2, label = bquote(atop(R^2 == .(round(uunifrac_adon[1,3],3)), p == .(round(uunifrac_adon[1,5],3)))))


wunifrac.plot.nmds <- plot_ordination(base_b12, wunifrac.ord.nmds, color = "b12_status", shape = "sexf") + 
              labs(title="Weighted UniFrac", color = "Timepoint", shape = "Intervention b12_status") + 
              geom_point(size=4) + 
              scale_color_brewer(palette = "Set2") +
              theme(legend.title= element_text(size = fontsize, face='bold', color='black'), plot.title = element_text(size = 12)) +
              scale_shape_manual(values = c(16,17), labels = trt_labels)+ 
   annotate("text", x = Inf, y = Inf, hjust = 2, vjust = 2, label = bquote(atop(R^2 == .(round(wunifrac_adon[1,3],3)), p == .(round(wunifrac_adon[1,5],3)))))

ggarrange(bray.plot.nmds, jac.plot.nmds, uunifrac.plot.nmds, wunifrac.plot.nmds,ncol = 2, nrow=2, common.legend = T, legend = "bottom", labels = "AUTO")

```

**β-diversity ordination plots comparing intervention b12_statuss at baseline and endpoint:** There were no differences in  community structure, nor dispersion of sequences by treatment group for any β-diversity metric measured (PERMDISP; p > 0.05).  However, overall, there was a small shift in community structure at endpoint compared to baseline (determined by PERMANOVA) for (A) Bray Curtis distances (F statistic: `r bray_adon$F[1]`; p = `r bray_adon$'Pr(>F)'[1]`), (B) Jaccard similarity (F statistic: `r jac_adon$F[1]`; p = `r jac_adon$'Pr(>F)'[1]`), (C) Unweighted UniFrac (F  statistic: `r uunifrac_adon$F[1]`; p = `r uunifrac_adon$'Pr(>F)'[1]`), and (D) Weighted UniFrac (F statistic: `r wunifrac_adon$F[1]`; p = `r wunifrac_adon$'Pr(>F)'[1]`).

:::

### Beta diversity summary tables

```{r}
#| eval: true
#| label: MS Table 2 cont.
#| tbl-cap: Beta diversity summary table
#| results: asis

dissimilarity_by_time_treatment <- list()

timepoints <- unique(sample_data(base_b12)$timepoint)

distance_methods <- c("bray", "jaccard", "uunifrac", "wunifrac")

# Calculate dissimilarity for each timepoint
for(t in timepoints) {
  index <- timepoints == t
  for (method in distance_methods) {
    # Print out current timepoint and distance method for debugging
    print(paste("Timepoint:", t, "Method:", method))

    # Subset the data by timepoint
    subset_physeq_time <- phyloseq::subset_samples(base_b12, index)
    
    # Get unique treatment groups within this timepoint
    treatments_in_timepoint <- unique(sample_data(subset_physeq_time)$b12_status)
    
    # Calculate Bray-Curtis dissimilarity for each treatment group within this timepoint
    for (treat in treatments_in_timepoint) {
      # Subset the data by treatment group within the timepoint
      subset_physeq_treatment <- phyloseq::subset_samples(subset_physeq_time, b12_status == treat) 
      
      # Calculate dissimilarity
      dist_subset <- distance(subset_physeq_treatment, method = method)
      
      # Store the dissimilarity matrix in the list
      dissimilarity_by_time_treatment[[paste(t, treat, method, sep = "_")]] <- as.numeric(dist_subset)
    }
  }
}

beta_df <- as.data.frame(do.call(cbind, dissimilarity_by_time_treatment))

beta_df$id <- seq_len(nrow(beta_df))

beta_df <- beta_df %>%
  pivot_longer(cols = -id,names_to = "variable",values_to = "value") %>%
  mutate(variable = str_remove_all(variable, "dissimilarity_by_time_treatment|\\.")) %>%
  separate(variable, into = c("timepoint", "b12_status", "metric"), sep = "_")%>%
  pivot_wider(names_from = metric) %>%
  dplyr::select(!c(id, timepoint))

saveRDS(beta_df, file.path(RESULTS, "B12-status-subgroup-analysis/base_b12-beta-div.rds"))

(p <- beta_df %>%  
        tbl_summary(
          by = b12_status,
          missing = "no", 
          statistic = list(
            all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
            all_categorical() ~ "{n} / {N}  \n({p}%)")) %>% # stats and format for categorical columns
      add_p(pvalue_fun = function(x) pvalue_format(x)) %>%
      add_q(pvalue_fun = function(x) pvalue_format(x)) %>%
      bold_p(q = T)%>%
		as_gt() %>% 
			fmt_markdown(columns=where(~is.character(.x))) %>%  
      fmt_number(columns = where(~is.numeric(.x)), drop_trailing_zeros = TRUE, use_seps = FALSE) %>% 
			fmt_number(columns = where(~is.numeric(.x)<0.0001), pattern = "<0.0001") %>% 
			fmt_number(columns = where(~is.numeric(.x)<0.001), pattern = "<0.001")%>% 
			fmt_number(columns = where(~is.numeric(.x)<0.01), pattern = "<0.01")%>% 
			fmt_number(columns = where(~is.numeric(.x)<0.05), pattern = "<0.05"))

htmltools::save_html(gt::as_raw_html(p), file.path(RESULTS, "B12-status-subgroup-analysis/baseline-beta-div.tbl"))

```

```{r}
#| eval: true
#| label: tbl-pairwise

tmp <- beta_df %>% 
            pivot_longer(cols=!b12_status, names_to="distance")

dunn <- tmp %>% 
          group_by(distance) %>%
          dunn_test(value ~ b12_status)

b12_vda <- data.frame()

for(p in unique(tmp$distance)) {
  tmp_vda <- tmp %>% dplyr::filter(distance==p) %>%
  calculate_pairwise_vda(., "value", "b12_status") %>% 
  mutate(distance=p) %>% 
  relocate(distance, .before=everything())

  b12_vda <- rbind(b12_vda,tmp_vda)
}

left_join(dunn, b12_vda, by =c("distance")) %>% kable()

```
