### B12 Status: endpoint Bacterial Relative Sums

```{r}
#| echo: false
#| eval: true
#| label: setup
#| include: false

# This initial code block loads custom utility functions and environment variables

# These are stored in separate files to keep the main analysis clean and organized
source(normalizePath("utils/r-utils.r"))
source(normalizePath("_environment.required"))
load(normalizePath("data/seqdata.RData"))

metadata <- read_delim(normalizePath(METADATA))

suppressPackageStartupMessages(library(emmeans))
suppressPackageStartupMessages(library(nlme))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ampvis2))

dir.create(file.path(RESULTS, "B12-status-subgroup-analysis"), showWarnings = FALSE, recursive = TRUE)
```

#### Data preparation

Relative abundances are calculated to show the proportion of each bacterial group in a sample.

Create an ampvis object from phyloseq object

```{r}
#| eval: true
#| label: create ampvis object

end_b12 <- phyloseq::subset_samples(seqdata_bac, !is.na(vitb12))
end_b12 <- phyloseq::subset_samples(end_b12, timepoint=="3Endpoint")

sample_data(end_b12) <- microbiome::meta(end_b12) %>% 
dplyr::mutate(
    sample_names = rownames(.),
    b12_status = case_when(
      vitb12 < 200 ~ "deficiency",
      vitb12 >= 200 & vitb12 <= 300 ~ "insufficiency",
      vitb12 >= 300 ~ "sufficiency",
      TRUE ~ "Unknown"
    ),
    sexf = case_when(
      sexf == "1" ~ "Female",
      sexf == "0" ~ "Male",
      TRUE ~ "Unknown"
    ),
    arm = case_when(
      trt == 1 ~ "FeZnCPM",
      trt == 0 ~ "CPM"
    )
  ) %>%
  relocate(sample_names, .before=everything())

b12_end_meta <- microbiome::meta(end_b12)

saveRDS(b12_end_meta, file.path(RESULTS, "B12-status-subgroup-analysis/b12-endpoint-metadata.rds"))

#slight modification to collapse Bacillota subgroups into one. As of Apr 2025 I cannot find any valid publication of this taxonomy. It appears to be based purely on the analysis of MAGs at GTDB
tax_table(end_b12) <- tax_df(end_b12) %>% 
                            mutate(across(where(is.character), str_trim, "both")) %>%
                            mutate(Phylum = str_remove(Phylum, "_[a-zA-Z]\\b")) %>%
                            mutate(Genus = str_remove(Genus, "_[a-zA-Z]\\b")) %>%
                            mutate(Species = str_remove(Species, "_[a-zA-Z]\\b")) %>%
                            mutate(Phylum = case_when(
                              Phylum == "" ~ "Unknown",
                              str_detect(Phylum, "\\d") ~ "Unclassified",
                              TRUE ~ Phylum)) %>%
                            mutate(Genus = case_when(
                              str_detect(Genus, "\\d") ~ "Unclassified",
                              Genus == "" ~ "Unknown",
                              TRUE ~ Genus)) %>%
                            as.matrix()

saveRDS(end_b12, file.path(RESULTS, "B12-status-subgroup-analysis/b12-endpoint-seqdata.rds"))

#Rows are OTU's, columns are samples
#metadata first column must contain sample ID's matching those in the otutable
d <- amp_load(otutable = t(otu_table(end_b12)), metadata = b12_end_meta, taxonomy = tax_table(end_b12))
```

#### Basic Characteristics of Subgroup

```{r}
#| eval: true
#| label: tbl-base-char
#| tbl-cap: Endpoint characterisitcs of subjects with vitamin B12 measurements available.

b12_end_meta %>% select(Sex = sexf, `Intervention Arm`= arm, `B12 Status` = b12_status, `B12 Concentration (pg/mL)` = vitb12) %>% tbl_summary() %>% as_kable()

```

```{r}
#| eval: true
#| label: tbl-base-char
#| tbl-cap: Endpoint characterisitcs by sex for subjects with vitamin B12 measurements available.

b12_end_meta %>% select(Sex = sexf, `Intervention Arm`= arm, `B12 Status` = b12_status, `B12 Concentration (pg/mL)` = vitb12) %>% tbl_summary(by=Sex) %>% as_kable()

```

#### endpoint Phyla Abundance

```{r}
#| eval: true
#| label: fig-end-b12-phyla-heatmap
#| fig-cap: Relative abundance of bacterial phyla at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL). Average group relative abundances are shown on the graph.

(heatmap <- amp_heatmap(d,
            group_by = "b12_status",
            tax_aggregate = "Phylum",
            tax_show = 20,
            plot_values = T,
            normalise = T, 
            plot_values_size = 8,
            round = 3,
            tax_empty="OTU",
            order_x_by = "cluster",
            order_y_by = "cluster") +
              theme(axis.text.x = element_text(angle = 45, size = 14, vjust = 1), axis.text.y = element_text(size = 14),  legend.position = "right", legend.title = element_text(color = "black", size = 10, face = "bold")))
```

```{r}
#| eval: true
#| label: tbl-b12-phyla-heatmap
#| tbl-cap: Relative abundance of bacterial phyla at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
heatmap$data %>% 
  group_by(Display, .Group) %>% 
  summarise(mean = round(mean(Sum), 4)) %>% 
  arrange(desc(mean)) %>%
  mutate(mean =  case_when(
    mean >= 0.01 ~ round(mean, 2), 
    TRUE ~ mean
  )) %>%
  mutate(mean = paste0(mean, "%")) %>%
  pivot_wider(names_from = .Group, values_from = mean) %>%
  kable()
```

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: Summary statistics of bacterial relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
df <- heatmap$data %>% 
  dplyr::select(phyla = Display, Sample , relAbund = Sum) %>% 
  mutate(timepoint = b12_end_meta$timepoint[match(Sample,b12_end_meta$sample_names)]) %>%
  mutate(b12_status = b12_end_meta$b12_status[match(Sample,b12_end_meta$sample_names)])
  
df %>%
  dplyr::select(phyla, b12_status, relAbund) %>%
  group_by(phyla, b12_status) %>%
  summarise(mean = mean(relAbund), median = median(relAbund), max = max(relAbund), min = min(relAbund))
```

```{r}
#| eval: true
#| label: fig-end-phyla-boxplot-base
#| fig-cap: Relative abundances of the top five most abundant bacterial phyla at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).

#default settings plot relative abundances (normalise = TRUE)
p <- amp_boxplot(d,
            normalise = TRUE,
            group_by = "b12_status",
            tax_aggregate = "Phylum",
            tax_show = 5,
            point_size = 1,
            plot_flip = T)

p + 
  scale_color_brewer(palette = "Set2")        
```

##### Statistical Comparison between B12 Status group Phyla Sum

::: h0

There is no differnece in the counts of bacterial phyla between vitamin B12 status groups at endpoint 

:::

::: h1

Bacterial phyla counts diverge between vitamin B12 status groups at endpoint

:::

```{r}
#| eval: true
#| label: tbl-b12-kw
#| tbl-cap: Kruskall-Wallis test results with computed $\eta^2_H$ effect size estimate for bacterial phyla relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
tmp <- amp_heatmap(d,
            group_by = "b12_status",
            tax_aggregate = "Phylum",
            tax_show = 20,
            plot_values = T,
            normalise = F, 
            tax_empty="OTU")

df <- tmp$data %>% dplyr::rename(phyla = Display, b12_status = `.Group`)

kw <- df %>% group_by(phyla) %>%
  kruskal_test(Sum ~ b12_status)

kw_eff<-df %>% group_by(phyla) %>% kruskal_effsize(Sum ~ b12_status)

left_join(kw,kw_eff, by="phyla") %>% kable()
```

```{r}
#| eval: true
#| label: tbl-b12-dunn-vda
#| tbl-cap: Kruskall-Wallis post hoc Dunn's test results with computed pairwise VDA effect size estimate for bacterial phyla relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
 
dunn <- df %>% group_by(phyla) %>%
  dunn_test(Sum ~ b12_status)

b12_vda <- data.frame()

for(p in unique(df$phyla)) {
  tmp_vda <- df %>% dplyr::filter(phyla==p) %>%
  calculate_pairwise_vda(., "Sum", "b12_status") %>% 
  mutate(phyla=p) %>% 
  relocate(phyla, .before=everything())

  b12_vda <- rbind(b12_vda,tmp_vda)
}

left_join(dunn, b12_vda, by =c("phyla", "group1", "group2")) %>% kable()
```

#### endpoint Genera Abundance

```{r}
#| eval: true
#| label: fig-end-b12-gen-heatmap
#| fig-cap: Relative abundance of top 20 bacterial genera at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL). Average group relative abundances are shown on the graph.

(heatmap <- amp_heatmap(d,
            group_by = "b12_status",
            tax_aggregate = "Genus",
            tax_show = 20,
            plot_values = T,
            normalise = T, 
            plot_values_size = 8,
            round = 3,
            tax_empty="OTU") +
              theme(axis.text.x = element_text(angle = 45, size = 14, vjust = 1), axis.text.y = element_text(size = 14, face="bold.italic"),  legend.position = "right", legend.title = element_text(color = "black", size = 10, face = "bold")))
```

```{r}
#| eval: true
#| label: tbl-b12-spp-heatmap
#| tbl-cap: Relative abundance of top 20 bacterial genera at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
heatmap$data %>% 
  group_by(Display, .Group) %>% 
  summarise(mean = round(mean(Sum), 4)) %>% 
  arrange(desc(mean)) %>%
  mutate(mean =  case_when(
    mean >= 0.01 ~ round(mean, 2), 
    TRUE ~ mean
  )) %>%
  mutate(mean = paste0(mean, "%")) %>%
  pivot_wider(names_from = .Group, values_from = mean) %>%
  kable()
```

###### Statistical Compariosn between B12 Status group  Genera Sum

::: h0

there is no differnece in count of genera between vitamin B12 status groups at endpoint 

:::

::: h1

genera count increases/decreases from vitamin B12 status groups at endpoint

:::

```{r}
#| eval: true
#| label: tbl-sum-gen
#| tbl-cap: Summary of counts

tmp <- amp_heatmap(d,
            group_by = "b12_status",
            tax_aggregate = "Genus",
            tax_show = 200,
            normalise = F, 
            tax_empty="OTU")

df <- tmp$data %>% dplyr::rename(genera = Display, b12_status = `.Group`) 

df %>%
  group_by(genera, b12_status) %>%
  summarise(mean = mean(Sum), median = median(Sum), max = max(Sum), min = min(Sum)) %>%
  arrange(desc(genera))
```

```{r}
#| eval: true
#| label: tbl-b12-kw
#| tbl-cap: Kruskall-Wallis test results with computed $\eta^2_H$ effect size estimate for bacterial phyla relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).

kw <- df %>% group_by(genera) %>%
  kruskal_test(Sum ~ b12_status)

kw_eff<-df %>% group_by(genera) %>% kruskal_effsize(Sum ~ b12_status)

left_join(kw,kw_eff, by="genera") %>% kable()
```

```{r}
#| eval: true
#| label: tbl-b12-dunn-vda
#| tbl-cap: Kruskall-Wallis post hoc Dunn's test results with computed pairwise VDA effect size estimate for bacterial genera relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).

kw_sig <- kw %>% filter(p<=0.05)

dunn <- df %>% filter(genera %in% kw_sig$genera) %>%
  group_by(genera) %>%
  dunn_test(Sum ~ b12_status)

b12_vda <- data.frame()

for(p in unique(df$genera)) {
  tmp_vda <- df %>% dplyr::filter(genera==p) %>%
  calculate_pairwise_vda(., "Sum", "b12_status") %>% 
  mutate(genera=p) %>% 
  relocate(genera, .before=everything())

  b12_vda <- rbind(b12_vda,tmp_vda)
}

left_join(dunn, b12_vda, by =c("genera", "group1", "group2")) %>% kable()
```

#### endpoint Species Abundance

```{r}
#| eval: true
#| label: fig-end-b12-spp-heatmap
#| fig-cap: Relative abundance of top 20 bacterial species at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL). Average group relative abundances are shown on the graph.

(heatmap <- amp_heatmap(d,
            group_by = "b12_status",
            tax_aggregate = "Species",
            tax_show = 20,
            plot_values = T,
            normalise = T, 
            plot_values_size = 8,
            round = 3,
            tax_empty="OTU") +
              theme(axis.text.x = element_text(angle = 45, size = 14, vjust = 1), axis.text.y = element_text(size = 14, face="bold.italic"),  legend.position = "right", legend.title = element_text(color = "black", size = 10, face = "bold")))
```

```{r}
#| eval: true
#| label: tbl-b12-spp-heatmap
#| tbl-cap: Relative abundance of top 20 bacterial species at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
heatmap$data %>% 
  group_by(Display, .Group) %>% 
  summarise(mean = round(mean(Sum), 4)) %>% 
  arrange(desc(mean)) %>%
  mutate(mean =  case_when(
    mean >= 0.01 ~ round(mean, 2), 
    TRUE ~ mean
  )) %>%
  mutate(mean = paste0(mean, "%")) %>%
  pivot_wider(names_from = .Group, values_from = mean) %>%
  kable()
```

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: Summary statistics of bacterial relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
df <- heatmap$data %>% 
  dplyr::select(species = Display, Sample , relAbund = Sum) %>% 
  mutate(timepoint = b12_end_meta$timepoint[match(Sample,b12_end_meta$sample_names)]) %>%
  mutate(b12_status = b12_end_meta$b12_status[match(Sample,b12_end_meta$sample_names)])

df %>%
  dplyr::select(species, b12_status, relAbund) %>%
  group_by(species, b12_status) %>%
  summarise(mean = mean(relAbund), median = median(relAbund), max = max(relAbund), min = min(relAbund))
```

###### Statistical Compariosn between B12 Status group  Species Sum

::: h0

there is no differnece in realtive abundance of species between vitamin B12 status groups at endpoint 

:::

::: h1

species realtive abundance increases/decreases from vitamin B12 status groups at endpoint

:::

```{r}
#| eval: true
#| label: tbl-sum-gen
#| tbl-cap: Summary of counts

tmp <- amp_heatmap(d,
            group_by = "b12_status",
            tax_aggregate = "Species",
            tax_show = 200,
            normalise = F, 
            tax_empty="OTU")

df <- tmp$data %>% dplyr::rename(species = Display, b12_status = `.Group`) 

df %>%
  group_by(species, b12_status) %>%
  summarise(mean = mean(Sum), median = median(Sum), max = max(Sum), min = min(Sum)) %>%
  arrange(desc(species)) %>% kable()
```

```{r}
#| eval: true
#| label: tbl-b12-kw
#| tbl-cap: Kruskall-Wallis test results with computed $\eta^2_H$ effect size estimate for bacterial phyla relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).

kw <- df %>% group_by(species) %>%
  kruskal_test(Sum ~ b12_status)

kw_eff<-df %>% group_by(species) %>% kruskal_effsize(Sum ~ b12_status)

left_join(kw,kw_eff, by="species") %>% kable()
```

```{r}
#| eval: true
#| label: tbl-b12-dunn-vda
#| tbl-cap: Kruskall-Wallis post hoc Dunn's test results with computed pairwise VDA effect size estimate for bacterial species relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).

kw_sig <- kw %>% filter(p<=0.05)

dunn <- df %>% filter(species %in% kw_sig$species) %>%
  group_by(species) %>%
  dunn_test(Sum ~ b12_status)

b12_vda <- data.frame()

for(p in unique(df$species)) {
  tmp_vda <- df %>% dplyr::filter(species==p) %>%
  calculate_pairwise_vda(., "Sum", "b12_status") %>% 
  mutate(species=p) %>% 
  relocate(species, .before=everything())

  b12_vda <- rbind(b12_vda,tmp_vda)
}

kable(dunn_spp_results <- left_join(dunn, b12_vda, by =c("species", "group1", "group2")))
```

```{r}
#| eval: true
#| label: fig-end-heatmap
#| fig-cap: Heatmap of phyla and species differences between 

ggplot(df, aes(x = b12_status, y = Sum)) +
  # Add boxplot
  geom_boxplot(aes(fill = b12_status), alpha = 0.5) +
  # Add points with size reflecting abundance
  geom_point(aes(size = Abundance), 
             position = position_jitterdodge(),
             alpha = 0.6) +
  # Facet by species
  facet_wrap(~species, scales = "free_y") +
  # Add significance levels from Dunn's test
  geom_text(data = dunn_spp_results, 
            aes(x = group1, y = max(df$Sum) * 1.1,
                label = p.adj.signif),
            size = 3) +
  # Customize appearance
  scale_fill_manual(values = c(
    "deficiency" = "#e78284",
    "insufficiency" = "#f4b8e4",
    "sufficiency" = "#8CAAEE"
  )) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  ) +
  labs(
    x = "B12 Status",
    y = "Sum",
    size = "Abundance",
    fill = "B12 Status",
    title = "Species Abundance by B12 Status",
    subtitle = "Point size indicates abundance; ns: not significant"
  )

#Bubble plot with significance connections
abundance_summary <- count_data %>%
  group_by(species, b12_status) %>%
  summarise(
    mean_abundance = mean(Abundance),
    mean_sum = mean(Sum)
  )
  
ggplot(abundance_summary, 
       aes(x = b12_status, y = species)) +
  # Add bubbles
  geom_point(aes(size = mean_sum, fill = mean_abundance),
             shape = 21, alpha = 0.7) +
  # Add significance lines
  geom_segment(data = dunns_results,
               aes(x = group1, xend = group2,
                   y = species, yend = species,
                   linetype = p.adj.signif),
               position = position_dodge(width = 0.9)) +
  scale_size_continuous(range = c(2, 20)) +
  scale_fill_gradient(
    low = "#8CAAEE",
    high = "#f4b8e4"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
