## B12 Status: Baseline Beta Diversity

This analysis examines how bacterial communities differ between samples, using multiple distance metrics to capture different aspects of community similarity.

Four different distance metrics were calculated:

1. *Bray-Curtis*: Considers abundance data, sensitive to dominant species
2. *Jaccard*: Based on presence/absence, treats all species equally
3. *Unweighted UniFrac*: Incorporates phylogenetic relationships, presence/absence
4. *Weighted UniFrac*: Combines phylogenetic relationships with abundance data

For each metric, the following is performed:

- Ordination (NMDS and MDS)
- Statistical testing (PERMANOVA, ANOSIM)
- Effect size calculations
- Pairwise comparisons

```{r}
#| echo: false
#| eval: true
#| label: setup
#| include: false

# This initial code block loads custom utility functions and environment variables

# These are stored in separate files to keep the main analysis clean and organized
source(normalizePath("utils/r-utils.r"))

# Source beta diversity analysis functions
source("scripts/beta-diversity-analysis.r")

source(normalizePath("_environment.required"))
load(normalizePath("data/seqdata.RData"))

b12_base <- readRDS(file.path(RESULTS, "B12-status-subgroup-analysis/b12-baseline-seqdata.rds"))

suppressPackageStartupMessages(library(vegan))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(pairwiseAdonis))

dir.create(file.path(RESULTS, "B12-status-subgroup-analysis"), showWarnings = FALSE, recursive = TRUE)
```

### B12 Tertiles

```{r}
#| eval: true
#| label: params

phyloseq_object <- b12_base
grouping_var <- "b12_tertiles"

```

#### Run Beta Diversity Pipeline

```{r}
#| eval: true
#| label: beta-diversity-analysis
beta_results <- beta_main(
  phyloseq_object = phyloseq_object,
  group_var = grouping_var,
  plot = FALSE, # Plots generated later
  debug = TRUE # Set to FALSE to suppress debug messages
)
```

#### Statistial Analysis Results: Differences between B12 Status Groups

Three complementary approaches were used to test for community differences:

::: {.grid}

::: {.g-col-4}

**ANOSIM (Analysis of Similarity)**

Tests whether between-group differences are larger than within-group differences

- R statistic ranges from -1 to 1
- R ≈ 0: no difference between groups
- R > 0: groups are more different than expected by chance
- Stratification by timepoint accounts for temporal effects

:::

::: {.g-col-4}

**PERMDISP (Permutational Analysis of Multivariate Dispersions)**

Tests for homogeneity of dispersions between groups

- Important assumption check for PERMANOVA
- Analogous to Levene's test for univariate data
- Significant results could indicate different within-group variability

:::

::: {.g-col-4}

**PERMANOVA (Permutational Multivariate Analysis of Variance)**

Tests for differences in community centroids between groups

- Accounts for both treatment and time effects
- More robust than ANOSIM
- Effect sizes calculated using partial eta-squared and omega-squared
- Requires homogeneous dispersions (tested by PERMDISP)

:::

:::

##### ANOSIM: Analysis of Similarity

ANOSIM is a permutation test which calculates an R statistic (-1,1) $\approx$ correlation coefficient $\approx$ effect size, where:

- R ≈ 0: Groups are not different
- R > 0: Groups are different
- R = 1: All samples within groups are more similar to each other than to any samples from other groups


::: h0

there is no difference between intervention b12_statuss in terms of beta diversity metrics ($R = 0$)

:::

::: h0

biofortified pearl millet differentlaly changes beta diversity metrics ($R \neq 0$)

:::

```{r}
#| eval: true
#| label: tbl-anosim
#| tbl-cap: "ANOSIM Results - Testing for differences between B12 tertiles"

anosim_summary <- lapply(beta_results$analysis_results$anosim, function(x) {
  data.frame(
    R_statistic = x$statistic,
    p_value = x$signif,
    Number_of_permutations = x$permutations
  )
}) %>% 
  bind_rows(.id = "Distance_metric")

knitr::kable(anosim_summary,
             digits = 3)
```

##### PERMDISP

::: {.callout-tip}
PERMDISP tests the homogeneity of dispersion among groups. A significant result (p < 0.05) indicates that groups have different levels of dispersion (spread) from their centroids, which could affect the interpretation of PERMANOVA results.
:::

PERMANOVA assumes homogeneity of multivariate dispersions across groups therefore a significant PERMANOVA result could be due to dispersion, location or both many permutational techniqures can be sensitive to dispersion (differences in centroids and/or differences in how data points in clusters scatter around their centroid). In univariate analysis Levene's test can test for this. PERMDISP is an extension of this that can be used in multivariate analysis technique involves calculating the distance from each data point to its group centroid and then testing whether those distances differ among the groups however, certain metrics are non-metric or semi-metric e.g. Bray-Curtis and centroids cannot be directly calculated -> PCoA + cantroid calc on the coordinates ([cite](https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/permdisp/))

```{r}
#| eval: true
#| label: tbl-permdisp
#| tbl-cap: "PERMDISP Results - Testing homogeneity of group dispersions"

permdisp_summary <- lapply(beta_results$analysis_results$permdisp, function(x) {
  if(!is.null(x)) {
    data.frame(
      F_statistic = x$anova$F[1],
      Df = paste(x$anova$Df[1], x$anova$Df[2], sep = ", "),
      p_value = x$anova$`Pr(>F)`[1]
    )
  }
}) %>% 
  bind_rows(.id = "Distance_metric")

knitr::kable(permdisp_summary,
             digits = 3)

```

```{r}
#| eval: true
#| label: tbl-permdisp-means
#| tbl-cap: "Average distances to group centroids"

# Display group-specific dispersions
permdisp_means <- lapply(names(beta_results$analysis_results$permdisp), function(method) {
  if(!is.null(beta_results$analysis_results$permdisp[[method]])) {
    data.frame(
      Distance_metric = method,
      Group = names(beta_results$analysis_results$permdisp[[method]]$disp$group.distances),
      Average_distance_to_centroid = beta_results$analysis_results$permdisp[[method]]$disp$group.distances
    )
  }
}) %>% 
  bind_rows()

knitr::kable(permdisp_means,
             digits = 3)
```

##### PERMANOVA

::: h0

there is no change in beta diversity metrics arosss time (baseline to endpoint)
:::

::: h1

beta diversity metrics change (increase) across time (baseline to endpoint)
:::

assumption: variance of dispersions are homogenous across groups

```{r}
#| eval: true
#| label: tbl-permanova
#| tbl-cap: PERMANOVA results table

# Format and display PERMANOVA results
permanova_summary <- lapply(beta_results$analysis_results$permanova, function(x) {
  data.frame(
    R2 = x$R2[1],
    F_stat = x$F[1],
    p_value = x$`Pr(>F)`[1]
  )
}) %>% 
  bind_rows(.id = "Distance_metric")

knitr::kable(permanova_summary,
             digits = 3)
```


##### Effect Size Calculations

###### Eta-Squared

PERMANOVA effect size is typically measured using the partial eta-squared ($\eta^2_p$) statistic. The partial eta-squared is a measure of the proportion of total variance in the multivariate data explained by the grouping variable (i.e., the factor of interest). In other words, $\eta^2_p$ calculations assess the magnitude of community differences.

::: {.callout-warn}

This measure tends to overestimate effect sizes in small samples

:::

$$
(\eta^2_p = \frac{SS_{\text{between}}}{SS_{\text{total}}})
$$

- Partial eta-squared ranges from 0 to 1.
- Values < 0.6 indicate small effect sizes
- A value of 0 indicates that the grouping variable explains none of the variance in the dependent variable.
- A value of 1 indicates that the grouping variable explains all of the variance in the dependent variable.
- Can be interpreted similarly to $R^2$ in linear regression
- In the context of PERMANOVA, $\eta^2_p$ indicates the proportion of total variance in the multivariate data (e.g., beta diversity distances) that is explained by the grouping variable.

###### Omega-squared

Omega-squared ($\omega^2$) provides a less biased measure of effect size for ANOVA-type analyses by accounting for the mean-squared error of the observed samples [@Olejnik2009]

$$
\omega^2 = \frac{SS_A - (a-1)\frac{SS_W}{(N-a)}}{SS_T +\frac{SS_W}{(N-a)}}
$$

Where:

- $SS_T$: total sum of squares
- $SS_W$: within-group sum of squares (sum of the squares of distances within groups divided by the number of subjects per group)
- $SS_A$: between-group sum of squares ($SS_A =SS_T − SS_W$)
- $(a − 1)$: degress of freedom of the grouping factor
- $(N - a)$: residual degrees of freedom

```{r}
#| eval: true
#| label: tbl-effect-size
#| tbl-cap: "Omega and Eta-squared effect size calculations for PERMANOVA results"

effect_sizes <- data.frame(
  Distance_metric = names(beta_results$analysis_results$effect_sizes$eta_squared),
  Eta_squared = unlist(beta_results$analysis_results$effect_sizes$eta_squared),
  Omega_squared = unlist(beta_results$analysis_results$effect_sizes$omega_squared)
)

knitr::kable(effect_sizes,
             digits = 3)
```

#### Ordination Plots with PERMANOVA $R^2$ and P-values

Two different ordination approaches were used to visualize community differences: MDS and NMDS.

##### Metric Multidimensional Scaling (MDS)

MDS creates a visualization where:

- Similar samples are placed closer together
- Distances between points approximate their actual dissimilarities
- `trymax = 100` ensures a stable solution by trying multiple starting configurations

::: {#fig-box}

```{r}
#| eval: true
#| label: fig-base-beta-mds
#| fig.width: 18
#| fig.cap: "Metric Multidimensional Scaling (MDS) ordination plots showing beta diversity patterns across B12 tertiles using different distance metrics"

# Generate ordination plots
plots <- plot_beta_diversity(
  results = beta_results$analysis_results,
  physeq = phyloseq_object,
  group_var = grouping_var,
  ordination_type = "MDS"
)

# Arrange and display plots
do.call(
  ggarrange, 
  c(plots, 
    list(ncol = 2, 
      nrow = 2, 
      common.legend = TRUE, 
      legend = "bottom", 
      labels = "AUTO")))
```

**β-diversity ordination plots comparing intervention b12_statuss at baseline and endpoint:** There were no differences in  community structure, nor dispersion of sequences by treatment group for any β-diversity metric measured (PERMDISP; p > 0.05).  However, overall, there was a small shift in community structure at endpoint compared to baseline (determined by PERMANOVA) for (A) Bray Curtis distances (F statistic: `r beta_results$analysis_results$permanova$bray$F[1]`; p = `r beta_results$analysis_results$permanova$bray$'Pr(>F)'[1]`), (B) Jaccard similarity (F statistic: `r beta_results$analysis_results$permanova$jaccard$F[1]`; p = `r beta_results$analysis_results$permanova$jaccard$'Pr(>F)'[1]`), (C) Unweighted UniFrac (F  statistic: `r beta_results$analysis_results$permanova$uunifrac$F[1]`; p = `r beta_results$analysis_results$permanova$uunifrac$'Pr(>F)'[1]`), and (D) Weighted UniFrac (F statistic: `r beta_results$analysis_results$permanova$wunifrac$F[1]`; p = `r beta_results$analysis_results$permanova$wunifrac$'Pr(>F)'[1]`).

:::

##### Non-metric Multidimensional Scaling (NMDS)

NMDS is an iterative rank based procedure that sets stress values rather than eigenvalues to determine the "success" of the ordination. It is typically to aim for stress values $\leq$ 0.1, which indicates a good representation of the relationship between values.

NMDS is often preferred for ecological data because:

- It works with rank-order distances rather than absolute distances
- More robust to non-linear relationships
- Stress value indicates how well the ordination represents the data

::: {fig-box}

```{r}
#| eval: true
#| label: fig-base-beta-nmds
#| fig.width: 18
#| fig.cap: "Non-metric Multidimensional Scaling (NMDS) ordination plots showing beta diversity patterns across B12 tertiles using different distance metrics"

# Generate ordination plots
plots <- plot_beta_diversity(
  results = beta_results$analysis_results,
  physeq = phyloseq_object,
  group_var = grouping_var,
  ordination_type = "NMDS"
)

# Arrange and display plots
do.call(
  ggarrange, 
  c(plots, 
    list(ncol = 2, 
      nrow = 2, 
      common.legend = TRUE, 
      legend = "bottom", 
      labels = "AUTO")))
```

**β-diversity ordination plots comparing intervention b12_statuss at baseline and endpoint:** There were no differences in  community structure, nor dispersion of sequences by treatment group for any β-diversity metric measured (PERMDISP; p > 0.05).  However, overall, there was a small shift in community structure at endpoint compared to baseline (determined by PERMANOVA) for (A) Bray Curtis distances (F statistic: `r beta_results$analysis_results$permanova$bray$F[1]`; p = `r beta_results$analysis_results$permanova$bray$'Pr(>F)'[1]`), (B) Jaccard similarity (F statistic: `r beta_results$analysis_results$permanova$jaccard$F[1]`; p = `r beta_results$analysis_results$permanova$jaccard$'Pr(>F)'[1]`), (C) Unweighted UniFrac (F  statistic: `r beta_results$analysis_results$permanova$uunifrac$F[1]`; p = `r beta_results$analysis_results$permanova$uunifrac$'Pr(>F)'[1]`), and (D) Weighted UniFrac (F statistic: `r beta_results$analysis_results$permanova$wunifrac$F[1]`; p = `r beta_results$analysis_results$permanova$wunifrac$'Pr(>F)'[1]`).

:::

#### Beta Diversity Summary Tables

```{r}
#| eval: true
#| label: tbl-ms-table2-base-beta
#| tbl-cap: "Beta diversity summary statistics by B12 tertile"
#| results: asis

if(!is.null(beta_results$beta_summary)) {
  knitr::kable(beta_results$beta_summary,
               digits = 3)
}

```

```{r}
#| eval: true
#| label: tbl-beta-sum
#| tbl-cap: Beta diversity summary table

# Create summary tables
beta_df <- beta_results$analysis_results$dissimilarity %>%
  dplyr::mutate(id=seq_len(nrow(.))) %>%
  pivot_longer(cols = -id,names_to = "variable",values_to = "value") %>%
  dplyr::mutate(variable = str_remove_all(variable, "dissimilarity_by_time_treatment|\\.")) %>%
  separate(variable, into = c("timepoint", "b12_status", "metric"), sep = "_")%>%
  pivot_wider(names_from = metric) %>%
  dplyr::select(!c(id, timepoint))

saveRDS(beta_df, file.path(RESULTS, "B12-status-subgroup-analysis", paste0(grouping_var, "-b12_base-beta-div.rds")))

(p <- beta_df %>% gt_summary(by = "b12_status"))

htmltools::save_html(gt::as_raw_html(p), file.path(RESULTS, "B12-status-subgroup-analysis", paste0(grouping_var, "-baseline-beta-div.tbl")))
```

##### Pairwise Comparisons

```{r}
#| eval: true
#| label: tbl-dunn
#| fig-cap: NA
#| tbl-cap: "Pairwise Comparisons Between B12 Tertiles (Dunn's Test)"

if(!is.null(beta_results$dunn_results)) {

  tmp <- beta_df %>% 
      pivot_longer(cols=!b12_status, names_to="metric")

  b12_vda <- data.frame()

  for(p in unique(tmp$metric)) {
    tmp_vda <- tmp %>% dplyr::filter(metric==p) %>%
    calculate_pairwise_vda(., "value", "b12_status") %>% 
    dplyr::mutate(metric=p) %>% 
    relocate(metric, .before=everything())

    b12_vda <- rbind(b12_vda,tmp_vda)
  }

  left_join(beta_results$dunn_results, b12_vda) %>% kable()
}
```

### Binary level varaible: Deficiency (<148 pmol/L)

```{r}
#| eval: true
#| label: params

phyloseq_object <- b12_base
grouping_var <- "b12_2Lev"

```

#### Run Beta Diversity Pipeline

```{r}
#| eval: true
#| label: beta-diversity-analysis
beta_results <- beta_main(
  phyloseq_object = phyloseq_object,
  group_var = grouping_var,
  plot = FALSE, # Plots generated later
  debug = FALSE # Set to FALSE to suppress debug messages
)
```

#### Statistial Analysis Results: Differences between B12 Status Groups

##### ANOSIM: Analysis of Similarity

```{r}
#| eval: true
#| label: tbl-anosim
#| tbl-cap: "ANOSIM Results - Testing for differences between B12 tertiles"

anosim_summary <- lapply(beta_results$analysis_results$anosim, function(x) {
  data.frame(
    R_statistic = x$statistic,
    p_value = x$signif,
    Number_of_permutations = x$permutations
  )
}) %>% 
  bind_rows(.id = "Distance_metric")

knitr::kable(anosim_summary,
             digits = 3)
```

##### PERMDISP

```{r}
#| eval: true
#| label: tbl-permdisp
#| tbl-cap: "PERMDISP Results - Testing homogeneity of group dispersions"

permdisp_summary <- lapply(beta_results$analysis_results$permdisp, function(x) {
  if(!is.null(x)) {
    data.frame(
      F_statistic = x$anova$F[1],
      Df = paste(x$anova$Df[1], x$anova$Df[2], sep = ", "),
      p_value = x$anova$`Pr(>F)`[1]
    )
  }
}) %>% 
  bind_rows(.id = "Distance_metric")

knitr::kable(permdisp_summary,
             digits = 3)

```

```{r}
#| eval: true
#| label: tbl-permdisp-means
#| tbl-cap: "Average distances to group centroids"

# Display group-specific dispersions
permdisp_means <- lapply(names(beta_results$analysis_results$permdisp), function(method) {
  if(!is.null(beta_results$analysis_results$permdisp[[method]])) {
    data.frame(
      Distance_metric = method,
      Group = names(beta_results$analysis_results$permdisp[[method]]$disp$group.distances),
      Average_distance_to_centroid = beta_results$analysis_results$permdisp[[method]]$disp$group.distances
    )
  }
}) %>% 
  bind_rows()

knitr::kable(permdisp_means,
             digits = 3)
```

##### PERMANOVA

```{r}
#| eval: true
#| label: tbl-permanova
#| tbl-cap: PERMANOVA results table

# Format and display PERMANOVA results
permanova_summary <- lapply(beta_results$analysis_results$permanova, function(x) {
  data.frame(
    R2 = x$R2[1],
    F_stat = x$F[1],
    p_value = x$`Pr(>F)`[1]
  )
}) %>% 
  bind_rows(.id = "Distance_metric")

knitr::kable(permanova_summary,
             digits = 3)
```

##### Effect Size Calculations

```{r}
#| eval: true
#| label: tbl-effect-size
#| tbl-cap: "Omega and Eta-squared effect size calculations for PERMANOVA results"

effect_sizes <- data.frame(
  Distance_metric = names(beta_results$analysis_results$effect_sizes$eta_squared),
  Eta_squared = unlist(beta_results$analysis_results$effect_sizes$eta_squared),
  Omega_squared = unlist(beta_results$analysis_results$effect_sizes$omega_squared)
)

knitr::kable(effect_sizes,
             digits = 3)
```

#### Ordination Plots with PERMANOVA $R^2$ and P-values

##### Metric Multidimensional Scaling (MDS)

::: {#fig-box}

```{r}
#| eval: true
#| label: fig-base-beta-mds
#| fig.width: 16
#| fig.cap: "Metric Multidimensional Scaling (MDS) ordination plots showing beta diversity patterns across B12 tertiles using different distance metrics"

# Generate ordination plots
plots <- plot_beta_diversity(
  results = beta_results$analysis_results,
  physeq = phyloseq_object,
  group_var = grouping_var,
  ordination_type = "MDS"
)

# Arrange and display plots
do.call(
  ggarrange, 
  c(plots, 
    list(ncol = 2, 
      nrow = 2, 
      common.legend = TRUE, 
      legend = "bottom", 
      labels = "AUTO")))
```

**β-diversity ordination plots comparing intervention b12_statuss at baseline and endpoint:** There were no differences in  community structure, nor dispersion of sequences by treatment group for any β-diversity metric measured (PERMDISP; p > 0.05).  However, overall, there was a small shift in community structure at endpoint compared to baseline (determined by PERMANOVA) for (A) Bray Curtis distances (F statistic: `r beta_results$analysis_results$permanova$bray$F[1]`; p = `r beta_results$analysis_results$permanova$bray$'Pr(>F)'[1]`), (B) Jaccard similarity (F statistic: `r beta_results$analysis_results$permanova$jaccard$F[1]`; p = `r beta_results$analysis_results$permanova$jaccard$'Pr(>F)'[1]`), (C) Unweighted UniFrac (F  statistic: `r beta_results$analysis_results$permanova$uunifrac$F[1]`; p = `r beta_results$analysis_results$permanova$uunifrac$'Pr(>F)'[1]`), and (D) Weighted UniFrac (F statistic: `r beta_results$analysis_results$permanova$wunifrac$F[1]`; p = `r beta_results$analysis_results$permanova$wunifrac$'Pr(>F)'[1]`).

:::

##### Non-metric Multidimensional Scaling (NMDS)

::: {fig-box}

```{r}
#| eval: true
#| label: fig-base-beta-nmds
#| fig.width: 16
#| fig.cap: "Non-metric Multidimensional Scaling (NMDS) ordination plots showing beta diversity patterns across B12 tertiles using different distance metrics"

# Generate ordination plots
plots <- plot_beta_diversity(
  results = beta_results$analysis_results,
  physeq = phyloseq_object,
  group_var = grouping_var,
  ordination_type = "NMDS"
)

# Arrange and display plots
do.call(
  ggarrange, 
  c(plots, 
    list(ncol = 2, 
      nrow = 2, 
      common.legend = TRUE, 
      legend = "bottom", 
      labels = "AUTO")))
```

**β-diversity ordination plots comparing intervention b12_statuss at baseline and endpoint:** There were no differences in  community structure, nor dispersion of sequences by treatment group for any β-diversity metric measured (PERMDISP; p > 0.05).  However, overall, there was a small shift in community structure at endpoint compared to baseline (determined by PERMANOVA) for (A) Bray Curtis distances (F statistic: `r beta_results$analysis_results$permanova$bray$F[1]`; p = `r beta_results$analysis_results$permanova$bray$'Pr(>F)'[1]`), (B) Jaccard similarity (F statistic: `r beta_results$analysis_results$permanova$jaccard$F[1]`; p = `r beta_results$analysis_results$permanova$jaccard$'Pr(>F)'[1]`), (C) Unweighted UniFrac (F  statistic: `r beta_results$analysis_results$permanova$uunifrac$F[1]`; p = `r beta_results$analysis_results$permanova$uunifrac$'Pr(>F)'[1]`), and (D) Weighted UniFrac (F statistic: `r beta_results$analysis_results$permanova$wunifrac$F[1]`; p = `r beta_results$analysis_results$permanova$wunifrac$'Pr(>F)'[1]`).

:::

#### Beta Diversity Summary Tables

```{r}
#| eval: true
#| label: tbl-ms-table2-base-beta
#| tbl-cap: "Beta diversity summary statistics by B12 tertile"
#| results: asis

if(!is.null(beta_results$beta_summary)) {
  knitr::kable(beta_results$beta_summary,
               digits = 3)
}

```

```{r}
#| eval: true
#| label: tbl-beta-sum
#| tbl-cap: Beta diversity summary table

# Create summary tables
beta_df <- beta_results$analysis_results$dissimilarity %>%
  dplyr::mutate(id=seq_len(nrow(.))) %>%
  pivot_longer(cols = -id,names_to = "variable",values_to = "value") %>%
  dplyr::mutate(variable = str_remove_all(variable, "dissimilarity_by_time_treatment|\\.")) %>%
  separate(variable, into = c("timepoint", "b12_status", "metric"), sep = "_")%>%
  pivot_wider(names_from = metric) %>%
  dplyr::select(!c(id, timepoint))

saveRDS(beta_df, file.path(RESULTS, "B12-status-subgroup-analysis", paste0(grouping_var, "-b12_base-beta-div.rds")))

(p <- beta_df %>% gt_summary(by = "b12_status"))

htmltools::save_html(gt::as_raw_html(p), file.path(RESULTS, "B12-status-subgroup-analysis", paste0(grouping_var, "-baseline-beta-div.tbl")))
```

##### Pairwise Comparisons

```{r}
#| eval: true
#| label: tbl-dunn
#| fig-cap: NA
#| tbl-cap: "Pairwise Comparisons Between B12 Tertiles (Dunn's Test)"

if(!is.null(beta_results$dunn_results)) {

  tmp <- beta_df %>% 
      pivot_longer(cols=!b12_status, names_to="metric")

  b12_vda <- data.frame()

  for(p in unique(tmp$metric)) {
    tmp_vda <- tmp %>% dplyr::filter(metric==p) %>%
    calculate_pairwise_vda(., "value", "b12_status") %>% 
    dplyr::mutate(metric=p) %>% 
    relocate(metric, .before=everything())

    b12_vda <- rbind(b12_vda,tmp_vda)
  }

  left_join(beta_results$dunn_results, b12_vda) %>% kable()
}
```


### Three level variable: Insufficiency (≥148 & <221) | 34 (26.8%)

```{r}
#| eval: true
#| label: params

phyloseq_object <- b12_base
grouping_var <- "b12_2Lev"

```

#### Run Beta Diversity Pipeline

```{r}
#| eval: true
#| label: beta-diversity-analysis
beta_results <- beta_main(
  phyloseq_object = phyloseq_object,
  group_var = grouping_var,
  plot = FALSE, # Plots generated later
  debug = FALSE # Set to FALSE to suppress debug messages
)
```

#### Statistial Analysis Results: Differences between B12 Status Groups

##### ANOSIM: Analysis of Similarity

```{r}
#| eval: true
#| label: tbl-anosim
#| tbl-cap: "ANOSIM Results - Testing for differences between B12 tertiles"

anosim_summary <- lapply(beta_results$analysis_results$anosim, function(x) {
  data.frame(
    R_statistic = x$statistic,
    p_value = x$signif,
    Number_of_permutations = x$permutations
  )
}) %>% 
  bind_rows(.id = "Distance_metric")

knitr::kable(anosim_summary,
             digits = 3)
```

##### PERMDISP

```{r}
#| eval: true
#| label: tbl-permdisp
#| tbl-cap: "PERMDISP Results - Testing homogeneity of group dispersions"

permdisp_summary <- lapply(beta_results$analysis_results$permdisp, function(x) {
  if(!is.null(x)) {
    data.frame(
      F_statistic = x$anova$F[1],
      Df = paste(x$anova$Df[1], x$anova$Df[2], sep = ", "),
      p_value = x$anova$`Pr(>F)`[1]
    )
  }
}) %>% 
  bind_rows(.id = "Distance_metric")

knitr::kable(permdisp_summary,
             digits = 3)

```

```{r}
#| eval: true
#| label: tbl-permdisp-means
#| tbl-cap: "Average distances to group centroids"

# Display group-specific dispersions
permdisp_means <- lapply(names(beta_results$analysis_results$permdisp), function(method) {
  if(!is.null(beta_results$analysis_results$permdisp[[method]])) {
    data.frame(
      Distance_metric = method,
      Group = names(beta_results$analysis_results$permdisp[[method]]$disp$group.distances),
      Average_distance_to_centroid = beta_results$analysis_results$permdisp[[method]]$disp$group.distances
    )
  }
}) %>% 
  bind_rows()

knitr::kable(permdisp_means,
             digits = 3)
```

##### PERMANOVA

```{r}
#| eval: true
#| label: tbl-permanova
#| tbl-cap: PERMANOVA results table

# Format and display PERMANOVA results
permanova_summary <- lapply(beta_results$analysis_results$permanova, function(x) {
  data.frame(
    R2 = x$R2[1],
    F_stat = x$F[1],
    p_value = x$`Pr(>F)`[1]
  )
}) %>% 
  bind_rows(.id = "Distance_metric")

knitr::kable(permanova_summary,
             digits = 3)
```

##### Effect Size Calculations

```{r}
#| eval: true
#| label: tbl-effect-size
#| tbl-cap: "Omega and Eta-squared effect size calculations for PERMANOVA results"

effect_sizes <- data.frame(
  Distance_metric = names(beta_results$analysis_results$effect_sizes$eta_squared),
  Eta_squared = unlist(beta_results$analysis_results$effect_sizes$eta_squared),
  Omega_squared = unlist(beta_results$analysis_results$effect_sizes$omega_squared)
)

knitr::kable(effect_sizes,
             digits = 3)
```

#### Ordination Plots with PERMANOVA $R^2$ and P-values

##### Metric Multidimensional Scaling (MDS)

::: {#fig-box}

```{r}
#| eval: true
#| label: fig-base-beta-mds
#| fig.width: 16
#| fig.cap: "Metric Multidimensional Scaling (MDS) ordination plots showing beta diversity patterns across B12 tertiles using different distance metrics"

# Generate ordination plots
plots <- plot_beta_diversity(
  results = beta_results$analysis_results,
  physeq = phyloseq_object,
  group_var = grouping_var,
  ordination_type = "MDS"
)

# Arrange and display plots
do.call(
  ggarrange, 
  c(plots, 
    list(ncol = 2, 
      nrow = 2, 
      common.legend = TRUE, 
      legend = "bottom", 
      labels = "AUTO")))
```

**β-diversity ordination plots comparing intervention b12_statuss at baseline and endpoint:** There were no differences in  community structure, nor dispersion of sequences by treatment group for any β-diversity metric measured (PERMDISP; p > 0.05).  However, overall, there was a small shift in community structure at endpoint compared to baseline (determined by PERMANOVA) for (A) Bray Curtis distances (F statistic: `r beta_results$analysis_results$permanova$bray$F[1]`; p = `r beta_results$analysis_results$permanova$bray$'Pr(>F)'[1]`), (B) Jaccard similarity (F statistic: `r beta_results$analysis_results$permanova$jaccard$F[1]`; p = `r beta_results$analysis_results$permanova$jaccard$'Pr(>F)'[1]`), (C) Unweighted UniFrac (F  statistic: `r beta_results$analysis_results$permanova$uunifrac$F[1]`; p = `r beta_results$analysis_results$permanova$uunifrac$'Pr(>F)'[1]`), and (D) Weighted UniFrac (F statistic: `r beta_results$analysis_results$permanova$wunifrac$F[1]`; p = `r beta_results$analysis_results$permanova$wunifrac$'Pr(>F)'[1]`).

:::

##### Non-metric Multidimensional Scaling (NMDS)

::: {fig-box}

```{r}
#| eval: true
#| label: fig-base-beta-nmds
#| fig.width: 16

#| fig.cap: "Non-metric Multidimensional Scaling (NMDS) ordination plots showing beta diversity patterns across B12 tertiles using different distance metrics"

# Generate ordination plots
plots <- plot_beta_diversity(
  results = beta_results$analysis_results,
  physeq = phyloseq_object,
  group_var = grouping_var,
  ordination_type = "NMDS"
)

# Arrange and display plots
do.call(
  ggarrange, 
  c(plots, 
    list(ncol = 2, 
      nrow = 2, 
      common.legend = TRUE, 
      legend = "bottom", 
      labels = "AUTO")))
```

**β-diversity ordination plots comparing intervention b12_statuss at baseline and endpoint:** There were no differences in  community structure, nor dispersion of sequences by treatment group for any β-diversity metric measured (PERMDISP; p > 0.05).  However, overall, there was a small shift in community structure at endpoint compared to baseline (determined by PERMANOVA) for (A) Bray Curtis distances (F statistic: `r beta_results$analysis_results$permanova$bray$F[1]`; p = `r beta_results$analysis_results$permanova$bray$'Pr(>F)'[1]`), (B) Jaccard similarity (F statistic: `r beta_results$analysis_results$permanova$jaccard$F[1]`; p = `r beta_results$analysis_results$permanova$jaccard$'Pr(>F)'[1]`), (C) Unweighted UniFrac (F  statistic: `r beta_results$analysis_results$permanova$uunifrac$F[1]`; p = `r beta_results$analysis_results$permanova$uunifrac$'Pr(>F)'[1]`), and (D) Weighted UniFrac (F statistic: `r beta_results$analysis_results$permanova$wunifrac$F[1]`; p = `r beta_results$analysis_results$permanova$wunifrac$'Pr(>F)'[1]`).

:::

#### Beta Diversity Summary Tables

```{r}
#| eval: true
#| label: tbl-ms-table2-base-beta
#| tbl-cap: "Beta diversity summary statistics by B12 tertile"
#| results: asis

if(!is.null(beta_results$beta_summary)) {
  knitr::kable(beta_results$beta_summary,
               digits = 3)
}

```

```{r}
#| eval: true
#| label: tbl-beta-sum
#| tbl-cap: Beta diversity summary table

# Create summary tables
beta_df <- beta_results$analysis_results$dissimilarity %>%
  dplyr::mutate(id=seq_len(nrow(.))) %>%
  pivot_longer(cols = -id,names_to = "variable",values_to = "value") %>%
  dplyr::mutate(variable = str_remove_all(variable, "dissimilarity_by_time_treatment|\\.")) %>%
  separate(variable, into = c("timepoint", "b12_status", "metric"), sep = "_")%>%
  pivot_wider(names_from = metric) %>%
  dplyr::select(!c(id, timepoint))

saveRDS(beta_df, file.path(RESULTS, "B12-status-subgroup-analysis", paste0(grouping_var, "-b12_base-beta-div.rds")))

(p <- beta_df %>% gt_summary(by = "b12_status"))

htmltools::save_html(gt::as_raw_html(p), file.path(RESULTS, "B12-status-subgroup-analysis", paste0(grouping_var, "-baseline-beta-div.tbl")))
```

##### Pairwise Comparisons

```{r}
#| eval: true
#| label: tbl-dunn
#| fig-cap: NA
#| tbl-cap: "Pairwise Comparisons Between B12 Tertiles (Dunn's Test)"

if(!is.null(beta_results$dunn_results)) {

  tmp <- beta_df %>% 
      pivot_longer(cols=!b12_status, names_to="metric")

  b12_vda <- data.frame()

  for(p in unique(tmp$metric)) {
    tmp_vda <- tmp %>% dplyr::filter(metric==p) %>%
    calculate_pairwise_vda(., "value", "b12_status") %>% 
    dplyr::mutate(metric=p) %>% 
    relocate(metric, .before=everything())

    b12_vda <- rbind(b12_vda,tmp_vda)
  }

  left_join(beta_results$dunn_results, b12_vda) %>% kable()
}
```

### Session Information

```{r}
#| eval: true
#| label: session-info

sessionInfo()
```