#### Bivariate analysis

Linear regression models can be used to examine associations between B12 status, and Alpha diversity metrics.

Regression should be used to investigate the interaction between host factors, such as B12 status, and the relative of abundance of specific taxa.

Preliminary bivariate analyses (correlations) to identify variables that have a significant relationship with alpha diversity metrics. Variables with strong relationships may be good candidates for inclusion in your model.

```{r}
#| eval: true
#| label: define B12 deficiency variable
#| tbl-cap: Number of baseline-endpoint paired samples with vitamin B12 measurements available
## ensure there are the same number of samples in each timepoint 
be_b12 %>%
  group_by(timepoint) %>%
  count() %>%
  kable()

diversity_stats <- data.frame(chao = as.numeric(be_b12$chao1), shannon = as.numeric(be_b12$diversity_shannon), evenness = as.numeric(be_b12$evenness_pielou), pd = as.numeric(be_b12$pd))

subject <- be_b12$displayid # categorical
arm_group <- as.factor(as.character(be_b12$arm)) # categorical factor
timepoint <- be_b12$timepoint # categorical
timepoint_group <- as.character(be_b12$visit_typeno) # categorical
time <- scale_values(as.numeric(be_b12$timesincefeedstart)) # numeric (days)
time <- ifelse(time < 0, 0, time)
age <- scale_values(as.numeric(be_b12$agedays_rounded)) # numeric (days)
age_B_unscaled <- be_b12 %>%
  group_by(displayid) %>%
  mutate(agedays_rounded = as.numeric(agedays_rounded)) %>%
  arrange(agedays_rounded, .by_group = TRUE) %>%
  mutate(ageAtBaseline = case_when(
    timepoint == "Baseline" ~ agedays_rounded,
    FALSE ~ NA
  )) %>%
  fill(ageAtBaseline) %>%
  ungroup() %>%
  dplyr::select(ageAtBaseline) %>%
  unlist() # turns df into numeric vector
age_B <- scale_values(age_B_unscaled)
sex <- be_b12$sexf
```

Pearson correlation is used because we want to aim for a linear relationship not simplty detect a montonic realtionship between variables (as with Spearman's rank correaltion coefficient)

```{r}
#| eval: true
#| label: fig-basic-plots
#| fig-cap: Investigating linear relationships between outcome and predictor variables by calculating the Pearson correlation coefficient

tmp <- be_b12 %>%
  dplyr::select(vitb12, observed, chao1, diversity_shannon, evenness_pielou, pd) %>%
  mutate_at(vars(everything()), as.numeric)

pairs(tmp, lower.panel = panel.smooth, upper.panel = panel.cor, cex.labels = 3)

tmp <- tmp %>% mutate(vitb12 = vitb12, observed = 1/observed, chao1 = 1/chao1, diversity_shannon = sqrt(diversity_shannon), evenness_pielou = sqrt(evenness_pielou), pd = log(pd)) 

straighten(tmp, y = "vitb12")

pairs(tmp, lower.panel = panel.smooth, upper.panel = panel.cor, cex.labels = 3)
```

```{r}
#| eval: true
#| warning: false
#| out-width: 100%
#| label: fig-pearson
#| fig-cap: B12 Spearman correlation heatmap
#| fig-height: 15
#| fig-width: 20

corr <- tmp %>%
  cor(method = "spearman", use = "pairwise.complete.obs")

cor_test_mat <- round(ggcorrplot::cor_pmat(tmp), 4)
cor_test_matFer <- cor_test_mat[, 1:6]
cor_test_matalpha <- cor_test_mat[, 7:ncol(cor_test_mat)]

#for use with ggplot tile or ggcorrplot
ggheatmap_palette <- c('#91bfdb','#ffffbf','#fc8d59')

corrPlot <- ggcorrplot::ggcorrplot(corr,
  p.mat = cor_test_mat,
  hc.order = FALSE,
  method = "square",
  type = "lower",
  colors = ggheatmap_palette,
  outline.color = "white",
  lab = TRUE,
  tl.cex = 15,
  tl.col = "black",
  tl.srt = 45,
  sig.level = 0.05,
  pch = 1,
  pch.cex = 15,
  insig = "pch",
  ggtheme = "theme_classic"
) +
  theme(legend.key.size = unit(1, "cm"))

ggarrange(
  annotate_figure(
    arrangeGrob(
      tableGrob(cor_test_matFer),
      tableGrob(cor_test_matalpha),
      nrow = 2,
      widths = 1,
      heights = NULL
    ),
    top = text_grob("Correlation Coefficient P-values",
      color = "black",
      face = "bold",
      size = 14,
      vjust = 3.5
    ),
    bottom = text_grob("Circles represent correlations that are NOT significant",
      color = "black",
      face = "italic",
      size = 14,
      vjust = -3
    )
  ),
  corrPlot,
  ncol = 2,
  nrow = 1,
  widths = c(1.25, 1),
  heights = 1
)
```

#### LMMs Alpha Diversity and B12 Status

```{r}
#| eval: true
#| label: using continuous Fe and Hb variables
b12_conc <- be_b12$vitb12

alpha_lmms_B12_conc <- list(
  # interaction between B12 status at timepoint and sex as fixed effect
  chao_lmm_B12 = lme(1/chao ~ arm_group * time + b12_conc * sex, random = ~ 1 | subject, method = "REML", data = diversity_stats),
  shannon_lmm_B12 = lme(shannon^2 ~ arm_group * time + b12_conc * sex, random = ~ 1 | subject, method = "REML", data = diversity_stats),
  evenness_lmm_B12 = lme(evenness^2 ~ arm_group * time + b12_conc * sex, random = ~ 1 | subject, method = "REML", data = diversity_stats),
  pd_lmm_B12 = lme(log(pd) ~ arm_group * time + b12_conc * sex, random = ~ 1 | subject, method = "REML", data = diversity_stats)
)
```

```{r}
#| eval: true
#| label: tbl-residuals-lmm
#| tbl-cap: Residual diagnostics on untransformed continuous B12 LMM
#check residuals for normal distribution
kable(residual_norm_test(alpha_lmms_B12_conc), format = "markdown")
```

```{r}
#| eval: true
#| label: tbl-star-B12
#| tbl-cap: Untransformed B12 continuous variable model summary statistics
#| results: asis

stargazer::stargazer(alpha_lmms_B12_conc, type = "html", star.cutoffs = c(0.05, 0.001, 0.0001), header = FALSE)
```

The precision of the correlation parameter estimation by the `alpha_lmms_B12_conc` model can be assessed with the `intervals()` function. This function provideds a basic summary of confidence intervals on the parameters, $\beta$, $\sigma$ and $\sigma_b$. *Should be used in conjuntion with `anova()` to access term significance.*

```{r}
#| eval: true
#| label: investigate precisiion of model
#| error: true
lapply(alpha_lmms_B12_conc, intervals, which = "all")
```

```{r}
#| eval: true
#| label: tbl-apVar
#| tbl-cap: apVar
do.call(rbind,lapply(alpha_lmms_B12_conc, function(x){
  x$apVar
}))
```

```{r}
#| eval: true
#| label: tbl-anova
#| tbl-cap: anova-term significance for continuous B12 LMM
tmp <- lapply(alpha_lmms_B12_conc, anova)

kable(anova_pvalue(tmp), format = "markdown")
```

```{r}
#| eval: true
#| label: fig-res-diagnose
#| fig-cap: Residual diagnostic plots for continuous B12 LMMs
mapply(residual_plots, model = alpha_lmms_B12_conc, model_name = names(alpha_lmms_B12_conc), render = TRUE, chol =T)
```

Calculation of estimated variances, standard deviations, and correlations between the random-effects. Also calculates within-group error variance and standard deviation [@Pinheiro2023].

```{r}
#| eval: true
#| label: variance correlation
lapply(alpha_lmms_B12_conc, function(x){
  kable(VarCorr(x))
})
```

```{r}
#| eval: true
#| label: tbl-heteroske-test
#| tbl-cap: Breusch-Pagan test for Heteroskedasticity
lapply(alpha_lmms_B12_conc, function(x){
  kable(performance::check_heteroskedasticity(x))
})
```

A p-value \< 0.05 indicates non-constant variance (heteroskedasticity)

```{r}
#| eval: true
#| label: fig-heteroske
#| fig-cap: Heteroskedacity plot of model residuals

library(see)

plots <- lapply(alpha_lmms_B12_conc, function(x){
  p <- performance::check_heteroscedasticity(x) 
  plot(p)
})

```

The RMSE is the square root of the variance of the residuals and indicates the absolute fit of the model to the data (difference between observed data to model’s predicted values). It can be interpreted as the standard deviation of the unexplained variance, and is in the same units as the response variable. Lower values indicate better model fit.

The normalized RMSE is the proportion of the RMSE related to the range of the response variable. Hence, lower values indicate less residual variance.

```{r}
#| eval: true
#| label: tbl-rmse
#| tbl-cap: Alpha diversity B12 model raw and normalized RMSE results.
roc <- lapply(alpha_lmms_B12_conc, function(x){
  performance::performance_rmse(x, normalized = FALSE)
})

roc_norm <- lapply(alpha_lmms_B12_conc, function(x){
  performance::performance_rmse(x, normalized = TRUE)
})

kable(data.frame(RMSE=do.call(rbind, roc), RMSE_normalized=do.call(rbind, roc_norm)))

```

##### Including an interaction with time

```{r}
#| eval: true
#| label: model time interaction
alpha_lmms_B12_conc <- list(
  # interaction between B12 status at timepoint and sex as fixed effect
  chao_lmm_B12 = lme(1/(chao) ~ arm_group * time + b12_conc * sex, random = ~ 1 | subject, method = "REML", data = diversity_stats),
  shannon_lmm_B12 = lme(shannon^2 ~ arm_group * time + b12_conc * sex, random = ~ 1 | subject, method = "REML", data = diversity_stats),
  evenness_lmm_B12 = lme(evenness^3 ~ arm_group * time + b12_conc * sex, random = ~ 1 | subject, method = "REML", data = diversity_stats),
  pd_lmm_B12 = lme(log(pd) ~ arm_group * time + b12_conc * sex, random = ~ 1 | subject, method = "REML", data = diversity_stats)
)

alpha_lmms_B12_conc_time <- lapply(alpha_lmms_B12_conc, update, fixed = varFixed(~ arm_group * time * b12_conc * sex))
```

```{r}
#| eval: true
#| label: variance
#| results: asis

lapply(alpha_lmms_B12_conc_time, function(x){
  kable(VarCorr(x))
})
```

##### Testing Model Assumptions

###### Residuals are normally distributed

```{r}
#| eval: true
#| label: tbl-res-b12-int
#| tbl-cap: Residual normality test results using continuous B12 concentration and time interaction

kable(residual_norm_test(alpha_lmms_B12_conc_time))
```

```{r}
#| eval: true
#| label: fig-res-diagnostics
#| fig-cap: Residual diagnostic plots for continuous B12 LMMs
#| message: false
#| warning: false
mapply(residual_plots, model = alpha_lmms_B12_conc_time, model_name = names(alpha_lmms_B12_conc_time), render = TRUE, chol =T)
```

###### Residual Heteroskedacity

Using this method for models converged with the REML method may be unreliable as the function expects ML to have been used. Best to go with visual plots.

```{r}
#| eval: true
#| label: tbl-res-febh-int-heteroske
#| tbl-cap: Breusch-Pagan test for Heteroskedasticity using continuous Fe and Hb variables with time interaction

lapply(alpha_lmms_B12_conc_time, function(x){
  kable(performance::check_heteroskedasticity(x), format = "markdown")
})

```

The to Breusch-Pagan test is highly sensitivity to sample size and small deviations from normal, often making it an unreliable test.

####### Pattern of heteroskedacity

```{r}
#| eval: true
#| label: fig-res-plot
#| fig-cap: Fitted values vs absolute residuals

p <- lapply(alpha_lmms_B12_conc_time, function(model){
  ggplot(data.frame(fitted = fitted(model), 
                 resid = residuals(model)), 
       aes(x = fitted, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted",
       x = "Fitted values", 
       y = "Residuals")
})

ggpubr::annotate_figure(ggpubr::ggarrange(plotlist = p, ncol = 2, nrow = 2, common.legend = TRUE))

```

####### Modelling heteroskedacity

```{r}
#| eval: true
#| label: tbl-res-b12-varIdent-int
#| tbl-cap: Residual normality test results using continuous Fe and Hb variables with time interaction with modelling heteroskedacity using varIdent.

varIdent <- lapply(alpha_lmms_B12_conc_time, function(x){
  update(x, weights = varIdent(form = ~fitted(.)))
})

residual_norm_test(varIdent)
```

```{r}
#| eval: true
#| label: fig-res-varIdent-plot
#| fig-cap: Fitted values vs absolute residuals with modelling heteroskedacity using varIdent

p <- lapply(varIdent, function(model){
  ggplot(data.frame(fitted = fitted(model), 
                 resid = residuals(model)), 
       aes(x = fitted, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted",
       x = "Fitted values", 
       y = "Residuals")
})

ggpubr::annotate_figure(ggpubr::ggarrange(plotlist = p, ncol = 2, nrow = 2, common.legend = TRUE))

```

####### Covaraite vs residual plots

```{r}
#| eval: true
#| label: fig-covar-res-plots
#| fig-cap: Covariables vs Residuals

covariates <- data.frame(subject, timepoint, arm_group,sex, b12_conc)

p <- lapply(alpha_lmms_B12_conc_time, function(model){
  fontsize = 8
  theme_set(
  theme_classic() +
    theme(
      axis.text.x = element_text(angle = 70, hjust = 1, vjust = 1, size = fontsize, face='bold', color='black'), 
      axis.text.y = element_text(size = fontsize, face='bold', color='black'),
      axis.title.x = element_text(size = fontsize, face='bold', color='black'),
      legend.title= element_blank(),
      legend.position = 'bottom',
      legend.location = "plot",
      legend.text=element_text(size = fontsize, face='bold', color='black'),
      legend.justification= "right",
      axis.title.y = element_text(size = fontsize, face='bold', color='black'),
      strip.text = element_text(size = fontsize, face = 'bold', color = 'black'),
      strip.background.x = element_rect(fill = "white"),
      strip.background.y = element_rect(fill = "lightgrey")))


  p1 <- covariates  %>% 
  dplyr::select(subject, timepoint, sex, arm_group) %>%
  mutate(resid = residuals(model)) %>%
  pivot_longer(cols=!c(subject, resid)) %>%
  group_by(name) %>%
    ggplot(aes(x = value, y = resid, fill = name)) +
    geom_boxplot() +  
      geom_jitter(width=0.1, alpha=0.5) +  
      geom_hline(yintercept=0, linetype="dashed")

   p2 <- covariates  %>% 
  dplyr::select(b12_conc) %>%
    mutate(resid = residuals(model)) %>%
  ggplot(aes(x = b12_conc, y = resid))+
  geom_point() +
      geom_hline(yintercept = 0, linetype = "dashed")

    p1+p2
})

ggpubr::annotate_figure(ggpubr::ggarrange(plotlist = p, ncol = 2, nrow = 4, common.legend = TRUE))

varConstPropFitted <- lapply(alpha_lmms_B12_conc_time, function(x){
  update(x, weights = varConstProp(form = ~fitted(.)))
})
residual_norm_test(varConstPropFitted)
p <- lapply(varConstPropFitted, function(model){
  ggplot(data.frame(fitted = fitted(model), 
                 resid = residuals(model)), 
       aes(x = fitted, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed")})

ggpubr::annotate_figure(ggpubr::ggarrange(plotlist = p, ncol = 2, nrow = 2, common.legend = TRUE))


varConstProp <- lapply(alpha_lmms_B12_conc_time, function(x){
  update(x, weights = varConstProp(form = ~b12_conc))
})
residual_norm_test(varConstProp)
p <- lapply(varConstProp, function(model){
  ggplot(data.frame(fitted = fitted(model), 
                 resid = residuals(model)), 
       aes(x = fitted, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed")})

ggpubr::annotate_figure(ggpubr::ggarrange(plotlist = p, ncol = 2, nrow = 2, common.legend = TRUE))
```

####### Comparision of models with different heteroskedacity variance functions

```{r}
#| eval: true
#| label: tbl-het-anova
#| tbl-cap: Comparison of models with different variance functions to model residual heteroskedacity

anova(alpha_lmms_B12_conc$chao_lmm_B12,alpha_lmms_B12_conc_time$chao_lmm_B12,
varConstPropFitted$chao_lmm_B12,
varConstProp$chao_lmm_B12)

anova(alpha_lmms_B12_conc$shannon_lmm_B12,alpha_lmms_B12_conc_time$shannon_lmm_B12,
varConstPropFitted$shannon_lmm_B12,
varConstProp$shannon_lmm_B12)

anova(alpha_lmms_B12_conc$evenness_lmm_B12,alpha_lmms_B12_conc_time$evenness_lmm_B12,
varConstPropFitted$evenness_lmm_B12,
varConstProp$evenness_lmm_B12)

anova(alpha_lmms_B12_conc$pd_lmm_B12,alpha_lmms_B12_conc_time$pd_lmm_B12,
varConstPropFitted$pd_lmm_B12,
varConstProp$pd_lmm_B12)

```

###### Finalised Model

$$
y \sim armGroup * time * b12Conc * sex, random = ~ 1 | subject
$$

```{r}
#| eval: true
#| label: tbl-res-b12-star
#| tbl-cap: Model summary results 
#| results: asis

stargazer::stargazer(alpha_lmms_B12_conc_time, type = "html", star.cutoffs = c(0.05, 0.001, 0.0001), header = FALSE)
```

```{r}
#| eval: true
#| label: tbl-res-b12-anova
#| tbl-cap: ANOVA term importance test results with accompanying partial omega squared effect size estimate.

tmp <- lapply(alpha_lmms_B12_conc_time, function(x){
  anov <- anova(x) 
  eff <- effectsize::omega_squared(anov)
  left_join(rownames_to_column(anov),eff, by = c("rowname"="Parameter"))

})

lapply(tmp, kable)

```

```{r}
#| eval: true
#| label: tbl-res-b12-anova-significant
#| tbl-cap: Terms with p < 0.05 or partial omega squared > 0.01

lapply(tmp, function(x){
  x %>%
  dplyr::filter(`p-value` <=0.05 | Omega2_partial > 0.01) %>%
  kable()
})

```

Chao and PD change over time is influenced by vitamin B12 concentration

There is a significant improvement of the model (except for Shannon) when time is included as an interaction term with all other fixed variables. Check precision and term significance:

```{r}
#| eval: true
#| label: investigate precisiion of the time interaction model
#| error: true

lapply(alpha_lmms_B12_conc_time, intervals, which = "all") 

lapply(alpha_lmms_B12_conc_time, function(x){
  kable(x$apVar)
})
```

::: callout-note
##### Summary of final model

-   Variable name: `alpha_lmms_B12_conc_time`
-   Formula: `y ~ arm_group * time * b12_conc * sex, random = ~ 1 | subject`
-   Method: REML
-   Within-group correlation structure: default
-   Residuals: normal, homoskedastic
-   Significant terms: Time and time\*B12_conc
:::

#### EMMs

Back transformation is possible by including the transformation on the response variable in the model formula, and using `type = "response"` to return results on the original scale.

The `time` variable is on a continuous scale between 0 and 1. As we are interested in the difference between baseline and endpoint, the estimated marginal means can be calculated at T0 and T1 (`at = list(time = c(0, 1))`). Weights are sent to as `proportional` (weight in proportion to the frequencies (in the original data) of the factor combinations that are averaged over) to adjust for any imbalances in the data. Our data is unbalanced and therefore each categorical predictor variable should not be weighted equally. For example, the B12 status at baseline is unbalanced towards being sufficient:

```{r}
#| eval: true
#| label: tbl-data-balance
#| tbl-cap: accessing the balance of the data
source(normalizePath("utils/r-utils.r"))

be_b12 %>%
  dplyr::select(b12_status, sexf) %>%
  tbl_summary(
    by = sexf,
    missing = "ifany",
    type = list(
      all_categorical() ~ "categorical"),
        statistic = list(
          all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
          all_categorical() ~ "{n} / {N}  \n({p}%)")) %>% # stats and format for categorical columns
        add_n() %>%
        add_p(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
        add_q(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
        bold_p(q = T) %>%
		as_gt() %>% fmt_markdown(columns=where(~is.character(.x))) %>% fmt_number(columns = everything(), drop_trailing_zeros = TRUE) %>% fmt_number(columns = where(~is.numeric(.x)<0.0001), pattern = "<0.0001") %>% fmt_number(columns = where(~is.numeric(.x)<0.001), pattern = "<0.001")%>% fmt_number(columns = where(~is.numeric(.x)<0.01), pattern = "<0.01")%>% fmt_integer(columns = where(~is.numeric(.x)<0.05), pattern = "<0.05")

```

##### Timepoint differences in alpha diveristy EMMs

::: question
Does alpha diversity differ between intervention arm within each timepoint?
:::

::: h0
There is no difference bwetween arms at eithe baseline or endpoint.

$$
\begin{array}{c}
H_{0a}: B12_{high}CPM - B12_{low}CPM = 0 \\
\\
H_{0b}: B12_{high}FeZnPM - B12_{low}FeZnPM = 0
\end{array}
$$
:::

::: callout-tip
$y \sim x \vert z$

The effect of $x$ on $y$ varies across all levels of $z$ i.e. it depends on the value $z$.
:::

The below estimates the mean alpha diveristy for each `arm_group` at each `timepoint`, controlling for sex and vitamin B12 concentration.

```{r}
#| eval: true
#| label: tbl-arm
#| tbl-cap: Alpha diversity marginal mean estimates (EMMs) between intervention arm at baseline and endpoint among individuals who had vitamin B12 measurements available (n = 63 paired samples)

emm <- lapply(alpha_lmms_B12_conc_time, function(x){
  # Get EMMs for arm_group at each time point
  emmeans(x, ~ arm_group | time, 
    at = list(time = c(0, 1)), 
    adjust = "sidak",
    weights = "proportional",
    type= "scale")
})

emm_bind(emm)
```

```{r}
#| eval: true
#| label: tbl-emm-pairs
#| tbl-cap: Comparison of alpha diversity marginal mean estimates (EMMs) betweeen intervention arm at baseline and endpoint among individuals who had vitamin B12 measurements available (n = 63 paired samples)

emm_bind(emm_pairs <- lapply(emm, pairs))
```

###### Alpha diveristy diversity changes over time within each intervention arm

```{r}
#| eval: true
#| label: tbl-emm-arm-trend
#| tbl-cap: Rate of change per unit increase of alpha diversity marginal mean estimates over time. The time.trend column represents the slope of the relationship with time.

emm_bind(trend <- lapply(alpha_lmms_B12_conc_time, function(x){
  emtrends(x, ~ arm_group | time, var ="time")
}))
```

```{r}
#| eval: true
#| label: tbl-emm-arm-trend-compare
#| tbl-cap: Comparison of the rates of change from baseline to endpoint in alpha diversity EMMs between intervention arms.

emm_bind(slopes <- lapply(trend, pairs))
```

Alpha diversity does not differ between intervention arm at baseline or endpoint, nor does intervention group impact the rate at which alpha diversity hchages.

##### Influence of B12 status on alpha diversity EMM differences between intervention arm at each timepoint

::: h0
$$
\begin{array}{cc}
\frac{B12_{low}Control + B12_{low}FeZnPM}{2} = \frac{B12_{high}Control + B12_{high}FeZnPM}{2} \\
\Rightarrow \frac{B12_{low}Control}{2} + \frac{B12_{low}FeZnPM}{2} - \frac{B12_{high}Control}{2} - \frac{B12_{high}FeZnPM}{2} = 0 \\
\Rightarrow \frac{B12_{high}Control}{2} + \frac{B12_{high}FeZnPM}{2} - \frac{B12_{low}Control}{2} - \frac{B12_{low}FeZnPM}{2} = 0
\end{array}
$$
:::

```{r}
#| eval: true
#| label: tbl-arm
#| tbl-cap: Alpha diversity marginal mean estimates (EMMs) within intervention arm at baseline and endpoint among individuals who had vitamin B12 measurements available (n = 63 paired samples)

bmean <- mean(b12_conc)
bsd <- sd(b12_conc)

emm_bind(emm_b12 <- lapply(alpha_lmms_B12_conc_time, function(x){
  emmeans(x, ~ (arm_group | time) | b12_conc, at = list(b12_conc = c(bmean-bsd, bmean, bmean+bsd), time = c(0,1)), adjust = "sidak", weights = "proportional")
}))

```

```{r}
#| eval: true
#| label: tbl-emm-b12-pairs
#| tbl-cap: Comparison of alpha diversity marginal mean estimates (EMMs) betweeen intervention arm at baseline and endpoint among individuals who had vitamin B12 measurements available (n = 63 paired samples)
emm_bind(lapply(emm_b12, pairs))
```

##### Alpha diveristy diversity changes over time within intervention arm with respect to vitamin B12 status

```{r}
#| eval: true
#| label: tbl-emm-arm-trend
#| tbl-cap: Rate of change per unit increase of alpha diversity marginal mean estimates over time. The b12_conc.trend column represents the slope of the relationship with B12_conc.

trend <- lapply(alpha_lmms_B12_conc_time, function(x){
  emtrends(x, ~ (arm_group | time) | b12_conc, var ="b12_conc")
})

emm_bind(trend)

```

```{r}
#| eval: true
#| label: tbl-emm-arm-trend-compare
#| tbl-cap: The rate of change in alpha diversity indices marginal mean estimates with respect to vitamin B12 concentration 

emm_bind(lapply(trend,pairs))
```

###### Influence without intervention arm

As there was no difference between the effect of vitamon B12 status alpha diversity statistics between arms, the data is collapsed.

::: question
Do alpha diveristy statistics differ between groups that have sufficient vitamin B12 levels (\> 300 pg/mL) vs those who do not (Deficiency: \<148 pmol/L (200 pg/mL); Insufficiency: \<221 pmol/L (300 pg/mL))?
:::

```{r}
#| eval: true
#| label: tbl-emm-pairs
#| tbl-cap: Alpha diversity marginal mean estimates (EMMs) according to B12 concentration at baseline and endpoint among individuals who had vitamin B12 measurements available (n = 63 paired samples).

emm_b12 <- lapply(alpha_lmms_B12_conc_time, function(x){
  # Get EMMs at different B12 concentrations
  emmeans(x, ~ b12_conc | time, at = list(b12_conc = c(bmean-bsd, bmean, bmean+bsd), time = c(0,1))) 
})

emm_bind(emm_b12)
```

```{r}
#| eval: true
#| label: tbl-emm-b12-pairs
#| tbl-cap: Comparison of alpha diversity marginal mean estimates (EMMs) according to vitamin B12 status at baseline and endpoint among individuals who had vitamin B12 measurements available (n = 63 paired samples)

emm_bind(lapply(emm_b12, pairs))
```

###### Alpha diversity changes over time with respect to vitamin B12 status (?)

```{r}
#| eval: true
#| label: tbl-emm-arm-trend
#| tbl-cap: Rate of change per unit increase of alpha diversity marginal mean estimates over time. The b12_conc.trend column represents the slope of the relationship with B12_conc.

trend <- lapply(alpha_lmms_B12_conc_time, function(x){
  emtrends(x, ~ b12_conc | time, var ="b12_conc",at = list(b12_conc = c(bmean-bsd, bmean, bmean+bsd), time = c(0,1)))
})

emm_bind(trend)
```

```{r}
#| eval: true
#| label: tbl-emm-arm-trend-compare
#| tbl-cap: The rate of change in alpha diversity indices marginal mean estimates with respect to vitamin B12 concentration 

emm_bind(lapply(trend,pairs))
```

```{r}
#| eval: true
#| label: fig-emm-sex
#| fig-cap: emmeans plots- sex differences
## Create an empty list to store plots

tmp <- lapply(alpha_lmms_B12_conc_time, function(x){
  # Get EMMs at different B12 concentrations
  emmeans(x, ~ b12_conc | time, at = list(b12_conc = c(200,300,bmean+bsd), time = c(0,1))) 
})

plot_list <- list()

## Loop through each item in the list
for (i in names(tmp)) {
  # Create a plot for each item (replace with your actual plot creation code)
  plot <- emmip(tmp[[i]], b12_conc ~ time, CIs = TRUE) + ggtitle(paste(str_extract(i, "^[^_]*"))) + theme(legend.position = "bottom", plot.title = element_text(size = 10, face = "bold"))

  # Add the plot to the list
  plot_list[[i]] <- plot
}

## Arrange and display the plots using ggarrange
ggpubr::annotate_figure(ggpubr::ggarrange(plotlist = plot_list, ncol = 2, nrow = 2, common.legend = TRUE), top = "Change in diversity statistics")
```

From the resulting graph, we can conclude that there is likely no/little difference between alpha diversity changes in males and females. However, does time have a different trend (slope) depending on sex?

```{r}
#| eval: true
#| label: fig-emm-sex
#| fig-cap: emmeans plots- sex differences
## Create an empty list to store plots
plot_list <- list()

## Loop through each item in the list
for (i in names(emm_b12)) {
  # Create a plot for each item (replace with your actual plot creation code)
  plot <- emmip(emm_b12[[i]], b12_conc ~ time, CIs = TRUE) + ggtitle(paste(str_extract(i, "^[^_]*"))) + theme(legend.position = "bottom", plot.title = element_text(size = 10, face = "bold"))

  # Add the plot to the list
  plot_list[[i]] <- plot
}

## Arrange and display the plots using ggarrange
ggpubr::annotate_figure(ggpubr::ggarrange(plotlist = plot_list, ncol = 2, nrow = 2, common.legend = TRUE), top = "Change in diversity statistics")
```


*Hypothesis 1:*

::: h0
$H_0$: The rate of changes in alpha diversity statistics are not different between males and females.
:::

::: h1
-   $H_1$: The rate of changes in alpha diversity statistics are different between males and females.
    -   $H_1a$: Alpha diversity increases faster among males than it deos females
    -   $H_1b$: Alpha diversity increases faster among females than it deos males

i.e. `time` has a different trend depending on `sex`
:::

#TODO change all correlations to spearman -\> doesnt assume variables are normally distributed

```{r}
#| eval: true
#| label: fig-emm-trends-sex
#| fig-cap: emmeans trends- sex differences

lapply(alpha_lmms_B12_conc, function(x){
  trend <- emtrends(x, ~ sex, var = "time")
  print(kable(trend))
  print(kable(pairs(trend)))
})
```

Apparently, in order for this to be useful, `var` should be a numeric predictor that interacts with at least one factor in `specs`. So I have simplified the model to have a look at this.

```{r}
#| eval: true
#| label: time*sex differences

tmp <- list(
  # interaction between B12 status at timepoint and sex as fixed effect
  chao_lmm_B12 = lme(chao ~  time*  sex, random = ~ 1 | subject, method = "REML", data = diversity_stats),
  shannon_lmm_B12 = lme(shannon^2 ~  time*  sex, random = ~ 1 | subject, method = "REML", data = diversity_stats),
  evenness_lmm_B12 = lme(evenness^2 ~  time*  sex, random = ~ 1 | subject, method = "REML", data = diversity_stats),
  pd_lmm_B12 = lme(log(pd) ~ time *sex, random = ~ 1 | subject, method = "REML", data = diversity_stats)
)
```

```{r}
#| eval: true
#| label: tbl-time-sex-res
#| tbl-cap: Residual normality test results for models investigating sex differences over time
kable(residual_norm_test(tmp), format = "markdown")
```

```{r}
#| eval: true
#| label: tbl-time-sex-conf
#| tbl-cap: Confidence intervals results for models investigating sex differences over time
lapply(tmp, intervals, which = "fixed")
```

```{r}
#| eval: true
#| label: tbl-time-sex-var
#| tbl-cap: Estimate covariance structure for models investigating sex differences over time
lapply(tmp, function(x){
  print(x$apVar)
})
```

```{r}
#| eval: true
#| label: tbl-time-sex-trend
#| tbl-cap: EMM trend results for models investigating sex differences over time
lapply(tmp, function(x){
  trend <- emtrends(x, ~ sex, var = "time")
  print(kable(trend, format = "markdown"))
  kable(pairs(trend), format = "markdown")
})
```

Both tests, and the plot suggest that there is no difference between the rates of changes in alpha diversity statistics among male and females. Therefore, we accept the NULL hypothesis ($H_0$) and reject the alternative hypothesis ($H_1$)

The formula `b12_conc * sex ~ time` specifies that you want to examine the effect of the interaction between `b12_conc * sex` on the outcome variable (alpha diversity statistic) across different levels of time. Confidence intervals are included to help assess estimate precision.

```{r}
#| eval: true
#| label: fig-emm-B12
#| fig-cap: emmeans plots- B12 level differences

# Create an empty list to store plots
plot_list <- list()

## Loop through each item in the list
for (i in names(emm)) {
  # Create a plot for each item (replace with your actual plot creation code)
  plot <- emmip(emm[[i]], b12_conc * sex ~ time, CIs = TRUE) + ggtitle(paste(str_extract(i, "^[^_]*"))) + theme(legend.position = "bottom", plot.title = element_text(size = 10, face = "bold"))

  # Add the plot to the list
  plot_list[[i]] <- plot
}

## Arrange and display the plots using ggarrange
ggpubr::annotate_figure(ggpubr::ggarrange(plotlist = plot_list, ncol = 2, nrow = 2, common.legend = TRUE), top = "Change in diversity statistics")
```

From the resulting graph, we can conclude that there is no/little difference between alpha diversity changes in males and females.

##### Alpha diversity changes over time within each sex

```{r}
#| eval: true
#| label: tbl-emm-sex
#| tbl-cap: Sex differences in alpha diversity marginal mean estimates (EMMs) at baseline and endpoint (n = 63 paired samples).

emm_sex <- lapply(alpha_lmms_B12_conc_time, function(x){
  # Get EMMs at different B12 concentrations
  emmeans(x, ~ sex | time, at = list(time = c(0,1))) 
})

emm_bind(emm_sex)
```

```{r}
#| eval: true
#| label: tbl-emm-sex-pairs
#| tbl-cap: Comparison of alpha diversity marginal mean estimates (EMMs) between the sexes at baseline and endpoint (n = 63 paired samples)

emm_bind(lapply(emm_sex, pairs))
```

There are no differences in alpha diversity at either baseline between male and female subjects.

###### Alpha diversity changes over time with respect to sex

```{r}
#| eval: true
#| label: tbl-emm-arm-trend
#| tbl-cap: Rate of change per unit increase of alpha diversity marginal mean estimates over time. The b12_conc.trend column represents the slope of the relationship with B12_conc.

trend <- lapply(alpha_lmms_B12_conc_time, function(x){
  emtrends(x, ~ sex | time, var ="time",at = list(time = c(0,1)))
})

emm_bind(trend)
```

```{r}
#| eval: true
#| label: tbl-emm-arm-trend-compare
#| tbl-cap: The rate of change in alpha diversity indices marginal mean estimates with respect to vitamin B12 concentration 

emm_bind(lapply(trend,pairs))
```

Alpha diversity rates of change over time do not differ between males and females.

##### Differences in alpha diversity statistics in each timepoint within sex and B12 status

The formula `b12_conc * sex ~ time` specifies that you want to examine the effect of the interaction between `b12_conc * sex` on the outcome variable (alpha diversity statistic) across different levels of time. Confidence intervals are included to help assess estimate precision.

```{r}
#| eval: true
#| label: tbl-emm-sex-b12
#| tbl-cap: Sex differences in alpha diversity marginal mean estimates (EMMs) at baseline and endpoint (n = 63 paired samples).

emm_sex <- lapply(alpha_lmms_B12_conc_time, function(x){
  # estimate b12_conc means within each combination of sex and time
  emmeans(x, ~ b12_conc | sex | time, at = list(time = c(0,1), b12_conc = c(bmean-bsd, bmean, bmean+bsd)))
})

emm_bind(emm_sex)
```

```{r}
#| eval: true
#| label: fig-emmip-sex
#| fig-cap: Visualisation of interations between B12 status group, se and time.

#y ~ b12_conc | sex | time
tmp <- alpha_lmms_B12_conc_time[[1]]

emmip(tmp, sex ~time |b12_conc, at = list(time = c(0,1)))

```


```{r}
#| eval: true
#| label: tbl-emm-sex-b12-pairs
#| tbl-cap: Comparison of alpha diversity marginal mean estimates (EMMs) between the sexes with different B12 status at baseline and endpoint (n = 63 paired samples)

emm_bind(lapply(emm_sex, pairs))
```

There are no differences in alpha diversity at either baseline between male and female subjects.

##### Alpha diversity changes over time with respect to sex and B12 status

```{r}
#| eval: true
#| label: tbl-emm-sex-b12-trend
#| tbl-cap: Rate of change per unit increase of alpha diversity marginal mean estimates over time. The b12_conc.trend column represents the slope of the relationship with B12_conc.

trend <- lapply(alpha_lmms_B12_conc_time, function(x){
  emtrends(x, ~ b12_conc | sex | time, var ="time",at = list(time = c(0,1), b12_conc = c(bmean-bsd, bmean, bmean+bsd)))
})

emm_bind(trend)
```

```{r}
#| eval: true
#| label: tbl-emm-sex-b12-trend-compare
#| tbl-cap: The rate of change in alpha diversity indices marginal mean estimates with respect to vitamin B12 concentration 

emm_bind(lapply(trend,pairs))
```

```{r}
#| eval: true
#| label: fig-emm-sex-b12-plot
#| fig-cap: emmeans plots- sex differences
## Create an empty list to store plots
plot_list <- list()

## Loop through each item in the list
for (i in names(emm_sex)) {
  # Create a plot for each item
  plot <- 
    emmip(emm_sex[[i]], b12_conc * sex ~ time, CIs = TRUE, plotit=FALSE) %>%
      ggplot(aes(x=time, y=yvar)) +
      geom_line(aes(color=sex))+
      geom_errorbar(aes(ymin=LCL, ymax=UCL, color=sex), width=0.2) +
      ggtitle(paste(str_extract(i, "^[^_]*"))) + 
      theme(legend.position = "bottom", plot.title = element_text(size = 10, face = "bold")) +
      facet_grid(~ b12_conc, scale= "free")

  # Add the plot to the list
  plot_list[[i]] <- plot
}


## Arrange and display the plots using ggarrange
ggpubr::annotate_figure(ggpubr::ggarrange(plotlist = plot_list, ncol = 2, nrow = 2, common.legend = TRUE), top = "Change in diversity statistics")
```

From the resulting graph, we can conclude that there is no/little difference between alpha diversity changes in males and females.

