### B12 Status: Baseline to Endpoint Bacterial Relative Abundances

```{r}
#| echo: false
#| eval: true
#| label: setup
#| include: false

# This initial code block loads custom utility functions and environment variables

# These are stored in separate files to keep the main analysis clean and organized
source(normalizePath("utils/r-utils.r"))
source(normalizePath("_environment.required"))
load(normalizePath("data/seqdata.RData"))

metadata <- read_delim(normalizePath(METADATA))
bac_base_end_meta <- readRDS(normalizePath("data/metadata/base-end-metadata.rds"))
baseline_metadata <- readRDS(normalizePath("data/metadata/baseline-metadata.rds"))

suppressPackageStartupMessages(library(emmeans))
suppressPackageStartupMessages(library(nlme))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ampvis2))


# source(normalizePath("../../utils/r-utils.r"))
# source(normalizePath("../../_envb12ment.required"))
# load(normalizePath("../../data/seqdata.RData"))

# #metadata <- read_delim(normalizePath(METADATA))
# bac_base_end_meta <- readRDS(normalizePath("../../data/metadata/base-end-metadata.rds"))
# baseline_metadata <- readRDS(normalizePath("../../data/metadata/baseline-metadata.rds"))

dir.create(file.path(RESULTS, "B12-status-subgroup-analysis"), showWarnings = FALSE, recursive = TRUE)
```

#### Data preparation

```{r}
#| eval: true
#| label: filter by b12 results

be_b12 <- bac_base_end_meta %>%
  mutate(
    b12_status = case_when(
      vitb12 < 200 ~ "deficiency",
      vitb12 >= 200 & vitb12 <= 300 ~ "insufficiency",
      vitb12 >= 300 ~ "sufficiency",
      TRUE ~ "Unknown"
    ),
    sexf = case_when(
      sexf == 1 ~ "Female",
      sexf == 0 ~ "Male"
    )
  ) %>%
  group_by(displayid) %>%
  dplyr::filter(!is.na(vitb12)) %>%
  dplyr::filter(all(c("Baseline", "Endpoint") %in% timepoint)) %>%
  ungroup() %>%
  as.data.frame()

rownames(be_b12) <- be_b12$sample_names

saveRDS(be_b12,file.path(RESULTS, "B12-status-subgroup-analysis/b12-BvsE-subset.rds"))
```

Create an ampvis object from phyloseq object

```{r}
#| eval: true
#| label: create ampvis object

paired_b12_samples <- 
  microbiome::meta(seqdata_bac) %>% 
  mutate(displayid = str_trim(displayid, side = "both")) %>% 
  group_by(displayid) %>%
  dplyr::filter(!is.na(vitb12)) %>%
  dplyr::filter(all(c("1Baseline", "3Endpoint") %in% timepoint)) %>%
  ungroup() %>%
  dplyr::filter(timepoint != "2Midpoint")

bac_b12 <- phyloseq::subset_samples(seqdata_bac, sample_names %in% paired_b12_samples$sample_names)

sample_data(bac_b12) <- be_b12

d <- amp_load(otutable = t(otu_table(bac_b12)), be_b12, taxonomy = tax_table(bac_b12))
```

#### Basic Characteristics of Subgroup

```{r}
#| eval: true
#| label: tbl-base-char
#| tbl-cap: Baseline characterisitcs of subgroup with vitamin B12 measurements available

be_b12 %>% select(sexf, arm, b12_status, vitb12) %>% tbl_summary() %>% as_kable()

```

#### Within timepoints

##### Baseline and Endpoint Phyla Abundance

```{r}
#| eval: true
#| label: fig-b12-phyla-heatmap
#| fig-cap: Relative abundance of bacterial phyla at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL). Average group relative abundances are shown on the graph.

(heatmap <- amp_heatmap(d,
            group_by = "b12_status",
            facet_by="timepoint",
            tax_aggregate = "Phylum",
            tax_show = 20,
            plot_values = T,
            normalise = T, 
            plot_values_size = 8,
            round = 3,
            tax_empty="OTU",
            order_x_by = "cluster",
            order_y_by = "cluster") +
              theme(axis.text.x = element_text(angle = 45, size = 14, vjust = 1), axis.text.y = element_text(size = 14),  legend.position = "right", legend.title = element_text(color = "black", size = 10, face = "bold")))
```

```{r}
#| eval: true
#| label: tbl-b12-phyla-heatmap
#| tbl-cap: Relative abundance of bacterial phyla at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
heatmap$data %>% 
  group_by(Display, .Group, timepoint) %>% 
  summarise(mean = round(mean(Abundance) *100, 4)) %>% 
  arrange(desc(mean)) %>%
  mutate(mean =  case_when(
    mean >= 0.01 ~ round(mean, 2), 
    TRUE ~ mean
  )) %>%
  mutate(mean = paste0(mean, "%")) %>%
  pivot_wider(names_from = .Group, values_from = mean) %>%
  kable()
```

###### Statistical Comparison between B12 Status group Phyla Abundance within each timepoint

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: Summary statistics of bacterial relative abundances at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
df_p <- heatmap$data %>% 
  dplyr::select(phyla = Display, Sample , relAbund = Abundance) %>% 
  mutate(timepoint = be_b12$timepoint[match(Sample,be_b12$sample_names)]) %>%
  mutate(b12_status = be_b12$b12_status[match(Sample,be_b12$sample_names)])
  
df_p %>%
  group_by(phyla, b12_status, timepoint) %>%
  summarise(mean = mean(relAbund), median = median(relAbund), max = max(relAbund), min = min(relAbund))
```

Non-parametric tests like the Kruskal-Wallis Rank Sum test do not take the compositional nature of relative abundance data into account

```{r}
#| eval: true
#| label: tbl-b12-kw
#| tbl-cap: Kruskall-Wallis test results with computed $\eta^2_H$ effect size estimate for bacterial phyla at baseline and endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).

tmp <- amp_heatmap(d,
            group_by = "b12_status",
            facet_by="timepoint",
            tax_aggregate = "Phylum",
            tax_show = 20,
            plot_values = T,
            normalise = F, 
            plot_values_size = 8,
            round = 3,
            tax_empty="OTU",
            order_x_by = "cluster",
            order_y_by = "cluster") +
              theme(axis.text.x = element_text(angle = 45, size = 14, vjust = 1), axis.text.y = element_text(size = 14),  legend.position = "right", legend.title = element_text(color = "black", size = 10, face = "bold"))

df_p <- tmp$data %>% 
  dplyr::rename(phyla = Display)

kw <- df_p %>% group_by(phyla, timepoint) %>%
  kruskal_test(Sum ~ b12_status)

kw_eff <- df_p %>% 
group_by(phyla, timepoint) %>% 
kruskal_effsize(Sum ~ b12_status)

left_join(kw,kw_eff, by=c("phyla","timepoint",".y.")) %>% kable()
```

```{r}
#| eval: true
#| label: tbl-b12-dunn-vda
#| tbl-cap: Kruskall-Wallis post hoc Dunn's test results with computed pairwise VDA effect size estimate for bacterial phyla read counts at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
 
dunn <- df_p %>% group_by(phyla,timepoint) %>%
  dunn_test(Sum ~ b12_status, p.adjust.method = "bonferroni")

b12_vda <- data.frame()

for(t in unique(df_p$timepoint)) {
  tmp <- df_p %>% filter(timepoint==t)
for(p in unique(df_p$phyla)) {
  tmp_vda <- tmp %>% dplyr::filter(phyla==p) %>%
  calculate_pairwise_vda(., "Sum", "b12_status") %>% 
  mutate(phyla=p, timepoint = t) %>% 
  relocate(phyla, .before=everything())

  b12_vda <- rbind(b12_vda, tmp_vda)
}
}

left_join(dunn, b12_vda, by =c("phyla", "group1", "group2", "timepoint")) %>% kable()
```

##### Baseline and Endpoint Species Abundance

```{r}
#| eval: true
#| label: fig-b12-spp-heatmap
#| fig-cap: Relative abundance of top 20 bacterial species at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL). Average group relative abundances are shown on the graph.

(heatmap <- amp_heatmap(d,
            group_by = "b12_status",
            facet_by="timepoint",
            tax_aggregate = "Species",
            tax_show = 20,
            plot_values = T,
            normalise = TRUE, 
            plot_values_size = 8,
            round = 3,
            tax_empty="OTU") +
              theme(axis.text.x = element_text(angle = 45, size = 14, vjust = 1), axis.text.y = element_text(size = 14),  legend.position = "right", legend.title = element_text(color = "black", size = 10, face = "bold")))
```

```{r}
#| eval: true
#| label: tbl-b12-spp-heatmap
#| tbl-cap: Relative abundance of top 20 bacterial species at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
heatmap$data %>% 
  group_by(Display, .Group,timepoint) %>% 
  summarise(mean = round(mean(Abundance) *100, 4)) %>% 
  arrange(desc(mean)) %>%
  mutate(mean =  case_when(
    mean >= 0.01 ~ round(mean, 2), 
    TRUE ~ mean
  )) %>%
  mutate(mean = paste0(mean, "%")) %>%
  pivot_wider(names_from = .Group, values_from = mean) %>%
  kable()
```

###### Statistical Compariosn between B12 Status group Species Abundance

::: h0

H0: there is no differnece in abundance of species between vitamin B12 status groups at baseline 

:::

::: h1

H1: species abundance increases/decreases from vitamin B12 status groups at baseline

:::

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: Summary statistics of bacterial relative abundances at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
tmp <- amp_heatmap(d,
            group_by = "b12_status",
            facet_by="timepoint",
            tax_aggregate = "Species",
            tax_show = 20,
            plot_values = T,
            normalise = F, 
            plot_values_size = 8,
            round = 3,
            tax_empty="OTU",
            order_x_by = "cluster",
            order_y_by = "cluster") +
              theme(axis.text.x = element_text(angle = 45, size = 14, vjust = 1), axis.text.y = element_text(size = 14),  legend.position = "right", legend.title = element_text(color = "black", size = 10, face = "bold"))

df_s <- tmp$data %>% 
  dplyr::rename(species = Display)

df_s %>%
  group_by(species, b12_status, timepoint) %>%
  summarise(mean = mean(Sum), median = median(Sum), max = max(Sum), min = min(Sum))
```

```{r}
#| eval: true
#| label: tbl-b12-kw
#| tbl-cap: Kruskall-Wallis test results with computed $\eta^2_H$ effect size estimate for bacterial phyla relative abundances at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
 
kw <- df_s %>% group_by(species, timepoint) %>%
  kruskal_test(Sum ~ b12_status)

kw_eff<-df_s %>% group_by(species, timepoint) %>% kruskal_effsize(Sum ~ b12_status)

left_join(kw,kw_eff, by=c("species","timepoint",".y.")) %>% kable()
```

```{r}
#| eval: true
#| label: tbl-b12-dunn-vda
#| tbl-cap: Kruskall-Wallis post hoc Dunn's test results with computed pairwise VDA effect size estimate for bacterial species relative abundances at baseline among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
 
dunn <- df_s %>% group_by(species,timepoint) %>%
  dunn_test(Sum ~ b12_status, p.adjust.method = "bonferroni")

b12_vda <- data.frame()

for(t in unique(df_s$timepoint)) {
  tmp <- df_s %>% filter(timepoint==t)
for(p in unique(df_s$species)) {
  tmp_vda <- tmp %>% dplyr::filter(species==p) %>%
  calculate_pairwise_vda(., "Sum", "b12_status") %>% 
  mutate(species=p, timepoint = t) %>% 
  relocate(species, .before=everything())

  b12_vda <- rbind(b12_vda, tmp_vda)
}
}

left_join(dunn, b12_vda, by =c("species", "group1", "group2", "timepoint")) %>% kable()
```

#### BvsE: Between timepoints

At present, the sample size is not large enough to conduct longitudinal analysis between B12 status groups as there are not enough individuals in the following groups:

```{r}
#| eval: true
#| label: tbl-change-percent
#| tbl-cap: Number and percentages of B12 status changes from baseline to endpoint

be_b12 %>% 
select(subject=displayid, timepoint, b12_status) %>%
  pivot_wider(names_from = timepoint,
             values_from = b12_status) %>%
  mutate(status_change = case_when(
    Baseline == Endpoint ~ "No Change",
    TRUE ~ "Changed"
  )) %>%
  count(Baseline, Endpoint) %>%
  arrange(Baseline) %>%
  mutate(percentage = (n/sum(n))*100)
```


### Alluvial Diagrams of B12 Status Changes

```{r}
#| eval: true
#| label: fig-sankey
#| fig-cap: "Network flow diagram of vitamin B12 status transitions. Sankey diagram depicting the movement of subjects between B12 status categories from baseline to endpoint. Node height represents the number of subjects in each category, while edge width represents the number of subjects transitioning between categories. Interactive visualization allows exploration of specific transition quantities."

library(networkD3)

links <- be_b12 %>%
  select(subject=displayid, timepoint, b12_status) %>%
  pivot_wider(names_from = timepoint,
             values_from = b12_status,
             names_prefix = "time_") %>%
  count(time_Baseline, time_Endpoint) %>%
  rename(source = time_Baseline,
         target = time_Endpoint,
         value = n)

# Create nodes dataframe
nodes <- data.frame(
  name = unique(c(links$source, links$target))
)

# Convert source and target to zero-based index
links$source <- match(links$source, nodes$name) - 1
links$target <- match(links$target, nodes$name) - 1
links$label <- paste0("n=", links$value)

# Create Sankey diagram
sankeyNetwork(Links = links, 
             Nodes = nodes,
             Source = "source",
             Target = "target",
             Value = "value",
             NodeID = "name",
             LinkGroup = "source",  # color flows by source
             nodeWidth = 30,
             fontSize = 12,
             sinksRight = FALSE) %>%
  htmlwidgets::onRender("
    function(el) {
      // Add labels to the links
      var links = d3.select(el)
        .selectAll('.link');
      
      links.append('title')
        .text(function(d) { 
          return 'n=' + d.value;
        });
    }
  ")

```

```{r}
#| eval: true
#| label: fig-alluvial
#| fig-cap: "Longitudinal changes in vitamin B12 status between baseline and endpoint. Alluvial diagram showing transitions between B12 status categories (deficiency: <200 pg/mL; insufficiency: 200-300 pg/mL; sufficiency: >300 pg/mL). Each flow represents subjects' movements between status categories, with band width proportional to the number of subjects. Vertical bar height indicates the relative proportion of subjects in each category at each timepoint."

# Alluvial plot
library(ggalluvial)

ggplot(be_b12,
       aes(x = timepoint, 
           stratum = b12_status, 
           alluvium = displayid,
           fill = b12_status,
           label = b12_status)) +
  geom_flow() +
  geom_stratum(alpha = .85) + # change strata text/colour alpha
  geom_text(stat = "stratum",
            aes(label = paste0(after_stat(stratum), "\nn=", after_stat(n))),
            size = 6) +
  scale_fill_brewer(palette="Set2") +
  labs(y = "Number of Subjects") +
  xlab(NULL) +
  theme(plot.caption = element_text(size = (fontsize-3)), legend.position = "none", axis.text.x = element_text(angle = 0, size = fontsize, face='bold', color='black'), 
      axis.text.y = element_text(size = fontsize, face='bold', color='black')) +
 geom_text(aes(label = ifelse(..flow.. == "to", "", after_stat(n)),size = 5, vjust=0.5), stat = "flow", nudge_x = 0.19) #add single instance of counts to flows between strata
 
 #geom_label(aes(label = ifelse(..flow.. == "to", "", after_stat(n)),size = 5, vjust=0.5, fill = ifelse(..flow.. == "to", "", "#B3B3B3")), stat = "flow", nudge_x = 0.185) #add single instance of counts to flows between strata

```

```{r}
#| eval: true
#| label: data prep

be_200 <- be_b12 %>% 
      mutate(
        b12_status = case_when(
          vitb12 < 200 ~ "deficiency",
          vitb12 >= 200 ~ "sufficiency",
          TRUE ~ "Unknown"
        )) %>%
      select(timepoint, b12_status, displayid)

be_300 <- be_b12 %>% 
      mutate(
        b12_status = case_when(
          vitb12 < 300 ~ "insufficiency",
          vitb12 >= 300 ~ "sufficiency",
          TRUE ~ "Unknown"
        )) %>%
      select(timepoint, b12_status, displayid)

```

```{r}
#| eval: true
#| label: tbl-change-percent
#| tbl-cap: Number and percentages of B12 status changes between B12 deficiency (<200 pg/mL) and B12 sufficiency (>200 pg/mL) from baseline to endpoint.

be_200 %>% pivot_wider(names_from = timepoint, values_from = b12_status) %>%
  mutate(status_change = case_when(
    Baseline == Endpoint ~ "No Change",
    TRUE ~ "Changed"
  )) %>%
  count(Baseline, Endpoint) %>%
  arrange(Baseline) %>%
  mutate(label = paste0("n=", n), percentage = (n/sum(n))*100)
```

```{r}
#| eval: true
#| label: tbl-change-percent
#| tbl-cap: Number and percentages of B12 status changes between B12 insufficiency (<300 pg/mL) and B12 sufficiency (>300 pg/mL) from baseline to endpoint.

be300 %>% pivot_wider(names_from = timepoint, values_from = b12_status) %>%
  mutate(status_change = case_when(
    Baseline == Endpoint ~ "No Change",
    TRUE ~ "Changed"
  )) %>%
  count(Baseline, Endpoint) %>%
  arrange(Baseline) %>%
  mutate(label = paste0("n=", n), percentage = (n/sum(n))*100)
```

```{r}
#| eval: true
#| label: fig-alluvial-200vs300
#| fig-cap: "Longitudinal changes in vitamin B12 status between baseline and endpoint. Alluvial diagram showing transitions between A) B12 deficiency (<200 pg/mL) and B12 sufficiency (>200 pg/mL), and B) B12 insufficiency (<300 pg/mL) and B12 sufficiency (>300 pg/mL). Each flow represents subjects' movements between status categories, with band width proportional to the number of subjects. Vertical bar height indicates the relative proportion of subjects in each category at each timepoint."
#| fig-width: 18
#| fig-length: 16

p200 <- ggplot(be_200,
       aes(x = timepoint, 
           stratum = b12_status, 
           alluvium = displayid,
           fill = b12_status,
           label = b12_status)) +
  geom_flow() +
  geom_stratum(alpha = .85) + # change strata text/colour alpha
  geom_text(stat = "stratum",
            aes(label = paste0(after_stat(stratum), "\nn=", after_stat(n))),
            size = 5) +
  scale_fill_brewer(palette="Set2") +
  labs(y = "Number of Subjects") +
  xlab(NULL) +
  theme(plot.caption = element_text(size = (fontsize-3)), legend.position = "none", axis.text.x = element_text(angle = 0, size = fontsize, face='bold', color='black'), 
      axis.text.y = element_text(size = fontsize, face='bold', color='black')) +
 geom_text(aes(label = ifelse(..flow.. == "to", "", after_stat(n)),size = 5, vjust=0.5), stat = "flow", nudge_x = 0.22) #add single instance of counts to flows between strata

p300 <- ggplot(be_300,
       aes(x = timepoint, 
           stratum = b12_status, 
           alluvium = displayid,
           fill = b12_status,
           label = b12_status)) +
  geom_flow() +
  geom_stratum(alpha = .85) + # change strata text/colour alpha
  geom_text(stat = "stratum",
            aes(label = paste0(after_stat(stratum), "\nn=", after_stat(n))),
            size = 5) +
  scale_fill_brewer(palette="Set2") +
  labs(y = "Number of Subjects") +
  xlab(NULL) +
  theme(plot.caption = element_text(size = (fontsize-3)), legend.position = "none", axis.text.x = element_text(angle = 0, size = fontsize, face='bold', color='black'), 
      axis.title.y = element_text(size = fontsize, face='bold', color='black'),
      axis.text.y = element_text(size = fontsize, face='bold', color='black'),) +
 geom_text(aes(label = ifelse(..flow.. == "to", "", after_stat(n)),size = 5, vjust=0.5), stat = "flow", nudge_x = 0.22) #add single instance of counts to flows between strata

ggarrange(p200,p300,ncol=2, labels=c("A) Transitions between B12 deficiency (<200 pg/mL) and B12 sufficiency (>200 pg/mL).", "B) Transitions between B12 insufficiency (<300 pg/mL) and B12 sufficiency (>300 pg/mL)."), hjust=-0.2, vjust=1.75)

```