### B12 Status: Baseline to Endpoint Bacterial Relative Abundances

```{r}
#| echo: false
#| eval: true
#| label: setup
#| include: false

# This initial code block loads custom utility functions and environment variables

# These are stored in separate files to keep the main analysis clean and organized
source(normalizePath("utils/r-utils.r"))
source(normalizePath("_environment.required"))
load(normalizePath("data/seqdata.RData"))

metadata <- read_delim(normalizePath(METADATA))
bac_base_end_meta <- readRDS(normalizePath("data/metadata/base-end-metadata.rds"))
baseline_metadata <- readRDS(normalizePath("data/metadata/baseline-metadata.rds"))
be_b12 <- readRDS(file.path(RESULTS, "B12-status-subgroup-analysis/b12-BvsE-subset.rds"))


suppressPackageStartupMessages(library(emmeans))
suppressPackageStartupMessages(library(nlme))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ampvis2))


# source(normalizePath("../../utils/r-utils.r"))
# source(normalizePath("../../_envb12ment.required"))
# load(normalizePath("../../data/seqdata.RData"))

# #metadata <- read_delim(normalizePath(METADATA))
# bac_base_end_meta <- readRDS(normalizePath("../../data/metadata/base-end-metadata.rds"))
# baseline_metadata <- readRDS(normalizePath("../../data/metadata/baseline-metadata.rds"))

dir.create(file.path(RESULTS, "B12-status-subgroup-analysis"), showWarnings = FALSE, recursive = TRUE)
```

#### Data preparation

```{r}
#| eval: true

b12_df <-
  be_b12 %>%
  dplyr::select(displayid, timepoint, arm, vitb12, observed, chao1, diversity_shannon, evenness_pielou, pd) %>%
  pivot_longer(cols = !c(displayid, timepoint, arm, vitb12), names_to = "diversity_metric") %>%
  group_by(displayid) %>%
  dplyr::filter(all(c("Baseline", "Endpoint") %in% timepoint)) %>%
  ungroup() %>%
  as.data.frame()

write.csv(b12_df, file.path(RESULTS, "B12-status-subgroup-analysis/alpha-diveristy-by-B12-status.csv"), row.names = FALSE)
```


```{r}
#| eval: true
#| label: tbl-alpha-sum
#| tbl-cap: Alpha diversity summary stats by timepoint

b12_df %>%
  group_by(timepoint, diversity_metric) %>%
  get_summary_stats(value, type = "common") %>%
  kable()
```

There are `r length(unique(b12_df$displayid))` individuals with baseline and endpoint samples available.

```{r}
#| eval: true
#| label: tbl-arm-n
#| tbl-cap: Number of individuals with baseline and endpoint samples available in each arm

be_b12 %>% 
  group_by(arm,timepoint) %>%
  summarise(n=n()) %>%
  kable()
```

```{r}
#| eval: true
#| label: tbl-sex-n
#| tbl-cap: Number of male and female individuals with baseline and endpoint samples available in each arm. Mean vitamin B12 concentration is also given 

be_b12 %>% 
  group_by(sexf,arm,timepoint) %>%
  summarise(n=n(), b12_conc = mean(vitb12)) %>%
  kable()
```

```{r}
#| eval: true
#| label: tbl-sex-n
#| tbl-cap: Differences in male and female mean vitamin B12 concentration at endpoint compared to baseline (E-B)

sex_stats <- be_b12 %>% 
  group_by(sexf,arm,timepoint) %>%
  select(displayid, sexf,arm,timepoint,vitb12) %>%
  pivot_wider(id_cols = c(displayid, sexf,arm), names_from=timepoint, values_from =vitb12) %>%
  mutate(deltaB12 = Endpoint-Baseline) %>% 
  summarize(across(c(Baseline,Endpoint, deltaB12), 
    list(n=~length(.), mean=~mean(.),sd=~sd(.),se = ~sd(., na.rm = TRUE)/sqrt(sum(!is.na(.))), median = median, min=min,max=max,IQR=IQR
        ))) %>%
  pivot_longer(!c(arm,sexf), names_sep='_', names_to=c('variable', '.value')) %>%
  mutate(lower = mean-qt(0.975, n - 1) * se, upper=mean+qt(0.975, n - 1) * se)

knitr::kable(sex_stats)

be_b12 %>% 
  group_by(sexf,arm,timepoint) %>%
  select(displayid, sexf,arm,timepoint,vitb12) %>%
  pivot_wider(id_cols = c(displayid, sexf,arm), names_from=timepoint, values_from =vitb12) %>%
  mutate(deltaB12 = Endpoint-Baseline) %>%
  ggplot(aes(x=sexf, y = deltaB12)) +
    geom_boxplot(aes(fill=arm)) +
    geom_point(aes(group = sexf, fill = arm), position = position_jitterdodge(dodge.width = 1, jitter.width = 0.5))

```


```{r}
#| eval: true
#| label: CRP

be_b12 %>% group_by(b12_status) %>% dplyr::select(b12_status, contains("CRP")) %>% summarise(across(contains("CRP"), list(n=~n(.na.rm=TRUE),mean=~mean(.,na.rm = TRUE))))

#random imputation fills in gaps
nrow(be_b12 %>% group_by(b12_status) %>% dplyr::select(b12_status, contains("CRP")) %>% dplyr::filter(crp>=3))

be_b12 %>% group_by(b12_status,timepoint) %>% dplyr::select(timepoint,b12_status, contains("CRP")) %>% dplyr::mutate_at(vars(everything()), type.convert) %>% pivot_longer(!c(b12_status,timepoint)) %>% pairwise_wilcox_test(value ~ b12_status, paired = FALSE, p.adjust.method = "BH")

```

#### Comparison of alpha diversity between intervention arms at either timepoint

::: question
Do B12 levels differ significantly between timepoints i.e. are the groups balanced?
:::

```{r}
#| eval: true
#| label: tbl-be-vitb12
#| tbl-cap: Comparison of B12 concentrations between intervention arms at baseline and endpoint using the Wilcoxon rank sign test.
(p <- be_b12 %>%
  dplyr::select(timepoint, arm, vitb12) %>%
  tbl_strata(
    strata = timepoint, .combine_with = c("tbl_stack"), .tbl_fun = ~ .x %>%
      tbl_summary(
        by = arm,
        missing = "no",
        statistic = list(
          all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
          all_categorical() ~ "{n} / {N}  \n({p}%)")) %>% # stats and format for categorical columns
        add_n() %>%
        add_p(everything() ~ "wilcox.test", pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
        add_q(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
        bold_p(q = T) %>%
        add_stat(fns = all_continuous() ~ vda) %>%
        modify_fmt_fun(list(vda_magnitude ~ as.character)) %>%
        modify_header(list(vda_estimate ~ "**Effect Size Estimate**", vda_magnitude ~ "**Magnitude**")),
    .header = "**{strata}**, N = {n}") %>%
		as_gt() %>% fmt_markdown(columns=where(~is.character(.x))) %>% fmt_number(columns = everything(), drop_trailing_zeros = TRUE) %>% fmt_number(columns = where(~is.numeric(.x)<0.0001), pattern = "<0.0001") %>% fmt_number(columns = where(~is.numeric(.x)<0.001), pattern = "<0.001")%>% fmt_number(columns = where(~is.numeric(.x)<0.01), pattern = "<0.01")%>% fmt_integer(columns = where(~is.numeric(.x)<0.05), pattern = "<0.05"))

htmltools::save_html(gt::as_raw_html(p), file.path(RESULTS, "B12-status-subgroup-analysis/BvsE-ferritin-hemoglobin-between-arms.tbl"))
```

```{r}
#| eval: true
#| label: tbl-alpha-b12-be
#| tbl-cap: Comparison of alpha diversity statistics overtime within intervention arm.
(p <- be_b12 %>%
  dplyr::select(timepoint, arm, observed_features = observed, chao1, diversity_shannon, evenness_pielou, pd) %>%
  group_by(arm, timepoint) %>%
  tbl_strata(
    strata = arm, .combine_with = "tbl_stack", .tbl_fun = ~ .x %>%
      tbl_summary(
        by = timepoint,
        missing = "no",
        digits = everything() ~ 2,
        statistic = list(
          all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
          all_categorical() ~ "{n} / {N}  \n({p}%)")) %>% # stats and format for categorical columns
      add_n() %>%
      add_p(everything() ~ "wilcox.test", pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
      add_q(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
      bold_p(q = T)
  ) %>%
  bold_labels() %>%
  bold_levels() %>%
  as_gt() %>% fmt_markdown(columns=where(~is.character(.x))) %>% fmt_number(columns = everything(), drop_trailing_zeros = TRUE) %>% fmt_number(columns = where(~is.numeric(.x)<0.0001), pattern = "<0.0001") %>% fmt_number(columns = where(~is.numeric(.x)<0.001), pattern = "<0.001")%>% fmt_number(columns = where(~is.numeric(.x)<0.01), pattern = "<0.01")%>% fmt_integer(columns = where(~is.numeric(.x)<0.05), pattern = "<0.05"))


htmltools::save_html(gt::as_raw_html(p), file.path(RESULTS, "B12-status-subgroup-analysis/BvsE-ferritin-hemoglobin.tbl"))

```

```{r}
#| eval: true
#| label: tbl-be-wilcox-arm-time
#| tbl-cap: Wilcoxon sign rank test comparing B12 levels between intervention arms within timepoint

## unpaired between group comparisons: arm_base vs. ctrl_base & arm_end vs. ctrl_end
kable(unpaired_wilcox_endpoint <-
  b12_df %>%
  dplyr::select(timepoint, arm, vitb12) %>%
  group_by(timepoint) %>%
  pairwise_wilcox_test(vitb12 ~ arm, paired = FALSE, p.adjust.method = "BH") %>%
  add_xy_position(x = "timepoint", step.increase = 0))
```

```{r}
#| eval: true
#| label: tbl-be-wilcox-arm-overtime
#| tbl-cap: Wilcoxon sign rank test comparing B12 levels between timepoint within intervention arm

## is there a significant difference between baseline and endpoint within treatment group?
## paired within group comparisions: arm_base vs. arm_end & ctrl_base vs. ctrl_end
kable(paired_wilcox_overtime <-
  b12_df %>%
  dplyr::select(timepoint, arm, vitb12) %>%
  reshape2::melt(id.vars = c("timepoint", "arm"), variable.name = "biomarker") %>%
  group_by(arm, biomarker) %>%
  pairwise_wilcox_test(value ~ timepoint, paired = TRUE, p.adjust.method = "BH") %>%
  add_xy_position(x = "timepoint", step.increase = 0))
```

```{r}
#| eval: true
#| label: tbl-be-wilcox
#| tbl-cap: Wilcoxon sign rank test comparing B12 levels between timepoint (collapsed by intervention arm)
kable(overall_paired_wilcox_overtime <-
  b12_df %>%
  dplyr::select(timepoint, arm, vitb12) %>%
  reshape2::melt(id.vars = c("timepoint", "arm"), variable.name = "biomarker") %>%
  group_by(biomarker) %>%
  pairwise_wilcox_test(value ~ timepoint, paired = TRUE, p.adjust.method = "BH") %>%
  add_xy_position(x = "timepoint", step.increase = 0))
```

```{r}
#| eval: true
#| label: fig-be-vitb12
#| fig-cap: Vitamin B12 levels at baseline and endpoint
## visualise as boxplot
b12_df %>%
  dplyr::select(timepoint, arm, vitb12) %>%
  reshape2::melt(id.vars = c("timepoint", "arm"), variable.name = "biomarker") %>%
  ggplot(aes(x = timepoint, y = value)) +
  geom_boxplot(aes(fill = arm), outlier.shape = NA, position = position_dodge(width = 1)) +
  geom_point(aes(group = arm, fill = arm), position = position_jitterdodge(dodge.width = 1, jitter.width = 0.5)) +
  facet_wrap(. ~ biomarker, scales = "free") +
  stat_pvalue_manual(unpaired_wilcox_endpoint, label = "p = {p}", step.increase = 0.05, color = "timepoint") +
  stat_pvalue_manual(paired_wilcox_overtime, label = "p = {p}", step.increase = 0.08, color = "arm") +
  stat_pvalue_manual(overall_paired_wilcox_overtime, label = "p = {p}", step.increase = 0.11, color = "red") +
  xlab(NULL) +
  ylab(NULL) +
  scale_fill_manual(values = armColours) +
  scale_x_discrete(labels = c("Baseline", "Endpoint")) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1), strip.text = element_text(size = 14, face = "bold", color = "black"))
```

It appears that B12 status is different at endpoint compared to baseline.

