
### Differential abundance

```{r}
#| echo: false
#| eval: true
#| label: setup

source(normalizePath("utils/r-utils.r"))
source(normalizePath("_environment.required"))
load(normalizePath("data/seqdata.RData"))

metadata <- read_delim(normalizePath(METADATA))
bac_base_end_meta <- readRDS(normalizePath("data/metadata/base-end-metadata.rds"))
baseline_metadata <- readRDS(normalizePath("data/metadata/baseline-metadata.rds"))
be_b12 <- readRDS(file.path(RESULTS, "B12-status-subgroup-analysis/b12-BvsE-subset.rds"))

suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(pairwiseAdonis))
suppressPackageStartupMessages(library(Glimma))
```

Differential abundance analysis is a common analysis in transcriptomics but the same principals can be applied to differences in abundances rather than expression. The `edgeR` and `limma` R packages can also be used to perform linear modelling on feature tables.

#### Data Format Conversion

1.  Creates a DGEList object (originally designed for RNA-seq)
2.  Stores count data and associated metadata
3.  Used because microbiome count data shares similar properties with RNA-seq data:
    -   Discrete counts
    -   Overdispersion
    -   Multiple testing issues

```{r}
#| eval: true 
#| label: dge object preparation

allcounts <- DGEList(as.data.frame(phyloseq::otu_table(bac_abs_b12)), group = sample_data(bac_abs_b12)$b12_status)

allcounts$samples$timepoint=sample_data(bac_abs_b12)$timepoint
```

The total number of columns DGEList object reflects the number of count files read in
(in this case `lapply(allcounts, ncol)`).

Count columns need to be associated with metadata e.g. experimental variables (both
biological and technical) such as sample names, intervention arm, timepoint,
sequencing batch etc (i.e. anything that may have an effect on gene expression levels)

Within the DGEList data class object there is a "samples" dataframe that associated
metadata with sample names.

There are `length(unique(allcounts$samples$group))` timepoints associated with this
count data.

```{r}
#| eval: true
#| label: Adding taxonomic information

#assign OGU IDs to the DGE object
allcounts$OGU$ID <- rownames(allcounts$counts)

#assign taxonomy table to DGE object
allcounts$OGU$taxonomy <- tax_df(absAbund)
```

::: callout-note
#### Data Transformation

Generally, raw counts must be transformed in order to adjust for the effect of
sequencing depth. Libraries sequenced at a greater depth will obviously result in more
reads and would therefore skew raw read counts. However, the absolute abundaces used
here have already been rarefied and all library sizes are the same.
:::

#### Data Filtering and Quality Control

Removing OTUs (Operational Taxonomic Units) with zero counts across all samples helps
to reduce computational burden and focus on biologically relevant features.

```{r}
#| eval: true
#| label: remove OGUs unobserved across all samples

#a TRUE value is the number of features with no counts across all samples
allcounts.keep <- rowSums(allcounts$counts==0)==ncol(allcounts$counts)
allcounts <- allcounts[allcounts.keep == FALSE,]

```


```{r}
#| eval: true
#| label: tbl-mobi
#| tbl-cap: Mean absolute abundance of Mobiluncus curtisii in each timepoint

mobiAbs <- phyloseq::subset_taxa(seqdata_bac, Species =="Mobiluncus curtisii")

seq_tmp <- seqdata_bac
otu_table(seq_tmp) <- otu_table(microbiome::abundances(seq_tmp, "compositional"), taxa_are_rows =T)

mobiRel <- phyloseq::subset_taxa(seq_tmp, Species =="Mobiluncus curtisii")

mobi <- otu_df(mobiAbs) %>% 
  rownames_to_column()%>% 
  dplyr::rename(OGU_ID = rowname) %>% 
  left_join(., rownames_to_column(as.data.frame(tax_table(mobiAbs))), by = c("OGU_ID" = "rowname"))

mobi.meta <- microbiome::meta(mobiAbs) %>% rownames_to_column()

mobi %>%
  pivot_longer(cols = !c(OGU_ID, Kingdom, Phylum, Class, Order, Family, Genus, Species)) %>%
  mutate(timepoint=mobi.meta$timepoint[match(.$name, mobi.meta$rowname)]) %>%
  dplyr::select(OGU_ID, Species, timepoint, value) %>%
  group_by(timepoint, OGU_ID) %>% 
  summarise(mean = mean(value)) %>%
  kable()
```

```{r}
#| eval: true
#| label: fig-mobi
#| fig-cap: Changes in absolute and relative abundances of Mobiluncus curtisii over time
library(patchwork)

p <- phyloseq::plot_bar(mobiRel, x="timepoint", fill = "timepoint") + geom_bar(aes(fill=timepoint), stat = "identity", position="stack")

p2 <- phyloseq::plot_bar(mobiAbs, x="timepoint", fill = "timepoint") + geom_bar(aes(fill=timepoint), stat = "identity", position="stack")

p + p2
```

#### Exploring OGU Abundance Differences between Timepoints (Endpoint vs Baseline)

As there was no statistically significant differences between intervention arms at either timepoint, the groups were pooled,ignoring intervention arm and compared between timepoints. The code below follows the same procedure as above except contrasts are defined for comparing Endpoint vs Baseline (*EvsB*):

```{r}
#| eval: true
#| label: EvsB design

alldesign <- stats::model.matrix(~0+allcounts$samples$group * allcounts$samples$timepoint)

colnames(alldesign) <-c("deficiency", "insufficiency", "sufficiency", "Endpoint", "insufficiencyEndpoint", "sufficiencyEndpoint")

(allcontrasts.mat <- makeContrasts(
  EvsB = Endpoint-Baseline,
  levels = alldesign
))

allV <- voom(allcounts, alldesign, plot = T, normalize.method="none")

allVfit <- lmFit(allV, alldesign)
allVfit <- contrasts.fit(allVfit, contrasts = allcontrasts.mat)
allEfit <- eBayes(allVfit) #estimation of QL dispersions

```

#### Mean-variance plots

```{r}
#| eval: true
#| label: fig-mean-var-be
#| fig-cap: Mean-Variance trend plot for bacterial abundance data overtime. The plotSA (Sample-Average) plot shows the relationship between gene expression variability (standard deviation) and average abundance (log-CPM).

plotSA(allEfit)
```

#### Identifying differentially abundant OGUs

::: callout-note
##### Using the eBayes or treat function?

The key difference between the eBayes and treat functions lie in how they determine
statistical significance. Both rely on an empirical Bayes framework. While eBayes
tests whether a feature count is simply different from zero (any change), treat tests
whether the change exceeds a user-defined threshold of biological significance. This
makes treat more conservative but potentially more biologically meaningful, as it
reduces false positives from statistically significant but small changes that may not
be biologically relevant.

While the `treat()` function allows you to set the p-value and log fold change (lfc)
cutoffs simultaneously, authors of the `limma` package caution against this
[@Smyth2005]. logFC cutoffs tend to favor lowly expressed genes, resulting in results
that may not have biological significance. In addition, setting the lfc and p-value
cutoffs together can increase the family-wise error rate or false discovery rate above
the nominal level (unless the fold changes and p-values are very highly correlated).
Therefore, in the context of setting a minimum fold change threshold, it is
recommended to use the `treat()` function instead of the `eBayes()` function and
keeping the lfc value at its default value when using decideTests.
:::

A log fold change (logFC) that is significantly greater than 1 (i.e. 2 fold difference
in abundance of feature) -\> e.g. a gene upregulated by a factor of 2 has a
log2(ratio) of 1: $log_2(2) = 1$

The results colunm returns -1, 0 or 1 depending on whether each t-statistic is
classified as significantly negative, not significant or significantly positive.
Setting fc=1.2 or fc=1.5 will usually cause most differentially expressed genes to
have estimated fold-changes of 2-fold or greater, depending on the sample size and
precision of the experiment. Larger thresholds are usually overly conservative and
counter productive. In general, the fc threshold should be chosen sufficiently small
so that a worthwhile number of DE genes remain, otherwise the purpose of prioritizing
genes with larger fold-changes will be defeated.

`decideTest()` performs multiple comparison p-value adjustment on the t-statistic
p-values using the specified method (default = "BH")

##### Features significantly different to zero

> $$
> H_0: FC = 0
> $$

```{r}
#| eval: true
#| label: tbl-fdr5-all

(summary(allFDR5 <- decideTests(allEfit, adjust.method = "BH", p.value = 0.05, lfc = 0)) %>%
  as.data.frame())

#dir.create(file.path(RESULTS, "base-end-differential-abundance-analysis"), showWarnings = FALSE, recursive = TRUE)

#write.csv(summary(allFDR5), file.path(RESULTS, "base-end-differential-abundance-analysis", "base-end-eBayes-FDR5.tbl"))
```

#### Features significantly different to a fold change threshold of 1.2

> $$
> H_0: |FC| \leq 1.2
> $$

```{r}
#| eval: true
#| label: tbl-fc1.2
#| tbl-cap: Total number of statistically significant (q < 0.05) changes in feature counts based on a fold change of 1.2 or more. The Benjimani-Hochberg procedure was used to correct for multiple comparisons.

#EvB DA OGUs- Fold Change of 1.2 & significant P < 0.05
#log fold changes can be changed to desired threshold

#FDR 0.05 and FC of at least 1.2
(summary(FDR5.FC1.2 <- decideTests(treat(allVfit, fc = 1.2), adjust.method = "BH", p.value = 0.05)) %>%
  as.data.frame())

#write.csv(summary(FDR5.FC1.2), file.path(RESULTS, "base-end-differential-abundance-analysis", "base-end-treat-total-FC1.2-FDR5.tbl"))

#write.csv(left_join(rownames_to_column(as.data.frame(topTreat(fit = treat(allVfit, fc = 1.2), n = Inf, confint = 0.95))), rownames_to_column(data.frame(FDR5.FC1.2)), by = "rowname"), row.names = FALSE, file = file.path(RESULTS, "base-end-differential-abundance-analysis", "base-end-treat-counts-FC1.2-FDR5.tbl"))

```

#### Features significantly different to a fold change threshold of 2

> $$
> H_0: |FC| \leq 2
> $$

```{r}
#| eval: true
#| label: tbl-fc2
#| tbl-cap: Total number of statistically significant (q < 0.05) changes in feature counts based on a fold change of 2 or more. The Benjimani-Hochberg procedure was used to correct for multiple comparisons.

#EvB DA OGUs- Fold Change of 2 & significant P < 0.05
#log fold changes can be changed to desired threshold
#FDR 0.05 and log fold change of at least 1 (log2(FC) >= 1) i.e. FC = 2
(summary(FDR5.FC2 <- decideTests(treat(allVfit, fc = 2), adjust.method = "BH", p.value = 0.05)) %>%
  as.data.frame())

# which is the same as:
#summary(decideTests(treat(allVfit, lfc = 1), adjust.method = "BH", p.value = 0.05))

write.csv(summary(FDR5.FC2), file.path(RESULTS, "base-end-differential-abundance-analysis", "base-end-treat-total-FC2-FDR5.tbl"))

#use topTreat as it gives a more intuitave output that write.fit
write.csv(left_join(rownames_to_column(as.data.frame(topTreat(fit = treat(allVfit, fc = 2), n = Inf, confint = 0.95))), rownames_to_column(data.frame(FDR5.FC2)), by = "rowname"), row.names = FALSE, file = file.path(RESULTS, "base-end-differential-abundance-analysis", "base-end-treat-counts-FC2-FDR5.tbl"))

```

```{r}
#write.fit(fit = treat(allVfit, fc = 1.2), results = FDR5.FC1.2, sep = ",", file = paste("~/Work/05-Projects/01-Precision_nutrition/07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/bac_base_end_timepoint_up_down_reg_OGU_table_FDR5_FC1.2", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.csv", sep = "_"))

#write.fit(fit = treat(allVfit, fc = 2), results = FDR5.FC2, sep = ",", file = paste("~/Work/05-Projects/01-Precision_nutrition/07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/bac_base_end_timepoint_up_down_reg_OGU_table_FDR5_FC2", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.csv", sep = "_"))

```

#### Volcano Plots

The `glimma` package visualises data in `MArrayLM` format (limma linear model object)
as volcano plots, showing the OGUs that are significantly different in abundance.

```{r}
#| eval: true
#| label: fig-vol-fdr5
#| fig-cap: "Volcano plot showing the OGUs with increased (blue) and decreased (red) abundance at endpoint compared to baseline. Significant changes were determined using a fold change threshold of > 0. P-values were adjusted using the Benjimani-Hochberg procedure (FDR = 0.05)."
#FDR = 0.05
glimmaVolcano(allEfit,  
              status = allFDR5,
              main=colnames(allEfit), 
              side.main="OGU_ID", 
              counts=allcounts$counts,
              transform.counts='none',
              path= file.path(RESULTS, "base-end-differential-abundance-analysis"),
              html = "bac_base_end_timepoint_up_down_reg_OGUs_volcano_allFDR5.html",
              launch=F)
```

```{r}
#| eval: true
#| label: fig-vol-fc1.2
#| fig-cap: Volcano plot showing the OGUs with increased (blue) and decreased (red) abundance at endpoint compared to baseline. Significant changes were determined using a fold change threshold of at least 1.2. P-values were adjusted using the Benjimani-Hochberg procedure (FDR = 0.05).

#FDR = 0.05 AND fc = 1.2
glimmaVolcano(treat(allVfit, fc = 1.2),
              status = FDR5.FC1.2@.Data,
              main=colnames(allVfit), 
              side.main="OGU_ID", 
              counts = allcounts$counts,
              transform.counts='none',
              path= file.path(RESULTS, "base-end-differential-abundance-analysis"),
              html = "bac_base_end_timepoint_up_down_reg_OGUs_volcano_FDR5_FC1.2.html",
              launch=F)
```

```{r}
#| eval: true
#| label: fig-vol-fc2
#| fig-cap: Volcano plot showing the OGUs with increased (blue) and decreased (red) abundance at endpoint compared to baseline. Significant changes were determined using a fold change threshold of at least 1.2. P-values were adjusted using the Benjimani-Hochberg procedure (FDR = 0.05).

#FDR = 0.05 AND fc = 2
glimmaVolcano(treat(allVfit, fc = 2),
              status = FDR5.FC2,
              main=colnames(allVfit), 
              side.main="OGU_ID", 
              counts = allcounts$counts, 
              path= file.path(RESULTS, "base-end-differential-abundance-analysis"),
              html = "bac_base_end_timepoint_up_down_reg_OGUs_volcano_FDR5_FC2.html",
              launch=F)
```

#### Summary stats of the two thresholds

```{r}
#| eval: true
#| label: tbl-sum-comp
#| tbl-cap: "Comparison of differentially abundant features at endpoint between two fold change thresholds: 1.2 and 2."

tmp <- list(
FC1.2 = read.csv(file.path(RESULTS, "base-end-differential-abundance-analysis", "base-end-treat-counts-FC1.2-FDR5.tbl"), header = T),

FC2 = read.csv(file.path(RESULTS, "base-end-differential-abundance-analysis", "base-end-treat-counts-FC2-FDR5.tbl"), header = T)
)

lapply(1:length(tmp), function(x){
  df <- tmp[[x]] %>% dplyr::filter(EvsB != 0)
  summary <- df %>% group_by(EvsB) %>% summarise(total = n()) %>% as.data.frame()
  n <- nrow(df)

  return(paste0("The number of OGUs with a significant (p < 0.05 FDR) fold change of ", names(tmp)[x], ": ", n, "\n", summary))
})

lapply(tmp, summary)
```

#### Identifying top Differentially Abundant Species (Baseline vs Endpoint; FC \> 1.2)

::: callout-tip
Note that the following results are changes in differential abundance of OGUs at
Endpoint compared to Baseline samples. So, for example, if the coefficient = +2.5 then
that particular OGU increased log2(2.5) fold from baseline to endpoint
:::

```{r}
#| eval: true
#| label: fig-heatmap-spp-1.2
#| fig-cap: Heatmap of the top differentially abundant species at endpoint. Fold change threshold was set to 1.2. FDR = 0.05.

# Creating taxonomically-informed heatmaps of differential abundance
da_results <- read.csv(file.path(RESULTS, "base-end-differential-abundance-analysis", "base-end-treat-counts-FC1.2-FDR5.tbl"), header = T)
colnames(da_results)[colnames(da_results) == "EvsB"] <- "Results"

da_results <- da_results %>% 
  dplyr::filter(Results != 0) %>%
  group_by(Results) %>%
  arrange(desc(abs(logFC)), .by_group = T)

da <- da_results %>%
  ungroup() %>%
  dplyr::select(OGU_ID = rowname, EvsB = logFC) %>% 
  left_join(., rownames_to_column(as.data.frame(tax_table(absAbund))), by = c("OGU_ID" = "rowname")) %>%
  mutate(across(everything(), str_trim)) %>%
  dplyr::select(OGU_ID, EvsB, Phylum, Genus, Species) %>%
  mutate(Species = case_when(
    Genus == "g__" ~ OGU_ID,
    str_detect(Genus, "\\d") ~ OGU_ID,
    Species == "s__" ~ paste(str_remove(Genus, "_[a-zA-Z]\\b"), "sp."),
    str_detect(Species, "\\d") ~ str_remove(Genus, "_[a-zA-Z]\\b"),
    str_detect(Species, "_[a-zA-Z]\\b") ~ paste(str_remove(Species, "_[a-zA-Z]\\b"), "sp."),
    TRUE ~ Species)) %>%
  mutate(Species = str_remove_all(Species, "s__|p__|g__|_[a-zA-Z]\\b")) %>%
  mutate(EvsB = as.numeric(EvsB)) %>%
  mutate(Phylum = str_remove_all(Phylum, "p__|_[a-zA-Z]\\b")) %>%
  group_by(Phylum, Species) %>%
  dplyr::summarise(EvsB = sum(EvsB)) %>%
  as.data.frame()

paste0("Number of unique species: ", nrow(da), " (represented by ", nrow(da_results), " unique OGUs)")

da <- da %>% slice(1:20)

phylum_anno <- assign_unique_colors(as.vector(da$Phylum), phyla_colours)

row_annotation <- HeatmapAnnotation(Phylum = da$Phylum, show_legend=T, col = list(Phylum = phylum_anno), which = "row",annotation_legend_param = list(title_gp = gpar(fontsize = 15, fontface = "bold"), labels_gp = gpar(fontsize = 15), grid_height = unit(8, "mm"), grid_width = unit(4, "mm")))

da_mat <- da
rownames(da_mat) <- da_mat$Species
da_mat$Species <- NULL
da_mat$Phylum <-NULL
da_mat <- data.matrix(da_mat)

Heatmap(da_mat,
        width = unit(0.5, "npc"),
        cluster_rows = T,
        show_row_dend = F,
        col = c('#0571b0','#92c5de','#f4a582','#ca0020'),
        row_names_side = "left",
        row_names_gp=gpar(cex=1, fontface = "bold"),
        show_column_names = F,
        show_heatmap_legend = T,
        right_annotation = row_annotation,
        heatmap_legend_param = list(title="log2 fold change", title_gp = gpar(fontsize = 15, fontface = "bold"), labels_gp = gpar(fontsize = 15), grid_height = unit(8, "mm"), grid_width = unit(4, "mm")),
        cell_fun = function(j, i, x, y, width, height, fill) {
        grid.text(sprintf("%.1f", da_mat[i, j]), x, y, gp = gpar(fontsize = 10))}
        )
```

#### Identifying top Differentially Abundant Genera (Baseline vs Endpoint; FC \> 1.2)

```{r}
#| eval: true
#| label: fig-heatmap-genus-1.2
#| fig-cap: Heatmap of the top differentially abundant genera at endpoint. Fold change threshold was set to 1.2. FDR = 0.05.
da <- da_results %>%
  ungroup() %>%
  dplyr::select(OGU_ID = rowname, EvsB = logFC) %>% 
  left_join(., rownames_to_column(as.data.frame(tax_table(absAbund))), by = c("OGU_ID" = "rowname")) %>%
  mutate(across(everything(), str_trim)) %>%
  dplyr::select(OGU_ID, EvsB, Phylum, Genus, Species) %>%
  mutate(Genus = case_when(
    Genus == "g__" ~ OGU_ID,
    str_detect(Genus, "\\d") ~ OGU_ID,
    str_detect(Genus, "_[a-zA-Z]\\b") ~ str_remove(Genus, "_[a-zA-Z]\\b"),
    TRUE ~ Genus)) %>%
  mutate(EvsB = as.numeric(EvsB)) %>%
  mutate(Genus = str_remove_all(Genus, "g__")) %>%
  mutate(Phylum = str_remove_all(Phylum, "p__|_[a-zA-Z]\\b")) %>%
  group_by(Phylum, Genus) %>%
  dplyr::summarise(EvsB = sum(EvsB)) %>%
  as.data.frame()

paste0("Number of unique genera: ", nrow(da), " (represented by ", nrow(da_results), " unique OGUs)")

da <- da %>% slice(1:20)

phylum_anno <- assign_unique_colors(as.vector(da$Phylum),phyla_colours)

row_annotation <- HeatmapAnnotation(Phylum = da$Phylum, show_legend=T, col = list(Phylum = phylum_anno), which = "row",annotation_legend_param = list(title_gp = gpar(fontsize = 15, fontface = "bold"), labels_gp = gpar(fontsize = 15), grid_height = unit(8, "mm"), grid_width = unit(4, "mm")))

da_mat <- da
rownames(da_mat) <- da_mat$Genus
da_mat$Genus <- NULL
da_mat$Phylum <-NULL
da_mat <- data.matrix(da_mat)

#png(paste("07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/bac_base_end_timepoint_up_down_reg_genera_FDR5_FC2_heatmap", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.png", sep = "_"), 1200, 800)

Heatmap(da_mat,
        width = unit(0.5, "npc"),
        cluster_rows = T,
        show_row_dend = F,
        col = c('#0571b0','#92c5de','#f4a582','#ca0020'),
        row_names_side = "left",
        row_names_gp=gpar(cex=1, fontface = "bold"),
        show_column_names = F,
        show_heatmap_legend = T,
        right_annotation = row_annotation,
        heatmap_legend_param = list(title="log2 fold change", title_gp = gpar(fontsize = 15, fontface = "bold"), labels_gp = gpar(fontsize = 15), grid_height = unit(8, "mm"), grid_width = unit(4, "mm")),
        cell_fun = function(j, i, x, y, width, height, fill) {
        grid.text(sprintf("%.1f", da_mat[i, j]), x, y, gp = gpar(fontsize = 10))}
        )

#dev.off()
```

#### Identifying top Differentially Abundant Species (Baseline vs Endpoint; FC \> 2)

```{r}
#| eval: true
#| label: fig-heatmap-spp-2
#| fig-cap: Heatmap of the top differentially abundant species at endpoint. Fold change threshold was set to 2. FDR = 0.05.

da_results <- read.csv(file.path(RESULTS, "base-end-differential-abundance-analysis", "base-end-treat-counts-FC2-FDR5.tbl"), header = T)

colnames(da_results)[colnames(da_results) == "EvsB"] <- "Results"

da_results <- da_results %>% 
  dplyr::filter(Results != 0 & adj.P.Val <=0.05 & abs(logFC) >= 1) %>%
  group_by(Results) %>%
  arrange(desc(abs(logFC)), .by_group = T)

dim(da_results)

da <- da_results %>%
  ungroup() %>%
  dplyr::select(OGU_ID = rowname, EvsB = logFC) %>% 
  left_join(., rownames_to_column(as.data.frame(tax_table(absAbund))), by = c("OGU_ID" = "rowname")) %>%
  mutate(across(everything(), str_trim)) %>%
  dplyr::select(OGU_ID, EvsB, Phylum, Genus, Species) %>%
  mutate(Species = case_when(
    Genus == "g__" ~ OGU_ID,
    str_detect(Genus, "\\d") ~ OGU_ID,
    Species == "s__" ~ paste(str_remove(Genus, "_[a-zA-Z]\\b"), "sp."),
    str_detect(Species, "\\d") ~ str_remove(Genus, "_[a-zA-Z]\\b"),
    str_detect(Species, "_[a-zA-Z]\\b") ~ paste(str_remove(Species, "_[a-zA-Z]\\b"), "sp."),
    TRUE ~ Species)) %>%
  mutate(Species = str_remove_all(Species, "s__|p__|g__|_[a-zA-Z]\\b")) %>%
  mutate(EvsB = as.numeric(EvsB)) %>%
  mutate(Phylum = str_remove_all(Phylum, "p__|_[a-zA-Z]\\b")) %>%
  group_by(Phylum, Species) %>%
  dplyr::summarise(EvsB = sum(EvsB)) %>%
  as.data.frame()

paste0("Number of unique species: ", nrow(da), " (represented by ", nrow(da_results), " unique OGUs)")

da <- da %>% slice(1:20)

phylum_anno <- assign_unique_colors(as.vector(da$Phylum),phyla_colours)

row_annotation <- HeatmapAnnotation(Phylum = da$Phylum, show_legend=T, col = list(Phylum = phylum_anno), which = "row",annotation_legend_param = list(title_gp = gpar(fontsize = 15, fontface = "bold"), labels_gp = gpar(fontsize = 15), grid_height = unit(8, "mm"), grid_width = unit(4, "mm")))

da_mat <- da
rownames(da_mat) <- da_mat$Species
da_mat$Species <- NULL
da_mat$Phylum <-NULL
da_mat <- data.matrix(da_mat)

Heatmap(da_mat,
        width = unit(0.5, "npc"),
        cluster_rows = T,
        show_row_dend = F,
        col = c('#0571b0','#92c5de','#f4a582','#ca0020'),
        row_names_side = "left",
        row_names_gp=gpar(cex=1, fontface = "bold"),
        show_column_names = F,
        show_heatmap_legend = T,
        right_annotation = row_annotation,
        heatmap_legend_param = list(title="log2 fold change", title_gp = gpar(fontsize = 15, fontface = "bold"), labels_gp = gpar(fontsize = 15), grid_height = unit(8, "mm"), grid_width = unit(4, "mm")),
        cell_fun = function(j, i, x, y, width, height, fill) {
        grid.text(sprintf("%.1f", da_mat[i, j]), x, y, gp = gpar(fontsize = 10))}
        )

```

#### Identifying top Differentially Abundant Genera (Baseline vs Endpoint; FC \> 2)

```{r}
#| eval: true
#| label: fig-heatmap-genus-2
#| fig-cap: Heatmap of the top differentially abundant species at endpoint. Fold change threshold was set to 2. FDR = 0.05.
da <- da_results %>%
  ungroup() %>%
  dplyr::select(OGU_ID = rowname, EvsB = logFC) %>% 
  left_join(., rownames_to_column(as.data.frame(tax_table(absAbund))), by = c("OGU_ID" = "rowname")) %>%
  mutate(across(everything(), str_trim)) %>%
  dplyr::select(OGU_ID, EvsB, Phylum, Genus, Species) %>%
  mutate(Genus = case_when(
    Genus == "g__" ~ OGU_ID,
    str_detect(Genus, "\\d") ~ OGU_ID,
    str_detect(Genus, "_[a-zA-Z]\\b") ~ str_remove(Genus, "_[a-zA-Z]\\b"),
    TRUE ~ Genus)) %>%
  mutate(EvsB = as.numeric(EvsB)) %>%
  mutate(Genus = str_remove_all(Genus, "g__")) %>%
  mutate(Phylum = str_remove_all(Phylum, "p__|_[a-zA-Z]\\b")) %>%
  group_by(Phylum, Genus) %>%
  dplyr::summarise(EvsB = sum(EvsB)) %>%
  as.data.frame()

paste0("Number of unique genera: ", nrow(da), " (represented by ", nrow(da_results), " unique OGUs)")

da <- da %>% slice(1:20)

phylum_anno <- assign_unique_colors(as.vector(da$Phylum),phyla_colours)

row_annotation <- HeatmapAnnotation(Phylum = da$Phylum, show_legend=T, col = list(Phylum = phylum_anno), which = "row",annotation_legend_param = list(title_gp = gpar(fontsize = 15, fontface = "bold"), labels_gp = gpar(fontsize = 15), grid_height = unit(8, "mm"), grid_width = unit(4, "mm")))

da_mat <- da
rownames(da_mat) <- da_mat$Genus
da_mat$Genus <- NULL
da_mat$Phylum <-NULL
da_mat <- data.matrix(da_mat)

Heatmap(da_mat,
        width = unit(0.5, "npc"),
        cluster_rows = T,
        show_row_dend = F,
        col = c('#0571b0','#92c5de','#f4a582','#ca0020'),
        row_names_side = "left",
        row_names_gp=gpar(cex=1, fontface = "bold"),
        show_column_names = F,
        show_heatmap_legend = T,
        right_annotation = row_annotation,
        heatmap_legend_param = list(title="log2 fold change", title_gp = gpar(fontsize = 15, fontface = "bold"), labels_gp = gpar(fontsize = 15), grid_height = unit(8, "mm"), grid_width = unit(4, "mm")),
        cell_fun = function(j, i, x, y, width, height, fill) {
        grid.text(sprintf("%.1f", da_mat[i, j]), x, y, gp = gpar(fontsize = 10))}
        )
```

#### Comparision of Differentially Abundant Species and LMM results

TODO

```{r}
#| eval: false
#| label: FC2 + LMM result comparisons

paste("Number of signifcant fold changes greater than 2:", nrow(da_results))

bac_taxa_signif <- read.csv("~/Work/05-Projects/01-Precision_nutrition/07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/Base_End_Bacteria_LMM_top_significant_OGUs_Mumbai_ID14768_Feb_14_2024_04:40_NC.csv", header = T)

apply(bac_taxa_signif, 2, function(x) any(x <= 0.05, na.rm = TRUE)) #determine which column if any has significant p-values

paste(nrow(bac_taxa_signif %>% 
            filter(time_lmm_adj <= 0.05)), "taxa identifeid as changing over time in LMMs")

common <- bac_taxa_signif %>% 
            filter(time_lmm_adj <= 0.05) %>%
            arrange(time_lmm_adj) %>%
            left_join(., da_results, by=c("ogus_all" = "X")) %>%
            #select(ogus_all =rowname,!c(Kingdom,Phylum, Class,Order, Family, Genus,Species)) %>%
            filter(Results == 1 | Results ==-1)  %>%
            left_join(., rownames_to_column(as.data.frame(tax_table(absAbund))), by = c("ogus_all" = "rowname")) %>%
            mutate(across(everything(), str_trim)) %>%
            mutate(Species = case_when(
              Genus == "g__" ~ ogus_all,
              str_detect(Genus, "\\d") ~ ogus_all,
              Species == "s__" ~ paste(str_remove(Genus, "_[a-zA-Z]\\b"), "sp."),
              str_detect(Species, "sp\\d") ~ str_replace(Species, "sp\\d*", "sp."),
              str_detect(Species, "\\d") ~ str_remove(Genus, "_[a-zA-Z]\\b"),
              str_detect(Species, "_[a-zA-Z]\\b") ~ str_remove(Species, "_[a-zA-Z]\\b"),
              TRUE ~ Species)) %>%
            mutate(Species = str_remove_all(Species, "s__|p__|g__|_[a-zA-Z]\\b")) %>%
            mutate(logFC = as.numeric(logFC)) %>%
            mutate(Phylum = str_remove_all(Phylum, "p__|_[a-zA-Z]\\b")) %>%
            group_by(Results) %>%
            arrange(desc(abs(logFC)), .by_group = T)

#write.csv(common, paste("~/Work/05-Projects/01-Precision_nutrition/07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/bac_base_end_timepoint_up_down_reg_OGUs_FDR5_FC2_LMM", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.csv", sep = "_"), row.names = F)

paste0(nrow(common), " OGUs (",length(unique(common$Species))," unique species and ", length(unique(common$Genus)), " unique genera) are significantly changing over time between LMM and DA results i.e. are common between the two methods")

common_mat <- common %>%
  slice(1:30) %>%
  ungroup() %>%
  group_by(Phylum, Species) %>%
  dplyr::summarise(EvsB = sum(logFC), FC = fc(EvsB)) %>%
  arrange(desc(abs(EvsB)), .by_group = T) %>%
  as.data.frame()

#total number of species in each phylum
common_mat %>% group_by(Phylum) %>% mutate(count = 1) %>% summarise(sum=sum(count))

#write.csv(common_mat, paste("~/Work/05-Projects/01-Precision_nutrition/07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/bac_base_end_timepoint_up_down_reg_OGUs_FDR5_FC2_LMM_total_abundanceFC", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.csv", sep = "_"), row.names = F)

phylum_anno <- assign_unique_colors(as.vector(common_mat$Phylum),phyla_colours)

row_annotation <- HeatmapAnnotation(Phylum = common_mat$Phylum, show_legend=T, col = list(Phylum = phylum_anno), which = "row")

rownames(common_mat) <- common_mat$Species
common_mat$Species <- NULL
common_mat$Phylum <-NULL
common_mat$FC <- NULL
common_mat <- data.matrix(common_mat)

lgd1 <- Legend(col_fun = circlize::colorRamp2(c(-12,-6,0), rev(c('#bdd7e7','#6baed6','#2171b5'))), title = "Log2 Fold Change\nin OGU Abundance", title_position = "topcenter", legend_gp = gpar(fontface = "bold"))
lgd2 <- Legend(labels = c("0","1"), legend_gp = gpar(fill = c("#FC8D62","#8DA0CB"), fontface = "bold"), title = "Intervention\nArm", title_position = "topcenter")

#png(paste("~/Work/05-Projects/01-Precision_nutrition/07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/bac_base_end_timepoint_up_down_reg_OGUs_FDR5_FC1.2_LMM_heatmap", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.png", sep = "_"), 1200, 800)

Heatmap(common_mat,
        width = unit(0.1, "npc"),
        cluster_rows = T,
        show_row_dend = F,
        col = circlize::colorRamp2(c(-12,-6,0), rev(c('#bdd7e7','#6baed6','#2171b5'))),
        row_names_side = "left",
        row_names_gp=gpar(cex=1, fontface = "bold"),
        show_column_names = F,
        show_heatmap_legend = F,
        right_annotation = row_annotation,
        cell_fun = function(j, i, x, y, width, height, fill) {
        grid.text(sprintf("%.1f", common_mat[i, j]), x, y, gp = gpar(fontsize = 10))}
        )

draw(lgd1, x = unit(0.8, "npc"), y = unit(0.5, "npc"), just = "centre")

#dev.off()
```

```{r}
#| eval: false
#| label: abundance ranges for Fc2 + LMM results
abund_ranges <- do.call(rbind, apply(as.data.frame(v$E), 1, function(x){
  data.frame(mean = mean(x), median=median(x), range = paste(round(min(x), 3), "-", round(max(x),3)), IQR = max(x)-min(x))
}))
abund_ranges %>% arrange(mean)
```

Edit: 11-09-24 Reformating the LMM/DA common OGU table to retain the label of OGUs
with no taxonomic name assigned rather than reporting their OGU -\> for publication

```{r}
#| eval: false
#| label: reformating table for publication

common <- bac_taxa_signif %>% 
            filter(time_lmm_adj <= 0.05) %>%
            arrange(time_lmm_adj) %>%
            left_join(., da_results, by=c("ogus_all" = "X")) %>%
            #select(ogus_all =rowname,!c(Kingdom,Phylum, Class,Order, Family, Genus,Species)) %>%
            filter(Results == 1 | Results ==-1)  %>%
            left_join(., rownames_to_column(as.data.frame(tax_table(absAbund))), by = c("ogus_all" = "rowname")) %>%
            mutate(across(everything(), str_trim)) %>%
            mutate(Species = str_remove_all(Species, "s__|p__|g__|_[a-zA-Z]\\b")) %>%
            mutate(logFC = as.numeric(logFC)) %>%
            mutate(Phylum = str_remove_all(Phylum, "p__|_[a-zA-Z]\\b")) %>%
            mutate(Class = str_remove_all(Class, "c__|_[a-zA-Z]\\b")) %>%
            mutate(Order = str_remove_all(Order, "o__|_[a-zA-Z]\\b")) %>%
            mutate(Family = str_remove_all(Family, "f__|_[a-zA-Z]\\b")) %>%
            mutate(Genus = str_remove_all(Genus, "g__|_[a-zA-Z]\\b")) %>% 
            select(!Kingdom) %>%
            group_by(Results) %>%
            arrange(desc(abs(logFC)), .by_group = T)

#write.csv(common, paste("~/Work/05-Projects/01-Precision_nutrition/07-Research_Q_analysis/03-diversity_analysis/02-Longitudinal_analysis/01-Base_End_alpha_diversity/01-Bacteria/01-Differential_abundance/publication_table_bac_base_end_timepoint_up_down_reg_OGUs_FDR5_FC2_LMM", "Mumbai_ID14768", format(Sys.time(), "%b_%d_%Y_%I:%M"), "NC.csv", sep = "_"), row.names = F)

```


