### B12 Status: Endpoint Alpha Diversity Statistics

```{r}
#| echo: false
#| eval: true
#| label: setup
#| include: false

# This initial code block loads custom utility functions and environment variables

# These are stored in separate files to keep the main analysis clean and organized
source(normalizePath("utils/r-utils.r"))
source(normalizePath("_environment.required"))
load(normalizePath("data/seqdata.RData"))

metadata <- read_delim(normalizePath(METADATA))
bac_base_end_meta <- readRDS(normalizePath("data/metadata/base-end-metadata.rds"))
endpoint_metadata <- readRDS(normalizePath("data/metadata/endpoint-metadata.rds"))
b12_end_meta <-readRDS(file.path(RESULTS, "B12-status-subgroup-analysis/b12-endpoint-metadata.rds"))

suppressPackageStartupMessages(library(emmeans))
suppressPackageStartupMessages(library(nlme))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ampvis2))


dir.create(file.path(RESULTS, "B12-status-subgroup-analysis"), showWarnings = FALSE, recursive = TRUE)
```

#### Between Arms: Comparison of endpoint alpha diversity between intervention arms

```{r}
#| eval: true
#| warning: false
#| label: tbl-B12-end-arm
#| tbl-cap: Comparison of alpha diversity statistics at endpoint between intervention arms in each vitamin B12 status group.

(p <- b12_end_meta %>%
  dplyr::select(b12_2Lev, `Chao 1 Index` = chao1, `Shannon Diversity Index` = diversity_shannon, Evenness = evenness_pielou, `Faith's Phylogenetic Diversity` = pd) %>%
  gt_summary(by = "b12_2Lev"))

htmltools::save_html(gt::as_raw_html(p), file.path(RESULTS, "B12-status-subgroup-analysis/2Level-endpoint-B12-status-summary.tbl"))
```


##### Influence of B12 status on alpha diversity in each intervention arm

```{r}
#| eval: true
#| warning: false
#| label: tbl-B12-end
#| tbl-cap: compare B12 status at endpoint
(p <- b12_end_meta %>%
  dplyr::select(arm, b12_2Lev, b12_pmol, `Chao 1 Index` = chao1, `Shannon Diversity Index` = diversity_shannon, Evenness = evenness_pielou, `Faith's Phylogenetic Diversity` = pd) %>%
  gt_strata(strata = "arm", by = "b12_2Lev", vda = F))

htmltools::save_html(gt::as_raw_html(p), file.path(RESULTS, "B12-status-subgroup-analysis/endpoint-B12-status.tbl"))
```

The table shows that there is a significant difference between B12 status groups. Pairwise comparisons are made below.

```{r}
#| eval: true
#| label: tbl-b12-wilcox
#| tbl-cap: Wilcoxon rank sign compariosn test between B12 status groups in both intervention arm

#compare between intervention arms
kable(unpaired_intervention_wilcox <- b12_end_meta %>%
  dplyr::select(arm, b12_2Lev, `Chao 1 Index` = chao1, `Shannon Diversity Index` = diversity_shannon, Evenness = evenness_pielou, `Faith's Phylogenetic Diversity` = pd) %>%
  pivot_longer(!c(arm,b12_2Lev), names_to="diversity_stat") %>%
  group_by(diversity_stat,arm) %>%
  wilcox_test(value ~ b12_2Lev, paired = FALSE) %>%
  add_xy_position(x = "b12_2Lev",  step.increase = 0))
```

```{r}
#| eval: true
#| label: tbl-b12-wilcox-signif
#| tbl-cap: "Significant p-values: Wilcoxon rank sign compariosn test between B12 status groups in both intervention arm"

#compare between intervention arms
kable(b12_end_meta %>%
  dplyr::select(arm, b12_2Lev, `Chao 1 Index` = chao1, `Shannon Diversity Index` = diversity_shannon, Evenness = evenness_pielou, `Faith's Phylogenetic Diversity` = pd) %>%
  pivot_longer(!c(arm,b12_2Lev), names_to="diversity_stat") %>%
  group_by(diversity_stat,arm) %>%
  wilcox_test(value ~ b12_2Lev, paired = FALSE) %>%
  filter(p.adj <= 0.05))
```

```{r}
#| eval: true
#| label: fig-end-b12-arm
#| fig-cap: Alpha diversity statistics among vitamin B12 status groups at endpoint separated by intervention arm (n = 127).
#| fig-height: 14
b12_end_meta %>%
  dplyr::select(arm, b12_2Lev, `Chao 1 Index` = chao1, `Shannon Diversity Index` = diversity_shannon, Evenness = evenness_pielou, `Faith's Phylogenetic Diversity` = pd) %>%
  pivot_longer(!c(arm,b12_2Lev), names_to="diversity_stat") %>%
  group_by(diversity_stat) %>%
  ggplot(aes(x = b12_2Lev, y = value)) +
      geom_boxplot(aes(fill = arm), outlier.shape=NA) +
    geom_point(aes(group=arm),position=position_dodge(width=0.75)) +
      facet_wrap(. ~ factor(diversity_stat), 2 ,2, scales = "free") +
    stat_pvalue_manual(unpaired_intervention_wilcox, label = "p = {p}", step.increase = 0.1) +
      xlab(NULL) +
      ylab(NULL) +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.1)))

```

Significant differences between intervention arms at endpoint suggest that randomisation into arms was not unbiased.

#### Between B12 Status: Influence of B12 status on alpha diversity at endpoint

```{r}
#| eval: true
#| label: tbl-b12-alpha-end
#| tbl-cap: Influence of B12 status on alpha diversity at endpoint

(p <- b12_end_meta %>%
  dplyr::select(b12_2Lev, b12_pmol, `Chao 1 Index` = chao1, `Shannon Entropy` = diversity_shannon, Evenness = evenness_pielou, `Faith's Phylogenetic Diversity` = pd) %>%
    gt_summary(by = "b12_2Lev"))


htmltools::save_html(gt::as_raw_html(p), file.path(RESULTS, "B12-status-subgroup-analysis/endpoint-b12-differences-by-arm.tbl"))
```

#### Within Sex: Comparison of endpoint alpha diversity between B12 status groups within male and female groupings

```{r}
#| eval: true
#| warning: false
#| label: tbl-end-alpha-sex-B12
#| tbl-cap: "Within Sex: differences in alpha diversity by sex and B12 status"
(p <- b12_end_meta %>%
  dplyr::select(Sex = sexf, b12_2Lev, `Chao 1 Index` = chao1, `Shannon Diversity Index` = diversity_shannon, Evenness = evenness_pielou, `Faith's Phylogenetic Diversity` = pd) %>%
    gt_strata(strata = "Sex", by = "b12_2Lev", vda = FALSE))


htmltools::save_html(gt::as_raw_html(p), file.path(RESULTS, "B12-status-subgroup-analysis/endpoint-alpha-div-by-B12-status-sex.tbl"))

```

```{r}
#| eval: true
#| label: tbl-b12-sex-wilcox
#| tbl-cap: Pairwise comparisons between alpha diversity statistics in males and females with different B12 status

#compare between intervention arms
kable(unpaired_sex_wilcox <- b12_end_meta %>%
  dplyr::select(Sex=sexf, b12_2Lev, `Chao 1 Index` = chao1, `Shannon Diversity Index` = diversity_shannon, Evenness = evenness_pielou, `Faith's Phylogenetic Diversity` = pd) %>%
  pivot_longer(!c(Sex,b12_2Lev), names_to="diversity_stat") %>%
  group_by(diversity_stat,Sex) %>%
  wilcox_test(value ~ b12_2Lev, paired = FALSE) %>%
  add_xy_position(x = "b12_2Lev",  step.increase = 0))
```

#### Between Sex: Comparison of alpha diversity between males and females with the same B12 status

```{r}
#| eval: true
#| warning: false
#| label: tbl-end-alpha-sex-B12
#| tbl-cap: differences in alpha diversity by sex and B12 status
(p <- b12_end_meta %>%
  dplyr::select(Sex = sexf, b12_2Lev, `Chao 1 Index` = chao1, `Shannon Diversity Index` = diversity_shannon, Evenness = evenness_pielou, `Faith's Phylogenetic Diversity` = pd) %>%
      gt_strata(by = "Sex", strata = "b12_2Lev", vda = FALSE))


htmltools::save_html(gt::as_raw_html(p), file.path(RESULTS, "B12-status-subgroup-analysis/endpoint-alpha-div-between-sexes-B12.tbl"))

```

```{r}
#| eval: true
#| label: tbl-b12-sex-wilcox
#| tbl-cap: Pairwise comparisons between alpha diversity statistics in males and females with different B12 status

#compare between intervention arms
kable(unpaired_sex_wilcox <- b12_end_meta %>%
  dplyr::select(Sex=sexf, b12_2Lev, `Chao 1 Index` = chao1, `Shannon Diversity Index` = diversity_shannon, Evenness = evenness_pielou, `Faith's Phylogenetic Diversity` = pd) %>%
  pivot_longer(!c(Sex,b12_2Lev), names_to="diversity_stat") %>%
  group_by(diversity_stat,b12_2Lev) %>%
  wilcox_test(value ~ Sex, paired = FALSE) %>%
  add_xy_position(x = "Sex",  step.increase = 0), table.attr = "style = \"color: #C6D0F5;\"")
```

```{r}
#| eval: true
#| label: fig-end-b12-sex
#| fig-cap: Alpha diversity statistics among vitamin B12 status groups at endpoint separated by sex (n = 127).
#| fig-height: 14
b12_end_meta %>%
  dplyr::select(Sex = sexf, b12_2Lev, `Chao 1 Index` = chao1, `Shannon Diversity Index` = diversity_shannon, Evenness = evenness_pielou, `Faith's Phylogenetic Diversity` = pd) %>%
  pivot_longer(!c(Sex,b12_2Lev), names_to="diversity_stat") %>%
  group_by(diversity_stat) %>%
  ggplot(aes(x = b12_2Lev, y = value)) +
      geom_boxplot(aes(fill = Sex), outlier.shape=NA) +
    geom_point(aes(group=Sex),position=position_dodge(width=0.75)) +
      facet_wrap(. ~ factor(diversity_stat), 2 ,2, scales = "free") +
      xlab(NULL) +
      ylab(NULL) +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.1)))

```

