### B12 Status: Endpoint Bacterial Relative Sums

```{r}
#| echo: false
#| eval: true
#| label: setup
#| include: false

# This initial code block loads custom utility functions and environment variables

# These are stored in separate files to keep the main analysis clean and organized
source(normalizePath("utils/r-utils.r"))
source(normalizePath("_environment.required"))
load(normalizePath("data/seqdata.RData"))

metadata <- read_delim(normalizePath(METADATA))
endpoint_metadata <- readRDS(normalizePath("data/metadata/endpoint-metadata.rds"))

suppressPackageStartupMessages(library(emmeans))
suppressPackageStartupMessages(library(nlme))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ampvis2))

dir.create(file.path(RESULTS, "B12-status-subgroup-analysis"), showWarnings = FALSE, recursive = TRUE)
```

#### Data preparation

Serum vitamin B12 concentration (`b12_pmol`) is in pg/mL and is converted below to pmol/L using the 0.738 conversion factor.

| Characteristic                       | Median (IQR) or n (%) |
| :----------------------------------- | :-------------------- |
| Serum vitamin B12 (pmol/L)           | 5.45 (5.08, 5.93)     |
|   Tertile 1 (n=42)                     | 4.99 (4.79, 5.08)     |
|   Tertile 2 (n=43)                     | 5.45 (5.34, 5.62)     |
|   Tertile 3 (n=42)                     | 6.13 (5.93, 6.34)     |
| Binary: Deficiency (<148 pmol/L)     | 21 (16.5%)            |
| 3-level: Insufficiency (≥148 & <221) | 34 (26.8%)            |
| Binary: Insufficiency (<221 pmol/L)  | 55 (43.3%)            |
| Sufficient (≥221 pmol/L)             | 72 (56.7%)            |

```{r}
#| eval: true
#| label: create ampvis object
#| results: asis

#adjust metadata to contian new b12 varaibles
b12_end_meta <- endpoint_metadata %>%
mutate(b12_pmol = vitb12*0.738) %>%
  mutate(
    b12_tertiles = tertiles(b12_pmol),
    b12_2Lev = case_when(
      b12_pmol < 148 ~ "deficiency",
      b12_pmol >= 148 ~ "sufficiency",
      TRUE ~ "Unknown"
    ),
    b12_3Lev = case_when(
      b12_pmol < 148 ~ "deficiency",
      b12_pmol >= 148 & b12_pmol < 221 ~ "insufficiency",
      b12_pmol >= 221 ~ "sufficiency",
      TRUE ~ "Unknown"
    ),
    sexf = case_when(
      sexf == 1 ~ "Female",
      sexf == 0 ~ "Male"
    ),
    arm = case_when(
      trt == 1 ~ "FeZnCPM",
      trt == 0 ~ "CPM"
    )
  ) %>%
  relocate(sample_names, .before=everything()) %>%
  dplyr::filter(!is.na(b12_pmol)) 

rownames(b12_end_meta) <- b12_end_meta$sample_names

saveRDS(b12_end_meta, file.path(RESULTS, "B12-status-subgroup-analysis/b12-endpoint-metadata.rds"))

```

```{r}
#| eval: true

end_b12 <- subset_samples(seqdata_bac, !is.na(vitb12))
end_b12 <- subset_samples(end_b12, timepoint=="3Endpoint")

sample_data(end_b12) <- b12_end_meta

#slight modification to collapse Bacillota subgroups into one. As of Apr 2025 I cannot find any valid publication of this taxonomy. It appears to be based purely on the analysis of MAGs at GTDB
tax_table(end_b12) <- tax_df(end_b12) %>% 
                            mutate(across(where(is.character), str_trim, "both")) %>%
                            mutate(Phylum = str_remove(Phylum, "_[a-zA-Z]\\b")) %>%
                            mutate(Genus = str_remove(Genus, "_[a-zA-Z]\\b")) %>%
                            mutate(Species = str_remove(Species, "_[a-zA-Z]\\b")) %>%
                            mutate(Phylum = case_when(
                              Phylum == "" ~ "Unknown",
                              str_detect(Phylum, "\\d") ~ "Unclassified",
                              TRUE ~ Phylum)) %>%
                            mutate(Genus = case_when(
                              str_detect(Genus, "\\d") ~ "Unclassified",
                              Genus == "" ~ "Unknown",
                              TRUE ~ Genus)) %>%
                            as.matrix()

saveRDS(end_b12, file.path(RESULTS, "B12-status-subgroup-analysis/b12-endpoint-seqdata.rds"))

#Rows are OTU's, columns are samples
#metadata first column must contain sample ID's matching those in the otutable
d <- amp_load(otutable = t(otu_table(end_b12)), metadata = b12_end_meta, taxonomy = tax_table(end_b12))
```

#### Basic Characteristics of Subgroup

```{r}
#| eval: true
#| label: tbl-end-char
#| tbl-cap: Endpoint characterisitcs of subjects with vitamin B12 measurements available.

b12_end_meta %>% select(Sex = sexf, `Intervention Arm`= arm, b12_tertiles, b12_2Lev, b12_3Lev, `B12 Concentration (pg/mL)` = b12_pmol) %>% tbl_summary() %>% as_kable()

```

#### Endpoint Phyla Abundance

```{r}
#| eval: true
#| label: fig-end-b12-phyla-heatmap
#| fig-cap: Relative abundance of bacterial phyla at endpoint. Average group relative abundances are shown on the graph.

hmaps <- lapply(list("b12_tertiles", "b12_2Lev", "b12_3Lev"), function(x){
  phylum_heatmap(d, x)
})

names(hmaps) <- c("b12_tertiles", "b12_2Lev", "b12_3Lev")

ggarrange(plotlist = hmaps, ncol=3, common.legend = TRUE, labels = "AUTO")
```

```{r}
#| eval: true
#| label: tbl-b12-phyla-heatmap
#| tbl-cap: Relative abundance of bacterial phyla at endpoint.
#| results: asis

lapply(hmaps, function(heatmap){
heatmap$data %>% 
  group_by(Display, .Group) %>% 
  summarise(mean = round(mean(Sum), 4)) %>% 
  arrange(desc(mean)) %>%
  mutate(mean =  case_when(
    mean >= 0.01 ~ round(mean, 2), 
    TRUE ~ mean
  )) %>%
  mutate(mean = paste0(mean, "%")) %>%
  pivot_wider(names_from = .Group, values_from = mean) %>%
  kable()

})

```

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: Summary statistics of bacterial relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
#| results: asis

hmaps_sum <- map2(hmaps, c("b12_tertiles", "b12_2Lev", "b12_3Lev"), hm_data, metadata = b12_end_meta) %>% set_names(c("b12_tertiles", "b12_2Lev", "b12_3Lev")) 

lapply(hmaps_sum, function(i){
  i %>% select(Phylum = phyla, `B12 Status` = grouping_var, `Median (IQR; Min-Max)`) %>% pivot_wider(names_from = `B12 Status`, values_from = `Median (IQR; Min-Max)`) %>% arrange() %>% gt() %>% fmt_markdown(columns=where(~is.character(.x))) %>% fmt_number(columns = everything(), drop_trailing_zeros = TRUE) %>% fmt_number(columns = where(~is.numeric(.x)<0.0001), pattern = "<0.0001") %>% fmt_number(columns = where(~is.numeric(.x)<0.001), pattern = "<0.001")%>% fmt_number(columns = where(~is.numeric(.x)<0.01), pattern = "<0.01")%>% fmt_integer(columns = where(~is.numeric(.x)<0.05), pattern = "<0.05")
})

```

```{r}
#| eval: true
#| label: gtsummary table
#| fig-cap: NA
#| tbl-cap: NA

combined_df <- do.call(rbind, Map(function(df, name) {
  df$data$source <- name
  return(df$data)
}, hmaps, names(hmaps)))

hm_gtsum <- function(data){
  data %>%
    group_by(.Group) %>%
    select(!Sample) %>%
    tbl_strata(strata = .Group, .tbl_fun = ~ .x %>%
      tbl_summary(
        by = source,
        digits = everything() ~ 2,
        missing = "no",
        statistic = list(
  all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
  all_categorical() ~ "{n} / {N}  \n({p}%)")
      ) %>% # stats and format for categorical columns
      add_n() %>%
      add_overall() %>%
      add_p(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
      add_q(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
      bold_p(q = TRUE),
	  .header = "**{strata}**, N = {n}") %>%
        # add_stat(fns = all_continuous() ~ vda) %>%
        # modify_fmt_fun(list(vda_magnitude ~ as.character)) %>%
        # modify_header(list(vda_estimate ~ "**Effect Size Estimate**", vda_magnitude ~ "**Magnitude**")) %>%
      as_kable()
      # as_gt() %>% fmt_markdown(columns=where(~is.character(.x))) %>% fmt_number(columns = everything(), drop_trailing_zeros = TRUE) %>% fmt_number(columns = where(~is.numeric(.x)<0.0001), pattern = "<0.0001") %>% fmt_number(columns = where(~is.numeric(.x)<0.001), pattern = "<0.001")%>% fmt_number(columns = where(~is.numeric(.x)<0.01), pattern = "<0.01")%>% fmt_integer(columns = where(~is.numeric(.x)<0.05), pattern = "<0.05")
}

hm_gtsum(combined_df)


combined_df %>%
  select(!c(Sample, Abundance)) %>%
    tbl_strata(strata = Display, .tbl_fun = ~ .x %>%
      tbl_summary(

        by = .Group,
        digits = everything() ~ 2,
        statistic = list(
  all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
  all_categorical() ~ "{n} / {N}  \n({p}%)")
      ), # stats and format for categorical columns
	  .header = "**{strata}**, N = {n}")


hmaps$b12_tertiles$data %>%
  select(!c(Sample, Abundance)) %>%
  tbl_summary(
    type = everything() ~ "continuous",
        by = Display,
        digits = everything() ~ 2,
        statistic = list(
  all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
  all_categorical() ~ "{n} / {N}  \n({p}%)")
      ) %>% # stats and format for categorical columns
      add_n() %>%
      add_overall() %>%
      add_p(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
      add_q(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
      bold_p(q = TRUE)

```

```{r}
#| eval: true
#| label: fig-end-phyla-boxplot
#| fig-cap: Relative abundances of the bacterial phyla at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).

bplots <- lapply(c("b12_tertiles", "b12_2Lev", "b12_3Lev"), phyla_boxplot, ampvis_object = d) 

ggarrange(plotlist = bplots, ncol=3, common.legend = FALSE, labels = "AUTO")

```

##### Statistical Comparison between B12 Status group Phyla Sum

::: h0

There is no differnece in the counts of bacterial phyla between vitamin B12 status groups at endpoint 

:::

::: h1

Bacterial phyla counts diverge between vitamin B12 status groups at endpoint

:::

```{r}
#| eval: true
#| label: tbl-b12-kw
#| tbl-cap: Kruskall-Wallis test results with computed $\eta^2_H$ effect size estimate for bacterial phyla relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
#| results: asis

stat <- lapply(c("b12_tertiles", "b12_2Lev", "b12_3Lev"), function(x){

  tmp <- amp_heatmap(d,
              group_by = x,
              tax_aggregate = "Phylum",
              tax_show = 20,
              plot_values = T,
              normalise = F, 
              tax_empty="OTU")

  df <- tmp$data %>% dplyr::rename(phyla = Display, b12_status = `.Group`)

  kw <- df %>% group_by(phyla) %>%
    kruskal_test(Sum ~ b12_status)

  kw_eff<-df %>% group_by(phyla) %>% kruskal_effsize(Sum ~ b12_status)

  kw <- left_join(kw,kw_eff, by="phyla")

  # Post-hoc test
  dunn <- df %>% group_by(phyla) %>%
      dunn_test(Sum ~ b12_status)

    b12_vda <- data.frame()

    for(p in unique(df$phyla)) {
      tmp_vda <- df %>% dplyr::filter(phyla==p) %>%
      calculate_pairwise_vda(., "Sum", "b12_status") %>% 
      mutate(phyla=p) %>% 
      relocate(phyla, .before=everything())

      b12_vda <- rbind(b12_vda,tmp_vda)
    }

    dunn <-left_join(dunn, b12_vda, by =c("phyla", "group1", "group2"))

  return(list(data = df, `Kruskal-Wallis Test` = kw , `Post-hoc Dunn's Test` = dunn))
  
}) %>% set_names(c("b12_tertiles", "b12_2Lev", "b12_3Lev"))

lapply(stat, function(x){
  print(names(x))
  kable(x$`Kruskal-Wallis Test`)
  })

```

```{r}
#| eval: true
#| label: tbl-b12-dunn-vda
#| tbl-cap: Kruskall-Wallis post hoc Dunn's test results with computed pairwise VDA effect size estimate for bacterial phyla relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
#| results: asis

lapply(stat, function(x){
  print(names(x))
  kable(x$`Post-hoc Dunn's Test`)
})
```

###### Fusobacteria and B12

```{r}
#| eval: true
#| label: fig-end-fuso
#| fig-cap: Fusobacterium at endpoint

df <- df %>%
  mutate(b12_pmol = endpoint_metadata$b12_pmol[match(Sample, endpoint_metadata$sample_names)],
  sex = endpoint_metadata$sexf[match(Sample, endpoint_metadata$sample_names)]) 

library(corrplot)
library(Hmisc)

# Create correlation matrix
cor_matrix <- rcorr(as.matrix(df[c("Sum", "b12_status")]), type = "spearman")

# Visualize
corrplot(cor_matrix$r, type = "upper", 
         p.mat = cor_matrix$P, 
         sig.level = 0.05)

fuso <- df %>%
  filter(phyla == "Fusobacteriota")

fuso %>%
  ggplot(aes(x=b12_status, y = Sum)) +
    geom_boxplot() +
    geom_jitter() +
    coord_cartesian(ylim=c(0,3000))

```

```{r}
#| eval: true
#| label: fig-corr-plot

# Reshape data to wide format (each phyla becomes a column)
phyla_wide <- df %>%
  select(Sample, phyla, Sum, b12_status) %>%
  pivot_wider(names_from = phyla, 
              values_from = Sum) %>%
  select(-Sample)  # Remove Sample column if you don't need it

# Calculate correlation matrix
cor_matrix <- rcorr(as.matrix(phyla_wide), type = "spearman")

# Create corrplot
corrplot(cor_matrix$r,
         p.mat = cor_matrix$P,
         method = "color",    # can be "circle", "square", "number"
         type = "upper",      # show only upper triangle
         order = "hclust",    # cluster similar correlations together
         addCoef.col = "black", # add correlation coefficients
         tl.col = "black",    # text label color
         tl.srt = 45,         # rotate text labels
         sig.level = 0.05,    # significance level
         insig = "blank")     # hide insignificant correlations

# Alternative with significance stars
corrplot(cor_matrix$r,
         p.mat = cor_matrix$P,
         method = "color",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45,
         sig.level = c(0.001, 0.01, 0.05),
         pch.cex = 0.9,
         insig = "label_sig")  # show significance stars


phyla_summary <- df %>%
  group_by(phyla, b12_status) %>%
  summarise(total_abundance = sum(Sum)) %>%
  pivot_wider(names_from = b12_status, 
              values_from = total_abundance)

# Calculate correlations
cor_matrix <- cor(phyla_summary[,-1], method = "spearman")  # remove phyla names column
p_values <- cor.mtest(phyla_summary[,-1], method = "spearman")$p

# Create corrplot
corrplot(cor_matrix,
         method = "color",
         type = "full",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45,
         sig.level = 0.05,
         insig = "blank")

```


Poisson distribution and log link function used as this is count data.

```{r}
#| eval: true
#| label: tbl-end-glm-fuso

anova(glm(Sum ~ b12_pmol*sex, data=fuso, family = "poisson"))
```

#### Endpoint Genera Abundance

```{r}
#| eval: true
#| label: fig-end-b12-gen-heatmap
#| fig-cap: Relative abundance of top 20 bacterial genera at endpoint. Average group relative abundances are shown on the graph.

gen_maps <- lapply(list("b12_tertiles", "b12_2Lev", "b12_3Lev"), function(x){
  genus_heatmap(d, x)
})

names(gen_maps) <- c("b12_tertiles", "b12_2Lev", "b12_3Lev")

ggarrange(plotlist = gen_maps, ncol=3, common.legend = TRUE, labels = "AUTO")
```

```{r}
#| eval: true
#| label: tbl-b12-spp-heatmap
#| tbl-cap: Relative abundance of top 20 bacterial genera at endpoint.
#| results: asis

lapply(gen_maps, function(heatmap){
heatmap$data %>% 
  group_by(Display, .Group) %>% 
  summarise(mean = round(mean(Sum), 4)) %>% 
  arrange(desc(mean)) %>%
  mutate(mean =  case_when(
    mean >= 0.01 ~ round(mean, 2), 
    TRUE ~ mean
  )) %>%
  mutate(mean = paste0(mean, "%")) %>%
  pivot_wider(names_from = .Group, values_from = mean) %>%
  kable()

})
```

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: Summary statistics of bacterial relative abundances at endpoint.
#| results: asis

gen_maps_sum <- map2(gen_maps, c("b12_tertiles", "b12_2Lev", "b12_3Lev"), hm_data, metadata = b12_end_meta) %>% set_names(c("b12_tertiles", "b12_2Lev", "b12_3Lev")) 

lapply(gen_maps_sum, function(i){
  i %>% select(Phylum = phyla, `B12 Status` = grouping_var, `Median (IQR; Min-Max)`) %>% pivot_wider(names_from = `B12 Status`, values_from = `Median (IQR; Min-Max)`) %>% arrange() %>% gt() %>% fmt_markdown(columns=where(~is.character(.x))) %>% fmt_number(columns = everything(), drop_trailing_zeros = TRUE) %>% fmt_number(columns = where(~is.numeric(.x)<0.0001), pattern = "<0.0001") %>% fmt_number(columns = where(~is.numeric(.x)<0.001), pattern = "<0.001")%>% fmt_number(columns = where(~is.numeric(.x)<0.01), pattern = "<0.01")%>% fmt_integer(columns = where(~is.numeric(.x)<0.05), pattern = "<0.05")
})

```

```{r}
#| eval: true
#| label: gtsummary table
#| fig-cap: NA
#| tbl-cap: NA

combined_df <- do.call(rbind, Map(function(df, name) {
  df$data$source <- name
  return(df$data)
}, gen_maps, names(gen_maps)))

hm_gtsum <- function(data){
  data %>%
    group_by(.Group) %>%
    select(!Sample) %>%
    tbl_strata(strata = .Group, .tbl_fun = ~ .x %>%
      tbl_summary(
        by = source,
        digits = everything() ~ 2,
        missing = "no",
        statistic = list(
  all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
  all_categorical() ~ "{n} / {N}  \n({p}%)")
      ) %>% # stats and format for categorical columns
      add_n() %>%
      add_overall() %>%
      add_p(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
      add_q(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
      bold_p(q = TRUE),
	  .header = "**{strata}**, N = {n}") %>%
        # add_stat(fns = all_continuous() ~ vda) %>%
        # modify_fmt_fun(list(vda_magnitude ~ as.character)) %>%
        # modify_header(list(vda_estimate ~ "**Effect Size Estimate**", vda_magnitude ~ "**Magnitude**")) %>%
      as_kable()
      # as_gt() %>% fmt_markdown(columns=where(~is.character(.x))) %>% fmt_number(columns = everything(), drop_trailing_zeros = TRUE) %>% fmt_number(columns = where(~is.numeric(.x)<0.0001), pattern = "<0.0001") %>% fmt_number(columns = where(~is.numeric(.x)<0.001), pattern = "<0.001")%>% fmt_number(columns = where(~is.numeric(.x)<0.01), pattern = "<0.01")%>% fmt_integer(columns = where(~is.numeric(.x)<0.05), pattern = "<0.05")
}

hm_gtsum(combined_df)


combined_df %>%
  select(!c(Sample, Abundance)) %>%
    tbl_strata(strata = Display, .tbl_fun = ~ .x %>%
      tbl_summary(

        by = .Group,
        digits = everything() ~ 2,
        statistic = list(
  all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
  all_categorical() ~ "{n} / {N}  \n({p}%)")
      ), # stats and format for categorical columns
	  .header = "**{strata}**, N = {n}")


gen_maps$b12_tertiles$data %>%
  select(!c(Sample, Abundance)) %>%
  tbl_summary(
    type = everything() ~ "continuous",
        by = Display,
        digits = everything() ~ 2,
        statistic = list(
  all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
  all_categorical() ~ "{n} / {N}  \n({p}%)")
      ) %>% # stats and format for categorical columns
      add_n() %>%
      add_overall() %>%
      add_p(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
      add_q(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
      bold_p(q = TRUE)

```

```{r}
#| eval: true
#| label: fig-end-phyla-boxplot
#| fig-cap: Relative abundances of the top five most abundant bacterial phyla at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).

bplots <- lapply(c("b12_tertiles", "b12_2Lev", "b12_3Lev"), phyla_boxplot, ampvis_object = d) 

ggarrange(plotlist = bplots, ncol=3, common.legend = FALSE, labels = "AUTO")

```

###### Statistical Compariosn between B12 Status group  Genera Sum

::: h0

there is no differnece in count of genera between vitamin B12 status groups at endpoint 

:::

::: h1

genera count increases/decreases from vitamin B12 status groups at endpoint

:::

```{r}
#| eval: true
#| label: tbl-sum-gen
#| tbl-cap: Summary of counts
#| results: asis

stat <- lapply(c("b12_tertiles", "b12_2Lev", "b12_3Lev"), function(x){

  tmp <- amp_heatmap(d,
              group_by = x,
              tax_aggregate = "Genus",
              tax_show = 20,
              plot_values = T,
              normalise = F, 
              tax_empty="OTU")

  df <- tmp$data %>% dplyr::rename(genus = Display, b12_status = `.Group`)

  kw <- df %>% group_by(genus) %>%
    kruskal_test(Sum ~ b12_status)

  kw_eff<-df %>% group_by(genus) %>% kruskal_effsize(Sum ~ b12_status)

  kw <- left_join(kw,kw_eff, by="genus")

  # Post-hoc test
  dunn <- df %>% group_by(genus) %>%
      dunn_test(Sum ~ b12_status)

    b12_vda <- data.frame()

    for(p in unique(df$genus)) {
      tmp_vda <- df %>% dplyr::filter(genus==p) %>%
      calculate_pairwise_vda(., "Sum", "b12_status") %>% 
      mutate(genus=p) %>% 
      relocate(genus, .before=everything())

      b12_vda <- rbind(b12_vda,tmp_vda)
    }

    dunn <-left_join(dunn, b12_vda, by =c("genus", "group1", "group2"))

  return(list(data = df, `Kruskal-Wallis Test` = kw , `Post-hoc Dunn's Test` = dunn))
  
}) %>% set_names(c("b12_tertiles", "b12_2Lev", "b12_3Lev"))

lapply(stat, function(x){
  print(names(x))
  kable(x$`Kruskal-Wallis Test`)
  })
```

```{r}
#| eval: true
#| label: tbl-b12-dunn-vda
#| tbl-cap: Kruskall-Wallis post hoc Dunn's test results with computed pairwise VDA effect size estimate for bacterial phyla relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
#| results: asis

lapply(stat, function(x){
  print(names(x))
  kable(x$`Post-hoc Dunn's Test`)
})
```

#### Endpoint Species Abundance

```{r}
#| eval: true
#| label: fig-end-b12-phyla-heatmap
#| fig-cap: Relative abundance of bacterial phyla at endpoint. Average group relative abundances are shown on the graph.

spp_maps <- lapply(list("b12_tertiles", "b12_2Lev", "b12_3Lev"), function(x){
  species_heatmap(d, x)
})

names(spp_maps) <- c("b12_tertiles", "b12_2Lev", "b12_3Lev")

ggarrange(plotlist = spp_maps, ncol=3, common.legend = TRUE, labels = "AUTO")
```

```{r}
#| eval: true
#| label: tbl-b12-species-heatmap
#| tbl-cap: Relative abundance of bacterial species at endpoint.
#| results: asis

lapply(spp_maps, function(heatmap){
heatmap$data %>% 
  group_by(Display, .Group) %>% 
  summarise(mean = round(mean(Sum), 4)) %>% 
  arrange(desc(mean)) %>%
  mutate(mean =  case_when(
    mean >= 0.01 ~ round(mean, 2), 
    TRUE ~ mean
  )) %>%
  mutate(mean = paste0(mean, "%")) %>%
  pivot_wider(names_from = .Group, values_from = mean) %>%
  kable()

})

```

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: Summary statistics of bacterial relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
#| results: asis

spp_maps_sum <- map2(spp_maps, c("b12_tertiles", "b12_2Lev", "b12_3Lev"), hm_data, metadata = b12_end_meta) %>% set_names(c("b12_tertiles", "b12_2Lev", "b12_3Lev")) 

lapply(spp_maps_sum, function(i){
  i %>% select(Species = species, `B12 Status` = grouping_var, `Median (IQR; Min-Max)`) %>% pivot_wider(names_from = `B12 Status`, values_from = `Median (IQR; Min-Max)`) %>% arrange() %>% gt() %>% fmt_markdown(columns=where(~is.character(.x))) %>% fmt_number(columns = everything(), drop_trailing_zeros = TRUE) %>% fmt_number(columns = where(~is.numeric(.x)<0.0001), pattern = "<0.0001") %>% fmt_number(columns = where(~is.numeric(.x)<0.001), pattern = "<0.001")%>% fmt_number(columns = where(~is.numeric(.x)<0.01), pattern = "<0.01")%>% fmt_integer(columns = where(~is.numeric(.x)<0.05), pattern = "<0.05")
})

```

```{r}
#| eval: true
#| label: gtsummary table
#| fig-cap: NA
#| tbl-cap: NA

combined_df <- do.call(rbind, Map(function(df, name) {
  df$data$source <- name
  return(df$data)
}, spp_maps, names(spp_maps)))

hm_gtsum <- function(data){
  data %>%
    group_by(.Group) %>%
    select(!Sample) %>%
    tbl_strata(strata = .Group, .tbl_fun = ~ .x %>%
      tbl_summary(
        by = source,
        digits = everything() ~ 2,
        missing = "no",
        statistic = list(
  all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
  all_categorical() ~ "{n} / {N}  \n({p}%)")
      ) %>% # stats and format for categorical columns
      add_n() %>%
      add_overall() %>%
      add_p(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
      add_q(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
      bold_p(q = TRUE),
	  .header = "**{strata}**, N = {n}") %>%
        # add_stat(fns = all_continuous() ~ vda) %>%
        # modify_fmt_fun(list(vda_magnitude ~ as.character)) %>%
        # modify_header(list(vda_estimate ~ "**Effect Size Estimate**", vda_magnitude ~ "**Magnitude**")) %>%
      as_kable()
      # as_gt() %>% fmt_markdown(columns=where(~is.character(.x))) %>% fmt_number(columns = everything(), drop_trailing_zeros = TRUE) %>% fmt_number(columns = where(~is.numeric(.x)<0.0001), pattern = "<0.0001") %>% fmt_number(columns = where(~is.numeric(.x)<0.001), pattern = "<0.001")%>% fmt_number(columns = where(~is.numeric(.x)<0.01), pattern = "<0.01")%>% fmt_integer(columns = where(~is.numeric(.x)<0.05), pattern = "<0.05")
}

hm_gtsum(combined_df)


combined_df %>%
  select(!c(Sample, Abundance)) %>%
    tbl_strata(strata = Display, .tbl_fun = ~ .x %>%
      tbl_summary(

        by = .Group,
        digits = everything() ~ 2,
        statistic = list(
  all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
  all_categorical() ~ "{n} / {N}  \n({p}%)")
      ), # stats and format for categorical columns
	  .header = "**{strata}**, N = {n}")


spp_maps$b12_tertiles$data %>%
  select(!c(Sample, Abundance)) %>%
  tbl_summary(
    type = everything() ~ "continuous",
        by = Display,
        digits = everything() ~ 2,
        statistic = list(
  all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
  all_categorical() ~ "{n} / {N}  \n({p}%)")
      ) %>% # stats and format for categorical columns
      add_n() %>%
      add_overall() %>%
      add_p(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
      add_q(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
      bold_p(q = TRUE)

```

```{r}
#| eval: true
#| label: fig-end-species-boxplot
#| fig-cap: Relative abundances of the top five most abundant bacterial species at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).

bplots <- lapply(c("b12_tertiles", "b12_2Lev", "b12_3Lev"), species_boxplot, ampvis_object = d) 

ggarrange(plotlist = bplots, ncol=3, common.legend = FALSE, labels = "AUTO")

```

##### Statistical Comparison between B12 Status group Phyla Sum

::: h0

There is no differnece in the counts of bacterial phyla between vitamin B12 status groups at endpoint 

:::

::: h1

Bacterial phyla counts diverge between vitamin B12 status groups at endpoint

:::

```{r}
#| eval: true
#| label: tbl-b12-kw
#| tbl-cap: Kruskall-Wallis test results with computed $\eta^2_H$ effect size estimate for bacterial phyla relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
#| results: asis

stat <- lapply(c("b12_tertiles", "b12_2Lev", "b12_3Lev"), function(x){

  tmp <- amp_heatmap(d,
              group_by = x,
              tax_aggregate = "Phylum",
              tax_show = 20,
              plot_values = T,
              normalise = F, 
              tax_empty="OTU")

  df <- tmp$data %>% dplyr::rename(phyla = Display, b12_status = `.Group`)

  kw <- df %>% group_by(phyla) %>%
    kruskal_test(Sum ~ b12_status)

  kw_eff<-df %>% group_by(phyla) %>% kruskal_effsize(Sum ~ b12_status)

  kw <- left_join(kw,kw_eff, by="phyla")

  # Post-hoc test
  dunn <- df %>% group_by(phyla) %>%
      dunn_test(Sum ~ b12_status)

    b12_vda <- data.frame()

    for(p in unique(df$phyla)) {
      tmp_vda <- df %>% dplyr::filter(phyla==p) %>%
      calculate_pairwise_vda(., "Sum", "b12_status") %>% 
      mutate(phyla=p) %>% 
      relocate(phyla, .before=everything())

      b12_vda <- rbind(b12_vda,tmp_vda)
    }

    dunn <-left_join(dunn, b12_vda, by =c("phyla", "group1", "group2"))

  return(list(data = df, `Kruskal-Wallis Test` = kw , `Post-hoc Dunn's Test` = dunn))
  
}) %>% set_names(c("b12_tertiles", "b12_2Lev", "b12_3Lev"))

lapply(stat, function(x){
  print(names(x))
  kable(x$`Kruskal-Wallis Test`)
  })

```

```{r}
#| eval: true
#| label: tbl-b12-dunn-vda
#| tbl-cap: Kruskall-Wallis post hoc Dunn's test results with computed pairwise VDA effect size estimate for bacterial phyla relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
#| results: asis

lapply(stat, function(x){
  print(names(x))
  kable(x$`Post-hoc Dunn's Test`)
})
```

###### Statistical Compariosn between B12 Status group  Species Sum

::: h0

there is no differnece in realtive abundance of species between vitamin B12 status groups at endpoint 

:::

::: h1

species realtive abundance increases/decreases from vitamin B12 status groups at endpoint

:::
```{r}
#| eval: true
#| label: tbl-b12-kw
#| tbl-cap: Kruskall-Wallis test results with computed $\eta^2_H$ effect size estimate for bacterial species relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
#| results: asis

stat <- lapply(c("b12_tertiles", "b12_2Lev", "b12_3Lev"), function(x){

  tmp <- amp_heatmap(d,
              group_by = x,
              tax_aggregate = "Species",
              tax_show = 20,
              plot_values = T,
              normalise = F, 
              tax_empty="OTU")

  df <- tmp$data %>% dplyr::rename(species = Display, b12_status = `.Group`)

  kw <- df %>% group_by(species) %>%
    kruskal_test(Sum ~ b12_status)

  kw_eff<-df %>% group_by(species) %>% kruskal_effsize(Sum ~ b12_status)

  kw <- left_join(kw,kw_eff, by="species")

  # Post-hoc test
  dunn <- df %>% group_by(species) %>%
      dunn_test(Sum ~ b12_status)

    b12_vda <- data.frame()

    for(p in unique(df$species)) {
      tmp_vda <- df %>% dplyr::filter(species==p) %>%
      calculate_pairwise_vda(., "Sum", "b12_status") %>% 
      mutate(species=p) %>% 
      relocate(species, .before=everything())

      b12_vda <- rbind(b12_vda,tmp_vda)
    }

    dunn <-left_join(dunn, b12_vda, by =c("species", "group1", "group2"))

  return(list(data = df, `Kruskal-Wallis Test` = kw , `Post-hoc Dunn's Test` = dunn))
  
}) %>% set_names(c("b12_tertiles", "b12_2Lev", "b12_3Lev"))

lapply(stat, function(x){
  print(names(x))
  kable(x$`Kruskal-Wallis Test`)
  })

```

```{r}
#| eval: true
#| label: tbl-b12-dunn-vda
#| tbl-cap: Kruskall-Wallis post hoc Dunn's test results with computed pairwise VDA effect size estimate for bacterial species relative abundances at endpoint among those with B12 deficiency (<200 pg/mL), insufficiency (200-300 pg/mL), and sufficiency (>300 pg/mL).
#| results: asis

lapply(stat, function(x){
  print(names(x))
  kable(x$`Post-hoc Dunn's Test`)
})
```

```{r}
#| eval: true
#| label: fig-end-heatmap
#| fig-cap: Heatmap of phyla and species differences between 

ggplot(stat$df, aes(x = b12_status, y = Sum)) +
  # Add boxplot
  geom_boxplot(aes(fill = b12_status), alpha = 0.5) +
  # Add points with size reflecting abundance
  geom_point(aes(size = Abundance), 
             position = position_jitterdodge(),
             alpha = 0.6) +
  # Facet by species
  facet_wrap(~species, scales = "free_y") +
  # Add significance levels from Dunn's test
  geom_text(data = dunn_spp_results, 
            aes(x = group1, y = max(df$Sum) * 1.1,
                label = p.adj.signif),
            size = 3) +
  # Customize appearance
  scale_fill_manual(values = c(
    "deficiency" = "#e78284",
    "insufficiency" = "#f4b8e4",
    "sufficiency" = "#8CAAEE"
  )) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  ) +
  labs(
    x = "B12 Status",
    y = "Sum",
    size = "Abundance",
    fill = "B12 Status",
    title = "Species Abundance by B12 Status",
    subtitle = "Point size indicates abundance; ns: not significant"
  )

#Bubble plot with significance connections
abundance_summary <- count_data %>%
  group_by(species, b12_status) %>%
  summarise(
    mean_abundance = mean(Abundance),
    mean_sum = mean(Sum)
  )
  
ggplot(abundance_summary, 
       aes(x = b12_status, y = species)) +
  # Add bubbles
  geom_point(aes(size = mean_sum, fill = mean_abundance),
             shape = 21, alpha = 0.7) +
  # Add significance lines
  geom_segment(data = dunns_results,
               aes(x = group1, xend = group2,
                   y = species, yend = species,
                   linetype = p.adj.signif),
               position = position_dodge(width = 0.9)) +
  scale_size_continuous(range = c(2, 20)) +
  scale_fill_gradient(
    low = "#8CAAEE",
    high = "#f4b8e4"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
